---
title: "RiverRevitalisation_analysis"
author: "GEgloff"
date: "`r Sys.Date()`"
output: word_document
---

introduction video for R Markdown used as reference here (https://www.youtube.com/watch?v=K418swtFnik&t=253s&ab_channel=RforEcology)
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load packages
```{r load packages, results = 'hide'}
library(car)
library(corrplot)
library(DHARMa) #simulateresiduals
library(dplyr)
library(ggplot2)
library(ggeffects)
library(glmmTMB)
library(lme4)
library(patchwork) #plotting
library(posterior)
library(purrr)
library(RColorBrewer)
library(readxl)
library(stringr)
library(tidyr)
library(tidyverse)
library(vegan) #NMDS
```


# Data import and preparation

Import beedata and mapp_data, create metadata frame
```{r import data, include = FALSE}
# nolint start: nonportable_path_linter.
beedata <- read_excel(
  "C:/Users/grege/OneDrive - unine.ch/PhD/Chapters/Chapter 1 - River revitalisation/data/beedata_RRproject_2025.xlsx",
  skip = 1)

mappingdata <- read_excel(
  "C:/Users/grege/OneDrive - unine.ch/PhD/Chapters/Chapter 1 - River revitalisation/data/mappingdata_RRproject_2025.xlsx",
  skip = 1)

canopyhp <- read_excel(
  "C:/Users/grege/OneDrive - unine.ch/PhD/Chapters/Chapter 1 - River revitalisation/data/canopy_hp/canopydata_RRproject_2025.xlsx",
  skip = 1)
# nolint end
#add column with plotID

beedata <- beedata |>
  mutate(plotID = sub("\\.[^.]+(?=\\.[^.]+$)", "", surveyID, perl = TRUE))

mappingdata <- mappingdata |>
  mutate(plotID = sub("\\.[^.]+(?=\\.[^.]+$)", "", surveyID, perl = TRUE))

metadata <- mappingdata |>
  select(surveyID, plotID, site, region, visit, plot) |>
  distinct()
```

Prepare mapping data (without plants)
```{r mapp_data, habplot, habsite prep}
mapp_data <- mappingdata

# add canopy hp data 
mapp_data <- mapp_data |>
  left_join(canopyhp |> select(plotID, canopy_hp), by = "plotID")

# Subset the data frame to include plot scale mapping data (not including plant data)
columns_plot <- c("plotID", "plot", "site", "region", "Xcoord", "Ycoord",
                  "Zcoord", "canton", "date", "visit", "surveyID", "observer",
                  "municipality", "bareground", "vegcover", "vegheight", "herb",
                  "woody", "sand", "gravel", "cobble", "stone", "wood", "soil",
                  "flat", "sloped", "steep", "canopy", "canopy_hp",
                  "standingdeadwood", "lyingdeadwood", "branchpile",
                  "stonepile", "patchiness")

habplot_data <- mapp_data[, columns_plot]

# Subset the data frame to include site scale mapping data 
# (including site scale plant data)
columns_site <- c("plotID", "plot", "site", "region", "Xcoord", "Ycoord",
                  "Zcoord", "canton", "date", "visit", "surveyID", "observer",
                  "municipality", "sitebareground", "sitevegcover",
                  "sitevegheight", "siteherb", "sitewoody", "sitesand",
                  "sitegravel", "sitecobble", "sitestone", "sitesoil",
                  "siteflat", "sitesloped", "sitesteep", "sitecanopy",
                  "sitepatchiness", "sitecornis", "sitecomment", "siteacer",
                  "sitecornus", "sitecrataegus", "siteeuonymus", "sitefrangula",
                  "sitefraxinus", "ligustrum", "sitelonicera","siteprunus",
                  "siterosa", "sitesalix", "sitesambucus", "siteviburnum",
                  "plotID")


# Subset the data frame
habsite_data <- mapp_data[, columns_site]


# calculate proportion of each bareground variable to full plot area
cols <- c("sand", "gravel", "cobble", "stone", "wood", "soil")
mapp_data[cols] <- mapp_data[cols] * mapp_data$bareground / 100
```

Prepare sampling covariate data
```{r sampling covariates data prep}
# Subset the data frame to include sampling covariates
columns_sampling <- c("site","region", "visit", "plotID","surveyID", "date",
                      "site", "temp", "wind", "cloud", "shade", "distwater",
                      "timestart", "timestop")

samplcov_data <- mapp_data[, columns_sampling]
```

Prepare plant genus matrix
```{r plant data prep}
plant_data <- mapp_data[, which(names(mapp_data) == "achillea")
                        :which(names(mapp_data) == "vicia")]

# replace NA with 0 
plant_data[is.na(plant_data)] <- 0

# Calculate plant richness: number of genera present per row
plant_data$richness <- rowSums(plant_data > 0, na.rm = TRUE)

# Calculate relative abundance: sum of all genus abundances per row
plant_data$relabundance <- rowSums(plant_data, na.rm = TRUE)

#Calculate the avg. relative abundance per plant genus per plot
plant_data$avgrelabundance <- plant_data$relabundance/plant_data$richness

plant_data$site <- mapp_data$site

plant_data$visit <- mapp_data$visit

plant_data$region <- mapp_data$region

plant_data$plotID <- mapp_data$plotID

plant_data$surveyID <- mapp_data$surveyID
```

```{r plant diversity indices}
# Hill number for q = 0 (species richness)
plant_data$hill_q0 <- rowSums(plant_data[1:131] > 0, na.rm = TRUE)

# Hill number for q = 1 (exponential of Shannon)
plant_data$hill_q1 <- exp(diversity(plant_data[1:131], index = "shannon"))

# Hill number for q = 2 (inverse Simpson)
plant_data$hill_q2 <- 1 / diversity(plant_data[1:131], index = "invsimpson")


# Some rows have q2 = Inf where Simpson index is zero (no species recorded) - convert Inf to NA for computation
plant_data$hill_q2[is.infinite(plant_data$hill_q2)] <- NA

# Current plant_data has metadata at the end; let's reorder
plant_data <- plant_data[, c("site","region", "visit", "plotID",
                             "surveyID", "hill_q0", "hill_q1", "hill_q2",
                              setdiff(names(plant_data),
                                      c("site", "visit", "plotID", "region",
                                        "surveyID", "hill_q0", "hill_q1",
                                        "hill_q2")))]
```

Prepare bee data
```{r bee data prep}
glimpse(beedata)

# remove extra plots (plots #13) and Wasps
beedata <- beedata[!(beedata$plot == 13 | beedata$Gattung == "Wasp"), ]

# Combine genus and species into a single species name
beedata <- beedata |>
  mutate(bee_species = paste(Gattung, Art, sep = " "))


# Count number of individuals per species per surveyID
bee_abundance_matrix <- beedata |>
  group_by(surveyID, bee_species) |>
  summarise(bee_abundance = n(), .groups = "drop") |>
  pivot_wider(names_from = bee_species, values_from = bee_abundance, values_fill = 0)

# Reorder columns: keep 'surveyID' first, sort the rest alphabetically
bee_abundance_matrix <- bee_abundance_matrix[, c(
  "surveyID",
  sort(setdiff(names(bee_abundance_matrix), "surveyID"))
)]

# check if right number of rows
nrow(beedata) # 1871 bees total
sum(bee_abundance_matrix[,-1])

# Merge metadata with species abundance matrix
bee_data <- left_join(metadata, bee_abundance_matrix, by = "surveyID")

str(bee_data)

#add field ID morphospecies of Bombus and Xylocopa (only for abundance data
bee_data <- left_join(bee_data, mapp_data |> select(surveyID, fieldIDsum), by = "surveyID")

#Calculate bee species richness per plot
bee_data$bee_richness <- rowSums(bee_data[, !(names(bee_data) %in% c("surveyID", "plotID", "site", "region", "visit", "plot", "fieldIDsum"))] > 0, na.rm = TRUE)

# Calculate total abundance by summing all species columns (not including fieldID Bombus / Xylocopa)
bee_data$bee_abundance_nofieldID <- rowSums(bee_data[, !(names(bee_data) %in% c("surveyID", "plotID", "site", "region", "visit", "plot", "fieldIDsum", "bee_richness"))], na.rm = TRUE)

# Calculate total abundance by summing all species columns (including FieldIF)
bee_data$bee_abundance<- rowSums(bee_data[, !(names(bee_data) %in% c("surveyID", "plotID", "site", "region", "visit", "plot", "bee_abundance_nofieldID", "bee_richness"))], na.rm = TRUE)

# RL Abundance per survey (drop NA RLstatus)
rl_abundance <- beedata |>
  filter(!is.na(RLstatus)) |>
  group_by(surveyID, RLstatus) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(
    names_from = RLstatus,
    values_from = n,
    values_fill = 0,
    names_prefix = "RLabund_"
  )

# RL Richness per survey (drop NA RLstatus)
rl_richness <- beedata |>
  filter(!is.na(RLstatus)) |>
  mutate(bee_species = paste(Gattung, Art)) |>
  distinct(surveyID, RLstatus, bee_species) |> 
  count(surveyID, RLstatus) |>
  pivot_wider(
    names_from = RLstatus,
    values_from = n,
    values_fill = 0,
    names_prefix = "RLrich_"
  )

# Merge RL metrics into bee_data
bee_data <- bee_data |>
  left_join(rl_abundance, by = "surveyID") |>
  left_join(rl_richness, by = "surveyID")

# Ensure threatened columns exist
for (col in c("RLrich_VU", "RLrich_EN", "RLrich_CR")) {
  if (!col %in% names(bee_data)) bee_data[[col]] <- 0
}
for (col in c("RLabund_VU", "RLabund_EN", "RLabund_CR")) {
  if (!col %in% names(bee_data)) bee_data[[col]] <- 0
}

# Add threatened and total richness/abundance
bee_data <- bee_data |>
  mutate(
    RLrich_threatened = RLrich_VU + RLrich_EN + RLrich_CR,
    RLabund_threatened = RLabund_VU + RLabund_EN + RLabund_CR,
    RLrich_total = rowSums(across(starts_with("RLrich_")), na.rm = TRUE),
    RLabund_total = rowSums(across(starts_with("RLabund_")), na.rm = TRUE)
  )

# turn NA into 0 in species matrix
bee_data[is.na(bee_data)] <- 0

glimpse(bee_data)

```

```{r handle aggregates}

# Define species within aggregates
halictus_species <- c("Halictus simplex", "Halictus eurygnathus", "Halictus langobardicus")
bombus_species <- c("Bombus terrestris", "Bombus cryptarum")

# Define aggregate column names
halictus_aggr <- "Halictus simplex aggr."
bombus_aggr <- "Bombus terrestris aggr."

# Loop through rows and print affected ones
for (i in 1:nrow(bee_data)) {
  row <- bee_data[i, ]
  
  halictus_present <- any(row[halictus_species] > 0)
  halictus_aggr_present <- row[[halictus_aggr]] > 0
  
  bombus_present <- any(row[bombus_species] > 0)
  bombus_aggr_present <- row[[bombus_aggr]] > 0
  
  if ((halictus_present & halictus_aggr_present) | (bombus_present & bombus_aggr_present)) {
    cat("\n--- Row", i, "surveyID:", row$surveyID, "---\n")
    
    # Print Halictus species and aggregate if relevant
    if (halictus_present & halictus_aggr_present) {
      cat("Halictus species present:\n")
      print(row[halictus_species])
      cat("Halictus aggregate present:\n")
      print(row[halictus_aggr])
    }
    
    # Print Bombus species and aggregate if relevant
    if (bombus_present & bombus_aggr_present) {
      cat("Bombus species present:\n")
      print(row[bombus_species])
      cat("Bombus aggregate present:\n")
      print(row[bombus_aggr])
    }
  }
}

sum(bee_data$`Halictus simplex aggr.` > 0)

sum(bee_data$`Bombus terrestris aggr.` > 0)


# merge all species within the aggregates 
# Collapse Halictus
bee_data[[halictus_aggr]] <- bee_data[[halictus_aggr]] +
  rowSums(bee_data[halictus_species], na.rm = TRUE)
bee_data[halictus_species] <- NULL   # remove species-level columns

# Collapse Bombus
bee_data[[bombus_aggr]] <- bee_data[[bombus_aggr]] +
  rowSums(bee_data[bombus_species], na.rm = TRUE)
bee_data[bombus_species] <- NULL     # remove species-level columns

```


# Data exploration, descirptive analysis

```{r exploring bee data}
  # Convert to long format
  bee_abundance_long <- bee_abundance_matrix |>
    pivot_longer(cols = -surveyID, names_to = "species", values_to = "abundance")
  
  bee_abundance_long |> filter(abundance > 10)
  
  # Filter for abundance > 15
  bee_abundance_long |> filter(abundance > 15)
  
  # Filter for abundance > 20
  bee_abundance_long |> filter(abundance > 20)
```
should we reduce the abundance of species with >10 collected per survey to 10?

Descriptive analysis plant data
```{r plant data Wilcoxon tests Abundance and Richness}
## Richness 
# Wilcoxon test Richness overall site comparison (all visits combined)
sig_label_richness_all <- case_when(
  is.na(tryCatch(wilcox.test(richness ~ site, data = plant_data)$p.value, error = function(e) NA_real_)) ~ "NA",
  tryCatch(wilcox.test(richness ~ site, data = plant_data)$p.value, error = function(e) NA_real_) < 0.001 ~ "***",
  tryCatch(wilcox.test(richness ~ site, data = plant_data)$p.value, error = function(e) NA_real_) < 0.01 ~ "**",
  tryCatch(wilcox.test(richness ~ site, data = plant_data)$p.value, error = function(e) NA_real_) < 0.05 ~ "*",
  tryCatch(wilcox.test(richness ~ site, data = plant_data)$p.value, error = function(e) NA_real_) < 0.1 ~ "°",
  TRUE ~ "ns"
)

#  Wilcoxon tests for richness per visit
wilcox_results_richness_vis <- plant_data |>
  group_by(visit) |>
  summarise(
    p_value = tryCatch(
      wilcox.test(richness ~ site)$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) |>
  mutate(sig_label = case_when(
    is.na(p_value) ~ "NA",
    p_value < 0.001 ~ "***",
    p_value < 0.01 ~ "**",
    p_value < 0.05 ~ "*",
    p_value < 0.1 ~ "°",
    TRUE ~ "ns"
  ))

# Wilcoxon tests for richness by region and visit
wilcox_results_richness_regvis <- plant_data |>
  group_by(region, visit) |>
  summarise(
    p_value = tryCatch(
      wilcox.test(richness ~ site)$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) |>
  mutate(sig_label = case_when(
    is.na(p_value) ~ "NA",
    p_value < 0.001 ~ "***",
    p_value < 0.01 ~ "**",
    p_value < 0.05 ~ "*",
    p_value < 0.1 ~ "°",
    TRUE ~ "ns"
  ))


## Relative Abundance
# Wilcoxon test for overall site comparison (all visits combined)
sig_label_abundance_all <- case_when(
  is.na(tryCatch(wilcox.test(relabundance ~ site, data = plant_data)$p.value, error = function(e) NA_real_)) ~ "NA",
  tryCatch(wilcox.test(relabundance ~ site, data = plant_data)$p.value, error = function(e) NA_real_) < 0.001 ~ "***",
  tryCatch(wilcox.test(relabundance ~ site, data = plant_data)$p.value, error = function(e) NA_real_) < 0.01 ~ "**",
  tryCatch(wilcox.test(relabundance ~ site, data = plant_data)$p.value, error = function(e) NA_real_) < 0.05 ~ "*",
  tryCatch(wilcox.test(relabundance ~ site, data = plant_data)$p.value, error = function(e) NA_real_) < 0.1 ~ "°",
  TRUE ~ "ns"
)

# Wilcoxon tests for relative abundance per visit
wilcox_results_abundance_vis <- plant_data |>
  group_by(visit) |>
  summarise(
    p_value = tryCatch(
      wilcox.test(relabundance ~ site)$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) |>
  mutate(sig_label = case_when(
    is.na(p_value) ~ "NA",
    p_value < 0.001 ~ "***",
    p_value < 0.01 ~ "**",
    p_value < 0.05 ~ "*",
    p_value < 0.1 ~ "°",
    TRUE ~ "ns"
  ))


# Wilcoxon tests for relative abundance by region and visit
wilcox_results_abundance_regvis <- plant_data |>
  group_by(region, visit) |>
  summarise(
    p_value = tryCatch(
      wilcox.test(relabundance ~ site)$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) |>
  mutate(sig_label = case_when(
    is.na(p_value) ~ "NA",
    p_value < 0.001 ~ "***",
    p_value < 0.01 ~ "**",
    p_value < 0.05 ~ "*",
    p_value < 0.1 ~ "°",
    TRUE ~ "ns"
  ))

```

```{r plant data descriptive plots Abundance and Richness}
### Richness plots

# Common y-axis limits
common_ymax_richness <- max(plant_data$richness, na.rm = TRUE)

# Richness by site type 
ggplot(plant_data, aes(x = site, y = richness, fill = site)) +
  geom_boxplot() +
  labs(title = "Plant Genus Richness per Plot by Site Type",
       x = "Site Type", y = "Number of Genera (Richness)") +
  theme_minimal() +
  annotate("text", x = 1.5, y = common_ymax_richness - 1,
           label = paste("Wilcoxon:", sig_label_richness_all),
           size = 5)

# Richenss by visit and site type
ggplot(plant_data, aes(x = factor(visit), y = richness, fill = site)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(title = "Plant Genus Richness per Plot by Visit and Site Type",
       x = "Visit", y = "Number of Genera (Richness)") +
  theme_minimal() +
  geom_text(data = wilcox_results_richness_vis,
            aes(x = factor(visit), y = common_ymax_richness - 1, label = sig_label),
            inherit.aes = FALSE, size = 5)


# Richness by region, visit, and site type
# Aire
plot_Aire <- ggplot(subset(plant_data, region == "Aire"),
                    aes(x = factor(visit), y = richness, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_richness)) +
  labs(title = "Aire", x = "Visit", y = "Genus Richness per Plot") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal() +
  geom_text(data = subset(wilcox_results_richness_regvis, region == "Aire"),
            aes(x = factor(visit), y = common_ymax_richness - 1, label = sig_label),
            inherit.aes = FALSE, size = 5)

# Scheulte
plot_Scheulte <- ggplot(subset(plant_data, region == "Scheulte"),
                        aes(x = factor(visit), y = richness, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_richness)) +
  labs(title = "Scheulte", x = "Visit", y = "Genus Richness per Plot") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal() +
  geom_text(data = subset(wilcox_results_richness_regvis, region == "Scheulte"),
            aes(x = factor(visit), y = common_ymax_richness - 1, label = sig_label),
            inherit.aes = FALSE, size = 5)

# Thiele
plot_Thiele <- ggplot(subset(plant_data, region == "Thiele"),
                      aes(x = factor(visit), y = richness, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_richness)) +
  labs(title = "Thièle", x = "Visit", y = "Genus Richness per Plot") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal() +
  geom_text(data = subset(wilcox_results_richness_regvis, region == "Thiele"),
            aes(x = factor(visit), y = common_ymax_richness - 1, label = sig_label),
            inherit.aes = FALSE, size = 5)

# Schwarzbach
plot_Schwarzbach <- ggplot(subset(plant_data, region == "Schwarzbach"),
                            aes(x = factor(visit), y = richness, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_richness)) +
  labs(title = "Schwarzbach", x = "Visit", y = "Genus Richness per Plot") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal() +
  geom_text(data = subset(wilcox_results_richness_regvis, region == "Schwarzbach"),
            aes(x = factor(visit), y = common_ymax_richness - 1, label = sig_label),
            inherit.aes = FALSE, size = 5)

# Patchwork combine
(plot_Aire | plot_Scheulte) / (plot_Thiele | plot_Schwarzbach)


### Relative abundance plots
# Common y-axis limits
common_ymax_abundance <- max(plant_data$relabundance, na.rm = TRUE)

# Relative abundance by site type
ggplot(plant_data, aes(x = site, y = relabundance, fill = site)) +
  geom_boxplot() +
  labs(title = "Plant Genus Relative Abundance per Plot by Site Type",
       x = "Site Type", y = "Summed Relative Abundance") +
  theme_minimal() +
  annotate("text", x = 1.5, y = common_ymax_abundance - 1,
           label = paste("Wilcoxon:", sig_label_abundance_all),
           size = 5)

# Relative abundance by visit and site type
ggplot(plant_data, aes(x = factor(visit), y = relabundance, fill = site)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(title = "Plant Genus Relative Abundance per Plot by Visit and Site Type",
       x = "Visit", y = "Summed Relative Abundance") +
  theme_minimal() +
  geom_text(data = wilcox_results_abundance_vis,
            aes(x = factor(visit), y = common_ymax_abundance - 1, label = sig_label),
            inherit.aes = FALSE, size = 5)


# Relative abundance by region, visit, and site type
common_ymax_abundance <- max(plant_data$relabundance, na.rm = TRUE)

# Aire
plot_Aire <- ggplot(subset(plant_data, region == "Aire"),
                    aes(x = factor(visit), y = relabundance, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_abundance)) +
  labs(title = "Aire", x = "Visit", y = "Summed Relative Abundance") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal() +
  geom_text(data = subset(wilcox_results_abundance_regvis, region == "Aire"),
            aes(x = factor(visit), y = common_ymax_abundance - 1, label = sig_label),
            inherit.aes = FALSE, size = 5)

# Scheulte
plot_Scheulte <- ggplot(subset(plant_data, region == "Scheulte"),
                        aes(x = factor(visit), y = relabundance, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_abundance)) +
  labs(title = "Scheulte", x = "Visit", y = "Summed Relative Abundance") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal() +
  geom_text(data = subset(wilcox_results_abundance_regvis, region == "Scheulte"),
            aes(x = factor(visit), y = common_ymax_abundance - 1, label = sig_label),
            inherit.aes = FALSE, size = 5)

# Thiele
plot_Thiele <- ggplot(subset(plant_data, region == "Thiele"),
                      aes(x = factor(visit), y = relabundance, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_abundance)) +
  labs(title = "Thièle", x = "Visit", y = "Summed Relative Abundance") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal() +
  geom_text(data = subset(wilcox_results_abundance_regvis, region == "Thiele"),
            aes(x = factor(visit), y = common_ymax_abundance - 1, label = sig_label),
            inherit.aes = FALSE, size = 5)

# Schwarzbach
plot_Schwarzbach <- ggplot(subset(plant_data, region == "Schwarzbach"),
                            aes(x = factor(visit), y = relabundance, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_abundance)) +
  labs(title = "Schwarzbach", x = "Visit", y = "Summed Relative Abundance") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal() +
  geom_text(data = subset(wilcox_results_abundance_regvis, region == "Schwarzbach"),
            aes(x = factor(visit), y = common_ymax_abundance - 1, label = sig_label),
            inherit.aes = FALSE, size = 5)

# Patchwork combine
(plot_Aire | plot_Scheulte) / (plot_Thiele | plot_Schwarzbach)


```

```{r plant data Wilcoxon tests Hill indices}
# Wilcoxon tests for Hill q1 per region and visit
wilcox_results_hill_q1 <- plant_data |>
  group_by(region, visit) |>
  summarise(
    p_value = tryCatch(
      wilcox.test(hill_q1 ~ site)$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) |>
  mutate(sig_label = case_when(
    is.na(p_value) ~ "NA",
    p_value < 0.001 ~ "***",
    p_value < 0.01 ~ "**",
    p_value < 0.05 ~ "*",
    p_value < 0.1 ~ "°",
    TRUE ~ "ns"
  ))

# Wilcoxon tests for Hill q2 per region and visit
wilcox_results_hill_q2 <- plant_data |>
  group_by(region, visit) |>
  summarise(
    p_value = tryCatch(
      wilcox.test(hill_q2 ~ site)$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) |>
  mutate(sig_label = case_when(
    is.na(p_value) ~ "NA",
    p_value < 0.001 ~ "***",
    p_value < 0.01 ~ "**",
    p_value < 0.05 ~ "*",
    p_value < 0.1 ~ "°",
    TRUE ~ "ns"
  ))
```

```{r plant data descriptive plots Hill indices}
# Common y-axis limits for Hill q1 and q2
common_ymax_q1 <- max(plant_data$hill_q1, na.rm = TRUE)
common_ymax_q2 <- max(plant_data$hill_q2, na.rm = TRUE)

### Hill q1 plots by region
plot_Aire_q1 <- ggplot(subset(plant_data, region == "Aire"),
                       aes(x = factor(visit), y = hill_q1, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_q1)) +
  labs(title = "Aire", x = "Visit", y = "Hill q1 (exp Shannon)") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal() +
  geom_text(data = subset(wilcox_results_hill_q1, region == "Aire"),
            aes(x = factor(visit), y = common_ymax_q1 - 0.5, label = sig_label),
            inherit.aes = FALSE, size = 5)

plot_Scheulte_q1 <- ggplot(subset(plant_data, region == "Scheulte"),
                           aes(x = factor(visit), y = hill_q1, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_q1)) +
  labs(title = "Scheulte", x = "Visit", y = "Hill q1 (exp Shannon)") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal() +
  geom_text(data = subset(wilcox_results_hill_q1, region == "Scheulte"),
            aes(x = factor(visit), y = common_ymax_q1 - 0.5, label = sig_label),
            inherit.aes = FALSE, size = 5)

plot_Thiele_q1 <- ggplot(subset(plant_data, region == "Thiele"),
                         aes(x = factor(visit), y = hill_q1, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_q1)) +
  labs(title = "Thièle", x = "Visit", y = "Hill q1 (exp Shannon)") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal() +
  geom_text(data = subset(wilcox_results_hill_q1, region == "Thiele"),
            aes(x = factor(visit), y = common_ymax_q1 - 0.5, label = sig_label),
            inherit.aes = FALSE, size = 5)

plot_Schwarzbach_q1 <- ggplot(subset(plant_data, region == "Schwarzbach"),
                              aes(x = factor(visit), y = hill_q1, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_q1)) +
  labs(title = "Schwarzbach", x = "Visit", y = "Hill q1 (exp Shannon)") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal() +
  geom_text(data = subset(wilcox_results_hill_q1, region == "Schwarzbach"),
            aes(x = factor(visit), y = common_ymax_q1 - 0.5, label = sig_label),
            inherit.aes = FALSE, size = 5)

hill_q1_patchwork <- (plot_Aire_q1 | plot_Scheulte_q1) / (plot_Thiele_q1 | plot_Schwarzbach_q1)

### Hill q2 plots by region
plot_Aire_q2 <- ggplot(subset(plant_data, region == "Aire"),
                       aes(x = factor(visit), y = hill_q2, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_q2)) +
  labs(title = "Aire", x = "Visit", y = "Hill q2 (Inverse Simpson)") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal() +
  geom_text(data = subset(wilcox_results_hill_q2, region == "Aire"),
            aes(x = factor(visit), y = common_ymax_q2 - 0.5, label = sig_label),
            inherit.aes = FALSE, size = 5)

plot_Scheulte_q2 <- ggplot(subset(plant_data, region == "Scheulte"),
                       aes(x = factor(visit), y = hill_q2, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_q2)) +
  labs(title = "Aire", x = "Visit", y = "Hill q2 (Inverse Simpson)") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal() +
  geom_text(data = subset(wilcox_results_hill_q2, region == "Scheulte"),
            aes(x = factor(visit), y = common_ymax_q2 - 0.5, label = sig_label),
            inherit.aes = FALSE, size = 5)

plot_Thiele_q2 <- ggplot(subset(plant_data, region == "Thiele"),
                       aes(x = factor(visit), y = hill_q2, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_q2)) +
  labs(title = "Aire", x = "Visit", y = "Hill q2 (Inverse Simpson)") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal() +
  geom_text(data = subset(wilcox_results_hill_q2, region == "Thiele"),
            aes(x = factor(visit), y = common_ymax_q2 - 0.5, label = sig_label),
            inherit.aes = FALSE, size = 5)

plot_Schwarzbach_q2 <- ggplot(subset(plant_data, region == "Schwarzbach"),
                       aes(x = factor(visit), y = hill_q2, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_q2)) +
  labs(title = "Aire", x = "Visit", y = "Hill q2 (Inverse Simpson)") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal() +
  geom_text(data = subset(wilcox_results_hill_q2, region == "Schwarzbach"),
            aes(x = factor(visit), y = common_ymax_q2 - 0.5, label = sig_label),
            inherit.aes = FALSE, size = 5)

# Repeat same structure for Scheulte, Thiele, Schwarzbach
# Combine into patchwork
hill_q2_patchwork <- (plot_Aire_q2 | plot_Scheulte_q2) / (plot_Thiele_q2 | plot_Schwarzbach_q2)

# Display both patchworks
hill_q1_patchwork
hill_q2_patchwork

```

Descriptive analysis habitat mapping
```{r mappingdata patchiness plot}

safe_wilcox <- function(df) {
  # Expect columns: patchiness, site
  site_levels <- unique(df$site)
  if (length(site_levels) != 2) return(NA_real_)
  # At least one non-NA per group
  gcounts <- df |>
    filter(!is.na(patchiness)) |>
    count(site) |> pull(n)
  if (length(gcounts) != 2 || any(gcounts == 0)) return(NA_real_)
  tryCatch(
    stats::wilcox.test(patchiness ~ site, data = df)$p.value,
    error = function(e) NA_real_
  )
}

# Compute p-values & labels
wilcox_results_patchiness <- mappingdata |>
  group_by(region, visit) |>
  summarise(
    p_value = safe_wilcox(cur_data()),
    .groups = "drop"
  ) |>
  mutate(sig_label = case_when(
    is.na(p_value) ~ "NA",
    p_value < 0.001 ~ "***",
    p_value < 0.01 ~ "**",
    p_value < 0.05 ~ "*",
    p_value < 0.1 ~ "°",
    TRUE ~ "ns"
  ))

# Common ymax
common_ymax_patch <- max(mappingdata$patchiness, na.rm = TRUE)
ylab_patch <- "Patchiness"

# Region-specific plotting function
plot_region_patch <- function(reg) {
  ggplot(subset(mappingdata, region == reg),
         aes(x = factor(visit), y = patchiness, fill = site)) +
    geom_boxplot(position = position_dodge()) +
    coord_cartesian(ylim = c(min(mappingdata$patchiness, na.rm = TRUE), common_ymax_patch)) +
    labs(title = reg, x = "Visit", y = ylab_patch) +
    scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
    theme_minimal() +
    geom_text(
      data = subset(wilcox_results_patchiness, region == reg),
      aes(x = factor(visit), y = common_ymax_patch - 0.05 * (common_ymax_patch - min(mappingdata$patchiness, na.rm = TRUE)), 
          label = sig_label),
      inherit.aes = FALSE, size = 5
    )
}

regions_to_plot <- c("Aire", "Scheulte", "Thiele", "Schwarzbach")
plots <- map(regions_to_plot, plot_region_patch)

patchiness_patchwork <- (plots[[1]] | plots[[2]]) / (plots[[3]] | plots[[4]])
patchiness_patchwork

```



Descriptive analysis bee data
```{r bee data overview numbers}
# total species count
length(which(names(bee_data) == "Andrena aeneiventris"): which(names(bee_data) == "Stelis punctulatissima"))



# Derive species column block by names (robust to column reordering)
start <- match("Andrena aeneiventris", names(bee_data))
end   <- match("Stelis punctulatissima", names(bee_data))
species_cols <- start:end

# Logical presence (>=1) per site type
control_pres <- colSums(bee_data[bee_data$site == "control", species_cols] > 0, na.rm = TRUE) > 0
focal_pres   <- colSums(bee_data[bee_data$site == "focal",   species_cols] > 0, na.rm = TRUE) > 0

# Number of Bee species by site
sum(control_pres)
sum(focal_pres)

```

```{r bee data wilcoxon tests}

# Bee species richness - run Wilcoxon test across all visits and regions
wilcox_beeri_all <- wilcox.test(bee_richness ~ site, data = bee_data)

sig_label_beeri_all <- case_when(
  wilcox_beeri_all$p.value < 0.001 ~ "***",
  wilcox_beeri_all$p.value < 0.01 ~ "**",
  wilcox_beeri_all$p.value < 0.05 ~ "*",
  TRUE ~ "ns"
)

# Bee species richness - run Wilcoxon test per visit across all regions
wilcox_beeri_by_visit <- bee_data |>
  group_by(visit) |>
  summarise(
    p_value = tryCatch(wilcox.test(bee_richness ~ site)$p.value, error = function(e) NA_real_),
    sig_label = case_when(
      is.na(p_value) ~ "NA",
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      p_value < 0.1 ~ ".",
      TRUE ~ "ns"
    ),
    label = paste0(sig_label, "\n(p = ", round(p_value, 3), ")"),
    .groups = "drop"
  )

# Bee species richness - run Wilcoxon tests for each region and visit
wilcox_results_beeri <- bee_data |>
  group_by(region, visit) |>
  summarise(
    n_focal = sum(site == "focal"),
    n_control = sum(site == "control"),
    p_value = if (n_focal >= 1 & n_control >= 1) {
      tryCatch(
        wilcox.test(bee_richness ~ site)$p.value,
        error = function(e) NA_real_
      )
    } else {
      NA_real_
    },
    .groups = "drop"
  )

wilcox_results_beeri <- wilcox_results_beeri |>
  mutate(sig_label = case_when(
    is.na(p_value) ~ "NA",
    p_value < 0.001 ~ "***",
    p_value < 0.01 ~ "**",
    p_value < 0.05 ~ "*",
    p_value < 0.1 ~ ".",
    TRUE ~ "ns"
  ))

# Bee abundance - run Wilcoxon test across all visits and regions
wilcox_beeab_all <- wilcox.test(bee_abundance ~ site, data = bee_data)

sig_label_beeab_all <- case_when(
  wilcox_beeab_all$p.value < 0.001 ~ "***",
  wilcox_beeab_all$p.value < 0.01 ~ "**",
  wilcox_beeab_all$p.value < 0.05 ~ "*",
  TRUE ~ "ns"
)

# Bee abundance - run Wilcoxon test per visit across all regions
wilcox_beeab_by_visit <- bee_data |>
  group_by(visit) |>
  summarise(
    p_value = tryCatch(wilcox.test(bee_abundance ~ site)$p.value, error = function(e) NA_real_),
    sig_label = case_when(
      is.na(p_value) ~ "NA",
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      p_value < 0.1 ~ ".",
      TRUE ~ "ns"
    ),
    label = paste0(sig_label, "\n(p = ", round(p_value, 3), ")"),
    .groups = "drop"
  )

# Bee abudnance - run Wilcoxon tests for each region and visit
wilcox_results_beeab <- bee_data |>
  group_by(region, visit) |>
  summarise(
    n_focal = sum(site == "focal"),
    n_control = sum(site == "control"),
    p_value = if (n_focal >= 1 & n_control >= 1) {
      tryCatch(
        wilcox.test(bee_abundance ~ site)$p.value,
        error = function(e) NA_real_
      )
    } else {
      NA_real_
    },
    .groups = "drop"
  )

wilcox_results_beeab <- wilcox_results_beeab |>
  mutate(sig_label = case_when(
    is.na(p_value) ~ "NA",
    p_value < 0.001 ~ "***",
    p_value < 0.01 ~ "**",
    p_value < 0.05 ~ "*",
    p_value < 0.1 ~ ".",
    TRUE ~ "ns"
  ))
```

```{r bee data descriptive plots}
                   
#boxplot bee species richness by site type and visit
ggplot(bee_data, aes(x = factor(visit), y = bee_richness, fill = site)) +
  geom_boxplot() +
  labs(title = "Bee Species Richness per Plot by Visit and Site Type",
       x = "Visit",
       y = "Number of Bee Species (Richness)") +
  theme_minimal() +
  geom_text(data = wilcox_beeri_by_visit,
            aes(x = factor(visit), y = max(bee_data$bee_richness, na.rm = TRUE) - 1, label = sig_label),
            inherit.aes = FALSE,
            size = 5)

#boxplot bee species richness by site type (all visits combined)
ggplot(bee_data, aes(x = site, y = bee_richness, fill = site)) +
  geom_boxplot() +
  labs(title = "Bee Species Richness per Plot by Site Type",
       x = "Site Type",
       y = "Number of Bee Species (Richness)") +
  theme_minimal() +
  annotate("text", x = 1.5, y = max(bee_data$bee_richness, na.rm = TRUE) - 1,
           label = paste("Wilcoxon:", sig_label_beeri_all),
           size = 5)

## create patchwork plots for each region by site
common_ymax <- max(bee_data$bee_richness, na.rm = TRUE)

# Aire
plot_Aire <- ggplot(subset(bee_data, region == "Aire"),aes(x = factor(visit), y = bee_richness, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax)) +
  labs(title = "Aire", x = "Visit", y = "Bee Species Richness") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal()

aire_labels <- wilcox_results_beeri |> filter(region == "Aire") # Filter and add wilcox labels for Aire
plot_Aire <- plot_Aire +
  geom_text(data = aire_labels,
            aes(x = factor(visit), y = common_ymax - 1, label = sig_label),
            inherit.aes = FALSE,
            size = 5)

# Scheulte
plot_Scheulte <- ggplot(subset(bee_data, region == "Scheulte"),
                        aes(x = factor(visit), y = bee_richness, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax)) +
  labs(title = "Scheulte", x = "Visit", y = "Bee Species Richness") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal()

scheulte_labels <- wilcox_results_beeri |> filter(region == "Scheulte")
plot_Scheulte <- plot_Scheulte +
  geom_text(data = scheulte_labels,
            aes(x = factor(visit), y = common_ymax - 1, label = sig_label),
            inherit.aes = FALSE,
            size = 5)

# Thiele
plot_Thiele <- ggplot(subset(bee_data, region == "Thiele"),
                      aes(x = factor(visit), y = bee_richness, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax)) +
  labs(title = "Thièle", x = "Visit", y = "Bee Species Richness") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal()

thiele_labels <- wilcox_results_beeri |> filter(region == "Thiele")
plot_Thiele <- plot_Thiele +
  geom_text(data = thiele_labels,
            aes(x = factor(visit), y = common_ymax - 1, label = sig_label),
            inherit.aes = FALSE,
            size = 5)

# Schwarzbach
plot_Schwarzbach <- ggplot(subset(bee_data, region == "Schwarzbach"),
                           aes(x = factor(visit), y = bee_richness, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax)) +
  labs(title = "Schwarzbach", x = "Visit", y = "Bee Species Richness") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal()

schwarzbach_labels <- wilcox_results_beeri |> filter(region == "Schwarzbach")
plot_Schwarzbach <- plot_Schwarzbach +
  geom_text(data = schwarzbach_labels,
            aes(x = factor(visit), y = common_ymax - 1, label = sig_label),
            inherit.aes = FALSE,
            size = 5)

(plot_Aire | plot_Scheulte) / (plot_Thiele | plot_Schwarzbach)

#Bee abundance 
common_ymax_abund <- max(bee_data$bee_abundance, na.rm = TRUE)

ggplot(bee_data, aes(x = factor(visit), y = bee_abundance, fill = site)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(title = "Total Bee Abundance per Plot by Visit and Site Type",
       x = "Visit",
       y = "Total Bee Abundance",
       fill = "Site Type") +
  theme_minimal() +
  geom_text(data = wilcox_beeab_by_visit,
            aes(x = factor(visit), y = max(bee_data$bee_abundance, na.rm = TRUE) - 1, label = sig_label),
            inherit.aes = FALSE,
            size = 5)


ggplot(bee_data, aes(x = site, y = bee_abundance, fill = site)) +
  geom_boxplot() +
  labs(title = "Total Bee Abundance per Plot by Site Type",
       x = "Site Type",
       y = "Total Bee Abundance") +
  theme_minimal() +
  annotate("text", x = 1.5, y = max(bee_data$bee_abundance, na.rm = TRUE) - 1,
           label = paste("Wilcoxon:", sig_label_beeab_all),
           size = 5)

common_ymax_abund <- max(bee_data$bee_abundance, na.rm = TRUE)

# Aire
plot_Aire <- ggplot(subset(bee_data, region == "Aire"),
                    aes(x = factor(visit), y = bee_abundance, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_abund)) +
  labs(title = "Aire", x = "Visit", y = "Total Bee Abundance") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal()

aire_labels_abund <- wilcox_results_beeab |> filter(region == "Aire")
plot_Aire <- plot_Aire +
  geom_text(data = aire_labels_abund,
            aes(x = factor(visit), y = common_ymax_abund - 1, label = sig_label),
            inherit.aes = FALSE,
            size = 5)

# Scheulte
plot_Scheulte <- ggplot(subset(bee_data, region == "Scheulte"),
                        aes(x = factor(visit), y = bee_abundance, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_abund)) +
  labs(title = "Scheulte", x = "Visit", y = "Total Bee Abundance") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal()

scheulte_labels_abund <- wilcox_results_beeab |> filter(region == "Scheulte")
plot_Scheulte <- plot_Scheulte +
  geom_text(data = scheulte_labels_abund,
            aes(x = factor(visit), y = common_ymax_abund - 1, label = sig_label),
            inherit.aes = FALSE,
            size = 5)

# Thiele
plot_Thiele <- ggplot(subset(bee_data, region == "Thiele"),
                      aes(x = factor(visit), y = bee_abundance, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_abund)) +
  labs(title = "Thièle", x = "Visit", y = "Total Bee Abundance") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal()

thiele_labels_abund <- wilcox_results_beeab |> filter(region == "Thiele")
plot_Thiele <- plot_Thiele +
  geom_text(data = thiele_labels_abund,
            aes(x = factor(visit), y = common_ymax_abund - 1, label = sig_label),
            inherit.aes = FALSE,
            size = 5)

# Schwarzbach
plot_Schwarzbach <- ggplot(subset(bee_data, region == "Schwarzbach"),
                           aes(x = factor(visit), y = bee_abundance, fill = site)) +
  geom_boxplot(position = position_dodge()) +
  coord_cartesian(ylim = c(0, common_ymax_abund)) +
  labs(title = "Schwarzbach", x = "Visit", y = "Total Bee Abundance") +
  scale_fill_manual(values = c("focal" = "#1b9e77", "control" = "#d95f02")) +
  theme_minimal()

schwarzbach_labels_abund <- wilcox_results_beeab |> filter(region == "Schwarzbach")
plot_Schwarzbach <- plot_Schwarzbach +
  geom_text(data = schwarzbach_labels_abund,
            aes(x = factor(visit), y = common_ymax_abund - 1, label = sig_label),
            inherit.aes = FALSE,
            size = 5)

# Combine with patchwork
(plot_Aire | plot_Scheulte) / (plot_Thiele | plot_Schwarzbach)

```

```{r export for BAFU report}

#boxplot bee species richness by site type (all visits combined)
ggplot(bee_data, aes(x = site, y = bee_richness, fill = site)) +
  geom_boxplot() +
  labs(title = "Bee Species Richness per Plot by Site Type",
       x = "Site Type",
       y = "Number of Bee Species (Richness)") +
  theme_minimal() +
  annotate("text", x = 1.5, y = max(bee_data$bee_richness, na.rm = TRUE) - 1,
           label = paste(sig_label_beeri_all),
           size = 5)

#boxplot bee abundance by site type (all visits combined)
ggplot(bee_data, aes(x = site, y = bee_abundance, fill = site)) +
  geom_boxplot() +
  labs(title = "Total Bee Abundance per Plot by Site Type",
       x = "Site Type",
       y = "Total Bee Abundance") +
  theme_minimal() +
  annotate("text", x = 1.5, y = max(bee_data$bee_abundance, na.rm = TRUE) - 1,
           label = paste(sig_label_beeab_all),
           size = 5)





ggplot(bee_data, aes(x = site, y = bee_richness, fill = site)) +
  geom_boxplot(color = "black") +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 2,
               fill = "white", color = "black", stroke = 0.8,
               position = position_dodge(width = 0.75)) +
  labs(title = "Richesse spécifique des abeilles par site",
       x = "",
       y = "Richesse spécifique par placette") +
  scale_x_discrete(labels = c("focal" = "Revitalisé", "control" = "Non-Revitalisé")) +
  scale_fill_manual(values = c("focal" = "#8c8c8c", "control" = "#66c2a5")) +
  theme_classic() +
  theme(axis.text.x = element_text(size = 16, margin = margin(t = 5)),
        axis.text.y = element_text(size = 16, margin = margin(r = 5)),
        axis.title.y = element_text(size = 21, margin = margin(r = 15)),
        axis.title.x = element_text(size = 21),
        axis.ticks.length = unit(0.25, "cm"),
        plot.margin = margin(t = 20, r = 20, l = 5, b = 5),
        plot.title = element_text(size = 21, face = "bold", hjust = 0.5, vjust = 4),
        legend.position = "none",
        panel.background = element_rect(fill = "white", color = NA),
        panel.grid = element_blank()) +
  coord_cartesian(ylim = c(0, max(bee_data$bee_richness, na.rm = TRUE) + 2)) +
  annotate("text", x = 1.5, y = max(bee_data$bee_richness, na.rm = TRUE) + 1,
           label = paste(sig_label_beeri_all),
           size = 6, vjust = 1, hjust = 0.5)

# ggsave("bee_richness.jpg",
#        width  = 5333, height = 4105, units = "px", dpi = 800,
#        plot = ggplot(bee_data, aes(x = site, y = bee_richness, fill = site)) +
#          geom_boxplot(color = "black") +
#          stat_summary(fun = mean, geom = "point", shape = 21, size = 2,
#                       fill = "white", color = "black", stroke = 0.8,
#                       position = position_dodge(width = 0.75)) +
#          labs(title = "Richesse spécifique des abeilles par site",
#               x = "",
#               y = "Richesse spécifique par placette") +
#          scale_x_discrete(labels = c("focal" = "Revitalisé", "control" = "Non-Revitalisé")) +
#          scale_fill_manual(values = c("focal" = "#66c2a5", "control" = "#8c8c8c")) +
#          theme_classic() +
#          theme(axis.text.x = element_text(size = 16, margin = margin(t = 5)),
#                axis.text.y = element_text(size = 16, margin = margin(r = 5)),
#                axis.title.y = element_text(size = 21, margin = margin(r = 15)),
#                axis.title.x = element_text(size = 21),
#                axis.ticks.length = unit(0.25, "cm"),
#                plot.margin = margin(t = 20, r = 20, l = 5, b = 5),
#                plot.title = element_text(size = 21, face = "bold", hjust = 0.5, vjust = 4),
#                legend.position = "none",
#                panel.background = element_rect(fill = "white", color = NA),
#                panel.grid = element_blank()) +
#          coord_cartesian(ylim = c(0, max(bee_data$bee_richness, na.rm = TRUE) + 2)) +
#          annotate("text", x = 1.5, y = max(bee_data$bee_richness, na.rm = TRUE) + 1,
#                   label = paste(sig_label_beeri_all),
#                   size = 6, vjust = 1, hjust = 0.5))



ggplot(bee_data, aes(x = site, y = bee_abundance, fill = site)) +
    geom_boxplot(color = "black") +
    stat_summary(fun = mean, geom = "point", shape = 21, size = 2,
                 fill = "white", color = "black", stroke = 0.8,
                 position = position_dodge(width = 0.75)) +
    labs(title = "Abondance des abeilles par site",
         x = "",
         y = "Abondance par placette") +
    scale_x_discrete(labels = c("focal" = "Revitalisé", "control" = "Non-Revitalisé")) +
    scale_fill_manual(values = c("focal" = "#66c2a5", "control" = "#8c8c8c")) + # Revitalisé = green, Chenalisé = gray
    theme_classic() +
    theme(axis.text.x = element_text(size = 16, margin = margin(t = 5)),
          axis.text.y = element_text(size = 16, margin = margin(r = 5)),
          axis.title.y = element_text(size = 21, margin = margin(r = 15)),
          axis.title.x = element_text(size = 21),
          axis.ticks.length = unit(0.25, "cm"),
          plot.margin = margin(t = 20, r = 20, l = 5, b = 5),
          plot.title = element_text(size = 21, face = "bold", hjust = 0.5, vjust = 4),
          legend.position = "none",
          panel.background = element_rect(fill = "white", color = NA),
          panel.grid = element_blank()) +
    coord_cartesian(ylim = c(0, max(bee_data$bee_abundance, na.rm = TRUE) + 10)) +
    annotate("text", x = 1.5, y = max(bee_data$bee_abundance, na.rm = TRUE) + 5,
             label = paste(sig_label_beeab_all),
             size = 6, vjust = 1, hjust = 0.5)

# 
# ggsave(
#   filename = "bee_abundance.jpg",
#   width = 5333, height = 4105, units = "px", dpi = 800,
#   plot = ggplot(bee_data, aes(x = site, y = bee_abundance, fill = site)) +
#     geom_boxplot(color = "black") +
#     stat_summary(fun = mean, geom = "point", shape = 21, size = 2,
#                  fill = "white", color = "black", stroke = 0.8,
#                  position = position_dodge(width = 0.75)) +
#     labs(title = "Abondance des abeilles par site",
#          x = "",
#          y = "Abondance par placette") +
#     scale_x_discrete(labels = c("focal" = "Revitalisé", "control" = "Non-Revitalisé")) +
#     scale_fill_manual(values = c("focal" = "#66c2a5", "control" = "#8c8c8c")) + # Revitalisé = green, Chenalisé = gray
#     theme_classic() +
#     theme(axis.text.x = element_text(size = 16, margin = margin(t = 5)),
#           axis.text.y = element_text(size = 16, margin = margin(r = 5)),
#           axis.title.y = element_text(size = 21, margin = margin(r = 15)),
#           axis.title.x = element_text(size = 21),
#           axis.ticks.length = unit(0.25, "cm"),
#           plot.margin = margin(t = 20, r = 20, l = 5, b = 5),
#           plot.title = element_text(size = 21, face = "bold", hjust = 0.5, vjust = 4),
#           legend.position = "none",
#           panel.background = element_rect(fill = "white", color = NA),
#           panel.grid = element_blank()) +
#     coord_cartesian(ylim = c(0, max(bee_data$bee_abundance, na.rm = TRUE) + 10)) +
#     annotate("text", x = 1.5, y = max(bee_data$bee_abundance, na.rm = TRUE) + 5,
#              label = paste(sig_label_beeab_all),
#              size = 6, vjust = 1, hjust = 0.5)
# )

#and in german

ggsave(
  filename = "bee_abundance_de.jpg",
  width = 5333, height = 4105, units = "px", dpi = 800,
  plot = ggplot(bee_data, aes(x = site, y = bee_abundance, fill = site)) +
    geom_boxplot(color = "black") +
    stat_summary(fun = mean, geom = "point", shape = 21, size = 2,
                 fill = "white", color = "black", stroke = 0.8,
                 position = position_dodge(width = 0.75)) +
    labs(title = "Bienenabundanz pro Standort",
         x = "",
         y = "Abundanz pro Probefläche") +
    scale_x_discrete(labels = c("focal" = "Renaturiert", "control" = "Kontrollabschnitt")) +
    scale_fill_manual(values = c("focal" = "#66c2a5", "control" = "#8c8c8c")) + # Revitalisé = green, Chenalisé = gray
    theme_classic() +
    theme(axis.text.x = element_text(size = 16, margin = margin(t = 5)),
          axis.text.y = element_text(size = 16, margin = margin(r = 5)),
          axis.title.y = element_text(size = 21, margin = margin(r = 15)),
          axis.title.x = element_text(size = 21),
          axis.ticks.length = unit(0.25, "cm"),
          plot.margin = margin(t = 20, r = 20, l = 5, b = 5),
          plot.title = element_text(size = 21, face = "bold", hjust = 0.5, vjust = 4),
          legend.position = "none",
          panel.background = element_rect(fill = "white", color = NA),
          panel.grid = element_blank()) +
    coord_cartesian(ylim = c(0, max(bee_data$bee_abundance, na.rm = TRUE) + 10)) +
    annotate("text", x = 1.5, y = max(bee_data$bee_abundance, na.rm = TRUE) + 5,
             label = paste(sig_label_beeab_all),
             size = 6, vjust = 1, hjust = 0.5)
)

ggsave("bee_richness_de.jpg",
       width  = 5333, height = 4105, units = "px", dpi = 800,
       plot = ggplot(bee_data, aes(x = site, y = bee_richness, fill = site)) +
         geom_boxplot(color = "black") +
         stat_summary(fun = mean, geom = "point", shape = 21, size = 2,
                      fill = "white", color = "black", stroke = 0.8,
                      position = position_dodge(width = 0.75)) +
         labs(title = "Artenvielfalt der Bienen pro Standort",
              x = "",
              y = "Artenvielfalt pro Probefläche") +
         scale_x_discrete(labels = c("focal" = "Renaturiert", "control" = "Kontrollabschnitt")) +
         scale_fill_manual(values = c("focal" = "#66c2a5", "control" = "#8c8c8c")) +
         theme_classic() +
         theme(axis.text.x = element_text(size = 16, margin = margin(t = 5)),
               axis.text.y = element_text(size = 16, margin = margin(r = 5)),
               axis.title.y = element_text(size = 21, margin = margin(r = 15)),
               axis.title.x = element_text(size = 21),
               axis.ticks.length = unit(0.25, "cm"),
               plot.margin = margin(t = 20, r = 20, l = 5, b = 5),
               plot.title = element_text(size = 21, face = "bold", hjust = 0.5, vjust = 4),
               legend.position = "none",
               panel.background = element_rect(fill = "white", color = NA),
               panel.grid = element_blank()) +
         coord_cartesian(ylim = c(0, max(bee_data$bee_richness, na.rm = TRUE) + 2)) +
         annotate("text", x = 1.5, y = max(bee_data$bee_richness, na.rm = TRUE) + 1,
                  label = paste(sig_label_beeri_all),
                  size = 6, vjust = 1, hjust = 0.5))


```



```{r bee Red List Wilcoxon & plots}
# Wilcoxon test across all visits & sites (paired)
wilcox_rlrich_all <- wilcox.test(
  bee_data$RLrich_threatened,
  bee_data$RLrich_total - bee_data$RLrich_threatened,
  paired = TRUE
)

sig_label_rlrich_all <- case_when(
  wilcox_rlrich_all$p.value < 0.001 ~ "***",
  wilcox_rlrich_all$p.value < 0.01 ~ "**",
  wilcox_rlrich_all$p.value < 0.05 ~ "*",
  wilcox_rlrich_all$p.value < 0.1 ~ ".",
  TRUE ~ "ns"
)

# Prepare long format for plotting
rl_plotdata_rich <- bee_data |>
  mutate(RLrich_non_threatened = RLrich_total - RLrich_threatened) |>
  select(surveyID, site, region, visit, RLrich_threatened, RLrich_non_threatened) |>
  pivot_longer(cols = starts_with("RLrich_"),
               names_to = "RL_category",
               values_to = "value")

# Plot RL richness
ggplot(rl_plotdata_rich, aes(x = factor(visit), y = value, fill = RL_category)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(~site) +
  labs(
    title = "Red List Bee Richness (Threatened vs Non-threatened)",
    x = "Visit",
    y = "Number of Species",
    fill = "Category"
  ) +
  scale_fill_manual(values = c(
    "RLrich_threatened" = "#d95f02",
    "RLrich_non_threatened" = "#1b9e77"
  )) +
  theme_minimal() +
  annotate("text",
           x = 1.5, y = max(rl_plotdata_rich$value, na.rm = TRUE) - 1,
           label = paste("Wilcoxon:", sig_label_rlrich_all),
           size = 5)


# RL Abundance: Threatened vs Non-threatened
wilcox_rlabund_all <- wilcox.test(
  bee_data$RLabund_threatened,
  bee_data$RLabund_total - bee_data$RLabund_threatened,
  paired = TRUE
)

sig_label_rlabund_all <- case_when(
  wilcox_rlabund_all$p.value < 0.001 ~ "***",
  wilcox_rlabund_all$p.value < 0.01 ~ "**",
  wilcox_rlabund_all$p.value < 0.05 ~ "*",
  wilcox_rlabund_all$p.value < 0.1 ~ ".",
  TRUE ~ "ns"
)

# Prepare long format for plotting
rl_plotdata_abund <- bee_data |>
  mutate(RLabund_non_threatened = RLabund_total - RLabund_threatened) |>
  select(surveyID, site, region, visit, RLabund_threatened, RLabund_non_threatened) |>
  pivot_longer(cols = starts_with("RLabund_"),
               names_to = "RL_category",
               values_to = "value")

# Plot RL abundance
ggplot(rl_plotdata_abund, aes(x = factor(visit), y = value, fill = RL_category)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(~site) +
  labs(
    title = "Red List Bee Abundance (Threatened vs Non-threatened)",
    x = "Visit",
    y = "Number of Individuals",
    fill = "Category"
  ) +
  scale_fill_manual(values = c(
    "RLabund_threatened" = "#d95f02",
    "RLabund_non_threatened" = "#1b9e77"
  )) +
  theme_minimal() +
  annotate("text",
           x = 1.5, y = max(rl_plotdata_abund$value, na.rm = TRUE) - 1,
           label = paste("Wilcoxon:", sig_label_rlabund_all),
           size = 5)

# RL Richness per region & visit
wilcox_results_rlrich <- bee_data |>
  mutate(RLrich_non_threatened = RLrich_total - RLrich_threatened) |>
  group_by(region, visit) |>
  summarise(
    p_value = tryCatch(
      wilcox.test(RLrich_threatened, RLrich_non_threatened, paired = TRUE)$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) |>
  mutate(sig_label = case_when(
    is.na(p_value) ~ "NA",
    p_value < 0.001 ~ "***",
    p_value < 0.01 ~ "**",
    p_value < 0.05 ~ "*",
    p_value < 0.1 ~ ".",
    TRUE ~ "ns"
  ))

rl_plotdata_rich <- bee_data |>
  mutate(RLrich_non_threatened = RLrich_total - RLrich_threatened) |>
  select(surveyID, site, region, visit, RLrich_threatened, RLrich_non_threatened) |>
  pivot_longer(cols = starts_with("RLrich_"),
               names_to = "RL_category",
               values_to = "value")

common_ymax_rlrich <- max(rl_plotdata_rich$value, na.rm = TRUE)

plot_rl_rich <- ggplot(rl_plotdata_rich, aes(x = factor(visit), y = value, fill = RL_category)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(~region) +
  coord_cartesian(ylim = c(0, common_ymax_rlrich)) +
  labs(title = "RL Richness (Threatened vs Non-threatened)",
       x = "Visit", y = "Number of Species", fill = "Category") +
  scale_fill_manual(values = c("RLrich_threatened" = "#d95f02",
                               "RLrich_non_threatened" = "#1b9e77")) +
  theme_minimal() +
  geom_text(data = wilcox_results_rlrich,
            aes(x = factor(visit), y = common_ymax_rlrich - 1, label = sig_label),
            inherit.aes = FALSE, size = 5)



# RL Abundance per region & visit

wilcox_results_rlabund <- bee_data |>
  mutate(RLabund_non_threatened = RLabund_total - RLabund_threatened) |>
  group_by(region, visit) |>
  summarise(
    p_value = tryCatch(
      wilcox.test(RLabund_threatened, RLabund_non_threatened, paired = TRUE)$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) |>
  mutate(sig_label = case_when(
    is.na(p_value) ~ "NA",
    p_value < 0.001 ~ "***",
    p_value < 0.01 ~ "**",
    p_value < 0.05 ~ "*",
    p_value < 0.1 ~ ".",
    TRUE ~ "ns"
  ))

rl_plotdata_abund <- bee_data |>
  mutate(RLabund_non_threatened = RLabund_total - RLabund_threatened) |>
  select(surveyID, site, region, visit, RLabund_threatened, RLabund_non_threatened) |>
  pivot_longer(cols = starts_with("RLabund_"),
               names_to = "RL_category",
               values_to = "value")

common_ymax_rlabund <- max(rl_plotdata_abund$value, na.rm = TRUE)

plot_rl_abund <- ggplot(rl_plotdata_abund, aes(x = factor(visit), y = value, fill = RL_category)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(~region) +
  coord_cartesian(ylim = c(0, common_ymax_rlabund)) +
  labs(title = "RL Abundance (Threatened vs Non-threatened)",
       x = "Visit", y = "Number of Individuals", fill = "Category") +
  scale_fill_manual(values = c("RLabund_threatened" = "#d95f02",
                               "RLabund_non_threatened" = "#1b9e77")) +
  theme_minimal() +
  geom_text(data = wilcox_results_rlabund,
            aes(x = factor(visit), y = common_ymax_rlabund - 1, label = sig_label),
            inherit.aes = FALSE, size = 5)



# Combine with patchwork
plot_rl_rich / plot_rl_abund


# RL Richness per region & visit

wilcox_results_rlrich <- bee_data |>
  mutate(RLrich_non_threatened = RLrich_total - RLrich_threatened) |>
  group_by(region, visit) |>
  summarise(
    p_value = tryCatch(
      wilcox.test(RLrich_threatened, RLrich_non_threatened, paired = TRUE)$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) |>
  mutate(sig_label = case_when(
    is.na(p_value) ~ "NA",
    p_value < 0.001 ~ "***",
    p_value < 0.01 ~ "**",
    p_value < 0.05 ~ "*",
    p_value < 0.1 ~ ".",
    TRUE ~ "ns"
  ))

rl_plotdata_rich <- bee_data |>
  mutate(RLrich_non_threatened = RLrich_total - RLrich_threatened) |>
  select(surveyID, site, region, visit, RLrich_threatened, RLrich_non_threatened) |>
  pivot_longer(cols = starts_with("RLrich_"),
               names_to = "RL_category",
               values_to = "value")

common_ymax_rlrich <- max(rl_plotdata_rich$value, na.rm = TRUE)

plot_rl_rich <- ggplot(rl_plotdata_rich, aes(x = factor(visit), y = value, fill = RL_category)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(~region) +
  coord_cartesian(ylim = c(0, common_ymax_rlrich)) +
  labs(title = "RL Richness (Threatened vs Non-threatened)",
       x = "Visit", y = "Number of Species", fill = "Category") +
  scale_fill_manual(values = c("RLrich_threatened" = "#d95f02",
                               "RLrich_non_threatened" = "#1b9e77")) +
  theme_minimal() +
  geom_text(data = wilcox_results_rlrich,
            aes(x = factor(visit), y = common_ymax_rlrich - 1, label = sig_label),
            inherit.aes = FALSE, size = 5)


# RL Richness: Focal vs Control

rl_rich_fc <- bee_data |>
  select(site, starts_with("RLrich_")) |>
  pivot_longer(cols = starts_with("RLrich_"),
               names_to = "RL_category",
               values_to = "value") |>
  mutate(RL_category = str_replace(RL_category, "RLrich_", ""))

ggplot(rl_rich_fc, aes(x = site, y = value, fill = site)) +
  geom_boxplot() +
  facet_wrap(~RL_category, scales = "free_y") +
  labs(title = "RL Richness by Site Type",
       x = "Site Type", y = "Number of Species") +
  theme_minimal() +
  scale_fill_manual(values = c("control" = "#1b9e77", "focal" = "#d95f02"))


# RL Abundance: Focal vs Control

rl_abund_fc <- bee_data |>
  select(site, starts_with("RLabund_")) |>
  pivot_longer(cols = starts_with("RLabund_"),
               names_to = "RL_category",
               values_to = "value") |>
  mutate(RL_category = str_replace(RL_category, "RLabund_", ""))

ggplot(rl_abund_fc, aes(x = site, y = value, fill = site)) +
  geom_boxplot() +
  facet_wrap(~RL_category, scales = "free_y") +
  labs(title = "RL Abundance by Site Type",
       x = "Site Type", y = "Number of Individuals") +
  theme_minimal() +
  scale_fill_manual(values = c("control" = "#1b9e77", "focal" = "#d95f02"))

```


```{r bee Red List pie charts}

# Define Red List richness columns in the desired order
rl_cols <- c("RLrich_LC", "RLrich_NT", "RLrich_VU", "RLrich_EN", "RLrich_DD", "RLrich_CR")

# 2) Define a stable color palette and NAME the vector with the base category names
#    (we will keep fill mapping on the base category, not the label text)
base_colors <- brewer.pal(n = length(rl_cols), name = "Set2")
names(base_colors) <- rl_cols  # named vector ensures correct mapping

# 3) Helper to summarise counts and build a label table with percentages
prepare_summary <- function(data, site_label) {
  long <- data |>
    summarise(across(all_of(rl_cols), sum, na.rm = TRUE)) |>
    pivot_longer(cols = everything(),
                 names_to = "RL_category",
                 values_to = "count") |>
    mutate(site = site_label) |>
    mutate(RL_category = factor(RL_category, levels = rl_cols))  # lock order
  
  # compute percentages for legend labels
  total <- sum(long$count)
  long |>
    mutate(perc = round(100 * count / total, 1),
           legend_label = paste0(RL_category, " (", perc, "%)"))
}

# 4) Build summaries
focal_summary   <- prepare_summary(filter(bee_data, site == "focal"),   "Focal")
control_summary <- prepare_summary(filter(bee_data, site == "control"), "Control")

# 5) Create a lookup from category -> legend label (per plot)
focal_label_map   <- setNames(focal_summary$legend_label,   focal_summary$RL_category)
control_label_map <- setNames(control_summary$legend_label, control_summary$RL_category)

# 6) Build pies using RL_category as the fill aesthetic (factor with fixed levels)
p_focal <- ggplot(focal_summary, aes(x = "", y = count, fill = RL_category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(title = "Focal Sites", fill = "Red List Category") +
  theme_void() +
  # Ensure consistent colors and order via named 'values' + 'breaks'
  scale_fill_manual(
    values = base_colors,              # named by RLrich_* columns
    breaks = rl_cols,                  # ensure order
    labels = focal_label_map[rl_cols]  # percent labels per category for this plot
  )

p_control <- ggplot(control_summary, aes(x = "", y = count, fill = RL_category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(title = "Control Sites", fill = "Red List Category") +
  theme_void() +
  scale_fill_manual(
    values = base_colors,
    breaks = rl_cols,
    labels = control_label_map[rl_cols]
  )

# 7) Combine and collect a single shared legend
(p_focal | p_control) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

```

```{r bee accumulaion curve}
colnames(bee_data)

# Extract only the species columns

bee_matrix <- bee_data[, !(names(bee_data) %in% c(
  "surveyID", "plotID", "site", "region", "visit", "plot",
  "richness", "fieldIDsum", "bee_abundance_nofieldID", "bee_abundance",
  "RLabund_LC", "RLabund_NT", "RLabund_VU", "RLabund_EN", "RLabund_DD",
  "RLrich_LC", "RLrich_NT", "RLrich_VU", "RLrich_EN", "RLrich_DD",
  "RLrich_CR", "RLabund_CR", "RLrich_threatened", "RLabund_threatened",
  "RLrich_total", "RLabund_total"))]


# Create species accumulation curve - X-axis number of surveys
spec_accum <- specaccum(bee_matrix, method = "random")

site_cols <- list(
  "focal" = "#2C7FB8",
  "control"   = "#D95F09",
  "focal_light"   = "cadetblue3",
  "control_light" = "#D95"
  )

# Plot the curve
plot(spec_accum, 
     xlab = "Number of surveys (plot visits)", 
     ylab = "Accumulated species richness", 
     main = "Species Accumulation Curve",
     ci.type = "polygon", 
     col = "blue", 
     lwd = 2, 
     ci.col = "lightblue")


# species accumulation curve for focal / control separately
bee_matrix_control <- bee_matrix[bee_data$site == "control", ]
bee_matrix_focal   <- bee_matrix[bee_data$site == "focal", ]


# Create species accumulation curve by site- X-axis number of surveys
bee_accum_control <- specaccum(bee_matrix_control, method = "random")
bee_accum_focal   <- specaccum(bee_matrix_focal, method = "random")

# Plot both curves
plot(bee_accum_control, 
     xlab = "Number of surveys (plot visits)", 
     ylab = "Accumulated species richness", 
     main = "Species Accumulation: Control vs Focal Sites",
     ci.type = "polygon", 
     col = site_cols$control,
     lwd = 2, 
     ci.col = site_cols$control_light, 
     ylim = range(c(bee_accum_control$richness + bee_accum_control$sd, 
                    bee_accum_focal$richness + bee_accum_focal$sd)))

plot(bee_accum_focal, 
     add = TRUE, 
     col = site_cols$focal,
     ci.col = site_cols$focal_light, 
     ci.type = "polygon",
     lwd = 2)

legend("bottomright", legend = c("Control", "Focal"), 
       col = c(site_cols$focal, site_cols$control), lwd = 2)

```

```{r}
regions <- unique(bee_data$region)

accum_region <- lapply(regions, function(r) {
  bee_matrix[bee_data$region == r, ]
})

names(accum_region) <- regions

spec_accum_region <- lapply(accum_region, function(x) {
  specaccum(x, method = "random")
})

region_cols <- c(
  "region1" = "#4E79A7",
  "region2" = "#59A14F",
  "region3" = "#E15759",
  "region4" = "#B07AA1"
)

region_cols_light <- c(
  "region1" = "#A6C1E0",
  "region2" = "#A8D5A2",
  "region3" = "#F1B0AD",
  "region4" = "#D7C1E3"
)

accum_region_df <- lapply(regions, function(r) {
  
  mat <- bee_matrix[bee_data$region == r, ]
  
  acc <- specaccum(mat, method = "random")
  
  data.frame(
    surveys  = acc$sites,
    richness = acc$richness,
    sd       = acc$sd,
    region   = r
  )
})

accum_plot_data <- bind_rows(accum_region_df)


ggplot(
  accum_plot_data,
  aes(x = surveys, y = richness, colour = region, fill = region)
) +
  geom_ribbon(
    aes(ymin = richness - sd, ymax = richness + sd),
    alpha = 0.25,
    colour = NA
  ) +
  geom_line(linewidth = 1.2) +
  labs(
    x = "Number of surveys (plot visits)",
    y = "Accumulated species richness",
    title = "Species accumulation curves by region",
    colour = "Region",
    fill = "Region"
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

```


```{r}
regions <- unique(bee_data$region)

for (r in regions) {
  
  bee_reg <- bee_data[bee_data$region == r, ]
  
  bee_mat <- bee_reg[, !(names(bee_reg) %in% c(
    "surveyID", "plotID", "site", "region", "visit", "plot",
    "richness", "fieldIDsum", "bee_abundance_nofieldID", "bee_abundance",
    "RLabund_LC", "RLabund_NT", "RLabund_VU", "RLabund_EN", "RLabund_DD",
    "RLrich_LC", "RLrich_NT", "RLrich_VU", "RLrich_EN", "RLrich_DD",
    "RLrich_CR", "RLabund_CR", "RLrich_threatened", "RLabund_threatened",
    "RLrich_total", "RLabund_total"
  ))]
  
  mat_c <- bee_mat[bee_reg$site == "control", ]
  mat_f <- bee_mat[bee_reg$site == "focal", ]
  
  acc_c <- specaccum(mat_c, method = "random")
  acc_f <- specaccum(mat_f, method = "random")
  
  ymax <- max(
    acc_c$richness + acc_c$sd,
    acc_f$richness + acc_f$sd
  )
  
  plot(acc_c,
       ci.type = "polygon",
       col = "#2C7FB8",
       ci.col = adjustcolor("#2C7FB8", 0.3),
       lwd = 2.2,
       ylim = c(0, ymax),
       xlab = "Number of surveys",
       ylab = "Accumulated species richness",
       main = r,
       cex.main = 1.15)
  
  plot(acc_f,
       add = TRUE,
       col = "#D95F02",
       ci.col = adjustcolor("#D95F02", 0.3),
       lwd = 2.2)
  
  legend("bottomright",
         legend = c("Control", "Focal"),
         col = c("#2C7FB8", "#D95F02"),
         lwd = 2.2,
         bty = "n",
         cex = 0.95)
}
```

```{r bee cumulative richness across time}
abund_var <- "bee_abundance"

bee_abund_cum <- bee_data %>%
  arrange(plotID, visit) %>%
  group_by(site, plotID) %>%
  mutate(cum_abundance = cumsum(.data[[abund_var]])) %>%
  ungroup()

bee_abund_summary <- bee_abund_cum %>%
  group_by(site, visit) %>%
  summarise(
    mean_abund = mean(cum_abundance, na.rm = TRUE),
    se_abund   = sd(cum_abundance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

site_cols <- list(
  focal   = "#2C7FB8",
  control = "#D95F09"
)

plot(
  NULL,
  xlim = range(bee_abund_summary$visit),
  ylim = range(bee_abund_summary$mean_abund +
                 bee_abund_summary$se_abund),
  xlab = "Survey visit",
  ylab = "Cumulative bee abundance",
  main = "Cumulative Abundance Over Visits"
)

for (s in c("control", "focal")) {

  d <- bee_abund_summary %>% filter(site == s)

  lines(d$visit, d$mean_abund,
        col = site_cols[[s]],
        lwd = 2)

  polygon(
    c(d$visit, rev(d$visit)),
    c(d$mean_abund - d$se_abund,
      rev(d$mean_abund + d$se_abund)),
    col = adjustcolor(site_cols[[s]], alpha.f = 0.3),
    border = NA
  )
}

legend("topleft",
       legend = c("Control", "Focal"),
       col = c(site_cols$control, site_cols$focal),
       lwd = 2)


# within region
bee_abund_summary <- bee_abund_cum %>%
  group_by(region, visit) %>%  # group by region for aggregation
  summarise(
    mean_abund = mean(cum_abundance, na.rm = TRUE),
    se_abund   = sd(cum_abundance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )
# Optional: assign colors to regions
region_cols <- RColorBrewer::brewer.pal(n = length(unique(bee_abund_summary$region)), name = "Set2")
names(region_cols) <- unique(bee_abund_summary$region)

plot(
  NULL,
  xlim = range(bee_abund_summary$visit),
  ylim = range(bee_abund_summary$mean_abund + bee_abund_summary$se_abund),
  xlab = "Survey visit",
  ylab = "Cumulative bee abundance",
  main = "Cumulative Abundance by Region"
)

for (r in unique(bee_abund_summary$region)) {

  d <- bee_abund_summary %>% filter(region == r)

  lines(d$visit, d$mean_abund,
        col = region_cols[[r]],
        lwd = 2)

  polygon(
    c(d$visit, rev(d$visit)),
    c(d$mean_abund - d$se_abund,
      rev(d$mean_abund + d$se_abund)),
    col = adjustcolor(region_cols[[r]], alpha.f = 0.3),
    border = NA
  )
}

legend("topleft",
       legend = unique(bee_abund_summary$region),
       col = region_cols,
       lwd = 2)

```


<!-- NMDS plants attempt # 1: keeping all plos and visit separate. As many plots have value = 0, I add a dummy collumn. Stress value is high, however. -->
<!-- ```{r plant NMDS build: all visits and plot, eval = interactive()} -->
<!-- colnames(plant_data) -->

<!-- metadata_cols <- c("site", "visit", "region", "plotID", "surveyID") -->
<!-- diversity_cols <- c("hill_q1", "hill_q2", "richness", "relabundance", "avgrelabundance") -->

<!-- # Extract metadata and species matrix -->
<!-- plant_meta <- plant_data[, metadata_cols] -->
<!-- plant_species <- plant_data[, !(names(plant_data) %in% c(metadata_cols, diversity_cols))] -->

<!-- # Add a dummy species column (bray Curtis cannot handle plots without plant presence) -->
<!-- plant_species$dummy <- 0.001 -->

<!-- # Run NMDS -->
<!-- set.seed(123) -->
<!-- nmds_plant <- metaMDS(plant_species, distance = "bray", k = 2, trymax = 100) -->

<!-- ``` -->

<!-- ```{r plant NMDS plot: all visits and plots, eval = interactive()} -->

<!-- # Combine NMDS scores with metadata -->

<!-- plant_scores <- as.data.frame(scores(nmds_plant)$sites) -->
<!-- plant_scores <- cbind(plant_meta, plant_scores) -->

<!-- # === 4. Plot NMDS === -->
<!-- ggplot(plant_scores, aes(x = NMDS1, y = NMDS2, color = site)) + -->
<!--   geom_point(aes(shape = region), size = 3) + -->
<!--   theme_minimal() + -->
<!--   labs(title = "NMDS of Plant Communities", color = "Site", shape = "Region") -->

<!-- #adonis site -->
<!-- adonis2(plant_species ~ site, data = plant_meta, method = "bray") -->

<!-- # adonis interaction site - region -->
<!-- adonis2(plant_species ~ site * region, data = plant_meta, method = "bray") -->
<!-- ``` -->

<!-- NMDS plants attempt # 2: merge all plots across regions per site and visit -->
<!-- ```{r plant NMDS build: merged all surveys by region and site, eval = interactive()} -->

<!-- metadata_cols_merged_reg <- c("site", "visit") -->
<!-- diversity_cols <- c("hill_q1", "hill_q2", "richness", "relabundance", "avgrelabundance") -->

<!-- # Merge plots per site + visit by summing species abundances -->
<!-- plant_merged_reg <- plant_data |> -->
<!--   group_by(site, visit) |> -->
<!--   summarise(across(where(is.numeric), sum), .groups = "drop") -->

<!-- # Extract metadata and species matrix -->
<!-- plant_merged_reg_meta <- plant_merged_reg[, metadata_cols_merged_reg] -->
<!-- plant_merged_reg_species <- plant_merged_reg[, !(names(plant_merged_reg) %in% c(metadata_cols, diversity_cols))] -->

<!-- # Run NMDS on merged data -->
<!-- set.seed(123) -->
<!-- nmds_merged_reg <- metaMDS(plant_merged_reg_species, distance = "bray", k = 2, trymax = 100) -->

<!-- # View NMDS results -->
<!-- print(nmds_merged_reg) -->

<!-- # Run PERMANOVA -->
<!-- adonis2_result <- adonis2(plant_merged_reg_species ~ site, data = plant_merged_reg_meta, method = "bray") -->

<!-- # View results -->
<!-- print(adonis2_result) -->


<!-- ``` -->

<!-- ```{r plant NMDS build: merged all surveys by region and site plot, eval = interactive()} -->

<!-- # Extract NMDS scores -->
<!-- scores_df_nmds_merged_reg <- as.data.frame(scores(nmds_merged_reg, display = "sites")) -->
<!-- scores_df_nmds_merged_reg$site <- plant_merged_reg_meta$site -->
<!-- scores_df_nmds_merged_reg$visit <- plant_merged_reg_meta$visit -->

<!-- # Plot with ellipsoids and visit labels -->
<!-- ggplot(scores_df_nmds_merged_reg, aes(x = NMDS1, y = NMDS2, color = site)) + -->
<!--   geom_point(size = 3) + -->
<!--   geom_text(aes(label = visit), vjust = -1, size = 3) + -->
<!--   stat_ellipse(type = "norm", level = 0.95, linetype = "dashed") + -->
<!--   theme_minimal() + -->
<!--   labs(title = "NMDS with Site Ellipsoids and Visit Labels", color = "Site") -->


<!-- ``` -->

<!-- ```{r plant NMDS build: merge surveys within visit by region and site, eval = interactive()} -->

<!-- # Define metadata and diversity columns -->
<!-- metadata_cols_merged_plot <- c("site", "visit", "region") -->
<!-- diversity_cols <- c("hill_q1", "hill_q2", "richness", "relabundance", "avgrelabundance") -->

<!-- # Merge plots within each site per visit -->
<!-- plant_merged_plot <- plant_data |> -->
<!--   group_by(region, site, visit) |> -->
<!--   summarise(across(where(is.numeric), sum), .groups = "drop") -->

<!-- # Extract metadata and species matrix -->
<!-- meta_cols <- c("region", "site", "visit") -->
<!-- plant_merged_plot_meta <- plant_merged_plot[, meta_cols] -->
<!-- plant_merged_plot_species <- plant_merged_plot[, !(names(plant_merged_plot) %in% c(meta_cols, diversity_cols))] -->

<!-- # Add dummy species to avoid empty rows -->
<!-- plant_merged_plot_species$dummy <- 0.001 -->

<!-- # Run NMDS -->
<!-- set.seed(123) -->
<!-- nmds_merged_plot <- metaMDS(plant_merged_plot_species, distance = "bray", k = 2, trymax = 100) -->

<!-- # Extract NMDS scores -->
<!-- scores_df_nmds_merged_plot <- as.data.frame(scores(nmds_merged_plot, display = "sites")) -->
<!-- scores_df_nmds_merged_plot$site <- plant_merged_plot_meta$site -->
<!-- scores_df_nmds_merged_plot$visit <- plant_merged_plot_meta$visit -->
<!-- scores_df_nmds_merged_plot$region <- plant_merged_plot_meta$region -->

<!-- # Plot NMDS with ellipsoids per site and visit labels -->
<!-- ggplot(scores_df_nmds_merged_plot, aes(x = NMDS1, y = NMDS2, color = site)) + -->
<!--   geom_point(size = 3) + -->
<!--   geom_text(aes(label = visit), vjust = -1, size = 3) + -->
<!--   stat_ellipse(type = "norm", level = 0.95, linetype = "dashed") + -->
<!--   theme_minimal() + -->
<!--   labs(title = "NMDS with Site Ellipsoids and Visit Labels", color = "Site") -->

<!-- ``` -->

<!-- ```{r plant NMDS build and plot: separate plots by visit, eval = interactive()} -->
<!-- # Define metadata and species columns -->
<!-- metadata_cols <- c("site", "visit", "region", "plotID", "surveyID") -->
<!-- diversity_cols <- c("hill_q1", "hill_q2", "richness", "relabundance", "avgrelabundance") -->

<!-- # Extract metadata and species matrix -->
<!-- plant_meta <- plant_data[, metadata_cols] -->
<!-- plant_species <- plant_data[, !(names(plant_data) %in% c(metadata_cols, diversity_cols))] -->

<!-- # Add dummy species column -->
<!-- plant_species$dummy <- 0.001 -->

<!-- # Combine metadata and species -->
<!-- plant_full <- cbind(plant_meta, plant_species) -->

<!-- # Get unique visits -->
<!-- visits <- unique(plant_full$visit) -->

<!-- # Loop over visits and create NMDS plots -->
<!-- for (v in visits) { -->
<!--   # Subset data for the visit -->
<!--   visit_data <- plant_full |> filter(visit == v) -->

<!--   # Separate species and metadata -->
<!--   visit_species <- visit_data[, !(names(visit_data) %in% metadata_cols)] -->
<!--   visit_meta <- visit_data[, metadata_cols] -->

<!--   # Run NMDS -->
<!--   set.seed(123) -->
<!--   nmds <- metaMDS(visit_species, distance = "bray", k = 2, trymax = 100) -->

<!--   # Get NMDS scores -->
<!--   scores_df <- as.data.frame(scores(nmds)$sites) -->
<!--   scores_df <- cbind(visit_meta, scores_df) -->

<!--   # Plot with ellipsoids by site -->
<!--   p <- ggplot(scores_df, aes(x = NMDS1, y = NMDS2, color = site)) + -->
<!--     geom_point(aes(shape = region), size = 3) + -->
<!--     stat_ellipse(aes(group = site), type = "norm", linetype = "dashed") + -->
<!--     theme_minimal() + -->
<!--     labs(title = paste("NMDS of Plant Communities - Visit", v), -->
<!--          color = "Site", shape = "Region") -->

<!--   print(p) -->
<!-- } -->
<!-- ``` -->

<!-- NMDS bee data -->
<!-- ```{r NMDS bee data plotted separately by visit, eval = interactive()} -->
<!-- # Define metadata and species columns -->
<!-- metadata_cols_bee <- c("surveyID", "plotID", "site", "region", "visit", "plot") -->
<!-- diversity_cols_bee <- c("richness", "fieldIDsum", "bee_abundance_nofieldID", "bee_abundance", -->
<!--   "RLabund_LC", "RLabund_NT", "RLabund_VU", "RLabund_EN", "RLabund_DD", -->
<!--   "RLrich_LC", "RLrich_NT", "RLrich_VU", "RLrich_EN", "RLrich_DD", -->
<!--   "RLrich_CR", "RLabund_CR", "RLrich_threatened", "RLabund_threatened", -->
<!--   "RLrich_total", "RLabund_total") -->

<!-- # Extract metadata and species matrix -->
<!-- bee_meta <- bee_data[, metadata_cols_bee] -->
<!-- bee_species <- bee_data[, !(names(bee_data) %in% c(metadata_cols_bee, diversity_cols_bee))] -->

<!-- # Add dummy species column -->
<!-- bee_species$dummy <- 0.001 -->

<!-- # Combine metadata and species -->
<!-- bee_full <- cbind(bee_meta, bee_species) -->

<!-- # Get unique visits -->
<!-- visits <- unique(bee_full$visit) -->

<!-- # Loop over visits and create NMDS plots -->
<!-- for (v in visits) { -->
<!--   # Subset data for the visit -->
<!--   visit_data <- bee_full |> filter(visit == v) -->

<!--   # Separate species and metadata -->
<!--   visit_species <- visit_data[, !(names(visit_data) %in% metadata_cols_bee)] -->
<!--   visit_meta <- visit_data[, metadata_cols_bee] -->

<!--   # Run NMDS -->
<!--   set.seed(123) -->
<!--   nmds <- metaMDS(visit_species, distance = "bray", k = 2, trymax = 100) -->

<!--   # Get NMDS scores -->
<!--   scores_df <- as.data.frame(scores(nmds)$sites) -->
<!--   scores_df <- cbind(visit_meta, scores_df) -->

<!--   # Plot with ellipsoids by site -->
<!--   p <- ggplot(scores_df, aes(x = NMDS1, y = NMDS2, color = site)) + -->
<!--     geom_point(aes(shape = region), size = 3) + -->
<!--     stat_ellipse(aes(group = site), type = "norm", linetype = "dashed") + -->
<!--     theme_minimal() + -->
<!--     labs(title = paste("NMDS of Bee Communities - Visit", v), -->
<!--          color = "Site", shape = "Region") -->

<!--   print(p) -->
<!-- } -->
<!-- ``` -->

<!-- ```{r NMDS bee data plotted separately by visit - delete rows with 0, eval = interactive()} -->
<!-- # Define metadata and species columns -->
<!-- metadata_cols_bee <- c("surveyID", "plotID", "site", "region", "visit", "plot") -->
<!-- diversity_cols_bee <- c("richness", "fieldIDsum", "bee_abundance_nofieldID", "bee_abundance", -->
<!--   "RLabund_LC", "RLabund_NT", "RLabund_VU", "RLabund_EN", "RLabund_DD", -->
<!--   "RLrich_LC", "RLrich_NT", "RLrich_VU", "RLrich_EN", "RLrich_DD", -->
<!--   "RLrich_CR", "RLabund_CR", "RLrich_threatened", "RLabund_threatened", -->
<!--   "RLrich_total", "RLabund_total") -->

<!-- # Extract metadata and species matrix -->
<!-- bee_meta <- bee_data[, metadata_cols_bee] -->
<!-- bee_species <- bee_data[, !(names(bee_data) %in% c(metadata_cols_bee, diversity_cols_bee))] -->

<!-- # Remove rows where all species values are zero -->
<!-- bee_species_filtered <- bee_species[rowSums(bee_species) > 0, ] -->

<!-- # Match metadata to filtered species data -->
<!-- bee_meta_filtered <- bee_meta[rowSums(bee_species) > 0, ] -->

<!-- # Combine metadata and species -->
<!-- bee_full_filtered <- cbind(bee_meta_filtered, bee_species_filtered) -->

<!-- # Get unique visits -->
<!-- visits <- unique(bee_full_filtered$visit) -->

<!-- # store nmds scores in list -->
<!-- nmds_list <- list() -->
<!-- scores_list <- list() -->

<!-- for (v in visits) { -->
<!--   visit_data <- bee_full_filtered |> filter(visit == v) -->
<!--   visit_species <- visit_data[, !(names(visit_data) %in% metadata_cols_bee)] -->
<!--   visit_meta <- visit_data[, metadata_cols_bee] -->

<!--   set.seed(123) -->
<!--   nmds <- metaMDS(visit_species, distance = "bray", k = 2, trymax = 100) -->

<!--   scores_df <- as.data.frame(scores(nmds)$sites) -->
<!--   scores_df <- cbind(visit_meta, scores_df) -->

<!--   nmds_list[[v]] <- nmds -->
<!--   scores_list[[v]] <- scores_df -->
<!-- } -->

<!-- # plot with and without outliers -->
<!-- for (v in visits) { -->
<!--   scores_df <- scores_list[[v]] -->

<!--   # Identify outliers -->
<!--   scores_df$z_NMDS1 <- scale(scores_df$NMDS1) -->
<!--   scores_df$z_NMDS2 <- scale(scores_df$NMDS2) -->
<!--   scores_df$outlier <- abs(scores_df$z_NMDS1) > 2 | abs(scores_df$z_NMDS2) > 2 -->

<!--   # Plot with outliers -->
<!--   p1 <- ggplot(scores_df, aes(x = NMDS1, y = NMDS2, color = site)) + -->
<!--     geom_point(aes(shape = region), size = 3) + -->
<!--     stat_ellipse(aes(group = site), type = "norm", linetype = "dashed") + -->
<!--     theme_minimal() + -->
<!--     labs(title = paste("NMDS of Bee Communities - Visit", v, "(All Points)"), -->
<!--          color = "Site", shape = "Region") -->

<!--   print(p1) -->

<!--   # Plot without outliers -->
<!--   scores_df_filtered <- scores_df |> filter(!outlier) -->

<!--   p2 <- ggplot(scores_df_filtered, aes(x = NMDS1, y = NMDS2, color = site)) + -->
<!--     geom_point(aes(shape = region), size = 3) + -->
<!--     stat_ellipse(aes(group = site), type = "norm", linetype = "dashed") + -->
<!--     theme_minimal() + -->
<!--     labs(title = paste("NMDS of Bee Communities - Visit", v, "(Outliers Removed)"), -->
<!--          color = "Site", shape = "Region") -->

<!--   print(p2) -->
<!-- } -->

<!-- ``` -->


# Bee models

Create full dataset for modelling
```{r bee model data prep: data_all prep}
data_all <- merge(habplot_data, plant_data, by = "surveyID")
data_all <- merge(data_all, samplcov_data, by = "surveyID")
data_all <- merge(data_all, bee_data, by = "surveyID")


# Remove duplicated columns manually
data_all <- data_all |>
  select(-matches("\\.x$|\\.y$"))  # removes columns ending in .x or .y

# Add plot info from mapp_data
mapp_subset <- mapp_data |>
  select(surveyID, site, region, plot, plotID, visit)

# Join with data_all by surveyID
data_all <- left_join(mapp_subset, data_all, by = "surveyID")


data_all <- data_all |>
  mutate(
    site   = factor(site, levels = c("control", "focal")),
    region = factor(region),
    plotID = factor(plotID)
  )


View(data_all)
```

Scale data 
```{r bee model data prep: scale data}
data_all_scaled <- data_all

# Identify continuous predictors
scale_vars <- c("bareground", "vegheight", "herb", "sand",
              "gravel", "cobble", "stone", "wood", "soil", "sloped",
              "steep", "canopy", "branchpile",
              "stonepile", "standingdeadwood", "lyingdeadwood", "hill_q0", "hill_q1", "hill_q2")

# Scale them (center and scale)
#data_all_scaled <- data_all_scaled |>
#  mutate(across(all_of(scale_vars), ~ as.numeric(scale(.))))


data_all_scaled <- data_all |>
  mutate(across(all_of(scale_vars), scale)) 

```
scale patchiness and visit? no, they are ordinal variables

```{r bee model data prep: handling non-numeric covariates}
#creat ordinal patchiness and visit columns
data_all_scaled <- data_all_scaled |>
  mutate(patchiness_ord = ordered(patchiness, levels = c("0","1","2","3")), 
         visit_ord = ordered(visit, levels = c("1","2","3","4","5")))
```


```{r bee model data prep: save model dataframe}
saveRDS(data_all_scaled, file = "data_all_scaled.rds")
```

"We included random intercepts for region and plot to account for hierarchical structure and repeated measures. Region controls for large-scale environmental differences among study areas, while plot accounts for non-independence of repeated surveys within the same plot. The effect of site (focal vs control) was modeled as a fixed effect because it represents the main ecological contrast of interest and has only two levels, making it unsuitable as a random effect." - GPT formulation for use of random effect structure (1|region) + (1|plotID)

Abundance modelling - test for overdispersion / zero inflation


## Bee abundance models

Bee ab model prep:
```{r bee ab model prep: overdispersion test, response data distribution}
#baseline poisson model
m_pois <- glmer(
  bee_abundance ~ site + (1 | region) + (1 | plotID), 
  data = data_all_scaled,
  family = poisson
)

summary(m_pois)

#test for overdispersion
overdisp_fun <- function(model) {
  rdf <- df.residual(model)
  rp <- residuals(model, type = "pearson")
  Pearson.chisq <- sum(rp^2)
  ratio <- Pearson.chisq / rdf
  pval <- pchisq(Pearson.chisq, df = rdf, lower.tail = FALSE)
  c(chisq = Pearson.chisq, ratio = ratio, rdf = rdf, p = pval)
}

overdisp_fun(m_pois) #clear overdispersion

#negatvie binomial modlel
m_nb <- glmmTMB(
  bee_abundance ~ site + (1 | region) + (1 | plotID),
  family = nbinom2,   # Negative binomial
  data = data_all_scaled)

summary(m_nb)

# Check diagnostics
sim_nb <- simulateResiduals(m_nb, n = 1000)
plot(sim_nb)
testDispersion(sim_nb) # no overdispersion with negbin
testZeroInflation(sim_nb) # no zero inflation

#negative binomial is the distribution used from now on
```

```{r bee ab model prep: test for collinearity among habitat explanatory variables}
# List of numeric environmental variables
env_vars <- c("visit", "bareground", "vegcover", "vegheight", "herb", "woody", "sand",
              "gravel", "cobble", "stone", "wood", "soil", "flat", "sloped",
              "steep", "canopy", "canopy_hp", "branchpile",
              "stonepile", "patchiness", "standingdeadwood", "lyingdeadwood", "hill_q0", "hill_q1", "hill_q2")
#visit and patchiness, although included in coming models as ordinal, are treated as numeric for Spearman computation

# Compute Spearman correlation matrix
cor_matrix <- cor(
  data_all_scaled[env_vars],
  method = "spearman",
  use = "pairwise.complete.obs")

# Identify pairs with |correlation| > 0.7
high_corr <- which(abs(cor_matrix) > 0.7 & abs(cor_matrix) <= 1, arr.ind = TRUE)

# Create a data frame of correlated pairs
correlated_pairs <- data.frame(
  var1 = rownames(cor_matrix)[high_corr[, 1]],
  var2 = colnames(cor_matrix)[high_corr[, 2]],
  correlation = cor_matrix[high_corr]
)

# Remove duplicates (var1-var2 vs var2-var1)
correlated_pairs[correlated_pairs$var1 < correlated_pairs$var2, ]

# visualize correlation
corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  addCoef.col = "black",
  number.cex = 0.6
)
# New variable dataframe (merged covariates with spearman above 0.7)
# bareground - vegcover -> bareground
# herb - woody -> herb
# flat - sloped -> sloped
# hill_q0 - hill_q1 -> hill_q1

# canopy_hp excluded as constant across visits
env_vars <- c("bareground", "vegheight", "herb", "sand",
              "gravel", "cobble", "stone", "wood", "soil", "sloped",
              "steep", "canopy", "branchpile",
              "stonepile", "standingdeadwood", "lyingdeadwood", "hill_q1", "hill_q2")
```

Can we include site as explanatory variable in a multivariate model? - No, this chunk shows that too many of the habitat variables confound with site, and should therefore not be included in the same multivariate model.
```{r bee ab model prep: diagnostic site vs habitat covariates}
num_vars <- c("visit", "bareground", "vegcover", "vegheight", "herb", "woody", "sand",
      "gravel", "cobble", "stone", "wood", "soil", "flat", "sloped", "steep", "canopy", 
      "canopy_hp", "branchpile", "stonepile", "patchiness", "standingdeadwood", 
      "lyingdeadwood", "hill_q1", "hill_q2")

site_tests <- lapply(num_vars, function(v) {
  wt <- wilcox.test(data_all_scaled[[v]] ~ data_all_scaled$site)
  data.frame(
    variable = v,
    W = wt$statistic,
    p_value = wt$p.value
  )
})

site_tests <- do.call(rbind, site_tests)

site_tests$p_adj <- p.adjust(site_tests$p_value, method = "BH")

# these variables confound with site and cannot be included in a multivariate model together
confounded_vars <- site_tests |>
  subset(p_adj < 0.05) |>
  arrange(p_adj)

confounded_vars$variable

# these variables do not confound with site and could be included in a multivariate model together
independent_vars <- site_tests |>
  subset(p_adj >= 0.05)

independent_vars$variable
```

Run a univariate model to see the effect of site on bee abundance.
```{r bee ab model: site}
NBab1 <- glmmTMB(
  bee_abundance ~ site + (1 | region) + (1|plotID),
  family = nbinom2,   # Negative binomial
  data = data_all_scaled)

summary(NBab1)

# Get predicted values for 'site'
pred_site <- ggpredict(NBab1, terms = "site")

# Plot
plot(pred_site) +
  ggtitle("Predicted Bee Abundance by Site") +
  ylab("Predicted Abundance") +
  xlab("Site")

#add raw data jitter

site_cols <- c(
  "focal" = "#6B7C93",
  "control"   = "#D08C60"
)

ggplot() +
  # raw data: barely there
  geom_jitter(
    data = data_all_scaled,
    aes(x = site, y = bee_abundance, colour = site),
    width = 0.12,
    alpha = 0.15,
    size = 1.6,
    show.legend = FALSE
  ) +
  # predicted means
  geom_point(
    data = pred_site,
    aes(x = x, y = predicted, colour = x),
    size = 4
  ) +
  # confidence intervals
  geom_errorbar(
    data = pred_site,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.12,
    linewidth = 1
  ) +
  scale_colour_manual(values = site_cols) +
  labs(
    x = NULL,
    y = "Predicted bee abundance",
    title = "Predicted bee abundance by site, raw data included"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    axis.title.y = element_text(margin = margin(r = 8))
  )
```

Run univariate models for ordinal covariates
```{r bee ab model: univarite models for ordinal variables}
# patchiness
p_ord <- glmmTMB(bee_abundance ~ patchiness_ord + (1 | region) + (1|plotID),
                 family = nbinom2, data = data_all_scaled)
summary(p_ord)

#visit
v_ord <- glmmTMB(bee_abundance ~ visit_ord + (1 | region) + (1|plotID),
                 family = nbinom2, data = data_all_scaled)
summary(v_ord)


#plot 
pred <- ggpredict(p_ord, terms = "patchiness_ord")
ggplot(pred, aes(x = x, y = predicted)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  theme_minimal() +
  labs(title = "Effect of Patchiness (Ordinal) on Bee Abundance",
       x = "Patchiness Level", y = "Predicted Abundance")


pred <- ggpredict(v_ord, terms = "visit_ord")
ggplot(pred, aes(x = x, y = predicted)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  theme_minimal() +
  labs(title = "Effect of visit (Ordinal) on Bee Abundance",
       x = "Visit Number", y = "Predicted Abundance")

```

Try interactive model between site * visit 
```{r}
m_sv <- glmmTMB(bee_abundance ~ site*visit + (1 | region) + (1|plotID),
                 family = nbinom2, data = data_all_scaled)
summary(m_sv)

pred_sv <- ggpredict(m_sv, terms = c("visit", "site"))


plot(pred_sv) +
  ggtitle("Predicted Bee Abundance across Visits by Site") +
  xlab("Visit") +
  ylab("Predicted abundance (Negative Binomial)") +
  theme_minimal(base_size = 13)

# Custom color palette
site_cols <- c("focal" = "#6B7C93", "control" = "#D08C60")

ggplot() +
  # Raw data
  geom_jitter(
    data = data_all_scaled,
    aes(x = visit, y = bee_abundance, colour = site),
    width = 0.2, alpha = 0.2, size = 1.8
  ) +
  # Predicted values
  geom_line(
    data = pred_sv,
    aes(x = x, y = predicted, colour = group),
    size = 1.2
  ) +
  # Confidence intervals
  geom_ribbon(
    data = pred_sv,
    aes(x = x, ymin = conf.low, ymax = conf.high, fill = group),
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  scale_colour_manual(values = site_cols) +
  scale_fill_manual(values = site_cols) +
  labs(
    x = "Visit (ordinal)",
    y = "Predicted Bee Abundance",
    title = "Predicted Bee Abundance Across Visits by Site"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```


Run univariate models for numeric covariates
```{r bee ab model: run univariate model for numeric variabels}
model_um_abundance <- tibble(
  Variable = env_vars,
  Model = map(env_vars, ~ glmmTMB(
    as.formula(paste("bee_abundance ~ `", .x, "` + (1 | region) + (1|plotID)", sep = "")),
    family = nbinom2,
    data = data_all_scaled
  ))
)
```


```{r bee ab model: plot univarite model predictions}
model_df <- model_um_abundance

# Extract model stats
model_df <- model_df |>
  mutate(
    Summary = map(Model, ~ summary(.x)$coefficients$cond),
    CoefName = map_chr(Summary, ~ setdiff(rownames(.x), "(Intercept)")),
    Estimate = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Estimate"]),
    StdError = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Std. Error"]),
    p_value = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Pr(>|z|)"]),
    CI_lower = Estimate - 1.96 * StdError,
    CI_upper = Estimate + 1.96 * StdError,
    AIC = map_dbl(Model, AIC)
  )

print(model_df)

# Plot all model output estimates, mark significant ones in different color
# Sort by AIC for convenience
model_df_am <- model_df |> arrange(AIC)

# Add significance flag
model_df_am <- model_df_am |>
  mutate(Significant = ifelse(p_value < 0.05, "Significant", "Not Significant"))

# Forest plot with color by significance
ggplot(model_df_am, aes(x = reorder(Variable, Estimate), y = Estimate, color = Significant)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  ylab("Effect Size (Estimate)") +
  xlab("Environmental Variable") +
  ggtitle("Univariate Effects on Bee Abundance") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(values = c("Significant" = "#D55E00", "Not Significant" = "gray"))


# plot significant univariate models
# Filter significant models
significant_df <- model_df |>
  filter(!is.na(p_value), p_value < 0.05) |>
  arrange(AIC)

# View summary table
print(significant_df |> select(Variable, AIC, Estimate, CI_lower, CI_upper, p_value))

# Plot predicted effects using ggpredict
plots <- map2(significant_df$Model, significant_df$Variable, ~ {
  pred <- ggpredict(.x, terms = .y, bias_correction = TRUE, transform = TRUE)
  ggplot(pred, aes(x = x, y = predicted)) +
    geom_line() +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
    theme_minimal() +
    labs(title = paste("Effect of", .y, "on Bee Abundance"),
         x = .y, y = "Predicted Abundance")
})

ggplot(significant_df, aes(x = reorder(Variable, Estimate), y = Estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  ylab("Effect Size (Estimate)") +
  xlab("Environmental Variable") +
  ggtitle("Significant Predictors of Bee Abundance (p < 0.05)") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray")
```


Build multivariate model for bee abundance
```{r}
AB_glmm <- glmmTMB(
  bee_abundance ~ hill_q1 + hill_q2 + sand + soil + vegheight + patchiness_ord +
    (1 | region) + (1|plotID),
  family = nbinom2,
  data = data_all_scaled
)
print(summary(AB_glmm))

vif(lm(bee_abundance ~ hill_q1 + hill_q2 + sand + soil + vegheight + patchiness_ord,
       data = data_all_scaled))
#all values below 3 - no multicollinearity 

```

```{r}

# Extract tidy summary
coef_df <- broom.mixed::tidy(AB_glmm, conf.int = TRUE)

# Filter only fixed effects
coef_df <- coef_df |>
  filter(effect == "fixed") |>
  mutate(significant = ifelse(conf.low * conf.high > 0, "Significant", "Not Significant"))



# Plot
ggplot(coef_df, aes(x = estimate, y = term, color = significant)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  scale_color_manual(values = c("Significant" = "orange", "Not Significant" = "gray")) +
  theme_minimal() +
  labs(title = "Falcon Plot of GLMM Effects",
       x = "Estimate (with 95% CI)",
       y = "Predictor")

```


```{r}
AB_glmm <- glmmTMB(
  bee_abundance ~ site + hill_q1 + hill_q2 + sloped + sand + patchiness +
    (1 | region/site/plotID),
  family = nbinom2,
  data = data_all_scaled
)
print(summary(AB_glmm))

```






Modelling bee richness


Richness modelling - test for overdispersion / zero inflation
```{r bee models richness}
#baseline poisson model
m_pois <- glmer(
  bee_richness ~ site + (1 | region/site/plotID), 
  data = bee_data,
  family = poisson
)

summary(m_pois)

#test for overdispersion
overdisp_fun <- function(model) {
  rdf <- df.residual(model)
  rp <- residuals(model, type = "pearson")
  Pearson.chisq <- sum(rp^2)
  ratio <- Pearson.chisq / rdf
  pval <- pchisq(Pearson.chisq, df = rdf, lower.tail = FALSE)
  c(chisq = Pearson.chisq, ratio = ratio, rdf = rdf, p = pval)
}

overdisp_fun(m_pois) #clear overdispersion

#negatvie binomial modlel
m_nb <- glmmTMB(
  bee_richness ~ site + (1 | region/site/plotID),
  family = nbinom2,   # Negative binomial
  data = bee_data)

# Check diagnostics
sim_nb <- simulateResiduals(m_nb, n = 1000)
plot(sim_nb)
testDispersion(sim_nb) # no overdispersion with negbin
testZeroInflation(sim_nb) # no zero inflation

```
```{r}
# Run univariate models for bee richness

env_vars <- c("bareground", "vegheight", "herb", "sand",
              "gravel", "cobble", "stone", "wood", "soil", "sloped",
              "steep", "canopy", "branchpile",
              "stonepile", "patchiness", "standingdeadwood", "lyingdeadwood", "hill_q0", "hill_q1", "hill_q2")

bee_um_richness <- tibble(
  Variable = env_vars,
  Model = map(env_vars, ~ glmmTMB(
    as.formula(paste("bee_richness ~ `", .x, "` + (1 | region/site/plotID)", sep = "")),
    family = nbinom2,  # still appropriate if richness is count data
    data = data_all_scaled
  ))
)
```
Due to warning: Model convergence problem; non-positive-definite Hessian matrix - run diagnostics
```{r}
# Check convergence for each model
convergence_flags <- map_lgl(bee_um_richness$Model, ~ {
  any(grepl("Model convergence problem", .x$fit$optinfo$conv$lme4$messages))
})

# Combine with variable names
tibble(Variable = bee_um_richness$Variable, ConvergenceIssue = convergence_flags)
```
models themselves do not have convergence issues, we can proceed. 
(possible explanation: glmmTMB often issues a temporary warning if the Hessian is singular at some iteration, even if the optimizer later resolves it.)


```{r}
model_df_richness <- bee_um_richness

# Extract model stats
model_df_richness <- model_df_richness |>
  mutate(
    Summary = map(Model, ~ summary(.x)$coefficients$cond),
    CoefName = map_chr(Summary, ~ setdiff(rownames(.x), "(Intercept)")),
    Estimate = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Estimate"]),
    StdError = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Std. Error"]),
    p_value = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Pr(>|z|)"]),
    CI_lower = Estimate - 1.96 * StdError,
    CI_upper = Estimate + 1.96 * StdError,
    AIC = map_dbl(Model, AIC)
  )

# Filter significant models
significant_df_richness <- model_df_richness |>
  filter(!is.na(p_value), p_value < 0.05) |>
  arrange(AIC)

# View summary table
print(significant_df_richness |> select(Variable, AIC, Estimate, CI_lower, CI_upper, p_value))

# Plot predicted effects using ggpredict
plots_richness <- map2(significant_df_richness$Model, significant_df_richness$Variable, ~ {
  pred <- ggpredict(.x, terms = .y, bias_correction = TRUE, transform = TRUE)
  ggplot(pred, aes(x = x, y = predicted)) +
    geom_line() +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
    theme_minimal() +
    labs(title = paste("Effect of", .y, "on Bee Richness"),
         x = .y, y = "Predicted Richness")
})

plots_richness


ggplot(significant_df_richness, aes(x = reorder(Variable, Estimate), y = Estimate)) +
  geom_point(size = 3, color = "#D55E00") +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, color = "#D55E00") +
  coord_flip() +
  theme_minimal() +
  ylab("Effect Size (Estimate)") +
  xlab("Environmental Variable") +
  ggtitle("Significant Predictors of Bee Richness (p < 0.05)") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray")
```


```{r}
# Forest plot for all models
model_df_richness <- model_df_richness |>
  arrange(AIC) |>
  mutate(Significant = ifelse(p_value < 0.05, "Significant", "Not Significant"))

ggplot(model_df_richness, aes(x = reorder(Variable, Estimate), y = Estimate, color = Significant)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  ylab("Effect Size (Estimate)") +
  xlab("Environmental Variable") +
  ggtitle("Univariate Effects on Bee Richness") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(values = c("Significant" = "#D55E00", "Not Significant" = "gray"))
```





```{r visit and site interactive model}

m_nb <- glmmTMB(bee_abundance ~ site * visit + (1 | region) + (1 | plotID),
                family = nbinom2, data = data_all_scaled)

# Newdata grid over site × visit (months 1..5)
newdata <- expand.grid(
  site  = levels(data_all_scaled$site),
  visit = sort(unique(data_all_scaled$visit))
)

# Marginal predictions on response scale (exclude random effects)
pred <- predict(m_nb, newdata = newdata, type = "response", se.fit = TRUE, re.form = NA)

newdata <- newdata |>
  mutate(
    fit = as.numeric(pred$fit),
    se  = as.numeric(pred$se.fit),
    lwr = pmax(fit - 1.96 * se, 0),
    upr = fit + 1.96 * se
  )

# Plot: raw points (transparent) + predicted line + 95% CI ribbon
p <- ggplot() +
  geom_point(
    data = data_all_scaled,
    aes(x = visit, y = bee_abundance, color = site),
    alpha = 0.25, size = 2,
    position = position_jitter(width = 0.05, height = 0)
  ) +
  geom_ribbon(
    data = newdata,
    aes(x = visit, ymin = lwr, ymax = upr, fill = site, group = site),
    alpha = 0.20
  ) +
  geom_line(
    data = newdata,
    aes(x = visit, y = fit, color = site, group = site),
    linewidth = 1.2
  ) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    x = "Visit (month)",
    y = "Predicted bee abundance (Negative Binomial, nbinom2)",
    color = "Site",
    fill  = "Site",
    title = "Bee abundance: raw counts with glmmTMB predictions and 95% CI"
  ) +
  theme_minimal(base_size = 13)

print(p)

```






