---
title: "brms_bee"
author: "GEgloff"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Introduction into using brms:
https://ourcodingclub.github.io/tutorials/brms/#part1


Based on methodology provided by Gabriel Marcacci: 
```{r}
library(brms)
library(ggeffects)
library(ggplot2)
library(loo)
library(rstanarm)
```

```{r}
data_all_scaled <- readRDS("data_all_scaled.rds")
```

```{r base model}
# BM <- brm(bee_abundance ~ site + (1|region/site/plotID),
#                          data = data_all_scaled, family = negbinomial(), chains = 3,
#                          iter = 3000, warmup = 1000)
# 
# saveRDS(BM, file = "BM_model.rds")
```
```{r}
BM <- readRDS("BM_model.rds")

summary(BM)                     # look at Rhat and Bulk/Tail ESS
mcmc_plot(BM, type = "rhat")    # all ~ 1.00
mcmc_plot(BM, type = "neff")    # adequate effective sample sizes
summary(BM)
plot(BM)
```

```{r mulitvariate model}

# MM <- brm(bee_abundance ~ site + hill_q1 + hill_q2 + sloped + sand + patchiness_fac + (1 | region/site/plotID), 
#            data = data_all_scaled, family = negbinomial(), chains = 3, iter = 7000, warmup = 2000)
# 
# saveRDS(MM, file = "MM_model.rds")
```
```{r}
MM <- readRDS("MM_model.rds")

summary(MM)
plot(MM)

```
```{r}
start.time <- Sys.time()
# MM2 <- brm(bee_abundance ~ site + hill_q1 + hill_q2 + sloped + sand + patchiness_fac + (1 | region/site/plotID), 
#            data = data_all_scaled, family = negbinomial(), chains = 4, iter = 10000, warmup = 5000)
# 
# saveRDS(MM2, file = "MM2_model.rds")

(run.time <- Sys.time() - start.time)

```
```{r}
MM2 <- readRDS("MM2_model.rds")

summary(MM2)                 # Rhat < 1.01, bulk/tail ESS high
plot(MM2)                    # trace plots and densities
pp_check(MM2)                # overall distribution
pp_check(MM2, type = "bars") # counts & zeros
pp_check(MM2, type = "rootogram")
pp_check(MM2, ndraws = 100)

loo_MM2 <- loo(MM2)          # out-of-sample predictive ability

```
```{r}
# This creates curves of predicted mean abundance across Hill q1,
# split by site (focal vs control), holding other predictors at typical values.
ce <- conditional_effects(
  MM2,
  effects = "hill_q1:site",  # interaction-style display
  prob = 0.95
)

p_ce <- plot(ce)[[1]] +
  labs(
    x = "Hill q1 (standardized)",
    y = "Predicted bee abundance (mean)",
    title = "Predicted bee abundance vs Hill q1 by site"
  ) +
  theme_minimal(base_size = 12)

print(p_ce)
```


```{r}
#try with different covariates
start.time <- Sys.time()
# MM3 <- brm(bee_abundance ~ site + bareground + hill_q1 + hill_q2 + herb + sand + sloped + steep + patchiness_fac + (1 | region/site/plotID), 
#            data = data_all_scaled, family = negbinomial(), chains = 4, iter = 10000, warmup = 5000)
# 
# saveRDS(MM3, file = "MM3_model.rds")

(run.time <- Sys.time() - start.time)
```
```{r}
MM3 <- readRDS("MM3_model.rds")

summary(MM3)                 # Rhat < 1.01, bulk/tail ESS high
plot(MM3)                    # trace plots and densities
pp_check(MM3)                # overall distribution
pp_check(MM3, type = "bars") # counts & zeros
pp_check(MM3, type = "rootogram")

loo_MM3 <- loo(MM3)          # out-of-sample predictive ability

```
“We compared two hierarchical negative binomial models via PSIS‑LOO. The lean model (fewer predictors) had elpd_loo = −743.6 (SE 20.4) and looic = 1487.2 (SE 40.7), versus −745.5 (SE 20.5) and 1491.0 (SE 40.9) for the larger model. The effective number of parameters was lower for the lean model (p_loo ≈ 19.4 vs. 21.3), and all Pareto‑k diagnostics were below 0.7. Differences in elpd_loo were small relative to their uncertainty, indicating similar predictive performance; we therefore selected the leaner model for parsimony.”


```{r}
#try with different covariates
start.time <- Sys.time()
MM4 <- brm(bee_abundance ~ site + bareground + hill_q1 + hill_q2 + herb + sand + soil + sloped + steep + patchiness_ord + vegheight + visit_ord +
             (1|region) + (1|plotID),
           data = data_all_scaled, family = negbinomial(), chains = 4, iter = 10000, warmup = 5000)

saveRDS(MM4, file = "MM4_model.rds")

(run.time <- Sys.time() - start.time)
```
```{r}
MM4 <- readRDS("MM4_model.rds")

summary(MM4)                 # Rhat < 1.01, bulk/tail ESS high
plot(MM4)                    # trace plots and densities
pp_check(MM4)                # overall distribution
pp_check(MM4, type = "bars", draws = 200) # counts & zeros
pp_check(MM4, type = "rootogram")

loo_MM4 <- loo(MM4)          # out-of-sample predictive ability

```
```{r}
# Set up colors for sites
site_cols <- c(
  "focal" = "#6B7C93",
  "control"   = "#D08C60"
)


base_theme <- theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    legend.title = element_blank()
  )

## ============================
## 1️⃣ SITE EFFECT
## ============================
pred_site <- ggpredict(MM4, terms = "site")

p_site <- ggplot() +
  geom_jitter(
    data = data_all_scaled,
    aes(x = site, y = bee_abundance, color = site),
    width = 0.1, alpha = 0.25, size = 1.8
  ) +
  geom_point(
    data = pred_site,
    aes(x = x, y = predicted, color = x),
    size = 3
  ) +
  geom_errorbar(
    data = pred_site,
    aes(x = x, ymin = conf.low, ymax = conf.high, color = x),
    width = 0.1
  ) +
  scale_color_manual(values = site_cols) +
  labs(
    x = "Site",
    y = "Bee abundance",
    title = "Effect of site on bee abundance"
  ) +
  base_theme +
  theme(legend.position = "none")

## ============================
## 2️⃣ PATCHINESS (ORDINAL)
## ============================
pred_patch <- ggpredict(MM4, terms = "patchiness_ord")

p_patch <- ggplot() +
  geom_jitter(
    data = data_all_scaled,
    aes(x = patchiness_ord, y = bee_abundance),
    width = 0.15, alpha = 0.2, size = 1.6, color = "grey40"
  ) +
  geom_line(
    data = pred_patch,
    aes(x = x, y = predicted, group = 1),
    size = 1.2, color = "#2F3E46"
  ) +
  geom_ribbon(
    data = pred_patch,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    alpha = 0.2, fill = "#84A59D"
  ) +
  labs(
    x = "Patchiness (ordinal)",
    y = "Bee abundance",
    title = "Effect of habitat patchiness on bee abundance"
  ) +
  base_theme

## ============================
## 3️⃣ HILL DIVERSITY (q = 1)
## ============================
pred_hill1 <- ggpredict(MM4, terms = "hill_q1 [all]")

p_hill1 <- ggplot() +
  geom_jitter(
    data = data_all_scaled,
    aes(x = hill_q1, y = bee_abundance),
    width = 0, alpha = 0.2, size = 1.6, color = "grey40"
  ) +
  geom_line(
    data = pred_hill1,
    aes(x = x, y = predicted),
    size = 1.2, color = "#3A5A40"
  ) +
  geom_ribbon(
    data = pred_hill1,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    alpha = 0.2, fill = "#A3B18A"
  ) +
  labs(
    x = "Hill diversity (q = 1)",
    y = "Bee abundance",
    title = "Effect of plant Hill q1 diversity on bee abundance"
  ) +
  base_theme

## ============================
## 4️⃣ SOIL COVER
## ============================
pred_soil <- ggpredict(MM4, terms = "soil [all]")

p_soil <- ggplot() +
  geom_jitter(
    data = data_all_scaled,
    aes(x = soil, y = bee_abundance),
    width = 0, alpha = 0.2, size = 1.6, color = "grey40"
  ) +
  geom_line(
    data = pred_soil,
    aes(x = x, y = predicted),
    size = 1.2, color = "#6D4C41"
  ) +
  geom_ribbon(
    data = pred_soil,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    alpha = 0.2, fill = "#D7CCC8"
  ) +
  labs(
    x = "Soil cover",
    y = "Bee abundance",
    title = "Effect of soil cover on bee abundance"
  ) +
  base_theme

## ============================
## 5️⃣ VISIT (ORDINAL / SEASON)
## ============================
pred_visit <- ggpredict(MM4, terms = "visit_ord")

p_visit <- ggplot() +
  geom_jitter(
    data = data_all_scaled,
    aes(x = visit_ord, y = bee_abundance),
    width = 0.15, alpha = 0.25, size = 1.6, color = "grey40"
  ) +
  geom_line(
    data = pred_visit,
    aes(x = x, y = predicted, group = 1),
    size = 1.2, color = "#5E548E"
  ) +
  geom_ribbon(
    data = pred_visit,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    alpha = 0.2, fill = "#CDB4DB"
  ) +
  labs(
    x = "Visit (ordinal)",
    y = "Bee abundance",
    title = "Seasonal trend in bee abundance"
  ) +
  base_theme

## ============================
## DISPLAY PLOTS
## ============================
p_site
p_patch
p_hill1
p_soil
p_visit
```




```{r mulitvariate model using stan_glmer 2}

# # Same formula structure as in lme4/brms
# fit_rs <- stan_glmer(
#   bee_abundance ~ site + hill_q1 + hill_q2 + sloped + sand + patchiness_fac +
#     (1 | region/site/plotID),
#   data = data_all_scaled,
#   family = neg_binomial_2(link = "log"),   # ~ nbinom2 variance: Var = mu + mu^2 / phi
#   chains = 4, iter = 6000, warmup = 2000, seed = 123,
#   # Defaults in rstanarm are weakly-informative; you can also set custom priors (next section)
#   cores = parallel::detectCores()
# )
# 
# # Save model
# saveRDS(fit_rs, "MM_model_rstanarm.rds")
```

```{r}
# Summary & diagnostics

fit_rs <- readRDS("MM_model_rstanarm.rds")

print(summary(fit_rs), digits = 2)
summary(fit_rs)
pp_check(fit_rs, plotfun = "dens_overlay")  # posterior predictive check (overall)
pp_check(fit_rs, plotfun = "error_binned")  # dispersion check

draws <- as.array(fit_rs)
bayesplot::mcmc_trace(draws, pars = c("hill_q1", "hill_q2", "sand"))
```


