---
title: "02_clean_data"
author: Gregory B. Egloff
---
```{r load packages}
library(dplyr)
library(readr)
library(tidyr)
library(vegan)
```

```{r load data}
# Compact helper function to read raw CSVs
read_raw <- function(file) {
  read.csv(file.path("..", "data", "raw", file), stringsAsFactors = FALSE)
}

# Example usage
beedata      <- read_raw("beedata.csv")
mappingdata  <- read_raw("mappingdata.csv")
canopyhp     <- read_raw("canopyhp.csv")
```

Create linkable datasets through plotID and metadata dataframe
```{r homogenize datasets}
#add column with plotID
beedata <- beedata |>
  mutate(plotID = sub("\\.[^.]+(?=\\.[^.]+$)", "", surveyID, perl = TRUE))

mappingdata <- mappingdata |>
  mutate(plotID = sub("\\.[^.]+(?=\\.[^.]+$)", "", surveyID, perl = TRUE))

metadata <- mappingdata |>
  select(surveyID, plotID, site, region, visit, plot) |>
  distinct()
```

# Mapping data (excluding plants)
Prepare mapping data (without plants)
```{r mapp_data, habplot, habsite prep}
mapp_data <- mappingdata

# add canopy hp data
mapp_data <- mapp_data |>
  left_join(canopyhp |> select(plotID, canopy_hp), by = "plotID")

# Subset the data frame to include plot scale mapping data
# (not including plant data)
columns_plot <- c("plotID", "plot", "site", "region", "Xcoord", "Ycoord",
                  "Zcoord", "canton", "date", "visit", "surveyID", "observer",
                  "municipality", "bareground", "vegcover", "vegheight", "herb",
                  "woody", "sand", "gravel", "cobble", "stone", "wood", "soil",
                  "flat", "sloped", "steep", "canopy", "canopy_hp",
                  "standingdeadwood", "lyingdeadwood", "branchpile",
                  "stonepile", "patchiness")

habplot_data <- mapp_data[, columns_plot]

# Subset the data frame to include site scale mapping data
# (including site scale plant data)
# NOTE: site scale plant data not used, as this was a visual survey of
#     of flowering woody vegetation in the site, not a mapping of all flowering
#     plant genuses

columns_site <- c("plotID", "plot", "site", "region", "Xcoord", "Ycoord",
                  "Zcoord", "canton", "date", "visit", "surveyID", "observer",
                  "municipality", "sitebareground", "sitevegcover",
                  "sitevegheight", "siteherb", "sitewoody", "sitesand",
                  "sitegravel", "sitecobble", "sitestone", "sitesoil",
                  "siteflat", "sitesloped", "sitesteep", "sitecanopy",
                  "sitepatchiness", "sitecornis", "sitecomment", "siteacer",
                  "sitecornus", "sitecrataegus", "siteeuonymus", "sitefrangula",
                  "sitefraxinus", "ligustrum", "sitelonicera", "siteprunus",
                  "siterosa", "sitesalix", "sitesambucus", "siteviburnum",
                  "plotID")

# Subset the data frame
habsite_data <- mapp_data[, columns_site]


# calculate proportion of each bareground variable to full plot area
cols <- c("sand", "gravel", "cobble", "stone", "wood", "soil")
mapp_data[cols] <- mapp_data[cols] * mapp_data$bareground / 100
```

Prepare sampling covariate data
```{r sampling covariates data prep}
# Subset the data frame to include sampling covariates
columns_sampling <- c("site", "region", "visit", "plotID", "surveyID", "date",
                      "site", "temp", "wind", "cloud", "shade", "distwater",
                      "timestart", "timestop")

samplcov_data <- mapp_data[, columns_sampling]
```

# Plant data
Prepare plant genus matrix
```{r plant data prep}
plant_data <- mapp_data[, which(names(mapp_data) == "achillea")
                        : which(names(mapp_data) == "vicia")]

# replace NA with 0
plant_data[is.na(plant_data)] <- 0

# Calculate plant richness: number of genera present per row
plant_data$richness <- rowSums(plant_data > 0, na.rm = TRUE)

# Calculate relative abundance: sum of all genus abundances per row
plant_data$relabundance <- rowSums(plant_data, na.rm = TRUE)

#Calculate the avg. relative abundance per plant genus per plot
plant_data$avgrelabundance <- plant_data$relabundance / plant_data$richness

plant_data$site <- mapp_data$site

plant_data$visit <- mapp_data$visit

plant_data$region <- mapp_data$region

plant_data$plotID <- mapp_data$plotID

plant_data$surveyID <- mapp_data$surveyID
```

```{r plant diversity indices}
# Hill number for q = 0 (species richness)
plant_data$hill_q0 <- rowSums(plant_data[1:131] > 0, na.rm = TRUE)

# Hill number for q = 1 (exponential of Shannon)
plant_data$hill_q1 <- exp(diversity(plant_data[1:131], index = "shannon"))

# Hill number for q = 2 (inverse Simpson)
plant_data$hill_q2 <- diversity(plant_data[1:131], index = "invsimpson")
# Plots with 0 plant return Inf - convert to NA for interpretability
plant_data$hill_q2 <- ifelse(plant_data$hill_q0 == 0, NA, plant_data$hill_q2)

# Current plant_data has metadata at the end; let's reorder
plant_data <- plant_data[, c("site", "region", "visit", "plotID",
                             "surveyID", "hill_q0", "hill_q1", "hill_q2",
                             setdiff(names(plant_data),
                                     c("site", "visit", "plotID", "region",
                                       "surveyID", "hill_q0", "hill_q1",
                                       "hill_q2")))]
```

# Bee data
```{r clean bee data}
# PREP BEE DATA
# remove extra plots (plots #13) and Wasps
beedata <- beedata[!(beedata$plot == 13 | beedata$Gattung == "Wasp"), ]

# remove "cf." from species names
# (these are counted as their respective species ID in this analysis)
beedata <- beedata |> mutate(Art = gsub("^cf\\.\\s*", "", Art))

# Combine genus and species into a single species name
beedata <- beedata |>
  mutate(bee_species = paste(Gattung, Art, sep = " "))


# Count number of individuals per species per surveyID
bee_abundance_matrix <- beedata |>
  group_by(surveyID, bee_species) |>
  summarise(bee_abundance = n(), .groups = "drop") |>
  pivot_wider(names_from = bee_species,
              values_from = bee_abundance, values_fill = 0)

# Reorder columns: keep 'surveyID' first, sort the rest alphabetically
bee_abundance_matrix <- bee_abundance_matrix[, c(
  "surveyID",
  sort(setdiff(names(bee_abundance_matrix), "surveyID"))
)]

# check if teh number of rows corresponds to sum caluclated in excel
nrow(beedata) # 1871 bees total
sum(bee_abundance_matrix[, -1])

# Merge metadata with species abundance matrix
bee_data <- left_join(metadata, bee_abundance_matrix, by = "surveyID")

# turn NA into 0 in species matrix
bee_data[is.na(bee_data)] <- 0

# add field ID morphospecies of Bombus and Xylocopa (only for abundance data)
bee_data <- left_join(bee_data, mapp_data |>
                        select(surveyID, fieldIDsum), by = "surveyID")

# Handle bumbleebee aggregates
halictus_species <- c("Halictus simplex", "Halictus eurygnathus",
                      "Halictus langobardicus")
bombus_t_species <- c("Bombus terrestris", "Bombus cryptarum")
bombus_h_species <- c("Bombus hortorum", "Bombus ruderatus")
halictus_aggr    <- "Halictus simplex aggr."
bombus_t_aggr    <- "Bombus terrestris aggr."
bombus_h_aggr    <- "Bombus hortorum aggr."

# Diagnostic loop
for (i in 1:nrow(bee_data)) {
  row <- bee_data[i, ]
  halictus_present        <- any(row[halictus_species] > 0)
  halictus_aggr_present   <- row[[halictus_aggr]] > 0
  bombus_t_present        <- any(row[bombus_t_species] > 0)
  bombus_t_aggr_present   <- row[[bombus_t_aggr]] > 0
  bombus_h_present        <- any(row[bombus_h_species] > 0)
  bombus_h_aggr_present   <- row[[bombus_h_aggr]] > 0
  if ((halictus_present && halictus_aggr_present) ||
      (bombus_t_present && bombus_t_aggr_present) ||
      (bombus_h_present && bombus_h_aggr_present)) {
    cat("\n--- Row", i, "surveyID:", row$surveyID, "---\n")
    if (halictus_present && halictus_aggr_present) {
      cat("Halictus species present:\n"); print(row[halictus_species])
      cat("Halictus aggregate present:\n"); print(row[halictus_aggr])
    }
    if (bombus_t_present && bombus_t_aggr_present) {
      cat("Bombus terrestris species present:\n"); print(row[bombus_t_species])
      cat("Bombus terrestris aggregate present:\n"); print(row[bombus_t_aggr])
    }
    if (bombus_h_present && bombus_h_aggr_present) {
      cat("Bombus hortorum species present:\n"); print(row[bombus_h_species])
      cat("Bombus hortorum aggregate present:\n"); print(row[bombus_h_aggr])
    }
  }
}

sum(bee_data$`Halictus simplex aggr.` > 0)
sum(bee_data$`Bombus terrestris aggr.` > 0)
sum(bee_data$`Bombus hortorum aggr.` > 0)

# Collapse aggregates
bee_data[[halictus_aggr]] <- bee_data[[halictus_aggr]] +
  rowSums(bee_data[halictus_species], na.rm = TRUE)
bee_data[halictus_species] <- NULL

bee_data[[bombus_t_aggr]] <- bee_data[[bombus_t_aggr]] +
  rowSums(bee_data[bombus_t_species], na.rm = TRUE)
bee_data[bombus_t_species] <- NULL

bee_data[[bombus_h_aggr]] <- bee_data[[bombus_h_aggr]] +
  rowSums(bee_data[bombus_h_species], na.rm = TRUE)
bee_data[bombus_h_species] <- NULL
```


should we limit the count of each species per plot above 10, 15, 20? - not for now
```{r exploring bee data}
# Convert to long format
bee_abundance_long <- bee_abundance_matrix |>
  pivot_longer(cols = -surveyID, names_to = "species", values_to = "abundance")

bee_abundance_long |> filter(abundance > 10)

# Filter for abundance > 15
bee_abundance_long |> filter(abundance > 15)

# Filter for abundance > 20
bee_abundance_long |> filter(abundance > 20)
```

Compute Bee diversity values
```{r bee diversity indices}
bee_sp_range <-
  which(names(bee_data) == "Andrena aeneiventris"):
  which(names(bee_data) == "Stelis punctulatissima")

# Richness
bee_data$bee_richness <-
  rowSums(bee_data[, bee_sp_range] > 0, na.rm = TRUE)

# Abundance without fieldID
bee_data$bee_abundance_nofieldID <-
  rowSums(bee_data[, bee_sp_range], na.rm = TRUE)

# Abundance with fieldID
bee_data$bee_abundance <-
  bee_data$bee_abundance_nofieldID + bee_data$fieldIDsum

# Hill number for q = 1 (exponential of Shannon)
bee_data$bee_hill_q1 <-
  exp(diversity(bee_data[, bee_sp_range], index = "shannon"))

# Hill number for q = 2 (inverse Simpson)
bee_data$bee_hill_q2 <-
  diversity(bee_data[, bee_sp_range], index = "invsimpson")

# Plots with 0 bees return Inf - convert to NA for interpretability
bee_data$bee_hill_q2 <-
  ifelse(bee_data$bee_richness == 0, NA, bee_data$bee_hill_q2)
```

Compute Bee Red List Metrics
```{r Bee Red List Metrics}
# RL METRICS
rl_abundance <- beedata |>
  filter(!is.na(RLstatus)) |>
  group_by(surveyID, RLstatus) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(names_from = RLstatus, values_from = n,
              values_fill = 0, names_prefix = "RLabund_")

rl_richness <- beedata |>
  filter(!is.na(RLstatus)) |>
  mutate(bee_species = paste(Gattung, Art)) |>
  distinct(surveyID, RLstatus, bee_species) |>
  count(surveyID, RLstatus) |>
  pivot_wider(names_from = RLstatus, values_from = n,
              values_fill = 0, names_prefix = "RLrich_")

bee_data <- bee_data |>
  left_join(rl_abundance, by = "surveyID") |>
  left_join(rl_richness, by = "surveyID")

for (col in c("RLrich_VU", "RLrich_EN", "RLrich_CR")) {
  if (!col %in% names(bee_data)) bee_data[[col]] <- 0
}
for (col in c("RLabund_VU", "RLabund_EN", "RLabund_CR")) {
  if (!col %in% names(bee_data)) bee_data[[col]] <- 0
}

bee_data <- bee_data |>
  mutate(
    RLrich_threatened  = RLrich_VU + RLrich_EN + RLrich_CR,
    RLabund_threatened = RLabund_VU + RLabund_EN + RLabund_CR,
    RLrich_total       = rowSums(across(starts_with("RLrich_")),  na.rm = TRUE),
    RLabund_total      = rowSums(across(starts_with("RLabund_")), na.rm = TRUE)
  )
```





# Combine model data
Create full dataset for modelling, each plot regarded separately
```{r bee model data prep: data_all prep}
data_all <- merge(habplot_data, plant_data, by = "surveyID")
data_all <- merge(data_all, samplcov_data, by = "surveyID")
data_all <- merge(data_all, bee_data, by = "surveyID")


# Remove duplicated columns manually
data_all <- data_all |>
  select(-matches("\\.x$|\\.y$"))  # removes columns ending in .x or .y

# Add plot info from mapp_data
mapp_subset <- mapp_data |>
  select(surveyID, site, region, plot, plotID, visit)

# Join with data_all by surveyID
data_all <- left_join(mapp_subset, data_all, by = "surveyID")


data_all <- data_all |>
  mutate(
    site   = factor(site, levels = c("control", "focal")),
    region = factor(region),
    plotID = factor(plotID)
  )


View(data_all)
```

Scale data 
```{r bee model data prep: scale data}
data_all_scaled <- data_all

# Identify continuous predictors
scale_vars <- c("bareground", "vegheight", "herb", "sand",
                "gravel", "cobble", "stone", "wood", "soil", "sloped",
                "steep", "canopy", "branchpile",
                "stonepile", "standingdeadwood", "lyingdeadwood",
                "hill_q0", "hill_q1", "hill_q2")

# scale
data_all_scaled <- data_all |>
  mutate(across(all_of(scale_vars), scale))
```

scale patchiness and visit? no, they are ordinal variables
```{r bee model data prep: handling non-numeric covariates}
#creat ordinal patchiness and visit columns
data_all_scaled <- data_all_scaled |>
  mutate(patchiness_ord = ordered(patchiness, levels = c("0", "1", "2", "3")),
         visit_ord = ordered(visit, levels = c("1", "2", "3", "4", "5")))
```

# Save clean survey-level dataframes
```{r save dataframes}
# Save as CSV function
save_csv <- function(df, name) {
  out_dir <- file.path("..", "data", "cleaned")
  # Create cleaned folder if missing
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  readr::write_csv(
    as.data.frame(df),
    file.path(out_dir, paste0(name, ".csv"))
    )
}

# Save scaled modeelling data containing list columns as RDS
save_rds <- function(obj, name) {
  out_dir <- file.path("..", "data", "cleaned")
  # Create cleaned folder if missing
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  saveRDS(
    obj,
    file.path(out_dir, paste0(name, ".rds"))
  )
}

save_csv(mapp_data, "mapp_data")
save_csv(metadata, "metadata")
save_csv(habplot_data, "habplot_data")
save_csv(habsite_data, "habsite_data")
save_csv(plant_data, "plant_data")
save_csv(samplcov_data, "samplcov_data")
save_csv(bee_data, "bee_data")
save_csv(data_all, "data_all")

save_rds(data_all_scaled, "data_all_scaled")
```


# Merge data across visits for each plot

Bee data
```{r bee data merged across visits per plot}
bee_plot <- bee_data |>
  group_by(region, site, plotID) |>
  summarise(
    across(all_of(names(bee_data)[bee_sp_range]), sum, na.rm = TRUE),
    bee_abundance = sum(bee_abundance, na.rm = TRUE),
    .groups = "drop"
  )

species_cols <- setdiff(
  colnames(bee_plot),
  c("region", "site", "plotID", "bee_abundance")
)

# compute species richness
bee_plot <- bee_plot |>
  mutate(
    bee_richness = rowSums(across(all_of(species_cols)) > 0)
  )

# compute bee hill q1 and hill q2
bee_plot <- bee_plot |>
  mutate(
    bee_hill_q1 = exp(diversity(
      as.matrix(select(., all_of(species_cols))),
      index = "shannon")),
    bee_hill_q2 = diversity(
      as.matrix(select(., all_of(species_cols))),
      index = "invsimpson")
    )
```

```{r merge habitat and plant vairables across visits per plot}
# define mode function for patchiness
mode_fun <- function(x) {
  ux <- na.omit(unique(x))
  ux[which.max(tabulate(match(x, ux)))]
}

# aggregate habitat data
hab_plot <- habplot_data |>
  group_by(region, site, plotID) |>
  summarise(
    across(
      where(is.numeric) & !all_of("patchiness"),
      \(x) mean(x, na.rm = TRUE)
    ), # instead of averaging patchiness, extract mode value (most common per plot)
    patchiness_mode = mode_fun(patchiness),
    patchiness_sd = sd(patchiness, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    patchiness_ord = ordered(
      patchiness_mode,
      levels = c(0, 1, 2, 3)
    )
  )


# group plant diversity metrics across visits per plot
plant_plot <- plant_data |>
  group_by(region, site, plotID) |>
  summarise(
    hill_q0 = mean(hill_q0, na.rm = TRUE),
    hill_q1 = mean(hill_q1, na.rm = TRUE),
    hill_q2 = mean(hill_q2, na.rm = TRUE),
    .groups = "drop"
  )

# merge dataframes

data_plot <- hab_plot |>
  left_join(plant_plot, by = c("region", "site", "plotID")) |>
  left_join(bee_plot,   by = c("region", "site", "plotID"))

data_plot <- data_plot |>
  mutate(
    site   = factor(site, levels = c("control", "focal")),
    region = factor(region),
    plotID = factor(plotID)
  )

```

```{r scale continuous vairables}
# scale continuous variables
scale_vars <- c("bareground", "vegheight", "herb", "sand",
                "gravel", "cobble", "stone", "wood", "soil", "sloped",
                "steep", "canopy", "branchpile", "stonepile",
                "standingdeadwood", "patchiness_sd","lyingdeadwood",
                "hill_q0", "hill_q1", "hill_q2")

data_plot_scaled <- data_plot |>
  mutate(across(all_of(scale_vars), scale))```
```

# Save clean plot-level dataframes
```{r}
write.csv(data_plot,
          file = file.path("..", "data", "cleaned", "data_plot.csv"),
          row.names = FALSE)

saveRDS(data_plot_scaled,
        file = file.path("..", "data", "cleaned", "data_plot_scaled.rds"))
```

