---
title: "clean_data"
output: html_document
---
```{r}
library(dplyr)
library(readr)
library(rprojroot)
library(tidyr)
library(vegan)
```

```{r Load data}
# Compact helper function to read raw CSVs
read_raw <- function(file) {
  read.csv(file.path("..", "data", "raw", file), stringsAsFactors = FALSE)
}

# Example usage
beedata      <- read_raw("beedata.csv")
mappingdata  <- read_raw("mappingdata.csv")
canopyhp     <- read_raw("canopyhp.csv")
```

Create linkable datasets through plotID and metadata dataframe
```{r}
#add column with plotID
beedata <- beedata |>
  mutate(plotID = sub("\\.[^.]+(?=\\.[^.]+$)", "", surveyID, perl = TRUE))

mappingdata <- mappingdata |>
  mutate(plotID = sub("\\.[^.]+(?=\\.[^.]+$)", "", surveyID, perl = TRUE))

metadata <- mappingdata |>
  select(surveyID, plotID, site, region, visit, plot) |>
  distinct()
```

# Mapping data (excluding plants)
Prepare mapping data (without plants)
```{r mapp_data, habplot, habsite prep}
mapp_data <- mappingdata

# add canopy hp data
mapp_data <- mapp_data |>
  left_join(canopyhp |> select(plotID, canopy_hp), by = "plotID")

# Subset the data frame to include plot scale mapping data
# (not including plant data)
columns_plot <- c("plotID", "plot", "site", "region", "Xcoord", "Ycoord",
                  "Zcoord", "canton", "date", "visit", "surveyID", "observer",
                  "municipality", "bareground", "vegcover", "vegheight", "herb",
                  "woody", "sand", "gravel", "cobble", "stone", "wood", "soil",
                  "flat", "sloped", "steep", "canopy", "canopy_hp",
                  "standingdeadwood", "lyingdeadwood", "branchpile",
                  "stonepile", "patchiness")

habplot_data <- mapp_data[, columns_plot]

# Subset the data frame to include site scale mapping data
# (including site scale plant data)
columns_site <- c("plotID", "plot", "site", "region", "Xcoord", "Ycoord",
                  "Zcoord", "canton", "date", "visit", "surveyID", "observer",
                  "municipality", "sitebareground", "sitevegcover",
                  "sitevegheight", "siteherb", "sitewoody", "sitesand",
                  "sitegravel", "sitecobble", "sitestone", "sitesoil",
                  "siteflat", "sitesloped", "sitesteep", "sitecanopy",
                  "sitepatchiness", "sitecornis", "sitecomment", "siteacer",
                  "sitecornus", "sitecrataegus", "siteeuonymus", "sitefrangula",
                  "sitefraxinus", "ligustrum", "sitelonicera", "siteprunus",
                  "siterosa", "sitesalix", "sitesambucus", "siteviburnum",
                  "plotID")

# Subset the data frame
habsite_data <- mapp_data[, columns_site]


# calculate proportion of each bareground variable to full plot area
cols <- c("sand", "gravel", "cobble", "stone", "wood", "soil")
mapp_data[cols] <- mapp_data[cols] * mapp_data$bareground / 100
```

Prepare sampling covariate data
```{r sampling covariates data prep}
# Subset the data frame to include sampling covariates
columns_sampling <- c("site", "region", "visit", "plotID", "surveyID", "date",
                      "site", "temp", "wind", "cloud", "shade", "distwater",
                      "timestart", "timestop")

samplcov_data <- mapp_data[, columns_sampling]
```

# Plant data
Prepare plant genus matrix
```{r plant data prep}
plant_data <- mapp_data[, which(names(mapp_data) == "achillea")
                        :which(names(mapp_data) == "vicia")]

# replace NA with 0
plant_data[is.na(plant_data)] <- 0

# Calculate plant richness: number of genera present per row
plant_data$richness <- rowSums(plant_data > 0, na.rm = TRUE)

# Calculate relative abundance: sum of all genus abundances per row
plant_data$relabundance <- rowSums(plant_data, na.rm = TRUE)

#Calculate the avg. relative abundance per plant genus per plot
plant_data$avgrelabundance <- plant_data$relabundance / plant_data$richness

plant_data$site <- mapp_data$site

plant_data$visit <- mapp_data$visit

plant_data$region <- mapp_data$region

plant_data$plotID <- mapp_data$plotID

plant_data$surveyID <- mapp_data$surveyID
```

```{r plant diversity indices}
# Hill number for q = 0 (species richness)
plant_data$hill_q0 <- rowSums(plant_data[1:131] > 0, na.rm = TRUE)

# Hill number for q = 1 (exponential of Shannon)
plant_data$hill_q1 <- exp(diversity(plant_data[1:131], index = "shannon"))

# Hill number for q = 2 (inverse Simpson)
plant_data$hill_q2 <- 1 / diversity(plant_data[1:131], index = "invsimpson")

# Current plant_data has metadata at the end; let's reorder
plant_data <- plant_data[, c("site", "region", "visit", "plotID",
                             "surveyID", "hill_q0", "hill_q1", "hill_q2",
                             setdiff(names(plant_data),
                                     c("site", "visit", "plotID", "region",
                                       "surveyID", "hill_q0", "hill_q1",
                                       "hill_q2")))]
```

#Bee data
Prepare bee data
```{r bee data prep}
# remove extra plots (plots #13) and Wasps
beedata <- beedata[!(beedata$plot == 13 | beedata$Gattung == "Wasp"), ]

# Combine genus and species into a single species name
beedata <- beedata |>
  mutate(bee_species = paste(Gattung, Art, sep = " "))


# Count number of individuals per species per surveyID
bee_abundance_matrix <- beedata |>
  group_by(surveyID, bee_species) |>
  summarise(bee_abundance = n(), .groups = "drop") |>
  pivot_wider(names_from = bee_species,
              values_from = bee_abundance, values_fill = 0)

# Reorder columns: keep 'surveyID' first, sort the rest alphabetically
bee_abundance_matrix <- bee_abundance_matrix[, c(
  "surveyID",
  sort(setdiff(names(bee_abundance_matrix), "surveyID"))
)]

# check if right number of rows
nrow(beedata) # 1871 bees total
sum(bee_abundance_matrix[, -1])

# Merge metadata with species abundance matrix
bee_data <- left_join(metadata, bee_abundance_matrix, by = "surveyID")

str(bee_data)

#add field ID morphospecies of Bombus and Xylocopa (only for abundance data
bee_data <- left_join(bee_data, mapp_data |>
                        select(surveyID, fieldIDsum), by = "surveyID")

#Calculate bee species richness per plot
bee_data$bee_richness <- rowSums(
  bee_data[, !(names(bee_data) %in%
                 c("surveyID", "plotID", "site", "region",
                   "visit", "plot", "fieldIDsum"))] > 0, na.rm = TRUE)

# Calculate total abundance by summing all species columns
# (not including fieldID Bombus / Xylocopa)
bee_data$bee_abundance_nofieldID <- rowSums(
  bee_data[, !(names(bee_data) %in%
                 c("surveyID", "plotID", "site", "region", "visit", "plot",
                   "fieldIDsum", "bee_richness"))], na.rm = TRUE)

# Calculate total abundance by summing all species columns (including FieldID)
bee_data$bee_abundance <- rowSums(
  bee_data[, !(names(bee_data) %in%
                 c("surveyID", "plotID", "site", "region", "visit", "plot",
                   "bee_abundance_nofieldID", "bee_richness"))], na.rm = TRUE)

# RL Abundance per survey (drop NA RLstatus)
rl_abundance <- beedata |>
  filter(!is.na(RLstatus)) |>
  group_by(surveyID, RLstatus) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(
    names_from = RLstatus,
    values_from = n,
    values_fill = 0,
    names_prefix = "RLabund_"
  )

# RL Richness per survey (drop NA RLstatus)
rl_richness <- beedata |>
  filter(!is.na(RLstatus)) |>
  mutate(bee_species = paste(Gattung, Art)) |>
  distinct(surveyID, RLstatus, bee_species) |>
  count(surveyID, RLstatus) |>
  pivot_wider(
    names_from = RLstatus,
    values_from = n,
    values_fill = 0,
    names_prefix = "RLrich_"
  )

# Merge RL metrics into bee_data
bee_data <- bee_data |>
  left_join(rl_abundance, by = "surveyID") |>
  left_join(rl_richness, by = "surveyID")

# Ensure threatened columns exist
for (col in c("RLrich_VU", "RLrich_EN", "RLrich_CR")) {
  if (!col %in% names(bee_data)) bee_data[[col]] <- 0
}
for (col in c("RLabund_VU", "RLabund_EN", "RLabund_CR")) {
  if (!col %in% names(bee_data)) bee_data[[col]] <- 0
}

# Add threatened and total richness/abundance
bee_data <- bee_data |>
  mutate(
    RLrich_threatened = RLrich_VU + RLrich_EN + RLrich_CR,
    RLabund_threatened = RLabund_VU + RLabund_EN + RLabund_CR,
    RLrich_total = rowSums(across(starts_with("RLrich_")), na.rm = TRUE),
    RLabund_total = rowSums(across(starts_with("RLabund_")), na.rm = TRUE)
  )

# turn NA into 0 in species matrix
bee_data[is.na(bee_data)] <- 0
```

```{r handle aggregates}

# Define species within aggregates
halictus_species <- c("Halictus simplex",
                      "Halictus eurygnathus",
                      "Halictus langobardicus")

bombus_species <- c("Bombus terrestris",
                    "Bombus cryptarum")

# Define aggregate column names
halictus_aggr <- "Halictus simplex aggr."
bombus_aggr <- "Bombus terrestris aggr."

# Loop through rows and print affected ones
for (i in 1:nrow(bee_data)) {
  row <- bee_data[i, ]
  halictus_present <- any(row[halictus_species] > 0)
  halictus_aggr_present <- row[[halictus_aggr]] > 0
  bombus_present <- any(row[bombus_species] > 0)
  bombus_aggr_present <- row[[bombus_aggr]] > 0
  if ((halictus_present && halictus_aggr_present) ||
      (bombus_present && bombus_aggr_present)) {
    cat("\n--- Row", i, "surveyID:", row$surveyID, "---\n")
    # Print Halictus species and aggregate if relevant
    if (halictus_present & halictus_aggr_present) {
      cat("Halictus species present:\n")
      print(row[halictus_species])
      cat("Halictus aggregate present:\n")
      print(row[halictus_aggr])
    }
    # Print Bombus species and aggregate if relevant
    if (bombus_present & bombus_aggr_present) {
      cat("Bombus species present:\n")
      print(row[bombus_species])
      cat("Bombus aggregate present:\n")
      print(row[bombus_aggr])
    }
  }
}

sum(bee_data$`Halictus simplex aggr.` > 0)

sum(bee_data$`Bombus terrestris aggr.` > 0)


# merge all species within the aggregates 
# Collapse Halictus
bee_data[[halictus_aggr]] <- bee_data[[halictus_aggr]] +
  rowSums(bee_data[halictus_species], na.rm = TRUE)
bee_data[halictus_species] <- NULL   # remove species-level columns

# Collapse Bombus
bee_data[[bombus_aggr]] <- bee_data[[bombus_aggr]] +
  rowSums(bee_data[bombus_species], na.rm = TRUE)
bee_data[bombus_species] <- NULL     # remove species-level columns

```


should we limit the count of each species per plot above 10, 15, 20?

```{r exploring bee data}
  # Convert to long format
  bee_abundance_long <- bee_abundance_matrix |>
    pivot_longer(cols = -surveyID, names_to = "species", values_to = "abundance")
  
  bee_abundance_long |> filter(abundance > 10)
  
  # Filter for abundance > 15
  bee_abundance_long |> filter(abundance > 15)
  
  # Filter for abundance > 20
  bee_abundance_long |> filter(abundance > 20)
```

# Model data
Create full dataset for modelling
```{r bee model data prep: data_all prep}
data_all <- merge(habplot_data, plant_data, by = "surveyID")
data_all <- merge(data_all, samplcov_data, by = "surveyID")
data_all <- merge(data_all, bee_data, by = "surveyID")


# Remove duplicated columns manually
data_all <- data_all |>
  select(-matches("\\.x$|\\.y$"))  # removes columns ending in .x or .y

# Add plot info from mapp_data
mapp_subset <- mapp_data |>
  select(surveyID, site, region, plot, plotID, visit)

# Join with data_all by surveyID
data_all <- left_join(mapp_subset, data_all, by = "surveyID")


data_all <- data_all |>
  mutate(
    site   = factor(site, levels = c("control", "focal")),
    region = factor(region),
    plotID = factor(plotID)
  )


View(data_all)
```

Scale data 
```{r bee model data prep: scale data}
data_all_scaled <- data_all

# Identify continuous predictors
scale_vars <- c("bareground", "vegheight", "herb", "sand",
              "gravel", "cobble", "stone", "wood", "soil", "sloped",
              "steep", "canopy", "branchpile",
              "stonepile", "standingdeadwood", "lyingdeadwood", "hill_q0", "hill_q1", "hill_q2")

# Scale them (center and scale)
#data_all_scaled <- data_all_scaled |>
#  mutate(across(all_of(scale_vars), ~ as.numeric(scale(.))))


data_all_scaled <- data_all |>
  mutate(across(all_of(scale_vars), scale)) 

```
scale patchiness and visit? no, they are ordinal variables

```{r bee model data prep: handling non-numeric covariates}
#creat ordinal patchiness and visit columns
data_all_scaled <- data_all_scaled |>
  mutate(patchiness_ord = ordered(patchiness, levels = c("0","1","2","3")), 
         visit_ord = ordered(visit, levels = c("1","2","3","4","5")))
```


```{r bee model data prep: save model dataframe}
saveRDS(data_all_scaled, file = "data_all_scaled.rds")
```

"We included random intercepts for region and plot to account for hierarchical structure and repeated measures. Region controls for large-scale environmental differences among study areas, while plot accounts for non-independence of repeated surveys within the same plot. The effect of site (focal vs control) was modeled as a fixed effect because it represents the main ecological contrast of interest and has only two levels, making it unsuitable as a random effect." - GPT formulation for use of random effect structure (1|region) + (1|plotID)

Abundance modelling - test for overdispersion / zero inflation


# Save clean dataframes
```{r}

# Save flat datasets as CSV
save_csv <- function(df, name) {
  # Find project root
   # Create cleaned folder if missing
  out_dir <- file.path("..", "data", "cleaned")
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  
  # Save as CSV
  readr::write_csv(
    as.data.frame(df),
    file.path(out_dir, paste0(name, ".csv"))
  )
}

# Save complex datasets (list-columns, modelling data) as RDS
save_rds <- function(obj, name) {
  out_dir <- file.path("..", "data", "cleaned")
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  
  saveRDS(
    obj,
    file.path(out_dir, paste0(name, ".rds"))
  )
}

# Save core datasets (flat tables)
save_csv(mapp_data, "mapp_data")
save_csv(metadata, "metadata")
save_csv(habplot_data, "habplot_data")
save_csv(habsite_data, "habsite_data")
save_csv(plant_data, "plant_data")
save_csv(samplcov_data, "samplcov_data")
save_csv(bee_data, "bee_data")
save_csv(data_all, "data_all")

# Save modelling dataset (contain list-columns)
save_rds(data_all_scaled, "data_all_scaled")
```

