---
title: "04_bee_rarefaction"
output: html_document
---

```{r load packages}
library(emmeans)
library(gt)
library(iNEXT)
library(dplyr)
library(tidyr)
library(stringr)
library(ggeffects)
library(ggplot2)
library(glmmTMB)
library(performance)
```

Coverage‑based rarefaction (Chao et al. 2014, 2020) provides descriptive standardization, not inferential estimates, and its resulting diversity values do not follow the distributional assumptions required for GLMMs (e.g. Gotelli & Colwell 2001; Chao & Jost 2012). Rarefaction also collapses plot‑level replication, removes ecological variance, and prevents correct estimation of random effects.
Because our sampling design includes spatial and hierarchical structure (multiple visits per plot; plots nested within sites; sites nested within regions), a GLMM using plot‑level richness/abundance with region and plotID as random effects is the statistically appropriate inferential framework.

```{r Load data}
# Compact helper function to read raw CSVs
bee_data <- read.csv(file.path("..", "data", "cleaned", "bee_data.csv"),
           stringsAsFactors = FALSE,
           check.names = FALSE)
```

```{r colorblind friendly approach}

# Identify metadata and species columns
meta_cols <- c(
  "surveyID", "plotID", "site", "region", "visit", "plot",
  "bee_richness", "bee_hill_q1", "bee_hill_q2", "fieldIDsum",
  "bee_abundance_nofieldID", "bee_abundance",
  "RLabund_LC", "RLabund_NT", "RLabund_VU", "RLabund_EN", "RLabund_DD",
  "RLrich_LC", "RLrich_NT", "RLrich_VU", "RLrich_EN", "RLrich_DD",
  "RLrich_CR", "RLabund_CR",
  "RLrich_threatened", "RLabund_threatened",
  "RLrich_total", "RLabund_total"
)

species_cols <- setdiff(names(bee_data), meta_cols)

```

# Site-level rarefaction by region
```{r rarefy by site per region}
# Aggregate all plots per site
bee_site <- bee_data %>%
  group_by(region, site) %>%
  summarise(across(all_of(species_cols), sum), .groups = "drop")


# Prepare iNEXT input list (site-level)
inext_site <- setNames(
  lapply(seq_len(nrow(bee_site)), function(i) {
    as.numeric(bee_site[i, species_cols])
  }),
  paste(bee_site$region, bee_site$site, sep = "_")
)


# Run iNEXT for abundance data (q = 0 richness)
out_site <- iNEXT(
  inext_site,
  q = 0,
  datatype = "abundance",
  endpoint = max(sapply(inext_site, sum)) * 2
)
```

# Coverage-based rarefaction site scale per region
```{r coverage based rarefaction site scale per region}
# Extract coverage-based rarefaction data
df_site <- fortify(out_site, type = 3)


# Recover region and site labels
df_site <- df_site %>%
  separate(Assemblage, into = c("region", "site"), sep = "_")


# Plot coverage-based rarefaction curves (SITE level)
p_covraref_site <- ggplot(df_site,
                          aes(x = x, y = y, color = site, fill = site)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr),
              alpha = 0.25, color = NA) +
  facet_wrap(~ region) +
  labs(
    x = "Sample coverage",
    y = "Estimated species richness",
    title = "Coverage-based rarefaction by site",
    color = "Site",
    fill = "Site"
  ) +
  scale_color_manual(values = c("focal" = "#2C7FB8", "control" = "#D95F02")) +
  scale_fill_manual(values = c("focal" = "#A6C1E0", "control" = "#F1B0AD")) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

p_covraref_site
```

```{r plot coverage based rarefaction by site and region}
# Extract transition points (Observed)
transition_pts <- df_site %>%
  filter(Method == "Observed")

# Plot
p_covraref_site <- ggplot(df_site,
                          aes(x = x, y = y, color = site, fill = site)) +
  
  # Lines with linetype mapping
  geom_line(aes(linetype = Method), linewidth = 1.2) +
  
  # Confidence ribbon
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr),
              alpha = 0.25, color = NA) +
  
  # Transition point (Observed)
  geom_point(
    data = transition_pts,
    aes(x = x, y = y),
    size = 3,
    shape = 21,
    stroke = 1,
    color = "black",
    fill = "white"
  ) +
  
  # Facet per region
  facet_wrap(~ region) +
  
  labs(
    x = "Sample coverage",
    y = "Estimated species richness",
    title = "Coverage-based rarefaction by site",
    color = "Site",
    fill = "Site",
    linetype = "Sampling segment"
  ) +
  
  # Color scales
  scale_color_manual(values = c(
    "focal" = "#2C7FB8",
    "control" = "#D95F02"
  )) +
  
  scale_fill_manual(values = c(
    "focal" = "#A6C1E0",
    "control" = "#F1B0AD"
  )) +
  
  scale_linetype_manual(
     values = c(
     "Rarefaction" = "solid", "Extrapolation" = "dashed"),
     breaks = c("Rarefaction", "Extrapolation"), # remove Observed
     name = "Sampling segment") +
  
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

p_covraref_site
```


```{r export coverage rarefaction curves as tiff}
# tiff(
#   file.path("..", "output", "figures", "cov_rarefaction_site_600dpi.tiff"),
#   width = 3000, height = 2250, res = 600, compression = "lzw"
# )
# 
# print(p_covraref_site)
# dev.off()
```

#Size-based rarefaction site scale per region
```{r}
df_site <- out_site$iNextEst$size_based

df_site <- df_site %>%
  rename(
    x = m,
    y = qD,
    y.lwr = qD.LCL,
    y.upr = qD.UCL
  ) %>%
  separate(Assemblage, into = c("region", "site"), sep = "_")

# Observed points
transition_pts <- df_site %>%
  filter(Method == "Observed")

# Plot
p_sizeraref_site <- ggplot(df_site,
                          aes(x = x, y = y, color = site, fill = site)) +
  
  geom_line(aes(linetype = Method), linewidth = 1.2) +
  
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr),
              alpha = 0.25, color = NA) +
  
  geom_point(
    data = transition_pts,
    aes(x = x, y = y),
    size = 3,
    shape = 21,
    stroke = 1,
    color = "black",
    fill = "white"
  ) +
  
  facet_wrap(~ region,
             labeller = as_labeller(c("Aire" = "Aire", 
                                      "Schwarzbach" = "Schwarzbach",
                                      "Scheulte" = "Scheulte",
                                      "Thiele" = "Thièle"))) +
  
  labs(
    x = "Number of individuals",
    y = "Estimated species richness",
    color = NULL,
    fill = NULL,
    linetype = NULL
  ) +
  
  scale_color_manual(values = c(
    "focal" = "#2C7FB8",
    "control" = "#D95F02"),
  labels = c(
    "focal" = "Focal",
    "control" = "Control")) +
  
  scale_fill_manual(values = c(
    "focal" = "#A6C1E0",
    "control" = "#F1B0AD"),
  labels = c(
    "focal" = "Focal",
    "control" = "Control")) +
  
  scale_linetype_manual(
    values = c(
      "Rarefaction" = "solid",
      "Extrapolation" = "22"
    ),
    breaks = c("Rarefaction", "Extrapolation"),
  ) +
  
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
  )

p_sizeraref_site
```

```{r export size rarefaction curves as tiff}
tiff(
  file.path("..", "output", "figures", "size_rarefaction_site_600dpi.tiff"),
  width = 3500, height = 2250, res = 600, compression = "lzw"
)

print(p_sizeraref_site)
dev.off()
```

# Pooled site-level rarefaction
```{r rarefy pooled coverage based rarefaction by site}
# Pool to two assemblages: focal and control
bee_focal_pool   <- colSums(bee_data[bee_data$site == "focal", species_cols])
bee_control_pool <- colSums(bee_data[bee_data$site == "control", species_cols])

inext_pooled <- list(
  focal   = bee_focal_pool,
  control = bee_control_pool
)

out_pooled <- iNEXT(
  inext_pooled,
  q = 0,
  datatype = "abundance",
  endpoint = max(sapply(inext_pooled, sum)) * 2
)
```

# (Pooled size-based rarefaction by site)
```{r size and coverage-based rarefaction pooled focal vs control}
# Extract size-based and coverage-based data
df_pooled_size     <- fortify(out_pooled, type = 1)
df_pooled_coverage <- fortify(out_pooled, type = 3)

# Transition points
transition_size     <- df_pooled_size     %>% filter(Method == "Observed")
transition_coverage <- df_pooled_coverage %>% filter(Method == "Observed")

# SIZE-BASED PLOT
p_raref_size <- ggplot(df_pooled_size,
                       aes(x = x, y = y, color = Assemblage, fill = Assemblage)) +
  geom_line(aes(linetype = Method), linewidth = 1.2) +
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr),
              alpha = 0.25, color = NA) +
  geom_point(
    data = transition_size,
    aes(x = x, y = y),
    size = 3, shape = 21, stroke = 1,
    color = "black", fill = "white"
  ) +
  scale_color_manual(values = c("control" = "#D95F09", "focal" = "#2C7FB8")) +
  scale_fill_manual(values  = c("control" = "#F1B0AD", "focal" = "#A6C1E0")) +
  scale_linetype_manual(
    values = c("Rarefaction" = "solid", "Extrapolation" = "dashed"),
    breaks = c("Rarefaction", "Extrapolation"),
    name   = "Sampling segment"
  ) +
  labs(
    x     = "Number of individuals",
    y     = "Estimated species richness",
    title = "Size-based rarefaction: Control vs Focal",
    color = "Site", fill = "Site"
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

# COVERAGE-BASED PLOT
p_raref_coverage <- ggplot(df_pooled_coverage,
                           aes(x = x, y = y, color = Assemblage, fill = Assemblage)) +
  geom_line(aes(linetype = Method), linewidth = 1.2) +
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr),
              alpha = 0.25, color = NA) +
  geom_point(
    data = transition_coverage,
    aes(x = x, y = y),
    size = 3, shape = 21, stroke = 1,
    color = "black", fill = "white"
  ) +
  scale_color_manual(values = c("control" = "#D95F09", "focal" = "#2C7FB8")) +
  scale_fill_manual(values  = c("control" = "#F1B0AD", "focal" = "#A6C1E0")) +
  scale_linetype_manual(
    values = c("Rarefaction" = "solid", "Extrapolation" = "dashed"),
    breaks = c("Rarefaction", "Extrapolation"),
    name   = "Sampling segment"
  ) +
  labs(
    x     = "Sample coverage",
    y     = "Estimated species richness",
    title = "Coverage-based rarefaction: Control vs Focal",
    color = "Site", fill = "Site"
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

p_raref_size
p_raref_coverage
```



# Plot-level rarefaction
```{r rarefaction}
# Identify metadata and species columns
meta_cols <- c(
  "surveyID", "plotID", "site", "region", "visit", "plot",
  "bee_richness", "fieldIDsum",
  "bee_abundance_nofieldID", "bee_abundance", "bee_hill_q1", "bee_hill_q2", 
  "RLabund_LC", "RLabund_NT", "RLabund_VU", "RLabund_EN", "RLabund_DD",
  "RLrich_LC", "RLrich_NT", "RLrich_VU", "RLrich_EN", "RLrich_DD",
  "RLrich_CR", "RLabund_CR",
  "RLrich_threatened", "RLabund_threatened",
  "RLrich_total", "RLabund_total"
)

species_cols <- setdiff(names(bee_data), meta_cols)

# Aggregate visits within each plot
bee_plot <- bee_data %>%
  group_by(region, site, plotID) %>%
  summarise(across(all_of(species_cols), sum), .groups = "drop")

# Prepare iNEXT input list (per plot)
inext_list <- setNames(
  lapply(seq_len(nrow(bee_plot)), function(i) {
    as.numeric(bee_plot[i, species_cols])
  }),
  paste(bee_plot$region, bee_plot$site, bee_plot$plotID, sep = "_")
)
```

```{r run iNext plot scale}
# Run iNEXT: species richness (q=0), abundance data
out_inext_q0 <- iNEXT(
  inext_list,
  q = 0,
  datatype = "abundance",
  endpoint = max(sapply(inext_list, sum)) * 2
)

# Run iNEXT: Shannon diversity (q=1), abundance data
out_inext_q1 <- iNEXT(
  inext_list,
  q = 1,
  datatype = "abundance",
  endpoint = max(sapply(inext_list, sum)) * 2
)
```

# Coverage-based rarefaction plot-scale (visits merged)
## Richness
```{r coverage based rarefaction plot scale (visits merged)}
df_plot <- fortify(out_inext_q0, type = 3)

df_cov_max <- df_plot %>%
  group_by(Assemblage) %>%
  summarise(max_coverage = max(x), .groups = "drop")

summary(df_cov_max$max_coverage)

quantile(df_cov_max$max_coverage,
         probs = c(0.1, 0.2, 0.3, 0.4, 0.5,
                   0.6, 0.7, 0.8, 0.9, 0.95))

ggplot(df_cov_max, aes(x = max_coverage)) +
  geom_histogram(bins = 20, fill = "grey70", color = "black") +
  geom_vline(xintercept = 0.85,
             linetype = "dashed", color = "red") +
  labs(
    title = "Maximum achievable coverage across plots",
    x = "Coverage",
    y = "Number of plots"
  ) +
  theme_bw()

# Set target coverage
target_coverage <- 0.85

# Visual check
ggiNEXT(out_inext_q0, type = 3) +
  geom_vline(xintercept = target_coverage,
             linetype = "dashed",
             color = "red") +
  theme_minimal()


# use estimateD (exact interpolation)
df_ri_rarefied <- estimateD(
  inext_list,
  q = 0,
  base = "coverage",
  level = target_coverage
)

# Prepare dataset for modelling
df_ri_rarefied <- df_ri_rarefied %>%
  separate(Assemblage,
           into = c("region", "site", "plotID"),
           sep = "_") %>%
  rename(
    richness = qD,
    richness_lwr = qD.LCL,
    richness_upr = qD.UCL,
    coverage = SC
  )
```
This shows that coverage is very high in most plots (median = 0.999) - plots are well-sampled - threshold was selected based on histogram, and model run with these values. 


```{r Coverage-based rarefaction model}
RRIcov_site <- glmmTMB(
  richness ~ site + (1 | region),
  data = df_ri_rarefied,
  family = Gamma(link = "log")
)

summary(RRIcov_site)

check_model(RRIcov_site)

# Predicted values
pred_cov_rich <- emmeans(RRIcov_site, ~ site, type = "response")
contrast(pred_cov_rich, "revpairwise")
r2(RRIcov_site)
```

Focal plots supported approximately higher bee species richness comapred to control plots. Increasing the coverage threshold led to progressively stronger differences, with focal plots supporting 2.39 times and 2.52 times higher richness at coverage levels of 0.85 and 0.90, respectively. This consistent increase indicates that differences between site types are particularly pronounced in the rarer components of the community.


2.39 higher richness in focla vs control site
Effect ≈ 2.39× higher richness
~ +139% increase -> multiplicative effect

```{r plot predicted coverage-based rarefaction }
# Predicted marginal means
pred_site_rrich_cov <- ggpredict(RRIcov_site, terms = "site", type = "fixed")

# ensure plotting order
pred_site_rrich_cov$x <- factor(pred_site_rrich_cov$x, levels = c("control", "focal"))
df_ri_rarefied$site <- factor(df_ri_rarefied$site, levels = c("control", "focal"))

# Color palette
site_cols <- list(
  "focal" = "#2C7FB8",
  "control" = "#D95F09",
  "focal_light" = "cadetblue3",
  "control_light" = "#D95"
)

# Plot
p_rrich_cov <- ggplot() +
  geom_jitter(
    data = df_ri_rarefied,
    aes(x = site, y = richness, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +
  geom_point(
    data = pred_site_rrich_cov,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +
  geom_errorbar(
    data = pred_site_rrich_cov,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +
  scale_colour_manual(values = site_cols) +
  scale_x_discrete(labels = c("control" = "Control", "focal" = "Focal")) +
  labs(
    x = NULL,
    y = "Bee richness \n (coverage-standardised)",
    title = "Predicted Bee Richness per Plot by Site (coverage-based rarefaction)",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = standardised richness at equal coverage"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  )

p_rrich_cov
```

## Hill q1
```{r Coverage-based rarefaction plot-scale (visits merged) - Hill q1}
df_plot_q1 <- fortify(out_inext_q1, type = 3)

df_cov_max_q1 <- df_plot_q1 %>%
  group_by(Assemblage) %>%
  summarise(max_coverage = max(x), .groups = "drop")

summary(df_cov_max_q1$max_coverage)

quantile(df_cov_max_q1$max_coverage,
         probs = c(0.1, 0.2, 0.3, 0.4, 0.5,
                   0.6, 0.7, 0.8, 0.9, 0.95))

ggplot(df_cov_max_q1, aes(x = max_coverage)) +
  geom_histogram(bins = 20, fill = "grey70", color = "black") +
  geom_vline(xintercept = 0.85,
             linetype = "dashed", color = "red") +
  labs(
    title = "Maximum achievable coverage across plots (Hill q1)",
    x = "Coverage",
    y = "Number of plots"
  ) +
  theme_bw()

# Set target coverage
target_coverage <- 0.85

# Visual check
ggiNEXT(out_inext_q1, type = 3) +
  geom_vline(xintercept = target_coverage,
             linetype = "dashed",
             color = "red") +
  theme_minimal()

# use estimateD (exact interpolation)
df_q1_rarefied <- estimateD(
  inext_list,
  q = 1,
  datatype = "abundance",
  base = "coverage",
  level = target_coverage
)

# Prepare dataset for modelling
df_q1_rarefied <- df_q1_rarefied %>%
  separate(Assemblage,
           into = c("region", "site", "plotID"),
           sep = "_") %>%
  rename(
    hill_q1 = qD,
    hill_q1_lwr = qD.LCL,
    hill_q1_upr = qD.UCL,
    coverage = SC
  )

# check how many plots you KEEP now
length(unique(df_q1_rarefied$plotID))
```
```{r Coverage-based rarefaction model (Hill q1)}
RQ1cov_site <- glmmTMB(
  hill_q1 ~ site + (1 | region),
  data = df_q1_rarefied,
  family = gaussian()
)

summary(RQ1cov_site)

check_model(RQ1cov_site)

# Predicted values
pred_cov_q1 <- emmeans(RQ1cov_site, ~ site, type = "response")
contrast(pred_cov_q1, "revpairwise")
```

```{r Plot predicted Hill q1 (coverage-based rarefaction)}
# Predicted marginal means
pred_site_q1_cov <- ggpredict(RQ1cov_site, terms = "site", type = "fixed")

# ensure plotting order
pred_site_q1_cov$x <- factor(pred_site_q1_cov$x, levels = c("control", "focal"))
df_q1_rarefied$site <- factor(df_q1_rarefied$site, levels = c("control", "focal"))

# Color palette
site_cols <- list(
  "focal" = "#2C7FB8",
  "control" = "#D95F09",
  "focal_light" = "cadetblue3",
  "control_light" = "#D95"
)

# Plot
p_q1_cov <- ggplot() +
  geom_jitter(
    data = df_q1_rarefied,
    aes(x = site, y = hill_q1, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +
  geom_point(
    data = pred_site_q1_cov,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +
  geom_errorbar(
    data = pred_site_q1_cov,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +
  scale_colour_manual(values = site_cols) +
  scale_x_discrete(labels = c("control" = "Control", "focal" = "Focal")) +
  labs(
    x = NULL,
    y = "Bee Shannon diversity \n (coverage-standardised)",
    title = "Predicted Bee Diversity per Plot by Site (coverage-based rarefaction, q = 1)",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = standardised Hill q1 at equal coverage"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  )

p_q1_cov
```

## Model predictions patchwork plot
```{r model predictions patchwork plor coverage based rarefaction}
# Shared publication theme
theme_pub <- theme_minimal(base_size = 16) +
  theme(
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 18),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)
  )
# clean plots
p_rrich_cov_clean <- p_rrich_cov +
  labs(title = NULL, subtitle = NULL) +
  theme_pub

p_q1_cov_clean <- p_q1_cov +
  labs(title = NULL, subtitle = NULL) +
  theme_pub

# Combine plots
p_cov_combined <- p_rrich_cov_clean + p_q1_cov_clean +
  patchwork::plot_layout(ncol = 2) +
  patchwork::plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")"
  )

p_cov_combined

# save tiff
tiff(
  file.path("..", "output", "figures", "preds_bee_covgrar_600dpi.tiff"),
  width = 5000, height = 2500, res = 600, compression = "lzw"
)

print(p_cov_combined)

dev.off()
```
## Model outputs table
```{r Model ouputs table coverage based rarefaction}
# helper function
extract_site_table_rarefied <- function(model, response_name, effect_type) {
  
  em <- emmeans(model, ~ site, type = "response")
  em_df <- as.data.frame(em)
  
  mean_col  <- if ("response" %in% names(em_df)) "response" else "emmean"
  lower_col <- if ("asymp.LCL" %in% names(em_df)) "asymp.LCL" else "lower.CL"
  upper_col <- if ("asymp.UCL" %in% names(em_df)) "asymp.UCL" else "upper.CL"
  
  means_wide <- em_df %>%
    mutate(site = as.character(site)) %>%
    transmute(
      site,
      mean = .data[[mean_col]],
      lower = .data[[lower_col]],
      upper = .data[[upper_col]]
    ) %>%
    mutate(mean_ci = sprintf("%.2f (%.2f\u2013%.2f)", mean, lower, upper)) %>%
    select(site, mean_ci) %>%
    pivot_wider(names_from = site, values_from = mean_ci)
  
  # raw means (unformatted), needed to compute % change for additive-scale models
  control_mean <- em_df[[mean_col]][em_df$site == "control"]
  focal_mean   <- em_df[[mean_col]][em_df$site == "focal"]
  pct_change   <- (focal_mean - control_mean) / control_mean * 100
  
  contr <- contrast(em, "revpairwise") %>%
    summary(infer = TRUE, type = "response") %>%
    as.data.frame() %>%
    rename(
      lower = any_of(c("asymp.LCL", "lower.CL")),
      upper = any_of(c("asymp.UCL", "upper.CL"))
    )
  
  if (effect_type == "ratio") {
    contr_out <- contr %>%
      transmute(
        effect = sprintf("%.2f\u00d7 (%.2f\u2013%.2f)", ratio, lower, upper),
        p_value = ifelse(p.value < 0.001, "< 0.001", sprintf("%.3f", p.value))
      )
  } else {
    contr_out <- contr %>%
      transmute(
        effect = sprintf("+%.2f (%.2f\u2013%.2f); +%.0f%%", estimate, lower, upper, pct_change),
        p_value = ifelse(p.value < 0.001, "< 0.001", sprintf("%.3f", p.value))
      )
  }
  
  # R2 extraction
  r2_vals <- tryCatch(r2(model), error = function(e) NULL)
  r2_marg <- if (!is.null(r2_vals$R2_marginal)) sprintf("%.3f", r2_vals$R2_marginal) else NA
  r2_cond <- if (!is.null(r2_vals$R2_conditional)) sprintf("%.3f", r2_vals$R2_conditional) else NA
  
  bind_cols(means_wide, contr_out) %>%
    mutate(
      response = sprintf("%s (R\u00b2m = %s, R\u00b2c = %s)", response_name, r2_marg, r2_cond),
      .before = everything()
    )
}

tab_cov_rich <- extract_site_table_rarefied(RRIcov_site, "Richness", "ratio")
tab_cov_q1   <- extract_site_table_rarefied(RQ1cov_site, "Hill q1", "additive")

table_cov_rarefied <- bind_rows(tab_cov_rich, tab_cov_q1)

gt_table_cov <- table_cov_rarefied %>%
  gt() %>%
  cols_label(
    response = "Response",
    control  = "Control (mean \u00b1 95% CI)",
    focal    = "Focal (mean \u00b1 95% CI)",
    effect   = "Focal vs Control Difference (\u00b1 95% CI)",
    p_value  = "P-value"
  ) %>%
  cols_align(
    align = "center",
    columns = everything()
  )

gt_table_cov

gtsave(
  gt_table_cov,
  filename = "../output/tables/cov_rarefaction_effects.docx"
)
```

# Size-based rarefaction plot scale (visits merged)
## Richness
```{r Size-based rarefaction plot scale (visits merged)}
# Add total abundance per plot
bee_plot_n <- bee_plot %>%
  mutate(total_abundance = rowSums(across(all_of(species_cols))))

summary(bee_plot_n$total_abundance)

quantile(bee_plot_n$total_abundance, probs = c(0.1, 0.2, 0.3))

ggplot(bee_plot_n, aes(x = total_abundance)) +
  geom_histogram(bins = 20, fill = "grey70", color = "black") +
  labs(
    title = "Total abundance per plot",
    x = "Number of individuals",
    y = "Number of plots"
  ) +
  theme_bw()

# Set target sample size
target_n <- floor(quantile(bee_plot_n$total_abundance, 0.3))
target_n

# Exact size-based rarefaction using estimateD - richness
df_size_rarefied <- estimateD(
  inext_list,
  q = 0,
  datatype = "abundance",
  base = "size",
  level = target_n
)

# Add metadata
df_size_rarefied <- df_size_rarefied %>%
  separate(Assemblage, into = c("region", "site", "plotID"), sep = "_") %>%
  rename(
    sample_size = m,
    richness = qD,
    richness_lwr = qD.LCL,
    richness_upr = qD.UCL
  )
```


```{r size-based rarefaction model}
RRIsiz_site <- glmmTMB(
  richness ~ site + (1 | region),
  data = df_size_rarefied,
  family = gaussian()
)

summary(RRIsiz_site)


check_model(RRIsiz_site)

# Predicted values
pred_siz_rich <- emmeans(RRIsiz_site, ~ site, type = "response")
contrast(pred_siz_rich, "revpairwise")
```
As this is size-based, the estimate of 2.08 of focal is the absolute difference in species expected when sampling target_n = 11 individuals in a plot - focal sites have 2.05 more species
Effect = +2.06 species
absolute difference

```{r plot predicted size-based rarefaction }

# Predicted marginal means
pred_site_rrich_size <- ggpredict(RRIsiz_site, terms = "site", type = "fixed")

# ensure plotting order
pred_site_rrich_size$x <- factor(pred_site_rrich_size$x, levels = c("control", "focal"))
df_size_rarefied$site <- factor(df_size_rarefied$site, levels = c("control", "focal"))

# Color palette
site_cols <- list(
  "focal" = "#2C7FB8",
  "control" = "#D95F09",
  "focal_light" = "cadetblue3",
  "control_light" = "#D95"
)

# Plot
p_rrich_size <- ggplot() +
  geom_jitter(
    data = df_size_rarefied,
    aes(x = site, y = richness, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +
  geom_point(
    data = pred_site_rrich_size,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +
  geom_errorbar(
    data = pred_site_rrich_size,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +
  scale_colour_manual(values = site_cols) +
  scale_x_discrete(labels = c("control" = "Control", "focal" = "Focal")) +
  labs(
    x = NULL,
    y = "Bee richness \n (size-standardised)",
    title = "Predicted Bee Richness per Plot by Site (size-based rarefaction)",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = rarefied plot values"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  )

p_rrich_size
```

## Hill q1
```{r Size-based rarefaction (Hill q1)}
# Exact size-based rarefaction using estimateD
df_size_q1_rarefied <- estimateD(
  inext_list,
  q = 1,
  datatype = "abundance",
  base = "size",
  level = target_n
)

# Add metadata
df_size_q1_rarefied <- df_size_q1_rarefied %>%
  separate(Assemblage, into = c("region", "site", "plotID"), sep = "_") %>%
  rename(
    sample_size = m,
    hill_q1 = qD,
    hill_q1_lwr = qD.LCL,
    hill_q1_upr = qD.UCL
  )

# Check number of plots retained
length(unique(df_size_q1_rarefied$plotID))
```

```{r Size-based rarefaction model (Hill q1)}
RRIsiz_q1_site <- glmmTMB(
  hill_q1 ~ site + (1 | region),
  data = df_size_q1_rarefied,
  family = gaussian()
)

summary(RRIsiz_q1_site)

check_model(RRIsiz_q1_site)

# Predicted values
pred_siz_q1 <- emmeans(RRIsiz_q1_site, ~ site, type = "response")
contrast(pred_siz_rich, "revpairwise")
```

```{r Plot predicted size-based rarefaction (Hill q1)}
# Predicted marginal means
pred_site_q1_size <- ggpredict(RRIsiz_q1_site, terms = "site", type = "fixed")

# ensure plotting order
pred_site_q1_size$x <- factor(pred_site_q1_size$x, levels = c("control", "focal"))
df_size_q1_rarefied$site <- factor(df_size_q1_rarefied$site, levels = c("control", "focal"))

# Plot
p_q1_size <- ggplot() +
  geom_jitter(
    data = df_size_q1_rarefied,
    aes(x = site, y = hill_q1, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +
  geom_point(
    data = pred_site_q1_size,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +
  geom_errorbar(
    data = pred_site_q1_size,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +
  scale_colour_manual(values = site_cols) +
  scale_x_discrete(labels = c("control" = "Control", "focal" = "Focal")) +
  labs(
    x = NULL,
    y = "Bee Shannon diversity \n (size-standardised)",
    title = "Predicted Bee Diversity per Plot by Site (size-based rarefaction)",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = rarefied plot values"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  )

p_q1_size
```

## Model predictions patchwork plot
```{r model predictions patchwork plos size based rarefaction}
# Shared publication theme
theme_pub <- theme_minimal(base_size = 16) +
  theme(
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 18),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)
  )

# clean plots
p_rrich_size_clean <- p_rrich_size +
  labs(title = NULL, subtitle = NULL) +
  theme_pub

p_q1_size_clean <- p_q1_size +
  labs(title = NULL, subtitle = NULL) +
  theme_pub

# combine plots
p_size_combined <- p_rrich_size_clean + p_q1_size_clean +
  patchwork::plot_layout(ncol = 2) +
  patchwork::plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")"
  )

p_size_combined

# save tiff
tiff(
  file.path("..", "output", "figures", "preds_bee_sizerar_600dpi.tiff"),
  width = 5000, height = 2500, res = 600, compression = "lzw"
)
print(p_size_combined)
dev.off()
```

## Model outputs table
```{r Model outputs table size-based rarefation}
# helper function
extract_site_table_rarefied <- function(model, response_name, effect_type) {
  
  em <- emmeans(model, ~ site)
  em_df <- as.data.frame(em)
  
  lower_col <- if ("asymp.LCL" %in% names(em_df)) "asymp.LCL" else "lower.CL"
  upper_col <- if ("asymp.UCL" %in% names(em_df)) "asymp.UCL" else "upper.CL"
  
  means_wide <- em_df %>%
    mutate(site = as.character(site)) %>%
    transmute(
      site,
      mean = emmean,
      lower = .data[[lower_col]],
      upper = .data[[upper_col]]
    ) %>%
    mutate(
      mean_ci = sprintf(
        "%.2f (%.2f–%.2f)",
        mean,
        lower,
        upper
      )
    ) %>%
    select(site, mean_ci) %>%
    pivot_wider(
      names_from = site,
      values_from = mean_ci
    )
  
  # percentage increase
  control_mean <- em_df$emmean[em_df$site == "control"]
  focal_mean   <- em_df$emmean[em_df$site == "focal"]
  pct_change   <- (focal_mean - control_mean) / control_mean * 100
  
  if (effect_type == "ratio") {
    
    contr <- contrast(
      em,
      "revpairwise"
    ) %>%
      summary(
        infer = TRUE,
        type = "response"
      ) %>%
      as.data.frame()
    
    contr <- contr %>%
      rename(
        lower = any_of(c("asymp.LCL", "lower.CL")),
        upper = any_of(c("asymp.UCL", "upper.CL"))
      )
    
    contr_out <- contr %>%
      transmute(
        effect = sprintf(
          "%.2f× (%.2f–%.2f)",
          ratio,
          lower,
          upper
        ),
        p_value = ifelse(
          p.value < 0.001,
          "< 0.001",
          sprintf("%.3f", p.value)
        )
      )
    
  } else {
    
    contr <- pairs(em) %>%
      summary(infer = TRUE) %>%
      as.data.frame()
    
    contr_out <- contr %>%
      transmute(
        effect = sprintf(
          "+%.2f (%.2f–%.2f); +%.0f%%",
          -estimate,
          -upper.CL,
          -lower.CL,
          pct_change
        ),
        p_value = ifelse(
          p.value < 0.001,
          "< 0.001",
          sprintf("%.3f", p.value)
        )
      )
  }
  
  # R²
  r2_vals <- tryCatch(
    performance::r2(model),
    error = function(e) NULL
  )
  
  r2_marg <- if (!is.null(r2_vals$R2_marginal))
    sprintf("%.3f", r2_vals$R2_marginal)
  else
    NA
  
  r2_cond <- if (!is.null(r2_vals$R2_conditional))
    sprintf("%.3f", r2_vals$R2_conditional)
  else
    NA
  
  bind_cols(
    means_wide,
    contr_out
  ) %>%
    mutate(
      response = sprintf(
        "%s (R²m = %s, R²c = %s)",
        response_name,
        r2_marg,
        r2_cond
      ),
      .before = everything()
    )
}

tab_size_rich <- extract_site_table_rarefied(RRIsiz_site, "Richness", "absolute")
tab_size_q1   <- extract_site_table_rarefied(RRIsiz_q1_site, "Hill q1", "absolute")

table_size_rarefied <- bind_rows(tab_size_rich, tab_size_q1)

gt_table_size <- table_size_rarefied %>%
  gt() %>%
  cols_label(
    response = "Response",
    control  = "Control (mean ± 95% CI)",
    focal    = "Focal (mean ± 95% CI)",
    effect   = "Focal vs Control Difference (± 95% CI)",
    p_value  = "P-value"
  ) %>%
  cols_align(
    align = "center",
    everything()
  )

gtsave(
  gt_table_size,
  filename = "../output/tables/size_rarefaction_effects.docx"
)
```