---
title: "04_bee_rarefaction"
output: html_document
---

```{r}
library(iNEXT)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(glmmTMB)
library(performance)
```

Coverage‑based rarefaction (Chao et al. 2014, 2020) provides descriptive standardization, not inferential estimates, and its resulting diversity values do not follow the distributional assumptions required for GLMMs (e.g. Gotelli & Colwell 2001; Chao & Jost 2012). Rarefaction also collapses plot‑level replication, removes ecological variance, and prevents correct estimation of random effects.
Because our sampling design includes spatial and hierarchical structure (multiple visits per plot; plots nested within sites; sites nested within regions), a GLMM using plot‑level richness/abundance with region and plotID as random effects is the statistically appropriate inferential framework.

```{r Load data}
# Compact helper function to read raw CSVs
bee_data <- read.csv(file.path("..", "data", "cleaned", "be
                               e_data.csv"),
           stringsAsFactors = FALSE,
           check.names = FALSE)
```

```{r colorblind friendly approach}

# Identify metadata and species columns
meta_cols <- c(
  "surveyID", "plotID", "site", "region", "visit", "plot",
  "bee_richness", "bee_hill_q1", "bee_hill_q2", "fieldIDsum",
  "bee_abundance_nofieldID", "bee_abundance",
  "RLabund_LC", "RLabund_NT", "RLabund_VU", "RLabund_EN", "RLabund_DD",
  "RLrich_LC", "RLrich_NT", "RLrich_VU", "RLrich_EN", "RLrich_DD",
  "RLrich_CR", "RLabund_CR",
  "RLrich_threatened", "RLabund_threatened",
  "RLrich_total", "RLabund_total"
)

species_cols <- setdiff(names(bee_data), meta_cols)

```

# Site-level rarefaction 
```{r rarefy by site per region}
# Aggregate all plots per site
bee_site <- bee_data %>%
  group_by(region, site) %>%
  summarise(across(all_of(species_cols), sum), .groups = "drop")


# Prepare iNEXT input list (site-level)
inext_site <- setNames(
  lapply(seq_len(nrow(bee_site)), function(i) {
    as.numeric(bee_site[i, species_cols])
  }),
  paste(bee_site$region, bee_site$site, sep = "_")
)


# Run iNEXT for abundance data (q = 0 richness)
out_site <- iNEXT(
  inext_site,
  q = 0,
  datatype = "abundance",
  endpoint = max(sapply(inext_site, sum)) * 2
)
```

# Coverage-based rarefaction site scale per region
```{r}
# Extract coverage-based rarefaction data
df_site <- fortify(out_site, type = 3)


# Recover region and site labels
df_site <- df_site %>%
  separate(Assemblage, into = c("region", "site"), sep = "_")


# Plot coverage-based rarefaction curves (SITE level)
p_covraref_site <- ggplot(df_site,
                          aes(x = x, y = y, color = site, fill = site)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr),
              alpha = 0.25, color = NA) +
  facet_wrap(~ region) +
  labs(
    x = "Sample coverage",
    y = "Estimated species richness",
    title = "Coverage-based rarefaction by site",
    color = "Site",
    fill = "Site"
  ) +
  scale_color_manual(values = c("focal" = "#2C7FB8", "control" = "#D95F02")) +
  scale_fill_manual(values = c("focal" = "#A6C1E0", "control" = "#F1B0AD")) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

p_covraref_site
```
At low coverage, focal and control look similar → sampling effort not yet enough to reveal differences.
At moderate coverage (0.6–0.9), focal curves are consistently above control → higher richness when standardized for completeness.
Near full coverage, extrapolated ends diverge sharply, showing that focal assemblages have a larger “potential species pool”.
```{r plot }
# Extract transition points (Observed)
transition_pts <- df_site %>%
  filter(Method == "Observed")

# Plot
p_covraref_site <- ggplot(df_site,
                          aes(x = x, y = y, color = site, fill = site)) +
  
  # Lines with linetype mapping
  geom_line(aes(linetype = Method), linewidth = 1.2) +
  
  # Confidence ribbon
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr),
              alpha = 0.25, color = NA) +
  
  # Transition point (Observed)
  geom_point(
    data = transition_pts,
    aes(x = x, y = y),
    size = 3,
    shape = 21,
    stroke = 1,
    color = "black",
    fill = "white"
  ) +
  
  # Facet per region
  facet_wrap(~ region) +
  
  labs(
    x = "Sample coverage",
    y = "Estimated species richness",
    title = "Coverage-based rarefaction by site",
    color = "Site",
    fill = "Site",
    linetype = "Sampling segment"
  ) +
  
  # Color scales
  scale_color_manual(values = c(
    "focal" = "#2C7FB8",
    "control" = "#D95F02"
  )) +
  
  scale_fill_manual(values = c(
    "focal" = "#A6C1E0",
    "control" = "#F1B0AD"
  )) +
  
  scale_linetype_manual(
     values = c(
     "Rarefaction" = "solid", "Extrapolation" = "dashed"),
     breaks = c("Rarefaction", "Extrapolation"), # remove Observed
     name = "Sampling segment") +
  
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

p_covraref_site

```


The coverage‑based rarefaction curves allow us to compare focal and control sites at the same level of sampling completeness, even if neither group has been fully sampled and many bee species remain undetected. The x‑axis indicates sample coverage, meaning the proportion of the underlying community that has actually been observed. As we approach coverage = 1, the focal sites show a pronounced upward spike. This reflects the presence of many very rare species: with only a small amount of additional sampling, a disproportionately large number of new species would still be added. In contrast, the control sites lack such a sharp increase, suggesting fewer rare species and a community that is more fully sampled. Below about 0.8 coverage, the curves for focal and control sites largely overlap, indicating that the core, well‑sampled portion of the community is similar between treatments.”

gpt


```{r export rarefaction curves as tiff}
tiff(
  file.path("..", "output", "figures", "cov_rarefaction_site_600dpi.tiff"),
  width = 3000, height = 2250, res = 600, compression = "lzw"
)

print(p_covraref_site)  # <-- replace with your actual plot object
dev.off()
```

#Size-based rarefaction site scale
```{r}
df_site <- out_site$iNextEst$size_based

df_site <- df_site %>%
  rename(
    x = m,
    y = qD,
    y.lwr = qD.LCL,
    y.upr = qD.UCL
  ) %>%
  separate(Assemblage, into = c("region", "site"), sep = "_")

# Observed points
transition_pts <- df_site %>%
  filter(Method == "Observed")

# Plot
p_sizeraref_site <- ggplot(df_site,
                          aes(x = x, y = y, color = site, fill = site)) +
  
  geom_line(aes(linetype = Method), linewidth = 1.2) +
  
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr),
              alpha = 0.25, color = NA) +
  
  geom_point(
    data = transition_pts,
    aes(x = x, y = y),
    size = 3,
    shape = 21,
    stroke = 1,
    color = "black",
    fill = "white"
  ) +
  
  facet_wrap(~ region) +
  
  labs(
    x = "Number of individuals",   # ✅ KEY CHANGE
    y = "Estimated species richness",
    title = "Sample-size rarefaction by site",
    color = "Site",
    fill = "Site",
    linetype = "Sampling segment"
  ) +
  
  scale_color_manual(values = c(
    "focal" = "#2C7FB8",
    "control" = "#D95F02"
  )) +
  
  scale_fill_manual(values = c(
    "focal" = "#A6C1E0",
    "control" = "#F1B0AD"
  )) +
  
  scale_linetype_manual(
    values = c(
      "Rarefaction" = "solid",
      "Extrapolation" = "dashed"
    ),
    breaks = c("Rarefaction", "Extrapolation"),
    name = "Sampling segment"
  ) +
  
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

p_sizeraref_site
```

# Pooled site-level rarefaction
```{r rarefy pooled coverage based rarefaction by site}
# Pool to two assemblages: focal and control
bee_focal_pool   <- colSums(bee_data[bee_data$site == "focal", species_cols])
bee_control_pool <- colSums(bee_data[bee_data$site == "control", species_cols])

inext_pooled <- list(
  focal   = bee_focal_pool,
  control = bee_control_pool
)

out_pooled <- iNEXT(
  inext_pooled,
  q = 0,
  datatype = "abundance",
  endpoint = max(sapply(inext_pooled, sum)) * 2
)
```

# Pooled size-based rarefaction by site
```{r size and coverage-based rarefaction pooled focal vs control}
# Extract size-based and coverage-based data
df_pooled_size     <- fortify(out_pooled, type = 1)
df_pooled_coverage <- fortify(out_pooled, type = 3)

# Transition points
transition_size     <- df_pooled_size     %>% filter(Method == "Observed")
transition_coverage <- df_pooled_coverage %>% filter(Method == "Observed")

# SIZE-BASED PLOT
p_raref_size <- ggplot(df_pooled_size,
                       aes(x = x, y = y, color = Assemblage, fill = Assemblage)) +
  geom_line(aes(linetype = Method), linewidth = 1.2) +
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr),
              alpha = 0.25, color = NA) +
  geom_point(
    data = transition_size,
    aes(x = x, y = y),
    size = 3, shape = 21, stroke = 1,
    color = "black", fill = "white"
  ) +
  scale_color_manual(values = c("control" = "#D95F09", "focal" = "#2C7FB8")) +
  scale_fill_manual(values  = c("control" = "#F1B0AD", "focal" = "#A6C1E0")) +
  scale_linetype_manual(
    values = c("Rarefaction" = "solid", "Extrapolation" = "dashed"),
    breaks = c("Rarefaction", "Extrapolation"),
    name   = "Sampling segment"
  ) +
  labs(
    x     = "Number of individuals",
    y     = "Estimated species richness",
    title = "Size-based rarefaction: Control vs Focal",
    color = "Site", fill = "Site"
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

# COVERAGE-BASED PLOT
p_raref_coverage <- ggplot(df_pooled_coverage,
                           aes(x = x, y = y, color = Assemblage, fill = Assemblage)) +
  geom_line(aes(linetype = Method), linewidth = 1.2) +
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr),
              alpha = 0.25, color = NA) +
  geom_point(
    data = transition_coverage,
    aes(x = x, y = y),
    size = 3, shape = 21, stroke = 1,
    color = "black", fill = "white"
  ) +
  scale_color_manual(values = c("control" = "#D95F09", "focal" = "#2C7FB8")) +
  scale_fill_manual(values  = c("control" = "#F1B0AD", "focal" = "#A6C1E0")) +
  scale_linetype_manual(
    values = c("Rarefaction" = "solid", "Extrapolation" = "dashed"),
    breaks = c("Rarefaction", "Extrapolation"),
    name   = "Sampling segment"
  ) +
  labs(
    x     = "Sample coverage",
    y     = "Estimated species richness",
    title = "Coverage-based rarefaction: Control vs Focal",
    color = "Site", fill = "Site"
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

p_raref_size
p_raref_coverage
```

Both plots are produced from the same single `iNEXT` run. The only differences between the two plot objects are `df_pooled_size` vs `df_pooled_coverage`, the transition points, and the x-axis label.

# Plot-level rarefaction
```{r rarefaction}
# ==============================
# Identify metadata and species columns
# ==============================
meta_cols <- c(
  "surveyID", "plotID", "site", "region", "visit", "plot",
  "bee_richness", "fieldIDsum",
  "bee_abundance_nofieldID", "bee_abundance", "bee_hill_q1", "bee_hill_q2", 
  "RLabund_LC", "RLabund_NT", "RLabund_VU", "RLabund_EN", "RLabund_DD",
  "RLrich_LC", "RLrich_NT", "RLrich_VU", "RLrich_EN", "RLrich_DD",
  "RLrich_CR", "RLabund_CR",
  "RLrich_threatened", "RLabund_threatened",
  "RLrich_total", "RLabund_total"
)

species_cols <- setdiff(names(bee_data), meta_cols)

# ==============================
# Aggregate visits WITHIN each plot
# ==============================
bee_plot <- bee_data %>%
  group_by(region, site, plotID) %>%
  summarise(across(all_of(species_cols), sum), .groups = "drop")

# ==============================
# Prepare iNEXT input list (per plot)
# ==============================
inext_list <- setNames(
  lapply(seq_len(nrow(bee_plot)), function(i) {
    as.numeric(bee_plot[i, species_cols])
  }),
  paste(bee_plot$region, bee_plot$site, bee_plot$plotID, sep = "_")
)

# ==============================
# Run iNEXT: species richness (q=0), abundance data
# ==============================
out_inext <- iNEXT(
  inext_list,
  q = 0,
  datatype = "abundance",
  endpoint = max(sapply(inext_list, sum)) * 2
)

ggiNEXT(out_inext, type = 3)


sub_names <- unique(names(inext_list))[1:6]
```

# Coverage-based rarefaction plot-scale (visits merged)
```{r}
df_plot <- fortify(out_inext, type = 3)  # coverage-based curves

df_cov_max <- df_plot %>%
  group_by(Assemblage) %>%
  summarise(max_coverage = max(x), .groups = "drop")


summary(df_cov_max$max_coverage)
quantile(df_cov_max$max_coverage, probs = c(0.7, 0.8, 0.9, 0.95))


ggplot(df_cov_max, aes(x = max_coverage)) +
  geom_histogram(bins = 20, fill = "grey70", color = "black") +
  geom_vline(xintercept = 0.8, linetype = "dashed", color = "red") +
  labs(
    title = "Maximum achievable coverage across plots",
    x = "Coverage",
    y = "Number of plots"
  ) +
  theme_bw()

# set target coverage
target_coverage <- 0.85

ggiNEXT(out_inext, type = 3) +
  geom_vline(xintercept = target_coverage,
             linetype = "dashed",
             color = "red") +
  theme_minimal()
# extract rarefied richness
df_plot_rarefied <- df_plot %>%
  filter(Method != "Observed") %>%
  group_by(Assemblage) %>%
  slice(which.min(abs(x - target_coverage))) %>%
  ungroup()


# add metadata
df_plot_rarefied <- df_plot_rarefied %>%
  separate(Assemblage, into = c("region", "site", "plotID"), sep = "_") %>%
  rename(
    coverage = x,
    richness = y,
    richness_lwr = y.lwr,
    richness_upr = y.upr
  )

# filter extreme extrapolation
df_plot_rarefied <- df_plot_rarefied %>%
  filter(coverage >= target_coverage - 0.02)

# 11 plots removed in filtering steps
length(unique(df_plot$Assemblage))        # initial plots
length(unique(df_plot_rarefied$plotID))   # remaining plots

# model with rarefied values
rrich_plot <- glmmTMB(
  richness ~ site + (1 | region),
  data = df_plot_rarefied,
  family = Gamma(link = "log")
)

check_model(rrich_plot)

summary(rrich_plot)


exp(0.5568)
```
This shows that coverage is very high in most plots (median = 0.999) - plots are well-sampled - threshold was selected based on histogram, and model run with these values. 
However, there is a large difference in sample size between plots, it is worth trying sample size based rarefaction (especially because we have standardised our plot size - see Clara comment)

1.745 higher richness in focla vs control site
Effect ≈ 1.75× higher richness
~ +75% increase -> multiplicative effect


# Size-based rarefaction plot scale (visits merged)
```{r}
bee_plot_n <- bee_plot %>%
  mutate(total_abundance = rowSums(across(all_of(species_cols))))


summary(bee_plot_n$total_abundance)
quantile(bee_plot_n$total_abundance, probs = c(0.1, 0.2, 0.3))

ggplot(bee_plot_n, aes(x = total_abundance)) +
  geom_histogram(bins = 20, fill = "grey70", color = "black") +
  labs(
    title = "Total abundance per plot",
    x = "Number of individuals",
    y = "Number of plots"
  ) +
  theme_bw()

# set target n
target_n <- floor(quantile(bee_plot_n$total_abundance, 0.3))
target_n

# run INEXT
inext_list <- setNames(
  lapply(seq_len(nrow(bee_plot)), function(i) {
    as.numeric(bee_plot[i, species_cols])
  }),
  paste(bee_plot$region, bee_plot$site, bee_plot$plotID, sep = "_")
)


out_size <- iNEXT(
  inext_list,
  q = 0,
  datatype = "abundance",
  size = target_n
)

# extract rarefied richness
df_size <- fortify(out_size, type = 1)

# select sample size closets to targen_n
df_size_rarefied <- df_size %>%
  filter(Method != "Observed") %>%
  group_by(Assemblage) %>%
  slice(which.min(abs(x - target_n))) %>%
  ungroup()

# add metadata
df_size_rarefied <- df_size_rarefied %>%
  separate(Assemblage, into = c("region", "site", "plotID"), sep = "_") %>%
  rename(
    sample_size = x,
    richness = y,
    richness_lwr = y.lwr,
    richness_upr = y.upr
  )

# filter extreme extrapolation
df_size_rarefied <- df_size_rarefied %>%
  filter(sample_size >= target_n * 0.95)

# plots removed through filtering steps
length(unique(df_size$Assemblage))
length(unique(df_size_rarefied$plotID))


model_size <- glmmTMB(
  richness ~ site + (1 | region),
  data = df_size_rarefied,
  family = gaussian(
  )
)

summary(model_size)
check_model(model_size)
```
As this is size-based, the estimate of 2.08 of focal is the absolute difference in species expected when sampling target_n = 11 individuals in a plot - focal sites have 2.08 more species
Effect = +2.08 species
absolute difference

```{r}
RRI_site <- glmmTMB(
  richness ~ site + (1 | region),
  data = df_size_rarefied,
  family = gaussian()
)

summary(RRI_site)
```

```{r plot predicted size-based rarefaction }
# Predicted marginal means for site on the response scale
pred_site_rrich <- ggpredict(RRI_site, terms = "site", type = "fixed")

# ensure plotting order (control, focal)
pred_site_rrich$x <- factor(pred_site_rrich$x, levels = c("control", "focal"))
df_size_rarefied$site <- factor(df_size_rarefied$site, levels = c("control", "focal"))

# Color palette
site_cols <- list(
  "focal" = "#2C7FB8",
  "control"   = "#D95F09",
  "focal_light"   = "cadetblue3",
  "control_light" = "#D95"
  )

# Plot: predicted means ± 95% CI with raw jittered points

p_rrich <- ggplot() +
  # 1) Raw data (plot × visit), jittered in background
  geom_jitter(
    data = df_size_rarefied,
    aes(x = site, y = richness, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +

  # 2) Model-predicted means (ggeffects output; response scale)
  geom_point(
    data = pred_site_rrich,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +

  # 3) 95% confidence intervals
  geom_errorbar(
    data = pred_site_rrich,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +

  # Use identical site colours for both layers
  scale_colour_manual(values = site_cols) +

  # Labels
  labs(
    x = NULL,
    y = "Bee species richness",
    title = "Predicted Bee Richness per Plot by Site (pooled visits)",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = raw observations"
  ) +
  
  # Theme
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  )

p_rrich
```

```{r plot size based rarefaction across visits}

# Ensure factors are ordered nicely
df_size_rarefied <- df_size_rarefied %>%
  mutate(
    site = factor(site, levels = c("control", "focal")),
    region = factor(region)
  )

# Plot
p_rarefied <- ggplot(df_size_rarefied,
                     aes(x = site, y = richness, fill = site)) +
  
  # boxplot
  geom_boxplot(
    alpha = 0.6,
    width = 0.6,
    outlier.shape = NA,
    color = "black"
  ) +
  
  # jittered raw values
  geom_jitter(
    width = 0.12,
    size = 2,
    alpha = 0.6,
    color = "black"
  ) +
  
  # facet by region
  facet_wrap(~ region) +
  
  # colors
  scale_fill_manual(values = c(
    "control" = "#D95F09",
    "focal"   = "#2C7FB8"
  )) +
  
  # labels
  labs(
    title = "Bee richness at standardized sampling effort",
    subtitle = paste0("Rarefied to n = ", target_n, " individuals per plot"),
    x = NULL,
    y = "Rarefied species richness"
  ) +
  
  # theme (publication-style)
  theme_bw(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    strip.background = element_rect(fill = "grey90"),
    strip.text = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    plot.title = element_text(face = "bold", size = 15),
    legend.position = "none"
  )

# Print
p_rarefied

```

```{r}
saveRDS(df_size_rarefied, file = file.path("..","data","cleaned","rarefied_size_plot.rds"))
saveRDS(df_plot_rarefied, file = file.path("..","data","cleaned","rarefied_covg_plot.rds"))
```


