---
title: "04_bee_rarefaction"
output: html_document
---

```{r load packages}
library(emmeans)
library(gt)
library(iNEXT)
library(dplyr)
library(tidyr)
library(stringr)
library(ggeffects)
library(ggplot2)
library(glmmTMB)
library(performance)
```


```{r Load data}
# Compact helper function to read raw CSVs
bee_data <- read.csv(file.path("..", "data", "cleaned", "bee_data.csv"),
                     stringsAsFactors = FALSE,
                     check.names = FALSE)

```

```{r define metadata and speceis columns}
# Identify metadata and species columns
meta_cols <- c(
  "surveyID", "plotID", "site", "region", "visit", "plot",
  "bee_richness", "bee_hill_q1", "bee_hill_q2", "fieldIDsum",
  "bee_abundance_nofieldID", "bee_abundance",
  "RLabund_LC", "RLabund_NT", "RLabund_VU", "RLabund_EN", "RLabund_DD",
  "RLrich_LC", "RLrich_NT", "RLrich_VU", "RLrich_EN", "RLrich_DD",
  "RLrich_CR", "RLabund_CR",
  "RLrich_threatened", "RLabund_threatened",
  "RLrich_total", "RLabund_total"
)

species_cols <- setdiff(names(bee_data), meta_cols)
```

# Site-level rarefaction by region
```{r rarefy by site per region}
# Aggregate all plots per site
bee_site <- bee_data |>
  group_by(region, site) |>
  summarise(across(all_of(species_cols), sum), .groups = "drop")


# Prepare iNEXT input list (site-level)
inext_site <- setNames(
  lapply(seq_len(nrow(bee_site)), function(i) {
    as.numeric(bee_site[i, species_cols])
  }),
  paste(bee_site$region, bee_site$site, sep = "_")
)

# Run iNEXT for abundance data (q = 0 richness)
out_site <- iNEXT(
  inext_site,
  q = 0,
  datatype = "abundance",
  endpoint = max(sapply(inext_site, sum)) * 2
)
```

# Coverage-based rarefaction site scale per region
```{r coverage based rarefaction site scale per region}
# Extract coverage-based rarefaction data
df_site <- fortify(out_site, type = 3)

# Recover region and site labels
df_site <- df_site |>
  separate(Assemblage, into = c("region", "site"), sep = "_")

```

```{r plot coverage based rarefaction by site and region}
# Extract transition points (Observed)
transition_pts <- df_site |>
  filter(Method == "Observed")

# Plot
p_covraref_site <- ggplot(
  df_site, 
  aes(x = x, y = y, color = site, fill = site)) +
  geom_line(aes(linetype = Method), linewidth = 1.2) +
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr),
              alpha = 0.25, color = NA) +
    # Transition point (Observed)
  geom_point(
    data = transition_pts,
    aes(x = x, y = y),
    size = 3,
    shape = 21,
    stroke = 1,
    color = "black",
    fill = "white"
  ) +
    # Facet per region
  facet_wrap(~ region) +
    labs(
    x = "Sample coverage",
    y = "Estimated species richness",
    title = "Coverage-based rarefaction by site",
    color = "Site",
    fill = "Site",
    linetype = "Sampling segment"
  ) +
    # Color scales
  scale_color_manual(values = c(
    "focal" = "#2C7FB8",
    "control" = "#D95F02"
  )) +
    scale_fill_manual(values = c(
    "focal" = "#A6C1E0",
    "control" = "#F1B0AD"
  )) +
  
  scale_linetype_manual(
     values = c(
     "Rarefaction" = "solid", "Extrapolation" = "dashed"),
     breaks = c("Rarefaction", "Extrapolation"), # remove Observed
     name = "Sampling segment") +
  
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

p_covraref_site
```


```{r export coverage rarefaction curves as tiff}
# tiff(
#   file.path("..", "output", "figures", "cov_rarefaction_site_600dpi.tiff"),
#   width = 3000, height = 2250, res = 600, compression = "lzw"
# )
# 
# print(p_covraref_site)
# dev.off()
```

#Size-based rarefaction site scale per region
```{r }
df_site <- out_site$iNextEst$size_based

df_site <- df_site |>
  rename(
    x = m,
    y = qD,
    y.lwr = qD.LCL,
    y.upr = qD.UCL
  ) |>
  separate(Assemblage, into = c("region", "site"), sep = "_")

# Observed points
transition_pts <- df_site |>
  filter(Method == "Observed")

# Plot
p_sizeraref_site <- ggplot(df_site,
                          aes(x = x, y = y, color = site, fill = site)) +
  
  geom_line(aes(linetype = Method), linewidth = 1.2) +
  
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr),
              alpha = 0.25, color = NA) +
  
  geom_point(
    data = transition_pts,
    aes(x = x, y = y),
    size = 3,
    shape = 21,
    stroke = 1,
    color = "black",
    fill = "white"
  ) +
  
  facet_wrap(~ region,
             labeller = as_labeller(c("Aire" = "Aire", 
                                      "Schwarzbach" = "Schwarzbach",
                                      "Scheulte" = "Scheulte",
                                      "Thiele" = "Thièle"))) +
  
  labs(
    x = "Number of individuals",
    y = "Estimated species richness",
    color = NULL,
    fill = NULL,
    linetype = NULL
  ) +
  
  scale_color_manual(values = c(
    "focal" = "#2C7FB8",
    "control" = "#D95F02"),
  labels = c(
    "focal" = "Focal",
    "control" = "Control")) +
  
  scale_fill_manual(values = c(
    "focal" = "#A6C1E0",
    "control" = "#F1B0AD"),
  labels = c(
    "focal" = "Focal",
    "control" = "Control")) +
  
  scale_linetype_manual(
    values = c(
      "Rarefaction" = "solid",
      "Extrapolation" = "22"
    ),
    breaks = c("Rarefaction", "Extrapolation"),
  ) +
  
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
  )

p_sizeraref_site
```

```{r export size rarefaction curves as tiff}
tiff(
  file.path("..", "output", "figures", "size_rarefaction_site_600dpi.tiff"),
  width = 3500, height = 2250, res = 600, compression = "lzw"
)

print(p_sizeraref_site)
dev.off()
```

# (Pooled site-level rarefaction)
Not included in publication
```{r rarefy pooled coverage based rarefaction by site}
# Pool to two assemblages: focal and control
bee_focal_pool   <- colSums(bee_data[bee_data$site == "focal", species_cols])
bee_control_pool <- colSums(bee_data[bee_data$site == "control", species_cols])

inext_pooled <- list(
  focal   = bee_focal_pool,
  control = bee_control_pool
)

out_pooled <- iNEXT(
  inext_pooled,
  q = 0,
  datatype = "abundance",
  endpoint = max(sapply(inext_pooled, sum)) * 2
)
```

## Pooled size-based rarefaction by site
```{r size and coverage-based rarefaction pooled focal vs control}
# Extract size-based and coverage-based data
df_pooled_size     <- fortify(out_pooled, type = 1)
df_pooled_coverage <- fortify(out_pooled, type = 3)

# Transition points
transition_size     <- df_pooled_size     |> filter(Method == "Observed")
transition_coverage <- df_pooled_coverage |> filter(Method == "Observed")

# SIZE-BASED PLOT
p_raref_size <- ggplot(df_pooled_size,
                       aes(x = x, y = y, color = Assemblage, fill = Assemblage)) +
  geom_line(aes(linetype = Method), linewidth = 1.2) +
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr),
              alpha = 0.25, color = NA) +
  geom_point(
    data = transition_size,
    aes(x = x, y = y),
    size = 3, shape = 21, stroke = 1,
    color = "black", fill = "white"
  ) +
  scale_color_manual(values = c("control" = "#D95F09", "focal" = "#2C7FB8")) +
  scale_fill_manual(values  = c("control" = "#F1B0AD", "focal" = "#A6C1E0")) +
  scale_linetype_manual(
    values = c("Rarefaction" = "solid", "Extrapolation" = "dashed"),
    breaks = c("Rarefaction", "Extrapolation"),
    name   = "Sampling segment"
  ) +
  labs(
    x     = "Number of individuals",
    y     = "Estimated species richness",
    title = "Size-based rarefaction: Control vs Focal",
    color = "Site", fill = "Site"
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

# COVERAGE-BASED PLOT
p_raref_coverage <- ggplot(df_pooled_coverage,
                           aes(x = x, y = y, color = Assemblage, fill = Assemblage)) +
  geom_line(aes(linetype = Method), linewidth = 1.2) +
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr),
              alpha = 0.25, color = NA) +
  geom_point(
    data = transition_coverage,
    aes(x = x, y = y),
    size = 3, shape = 21, stroke = 1,
    color = "black", fill = "white"
  ) +
  scale_color_manual(values = c("control" = "#D95F09", "focal" = "#2C7FB8")) +
  scale_fill_manual(values  = c("control" = "#F1B0AD", "focal" = "#A6C1E0")) +
  scale_linetype_manual(
    values = c("Rarefaction" = "solid", "Extrapolation" = "dashed"),
    breaks = c("Rarefaction", "Extrapolation"),
    name   = "Sampling segment"
  ) +
  labs(
    x     = "Sample coverage",
    y     = "Estimated species richness",
    title = "Coverage-based rarefaction: Control vs Focal",
    color = "Site", fill = "Site"
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

p_raref_size
p_raref_coverage
```


# Plot-level rarefaction
```{r rarefaction}
# Identify metadata and species columns
meta_cols <- c(
  "surveyID", "plotID", "site", "region", "visit", "plot",
  "bee_richness", "fieldIDsum",
  "bee_abundance_nofieldID", "bee_abundance", "bee_hill_q1", "bee_hill_q2", 
  "RLabund_LC", "RLabund_NT", "RLabund_VU", "RLabund_EN", "RLabund_DD",
  "RLrich_LC", "RLrich_NT", "RLrich_VU", "RLrich_EN", "RLrich_DD",
  "RLrich_CR", "RLabund_CR",
  "RLrich_threatened", "RLabund_threatened",
  "RLrich_total", "RLabund_total"
)

species_cols <- setdiff(names(bee_data), meta_cols)

# Aggregate visits within each plot
bee_plot <- bee_data |>
  group_by(region, site, plotID) |>
  summarise(across(all_of(species_cols), sum), .groups = "drop")

# Prepare iNEXT input list (per plot)
inext_list <- setNames(
  lapply(seq_len(nrow(bee_plot)), function(i) {
    as.numeric(bee_plot[i, species_cols])
  }),
  paste(bee_plot$region, bee_plot$site, bee_plot$plotID, sep = "_")
)
```

```{r run iNext plot scale}
# Run iNEXT: species richness (q=0), abundance data
out_inext_q0 <- iNEXT(
  inext_list,
  q = 0,
  datatype = "abundance",
  endpoint = max(sapply(inext_list, sum)) * 2
)

# Run iNEXT: Shannon diversity (q=1), abundance data
out_inext_q1 <- iNEXT(
  inext_list,
  q = 1,
  datatype = "abundance",
  endpoint = max(sapply(inext_list, sum)) * 2
)
```

# Coverage-based rarefaction plot-scale (visits merged)
## Richness
```{r coverage based rarefaction plot scale (visits merged)}
df_plot <- fortify(out_inext_q0, type = 3)

df_cov_max <- df_plot |>
  group_by(Assemblage) |>
  summarise(max_coverage = max(x), .groups = "drop")

summary(df_cov_max$max_coverage)

quantile(df_cov_max$max_coverage,
         probs = c(0.1, 0.2, 0.3, 0.4, 0.5,
                   0.6, 0.7, 0.8, 0.9, 0.95))

ggplot(df_cov_max, aes(x = max_coverage)) +
  geom_histogram(bins = 20, fill = "grey70", color = "black") +
  geom_vline(xintercept = 0.85,
             linetype = "dashed", color = "red") +
  labs(
    title = "Maximum achievable coverage across plots",
    x = "Coverage",
    y = "Number of plots"
  ) +
  theme_bw()

# Set target coverage
target_coverage <- 0.85

# Visual check
ggiNEXT(out_inext_q0, type = 3) +
  geom_vline(xintercept = target_coverage,
             linetype = "dashed",
             color = "red") +
  theme_minimal()


# use estimateD (exact interpolation)
rarefied_cov_rich <- estimateD(
  inext_list,
  q = 0,
  base = "coverage",
  level = target_coverage
)

# Prepare dataset for modelling
rarefied_cov_rich <- rarefied_cov_rich |>
  separate(Assemblage,
           into = c("region", "site", "plotID"),
           sep = "_") |>
  rename(
    richness = qD,
    richness_lwr = qD.LCL,
    richness_upr = qD.UCL,
    coverage = SC
  )
```

## Hill q1
```{r Coverage-based rarefaction plot-scale (visits merged) - Hill q1}
df_plot_q1 <- fortify(out_inext_q1, type = 3)

df_cov_max_q1 <- df_plot_q1 |>
  group_by(Assemblage) |>
  summarise(max_coverage = max(x), .groups = "drop")

summary(df_cov_max_q1$max_coverage)

quantile(df_cov_max_q1$max_coverage,
         probs = c(0.1, 0.2, 0.3, 0.4, 0.5,
                   0.6, 0.7, 0.8, 0.9, 0.95))

ggplot(df_cov_max_q1, aes(x = max_coverage)) +
  geom_histogram(bins = 20, fill = "grey70", color = "black") +
  geom_vline(xintercept = 0.85,
             linetype = "dashed", color = "red") +
  labs(
    title = "Maximum achievable coverage across plots (Hill q1)",
    x = "Coverage",
    y = "Number of plots"
  ) +
  theme_bw()

# Set target coverage
target_coverage <- 0.85

# Visual check
ggiNEXT(out_inext_q1, type = 3) +
  geom_vline(xintercept = target_coverage,
             linetype = "dashed",
             color = "red") +
  theme_minimal()

# use estimateD (exact interpolation)
rarefied_cov_q1 <- estimateD(
  inext_list,
  q = 1,
  datatype = "abundance",
  base = "coverage",
  level = target_coverage
)

# Prepare dataset for modelling
rarefied_cov_q1 <- rarefied_cov_q1 |>
  separate(Assemblage,
           into = c("region", "site", "plotID"),
           sep = "_") |>
  rename(
    hill_q1 = qD,
    hill_q1_lwr = qD.LCL,
    hill_q1_upr = qD.UCL,
    coverage = SC
  )

# check how many plots you KEEP now
length(unique(rarefied_cov_q1$plotID))
```

# Size-based rarefaction plot scale (visits merged)
## Richness
```{r Size-based rarefaction plot scale (visits merged)}
# Add total abundance per plot
bee_plot_n <- bee_plot |>
  mutate(total_abundance = rowSums(across(all_of(species_cols))))

summary(bee_plot_n$total_abundance)

quantile(bee_plot_n$total_abundance, probs = c(0.1, 0.2, 0.3))

ggplot(bee_plot_n, aes(x = total_abundance)) +
  geom_histogram(bins = 20, fill = "grey70", color = "black") +
  labs(
    title = "Total abundance per plot",
    x = "Number of individuals",
    y = "Number of plots"
  ) +
  theme_bw()

# Set target sample size
target_n <- floor(quantile(bee_plot_n$total_abundance, 0.3))
target_n

# Exact size-based rarefaction using estimateD - richness
rarefied_size_rich <- estimateD(
  inext_list,
  q = 0,
  datatype = "abundance",
  base = "size",
  level = target_n
)

# Add metadata
rarefied_size_rich <- rarefied_size_rich |>
  separate(Assemblage, into = c("region", "site", "plotID"), sep = "_") |>
  rename(
    sample_size = m,
    richness = qD,
    richness_lwr = qD.LCL,
    richness_upr = qD.UCL
  )
```


```{r Size-based rarefaction (Hill q1)}
# Exact size-based rarefaction using estimateD
rarefied_size_q1 <- estimateD(
  inext_list,
  q = 1,
  datatype = "abundance",
  base = "size",
  level = target_n
)

# Add metadata
rarefied_size_q1 <- rarefied_size_q1 |>
  separate(Assemblage, into = c("region", "site", "plotID"), sep = "_") |>
  rename(
    sample_size = m,
    hill_q1 = qD,
    hill_q1_lwr = qD.LCL,
    hill_q1_upr = qD.UCL
  )

# Check number of plots retained
length(unique(rarefied_size_q1$plotID))
```

# Save rarefied data
```{r Generate combined datasets}
# load base data
data_plot_scaled <- readRDS(file.path(
  "..", "data", "cleaned", "data_plot_scaled.rds"))

# create coverage-based rarefaction dataset (richness + hill q1)
data_model_cov <- data_plot_scaled |>
  left_join(
    rarefied_cov_rich |>
      select(region, site, plotID, richness_rarefied_cov = richness),
    by = c("region", "site", "plotID")
  ) |>
  left_join(
    rarefied_cov_q1 |>
      select(region, site, plotID, hill_q1_rarefied_cov = hill_q1),
    by = c("region", "site", "plotID")
  ) |>
  filter(!is.na(richness_rarefied_cov) | !is.na(hill_q1_rarefied_cov))

# create size-based rarefaction dataset (richness + hill q1)
data_model_size <- data_plot_scaled |>
  left_join(
    rarefied_size_rich |>
      select(region, site, plotID, richness_rarefied_size = richness),
    by = c("region", "site", "plotID")
  ) |>
  left_join(
    rarefied_size_q1 |>
      select(region, site, plotID, hill_q1_rarefied_size = hill_q1),
    by = c("region", "site", "plotID")
  ) |>
  filter(!is.na(richness_rarefied_size) | !is.na(hill_q1_rarefied_size))
```

```{r export datasets}
saveRDS(data_model_cov, file = file.path("..","data","cleaned","data_model_cov.rds"))
saveRDS(data_model_size, file = file.path("..","data","cleaned","data_model_size.rds"))
```

