---
title: "04_bee_rarefaction"
output: html_document
---

```{r}
library(iNEXT)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
```

Coverage‑based rarefaction (Chao et al. 2014, 2020) provides descriptive standardization, not inferential estimates, and its resulting diversity values do not follow the distributional assumptions required for GLMMs (e.g. Gotelli & Colwell 2001; Chao & Jost 2012). Rarefaction also collapses plot‑level replication, removes ecological variance, and prevents correct estimation of random effects.
Because our sampling design includes spatial and hierarchical structure (multiple visits per plot; plots nested within sites; sites nested within regions), a GLMM using plot‑level richness/abundance with region and plotID as random effects is the statistically appropriate inferential framework.

```{r Load data}
# Compact helper function to read raw CSVs
bee_data <- read.csv(file.path("..", "data", "cleaned", "bee_data.csv"),
           stringsAsFactors = FALSE,
           check.names = FALSE)
```

# Site-level rarefaction 
```{r colorblidn friendly approach}

# Identify metadata and species columns
meta_cols <- c(
  "surveyID", "plotID", "site", "region", "visit", "plot",
  "richness", "fieldIDsum",
  "bee_abundance_nofieldID", "bee_abundance",
  "RLabund_LC", "RLabund_NT", "RLabund_VU", "RLabund_EN", "RLabund_DD",
  "RLrich_LC", "RLrich_NT", "RLrich_VU", "RLrich_EN", "RLrich_DD",
  "RLrich_CR", "RLabund_CR",
  "RLrich_threatened", "RLabund_threatened",
  "RLrich_total", "RLabund_total"
)

species_cols <- setdiff(names(bee_data), meta_cols)


# Aggregate all plots per site
bee_site <- bee_data %>%
  group_by(region, site) %>%
  summarise(across(all_of(species_cols), sum), .groups = "drop")


# Prepare iNEXT input list (site-level)
inext_site <- setNames(
  lapply(seq_len(nrow(bee_site)), function(i) {
    as.numeric(bee_site[i, species_cols])
  }),
  paste(bee_site$region, bee_site$site, sep = "_")
)


# Run iNEXT for abundance data (q = 0 richness)
out_site <- iNEXT(
  inext_site,
  q = 0,
  datatype = "abundance",
  endpoint = max(sapply(inext_site, sum)) * 2
)
```

```{r}

# Extract coverage-based rarefaction data
df_site <- fortify(out_site, type = 3)


# Recover region and site labels
df_site <- df_site %>%
  separate(Assemblage, into = c("region", "site"), sep = "_")


# Plot coverage-based rarefaction curves (SITE level)
p_covraref_site <- ggplot(df_site,
                          aes(x = x, y = y, color = site, fill = site)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr),
              alpha = 0.25, color = NA) +
  facet_wrap(~ region) +
  labs(
    x = "Sample coverage",
    y = "Estimated species richness",
    title = "Coverage-based rarefaction by site",
    color = "Site",
    fill = "Site"
  ) +
  scale_color_manual(values = c("focal" = "#2C7FB8", "control" = "#D95F02")) +
  scale_fill_manual(values = c("focal" = "#A6C1E0", "control" = "#F1B0AD")) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

p_covraref_site

```
At low coverage, focal and control look similar → sampling effort not yet enough to reveal differences.
At moderate coverage (0.6–0.9), focal curves are consistently above control → higher richness when standardized for completeness.
Near full coverage, extrapolated ends diverge sharply, showing that focal assemblages have a larger “potential species pool”.


The coverage‑based rarefaction curves allow us to compare focal and control sites at the same level of sampling completeness, even if neither group has been fully sampled and many bee species remain undetected. The x‑axis indicates sample coverage, meaning the proportion of the underlying community that has actually been observed. As we approach coverage = 1, the focal sites show a pronounced upward spike. This reflects the presence of many very rare species: with only a small amount of additional sampling, a disproportionately large number of new species would still be added. In contrast, the control sites lack such a sharp increase, suggesting fewer rare species and a community that is more fully sampled. Below about 0.8 coverage, the curves for focal and control sites largely overlap, indicating that the core, well‑sampled portion of the community is similar between treatments.”

gpt


```{r export rarefaction curves as tiff}
tiff(
  file.path("..", "output", "figures", "cov_rarefaction_site_600dpi.tiff"),
  width = 3000, height = 2250, res = 600, compression = "lzw"
)

print(p_covraref_site)  # <-- replace with your actual plot object
dev.off()
```

```{r rarefaction values}

asymptotic_df <- out_inext$AsyEst %>%
  # keep species richness (q = 0)
  filter(Diversity %in% c("Species richness", "q = 0", "q=0")) %>%
  # split "region_site" back into columns
  separate(Assemblage, into = c("region", "site"), sep = "_", remove = FALSE) %>%
  transmute(
    region,
    site,
    Assemblage,
    S_est = Estimator,
    S_SE  = `s.e.`,
    S_lwr = LCL %||% (Estimator - 1.96 * `s.e.`),
    S_upr = UCL %||% (Estimator + 1.96 * `s.e.`)
  )

asymptotic_df


bee_abundance_by_site <- bee_data %>%
  group_by(region, site) %>%
  summarise(
    total_bee_abundance = sum(bee_abundance, na.rm = TRUE),
    n_plots = n()
  ) %>%
  arrange(region, site)

bee_abundance_by_site

DataInfo(inext_list, datatype = "abundance")

print(fortify(out_inext, type = 3))
```



# Plot-level rarefaction and model
```{r rarefaction}
# ==============================
# Identify metadata and species columns
# ==============================
meta_cols <- c(
  "surveyID", "plotID", "site", "region", "visit", "plot",
  "richness", "fieldIDsum",
  "bee_abundance_nofieldID", "bee_abundance",
  "RLabund_LC", "RLabund_NT", "RLabund_VU", "RLabund_EN", "RLabund_DD",
  "RLrich_LC", "RLrich_NT", "RLrich_VU", "RLrich_EN", "RLrich_DD",
  "RLrich_CR", "RLabund_CR",
  "RLrich_threatened", "RLabund_threatened",
  "RLrich_total", "RLabund_total"
)

species_cols <- setdiff(names(bee_data), meta_cols)

# ==============================
# Aggregate visits WITHIN each plot
# ==============================
bee_plot <- bee_data %>%
  group_by(region, site, plotID) %>%
  summarise(across(all_of(species_cols), sum), .groups = "drop")

# ==============================
# Prepare iNEXT input list (per plot)
# ==============================
inext_list <- setNames(
  lapply(seq_len(nrow(bee_plot)), function(i) {
    as.numeric(bee_plot[i, species_cols])
  }),
  paste(bee_plot$region, bee_plot$site, bee_plot$plotID, sep = "_")
)

# ==============================
# Run iNEXT: species richness (q=0), abundance data
# ==============================
out_inext <- iNEXT(
  inext_list,
  q = 0,
  datatype = "abundance",
  endpoint = max(sapply(inext_list, sum)) * 2
)
```

```{r plot}
# ==============================
# Extract coverage-based rarefaction data
# ==============================
df_plot <- fortify(out_inext, type = 3)  # type = 3 = coverage-based rarefaction

# Recover metadata
df_plot <- df_plot %>%
  separate(Assemblage, into = c("region", "site", "plotID"), sep = "_")

# ==============================
# Plot (per plot)
# ==============================
p_covraref <- ggplot(df_plot,
                      aes(x = x, y = y, color = site, group = plotID)) +
  geom_line(alpha = 0.7) +
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr, fill = site), alpha = 0.15) +
  facet_wrap(~ region) +
  labs(
    x = "Sample coverage",
    y = "Estimated species richness",
    title = "Coverage-based rarefaction by plot",
    color = "Site",
    fill = "Site"
  ) +
  theme_bw(base_size = 12) +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold"))

p_covraref
```
```{r Model with rarefaction values}
df_plot <- fortify(out_inext, type = 3)


df_plot <- df_plot %>%
  # separate(Assemblage, into = c("region", "site", "plotID"), sep = "_")

extract_rarefied <- function(df, target_coverage) {
  df %>%
    filter(Method != "Observed") %>%             # exclude observed-only point
    group_by(region, site, plotID) %>%
    slice(which.min(abs(x - target_coverage))) %>%  # closest to target SC
    ungroup() %>%
    select(
      region, site, plotID,
      coverage = x,
      richness = y,
      richness_lwr = y.lwr,
      richness_upr = y.upr,
      Method
    )
}


target_coverage <- 0.85   # or 0.90 depending on your choice

df_plot_rarefied <- extract_rarefied(df_plot, target_coverage)



bm_rich <- glmmTMB(
  richness ~ site + (1 | region),
  family = nbinom2,
  data = df_plot_rarefied
)

summary(bm_rich)

```
It’s better to model the raw survey data because it preserves the full hierarchical structure of your sampling—plots, visits, and regions—allowing correct estimation of variability and detection effects, which rarefaction discards. Coverage‑rarefied richness values are generated estimates with their own uncertainty, not independent observations, and using them in regressions can distort variance and bias inference (a point emphasized by Gotelli & Colwell 2001). Lastly, raw‑data GLMMs align with best‑practice recommendations for analyzing ecological count data (Bolker et al. 2009), whereas rarefaction is intended primarily for descriptive standardization, not as the primary response in statistical models.
