---
title: "04_bee_rarefaction"
output: html_document
---

```{r}
library(iNEXT)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
```

Coverage‑based rarefaction (Chao et al. 2014, 2020) provides descriptive standardization, not inferential estimates, and its resulting diversity values do not follow the distributional assumptions required for GLMMs (e.g. Gotelli & Colwell 2001; Chao & Jost 2012). Rarefaction also collapses plot‑level replication, removes ecological variance, and prevents correct estimation of random effects.
Because our sampling design includes spatial and hierarchical structure (multiple visits per plot; plots nested within sites; sites nested within regions), a GLMM using plot‑level richness/abundance with region and plotID as random effects is the statistically appropriate inferential framework.

```{r Load data}
# Compact helper function to read raw CSVs
bee_data <- read.csv(file.path("..", "data", "cleaned", "bee_data.csv"),
           stringsAsFactors = FALSE,
           check.names = FALSE)
```

```{r}

# ==============================
# Identify metadata and species columns
# ==============================
meta_cols <- c(
  "surveyID", "plotID", "site", "region", "visit", "plot",
  "richness", "fieldIDsum",
  "bee_abundance_nofieldID", "bee_abundance",
  "RLabund_LC", "RLabund_NT", "RLabund_VU", "RLabund_EN", "RLabund_DD",
  "RLrich_LC", "RLrich_NT", "RLrich_VU", "RLrich_EN", "RLrich_DD",
  "RLrich_CR", "RLabund_CR",
  "RLrich_threatened", "RLabund_threatened",
  "RLrich_total", "RLabund_total"
)

species_cols <- setdiff(names(bee_data), meta_cols)

# ==============================
# Aggregate all plots per site
# ==============================
bee_site <- bee_data %>%
  group_by(region, site) %>%
  summarise(across(all_of(species_cols), sum), .groups = "drop")

# ==============================
# Prepare iNEXT input list
# ==============================
inext_list <- setNames(
  lapply(seq_len(nrow(bee_site)), function(i) {
    as.numeric(bee_site[i, species_cols])
  }),
  paste(bee_site$region, bee_site$site, sep = "_")
)

# ==============================
# Run iNEXT for abundance data, species richness (q=0)
# Coverage-based rarefaction + extrapolation
# ==============================
out_inext <- iNEXT(
  inext_list,
  q = 0,                  # species richness
  datatype = "abundance",
  endpoint = max(sapply(inext_list, sum)) * 2  # extrapolate to double sample size
)
```

```{r}
# ==============================
# Extract coverage-based rarefaction data
# ==============================
df_plot <- fortify(out_inext, type = 3)  # type = 3 = sample coverage

# Recover region and site labels
df_plot <- df_plot %>%
  separate(Assemblage, into = c("region", "site"), sep = "_")

# ==============================
# Plot coverage-based rarefaction curves
# ==============================
ggplot(df_plot, aes(x = x, y = y, color = site, fill = site)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr), alpha = 0.25, color = NA) +
  facet_wrap(~ region) +
  labs(
    x = "Sample coverage",
    y = "Estimated species richness",
    title = "Coverage-based rarefaction by site",
    color = "Site",
    fill = "Site"
  ) +
  scale_color_manual(values = c("focal" = "#2C7FB8", "control" = "#D95F02")) +
  scale_fill_manual(values = c("focal" = "#A6C1E0", "control" = "#F1B0AD")) +
  theme_bw(base_size = 12) +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold"))

```
At low coverage, focal and control look similar → sampling effort not yet enough to reveal differences.
At moderate coverage (0.6–0.9), focal curves are consistently above control → higher richness when standardized for completeness.
Near full coverage, extrapolated ends diverge sharply, showing that focal assemblages have a larger “potential species pool”.
gpt

```{r}

asymptotic_df <- out_inext$AsyEst %>%
  # keep species richness (q = 0)
  filter(Diversity %in% c("Species richness", "q = 0", "q=0")) %>%
  # split "region_site" back into columns
  separate(Assemblage, into = c("region", "site"), sep = "_", remove = FALSE) %>%
  transmute(
    region,
    site,
    Assemblage,
    S_est = Estimator,
    S_SE  = `s.e.`,
    S_lwr = LCL %||% (Estimator - 1.96 * `s.e.`),
    S_upr = UCL %||% (Estimator + 1.96 * `s.e.`)
  )

asymptotic_df


bee_abundance_by_site <- bee_data %>%
  group_by(region, site) %>%
  summarise(
    total_bee_abundance = sum(bee_abundance, na.rm = TRUE),
    n_plots = n()
  ) %>%
  arrange(region, site)

bee_abundance_by_site

DataInfo(inext_list, datatype = "abundance")

print(fortify(out_inext, type = 3))
```

