---
title: "05_site_unimodel"
output: word_document
---

```{r Load packages}
library(DHARMa)
library(dplyr)
library(emmeans)
library(ggeffects)
library(ggplot2)
library(glmmTMB)
library(gt)
library(lmerTest)
library(performance)
library(tidyverse)
```


```{r Load data}
data_all_scaled <- readRDS(file.path("..", "data", "cleaned", "data_all_scaled.rds"))
```

Coverage‑based rarefaction (Chao et al. 2014, 2020) provides descriptive standardization, not inferential estimates, and its resulting diversity values do not follow the distributional assumptions required for GLMMs (e.g. Gotelli & Colwell 2001; Chao & Jost 2012). Rarefaction also collapses plot‑level replication, removes ecological variance, and prevents correct estimation of random effects.
Because our sampling design includes spatial and hierarchical structure (multiple visits per plot; plots nested within sites; sites nested within regions), a GLMM using plot‑level richness/abundance with region and plotID as random effect is the statistically appropriate inferential framework.


Reason behind random effects strutcure:
Bee richness depends on the site, but richness also varies from region to region and from plot to plot. Also, the way richness changes over visits is different for each plot. The data are counts with overdispersion.”

A random intercept for region.
Bee richness varies among regions.
Each region gets its own intercept—i.e., its baseline bee richness level can be higher or lower than the global mean.
You assume that observations within a region are correlated.

A random intercept AND random slope for visit, grouped by plot.
This means:
Each plot has its own baseline bee richness (random intercept).
The effect of visit (e.g., sampling round, date, etc.) varies among plots (random slope).
So each plot has:
its own starting point (intercept),
its own trend across visits (slope),
and usually a correlation between the intercept and the slope is also estimated (unless you specify ||).

gpt


# Richness
```{r fit richness model on nbinom2}
nb2_rich <- glmmTMB(
  bee_richness ~ site + (1 | plotID),
  family = nbinom2,
  data = data_all_scaled
)

summary(nb2_rich)

res_nb2 <- simulateResiduals(nb2_rich)
plot(res_nb2)
testDispersion(nb2_rich)
testZeroInflation(nb2_rich)

check_model(nb2_rich)
```

```{r fit richness model on nbinom1}
nb1_rich <- glmmTMB(
  bee_richness ~ site + (1 | plotID),
  family = nbinom1,
  data = data_all_scaled
)

summary(nb1_rich)

res_nb1 <- simulateResiduals(nb1_rich)
plot(res_nb1)
testDispersion(nb1_rich)
testZeroInflation(nb1_rich)

check_model(nb1_rich)
```


To analyse the effect of site on bee richness, we fitted generalized linear mixed models including site as a fixed effect and plot identity as a random intercept to account for repeated sampling. Negative binomial models with alternative variance–mean relationships (nbinom1 and nbinom2) were compared using identical fixed‑ and random‑effects structures. For this model, the nbinom1 parameterisation provided a better fit, as indicated by a lower AIC and comparable or improved residual diagnostics. The nbinom1 model showed no evidence of over‑ or underdispersion, zero inflation, or residual structure based on DHARMa diagnostics, and parameter estimates were stable and biologically interpretable. Given the simpler random‑effects structure and approximately linear scaling of variance with the mean, the nbinom1 family was retained for inference.
gpt


```{r richness site model}
RI_site <- glmmTMB(
  bee_richness ~ site + (1 | plotID),
  family = nbinom1,
  data = data_all_scaled
)

print(summary(RI_site))
```

```{r richness site model predicted values}
# Absolute predicted richness per site
pred_ri <- emmeans(RI_site, ~ site, type = "response")

# Absolute difference
contrast(pred_ri, "revpairwise")
```

```{r plot model predictions richness}
# Predicted marginal means for site on the response scale
pred_site_rich <- ggpredict(RI_site, terms = "site", type = "fixed")

# ensure plotting order (control, focal)
pred_site_rich$x <- factor(pred_site_rich$x, levels = c("control", "focal"))
data_all_scaled$site <- factor(data_all_scaled$site, levels = c("control", "focal"))

# Color palette
site_cols <- list(
  "focal" = "#2C7FB8",
  "control"   = "#D95F09",
  "focal_light"   = "cadetblue3",
  "control_light" = "#D95"
  )

# Plot: predicted means ± 95% CI with raw jittered points

p_rich <- ggplot() +
  # 1) Raw data (plot × visit), jittered in background
  geom_jitter(
    data = data_all_scaled,
    aes(x = site, y = bee_richness, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +

  # 2) Model-predicted means (ggeffects output; response scale)
  geom_point(
    data = pred_site_rich,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +

  # 3) 95% confidence intervals
  geom_errorbar(
    data = pred_site_rich,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +

  # Use identical site colours for both layers
  scale_colour_manual(values = site_cols) +

  # Labels
  labs(
    x = NULL,
    y = "Bee species richness",
    title = "Predicted Bee Richness per Plot by Site",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = raw observations"
  ) +
  
  # Theme
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  )

p_rich
```

```{r print richness plot to tiff}
tiff(
  file.path("..", "output", "figures", "pred_bee_ri_site_600dpi.tiff"),
  width = 3000, height = 2250, res = 600, compression = "lzw"
)

print(p_rich)
dev.off()
```


# Abundance
```{r fit abundance model on nbinom2}
nb2_abun <- glmmTMB(
  bee_abundance ~ site + (1 | plotID),
  family = nbinom2,
  data = data_all_scaled
)

summary(nb2_abun)

res_nb2 <- simulateResiduals(nb2_abun)
plot(res_nb2)
testDispersion(nb2_abun)
testZeroInflation(nb2_abun)

check_model(nb2_abun)
```

```{r fit abundance model on nbinom1}
nb1_abun <- glmmTMB(
  bee_abundance ~ site + (1 | plotID),
  family = nbinom1,
  data = data_all_scaled
)

summary(nb1_abun)

res_nb1 <- simulateResiduals(nb1_abun)
plot(res_nb1)
testDispersion(nb1_abun)
testZeroInflation(nb1_abun)

check_model(nb1_abun)
```

```{r abundance site model}
AB_site <- glmmTMB(
  bee_abundance ~ site + (1 | plotID),
  family = nbinom1,
  data = data_all_scaled
)

print(summary(AB_site))
```

```{r abundance site model predicted values}
# Absolute predicted richness per site
pred_ab <- emmeans(AB_site, ~ site, type = "response")

# Absolute difference
contrast(pred_ab, "revpairwise")
```

```{r plot model predictions abundance}

# Predicted marginal means for site on the response scale
pred_site_abun <- ggpredict(AB_site, terms = "site", type = "fixed")

# ensure plotting order (control, focal)
pred_site_abun$x <- factor(pred_site_abun$x, levels = c("control", "focal"))
data_all_scaled$site <- factor(data_all_scaled$site, levels = c("control", "focal"))

# Color palette
site_cols <- list(
  "focal" = "#2C7FB8",
  "control"   = "#D95F09",
  "focal_light"   = "cadetblue3",
  "control_light" = "#D95"
  )

# Plot predicted means ± 95% CI with raw jittered points

p_abun <- ggplot() +
  # 1) Raw data (plot × visit), jittered in background
  geom_jitter(
    data = data_all_scaled,
    aes(x = site, y = bee_abundance, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +

  # 2) Model-predicted means (ggeffects output; response scale)
  geom_point(
    data = pred_site_abun,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +

  # 3) 95% confidence intervals
  geom_errorbar(
    data = pred_site_abun,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +

  # Use identical site colours for both layers
  scale_colour_manual(values = site_cols) +

  # Labels
  labs(
    x = NULL,
    y = "Bee abundance",
    title = "Predicted Bee Abundance per Plot by Site",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = raw observations"
  ) +

  # Theme
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  )

p_abun
```

```{r print abundance plot to tiff}
tiff(
  file.path("..", "output", "figures", "pred_bee_ab_site_600dpi.tiff"),
  width = 3000, height = 2250, res = 600, compression = "lzw"
)

print(p_abun)
dev.off()
```

# Hill q1
```{r}
hist(data_all_scaled$bee_hill_q1)
```

```{r Hillq1 site model}
Q1_site <- glmmTMB(
  bee_hill_q1~ site + (1 | plotID),
  family = Gamma(link = "log"),
  data = data_all_scaled
)

summary(Q1_site)

res_Q1 <- simulateResiduals(Q1_site)
plot(res_Q1)
testDispersion(Q1_site)
testZeroInflation(Q1_site)

check_model(Q1_site)
```
Residual diagnostics indicated minor deviations from uniformity associated with differences in residual variance between site types. However, dispersion, outlier, and convergence diagnostics showed no evidence of model misspecification, and parameter estimates were stable. These patterns are consistent with strong mean differences under a Gamma error structure and do not affect inference.
gpt

```{r Hillq1 site model predicted values}
# Absolute predicted richness per site
pred_q1 <- emmeans(Q1_site, ~ site, type = "response")
contrast(pred_q1, "revpairwise")
```

```{r plot model predictions hillq1}
# --- Predicted marginal means for 'site' on the response scale ---
# (type = "fe" keeps predictions at average random effects; type = "response" back-transforms)
pred_site_q1 <- ggpredict(Q1_site, terms = "site", type = "fixed")
# pred_site_q1 has columns: x, predicted, conf.low, conf.high (on response scale)

# --- Optional: ensure plotting order (control, focal) ---
pred_site_q1$x <- factor(pred_site_q1$x, levels = c("control", "focal"))
data_all_scaled$site <- factor(data_all_scaled$site, levels = c("control", "focal"))

# --- Colors (your palette) ---
site_cols <- list(
  "focal" = "#2C7FB8",
  "control"   = "#D95F09",
  "focal_light"   = "cadetblue3",
  "control_light" = "#D95"
  )

# --- Plot: predicted means ± 95% CI with raw jittered points ---

p_q1 <- ggplot() +
  # 1) Raw data (plot × visit), jittered in background
  geom_jitter(
    data = data_all_scaled,
    aes(x = site, y = bee_hill_q1, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +

  # 2) Model-predicted means (ggeffects output; response scale)
  geom_point(
    data = pred_site_q1,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +

  # 3) 95% confidence intervals
  geom_errorbar(
    data = pred_site_q1,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +

  # Use identical site colours for both layers
  scale_colour_manual(values = site_cols) +

  # Labels
  labs(
    x = NULL,
    y = "Bee Shannon diversity",
    title = "Predicted Bee Shannon Diversity per Plot by Site",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = raw observations"
  ) +

  # Theme
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  )

p_q1
```


```{r print q1 plot to tiff}
tiff(
  file.path("..", "output", "figures", "pred_bee_q1_site_600dpi.tiff"),
  width = 3000, height = 2250, res = 600, compression = "lzw"
)

print(p_q1)
dev.off()
```

# Richness Site and region interative model for context specificity
```{r interactive site region}
RI_sitereg <- glmmTMB(
  bee_richness ~ region * site + 
    (1 | plotID),
  family = nbinom2,
  data = data_all_scaled
)

summary(RI_sitereg)
```
Fixed effects:

Bee richness differed significantly among regions (Table X), with strong evidence for lower richness in Region Thiele relative to the reference region (β = −1.55 ± 0.35, p < 0.001).
Across regions, the focal site had substantially higher bee richness than the control site (β = 1.16 ± 0.23, p < 0.001).
However, the effect of site depended on region, as indicated by a significant region × site interaction for Region Thiele (β = 1.54 ± 0.41, p < 0.001). This interaction reveals that the difference in bee richness between focal and control sites was markedly stronger in Region Thiele than in the reference region, whereas site effects did not differ significantly among the other regions.
And for the random effects:

Random effects accounted for repeated sampling at the plot level, with moderate variation in baseline bee richness among plots (SD = 0.18). This indicates that plots differed in overall richness independently of region and site effects and justifies the inclusion of plot identity as a random intercept.
gpt

```{r plot site region interactive model}
data_all_scaled$site   <- factor(data_all_scaled$site,   levels = c("control", "focal"))
data_all_scaled$region <- factor(data_all_scaled$region)

# Color palette
site_cols <- c(
  "control" = "#D95F09",
  "focal"   = "#2C7FB8"
)

# emmeans predictions
em_rs <- emmeans(RI_sitereg, ~ site | region, type = "response")
pred_rs <- as.data.frame(em_rs)

# rename to match your existing plot code
pred_rs <- pred_rs %>%
  rename(predicted = response,
         conf.low  = asymp.LCL,
         conf.high = asymp.UCL)

pred_rs$region <- factor(pred_rs$region, levels = levels(data_all_scaled$region))
pred_rs$site   <- factor(pred_rs$site,   levels = c("control", "focal"))

# Plot
p_rich_rs <- ggplot() +

  # raw data jitter
  geom_jitter(
    data = data_all_scaled,
    aes(x = site, y = bee_richness, colour = site),
    width = 0.12, alpha = 0.20, size = 1.6,
    show.legend = FALSE
  ) +

  # predicted means 
  geom_point(
    data = pred_rs,
    aes(x = site, y = predicted, colour = site),
    size = 3.8,
    show.legend = FALSE
  ) +

  # 95% CIs  
  geom_errorbar(
    data = pred_rs,
    aes(x = site, ymin = conf.low, ymax = conf.high, colour = site),
    width = 0.10, linewidth = 1,
    show.legend = FALSE
  ) +

  scale_colour_manual(values = site_cols) +

  # facet by region
  facet_wrap(~ region, nrow = 1) +

  labs(
    x = NULL,
    y = "Bee species richness",
    title = "Predicted Bee Richness by Region × Site",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = raw observations"
  ) +

  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8)),
    panel.spacing = unit(0,"lines")
  )

p_rich_rs

```
```{r print richness sitreg plot to tiff}
tiff(
  file.path("..", "output", "figures", "pred_bee_ri_sitexregion_600dpi.tiff"),
  width = 3000, height = 2250, res = 600, compression = "lzw"
)
print(p_rich_rs)
dev.off()
```

```{r predictions richeness site region}
emmeans(RI_sitereg, pairwise ~ site | region)
```
The effect size variation is still worth noting
Back-transforming the contrasts to multiplicative effects:

Aire  3.2× (2.0–5.0×)
Scheulte  2.6× (1.7–4.0×)
Schwarzbach 2.4× (1.5–3.7×)
Thiele1 4.9× (7.7–28.7×)

Focal sites supported significantly higher bee species richness than control sites in all four regions, with the magnitude of the effect varying strongly across regions (Figure X, Table X). Back‑transformed model contrasts indicate that bee richness at focal sites was approximately 2.9× higher in Aire, 2.6× higher in Scheulte, and 2.3× higher in Schwarzbach compared to their respective control sites.
In Region Thiele, however, the focal–control contrast was substantially larger, with focal sites supporting approximately 14× higher bee richness than control sites. This pronounced effect resulted in a significant region × site interaction and indicates that the strength of the focal site effect in Thiele was qualitatively different from that observed in the other regions.
Together, these results demonstrate a consistent positive effect of focal sites on bee richness across all regions, while also revealing substantial spatial variation in effect magnitude. The exceptionally strong response in Region Thiele suggests that local or landscape‑level factors may strongly amplify the effectiveness of focal sites there, potentially reflecting differences in initial habitat degradation, restoration intensity or quality, or broader landscape context.
gpt
```{r}
# interaction-style predictions
em_int <- emmeans(RI_sitereg, ~ region * site, type = "response")
pred_int <- as.data.frame(em_int)

# rename for plotting
pred_int <- pred_int %>%
  rename(
    predicted = response,
    conf.low  = asymp.LCL,
    conf.high = asymp.UCL
  )

# ensure factor order
pred_int$region <- factor(pred_int$region, levels = levels(data_all_scaled$region))
pred_int$site   <- factor(pred_int$site, levels = c("control", "focal"))

ggplot(pred_int,
       aes(x = region, y = predicted, color = site, group = site)) +
  
  # lines = interaction
  geom_line(linewidth = 1.2) +
  
  # points
  geom_point(size = 3) +
  
  # confidence intervals
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0.15
  ) +
  
  scale_color_manual(values = site_cols) +
  
  labs(
    x = "Region",
    y = "Predicted bee richness",
    title = "Interaction between region and site",
    subtitle = "Non-parallel lines indicate interaction"
  ) +
  
  theme_bw()
```

# Model output tables
```{r model output table univariate models}
# --- Helper function to extract table from a model ---
extract_site_table <- function(model, response_name) {
  
  # emmeans (response scale)
  em <- emmeans(model, ~ site, type = "response")
  em_df <- as.data.frame(em)
  
  # reshape means (control vs focal)
  means_wide <- em_df %>%
    mutate(site = as.character(site)) %>%
    transmute(
      site,
      mean = response,
      lower = asymp.LCL,
      upper = asymp.UCL
    ) %>%
    mutate(mean_ci = sprintf("%.2f (%.2f–%.2f)", mean, lower, upper)) %>%
    select(site, mean_ci) %>%
    pivot_wider(names_from = site, values_from = mean_ci)
  
  # contrasts (ratio focal vs control)
  contr <- contrast(em, "revpairwise") %>%
    summary(infer = TRUE, type = "response") %>%
    as.data.frame()
  
  contr_out <- contr %>%
    transmute(
      effect_ratio = sprintf("%.2f", ratio),
      conf_int = sprintf("%.2f–%.2f", asymp.LCL, asymp.UCL),
      p_value = signif(p.value, 3)
    )
  
  # combine
  out <- bind_cols(means_wide, contr_out) %>%
    mutate(response = response_name, .before = everything())
  
  return(out)
}

# --- Apply to your models ---
tab_abun <- extract_site_table(AB_site, "Abundance")
tab_rich <- extract_site_table(RI_site, "Richness")
tab_q1   <- extract_site_table(Q1_site, "Shannon (q=1)")

# combine
table_univsite <- bind_rows(tab_abun, tab_rich, tab_q1)

table_univsite

```

```{r save univsite model output table}
gt_table_univsite <- table_univsite %>%
  gt() %>%
  cols_label(
    response = "Response",
    control  = "Control (mean ± 95% CI)",
    focal    = "Focal (mean ± 95% CI)",
    effect_ratio = "Effect Ratio",
    conf_int = "95% CI",
    p_value = "p-value"
  )

gtsave(gt_table_univsite, "../output/tables/site_univ_effects.docx")
```

```{r Model output table interactive model}
# emmeans by region
em_rs <- emmeans(RI_sitereg, ~ site | region, type = "response")
em_rs_df <- as.data.frame(em_rs)

# means
means_rs <- em_rs_df %>%
  transmute(
    region,
    site,
    mean_ci = sprintf("%.2f (%.2f–%.2f)",
                      response, asymp.LCL, asymp.UCL)
  ) %>%
  pivot_wider(names_from = site, values_from = mean_ci)

# contrasts per region
contr_rs <- contrast(em_rs, "revpairwise") %>%
  summary(infer = TRUE, type = "response") %>%
  as.data.frame()

contr_rs_out <- contr_rs %>%
  transmute(
    region,
    effect_ratio = sprintf("%.2f", ratio),
    conf_int = sprintf("%.2f–%.2f", asymp.LCL, asymp.UCL),
    p_value = signif(p.value, 3)
  )

# final table
table_interactsite<- left_join(means_rs, contr_rs_out, by = "region")

table_interactsite
```

```{r save interactive model output table}
gt_table_interactsite <- table_interactsite %>%
  gt() %>%
  cols_label(
    region = "Region",
    control  = "Control (mean ± 95% CI)",
    focal    = "Focal (mean ± 95% CI)",
    effect_ratio = "Effect Ratio",
    conf_int = "95% CI",
    p_value = "p-value"
  )

gtsave(gt_table_interactsite, "../output/tables/site_interact_effects.docx")
```



# Merged visits models

Load data
```{r}
data_plot_scaled <- readRDS(
  file.path("..", "data", "cleaned", "data_plot_scaled.rds")
)
```

```{r fit richness model on nbinom2}
nb2_rich <- glmmTMB(
  bee_richness ~ site + (1 | region),
  family = nbinom2,
  data = data_plot_scaled
)

summary(nb2_rich)

res_nb2 <- simulateResiduals(nb2_rich)
plot(res_nb2)
testDispersion(nb2_rich)
testZeroInflation(nb2_rich)

check_model(nb2_rich)
```

```{r fit richness model on nbinom1}
nb1_rich <- glmmTMB(
  bee_richness ~ site + (1 | region),
  family = nbinom1,
  data = data_plot_scaled
)

summary(nb1_rich)

res_nb1 <- simulateResiduals(nb1_rich)
plot(res_nb1)
testDispersion(nb1_rich)
testZeroInflation(nb1_rich)

check_model(nb1_rich)
```

```{r richness pooled site model}
PRI_site <- glmmTMB(
  bee_richness ~ site + (1 | region),
  family = nbinom1,
  data = data_plot_scaled
)

print(summary(PRI_site))
```

```{r richness site model predicted values}
# Absolute predicted richness per site
pred_pri <- emmeans(PRI_site, ~ site, type = "response")

# Absolute difference
contrast(pred_pri, "revpairwise")
```

```{r plot model predictions richness}
# Predicted marginal means for site on the response scale
pred_site_prich <- ggpredict(PRI_site, terms = "site", type = "fixed")

# ensure plotting order (control, focal)
pred_site_prich$x <- factor(pred_site_prich$x, levels = c("control", "focal"))
data_plot_scaled$site <- factor(data_plot_scaled$site, levels = c("control", "focal"))

# Color palette
site_cols <- list(
  "focal" = "#2C7FB8",
  "control"   = "#D95F09",
  "focal_light"   = "cadetblue3",
  "control_light" = "#D95"
  )

# Plot: predicted means ± 95% CI with raw jittered points

p_prich <- ggplot() +
  # 1) Raw data (plot × visit), jittered in background
  geom_jitter(
    data = data_plot_scaled,
    aes(x = site, y = bee_richness, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +

  # 2) Model-predicted means (ggeffects output; response scale)
  geom_point(
    data = pred_site_prich,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +

  # 3) 95% confidence intervals
  geom_errorbar(
    data = pred_site_prich,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +

  # Use identical site colours for both layers
  scale_colour_manual(values = site_cols) +

  # Labels
  labs(
    x = NULL,
    y = "Bee species richness",
    title = "Predicted Bee Richness per Plot by Site (pooled visits)",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = raw observations"
  ) +
  
  # Theme
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  )

p_prich
```


# Plant diversity
```{r fit model}

nm_plant <- glmmTMB(
  richness ~ 1 + (1 | region) + (1 + visit | plotID),
  family = poisson,
  data = data_all_scaled
)

bm_plant <- glmmTMB(
  richness ~ site + (1 | region) + (1 + visit | plotID),
  family = poisson,
  data = data_all_scaled
)

anova(nm_plant, bm_plant)
# random-effect structure is stable enough to support inference
# the visit random slope is not causing singularity
# adding site explains substantial variation in richness
# the data contain enough replication for strong inference

res <- simulateResiduals(bm_plant)
plot(res)
# no evidence for overdispersion, no deviation from expected residualdistribution
# no outliers, no heterogeneity of variance between site types

testDispersion(bm_plant)
testZeroInflation(bm_plant)


summary(bm_plant)
confint(bm_plant)
emmeans(bm_plant, pairwise ~ site)

emm <- emmeans(bm_plant, ~ site, type = "response")
summary(emm)

# Absolute and relative difference
exp(1.62) - exp(1.00)  # absolute
exp(1.62) / exp(1.00)  # ratio
# or directly from contrast:
exp(0.62)              # ratio (focal/control)
```

```{r plot predicted plant diversity by site}
# Predicted marginal means for plant richness
pred_site_plant <- ggpredict(bm_plant, terms = "site", type = "fixed")
pred_site_plant$x <- factor(pred_site_plant$x, levels = c("control", "focal"))
data_all_scaled$site <- factor(data_all_scaled$site, levels = c("control", "focal"))

# Viridis-based colors (colorblind-safe)
site_cols_plant <- c("control" = "#440154", "focal" = "#35B779")

# Plot
p_plant <- ggplot() +
  geom_jitter(
    data = data_all_scaled,
    aes(x = site, y = richness, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +
  geom_point(
    data = pred_site_plant,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +
  geom_errorbar(
    data = pred_site_plant,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +
  scale_colour_manual(values = site_cols_plant) +
  labs(
    x = NULL,
    y = "Plant species richness",
    title = "Predicted Flowering Plant Richness per Plot by Site",
    subtitle = "Points = model-estimated means ± 95% CI\nJittered points = raw observations"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  )

p_plant
```
```{r print plot to tiff}
tiff(
  file.path("..", "output", "figures", "pred_plant_Q1_site_600dpi.tiff"),
  width = 3000, height = 2250, res = 600, compression = "lzw"
)
print(p_plant)
dev.off()
```

