---
title: "05_site_unimodel"
output: html_document
---

```{r Load packages}
library(DHARMa)
library(tidyverse)
library(lmerTest)
library(emmeans)
library(ggeffects)
library(ggplot2)
library(glmmTMB)
library(dplyr)
```

```{r Load data}
# Compact helper function to read raw CSVs
bee_data <- read.csv(file.path("..", "data", "cleaned", "bee_data.csv"),
           stringsAsFactors = FALSE,
           check.names = FALSE)
```

Coverage‑based rarefaction (Chao et al. 2014, 2020) provides descriptive standardization, not inferential estimates, and its resulting diversity values do not follow the distributional assumptions required for GLMMs (e.g. Gotelli & Colwell 2001; Chao & Jost 2012). Rarefaction also collapses plot‑level replication, removes ecological variance, and prevents correct estimation of random effects.
Because our sampling design includes spatial and hierarchical structure (multiple visits per plot; plots nested within sites; sites nested within regions), a GLMM using plot‑level richness/abundance with region and plotID as random effects is the statistically appropriate inferential framework.

```{r fit model}

nm_rich <- glmmTMB(
  bee_richness ~ 1 + (1 | region) + (1 + visit | plotID),
  family = nbinom2,
  data = data_all_scaled
)

bm_rich <- glmmTMB(
  bee_richness ~ site + (1 | region) + (1 + visit | plotID),
  family = nbinom2,
  data = data_all_scaled
)

anova(nm_rich, bm_rich)
# random-effect structure is stable enough to support inference
# the visit random slope is not causing singularity
# adding site explains substantial variation in richness
# the data contain enough replication for strong inference

res <- simulateResiduals(bm_rich)
plot(res)
# no evidence for overdispersion, no deviation from expected residualdistribution
# no outliers, no heterogeneity of variance between site types

testDispersion(bm_rich)
testZeroInflation(bm_rich)


summary(bm_rich)
confint(bm_rich)
emmeans(bm_rich, pairwise ~ site)

```
We compared Poisson and negative binomial GLMMs with the same fixed and random effects. Although both models passed DHARMa diagnostics (no dispersion or zero-inflation detected), the negative binomial model fit substantially better (ΔAIC = 45.5; ΔBIC = 41.7) and estimated an overdispersion parameter of 3.29, indicating variance exceeded the Poisson mean assumption. We therefore report inference from the negative binomial model. Results were robust to using a Poisson model, which yielded nearly identical fixed-effect estimates.
gpt

```{r}
# Ratio (focal/control)
ratio <- exp(fixef(bm_rich)$cond["sitefocal"])

# 95% CI for ratio
ci_ratio <- exp(confint(bm_rich)[ "sitefocal", ])

# Absolute predicted richness per site
pred <- emmeans(bm_rich, ~ site, type = "response")

# Absolute difference
contrast(pred, "revpairwise")
```

```{r plot model predictions}

# --- Predicted marginal means for 'site' on the response scale ---
# (type = "fe" keeps predictions at average random effects; type = "response" back-transforms)
pred_site_rich <- ggpredict(bm_rich, terms = "site", type = "fixed")
# pred_site_rich has columns: x, predicted, conf.low, conf.high (on response scale)

# --- Optional: ensure plotting order (control, focal) ---
pred_site_rich$x <- factor(pred_site_rich$x, levels = c("control", "focal"))
data_all_scaled$site <- factor(data_all_scaled$site, levels = c("control", "focal"))

# --- Colors (your palette) ---
site_cols <- list(
  "focal" = "#2C7FB8",
  "control"   = "#D95F09",
  "focal_light"   = "cadetblue3",
  "control_light" = "#D95"
  )

# --- Plot object (not printed) ---


# --- Plot: predicted means ± 95% CI with raw jittered points ---

p_rich <- ggplot() +
  # 1) Raw data (plot × visit), jittered in background
  geom_jitter(
    data = data_all_scaled,
    aes(x = site, y = bee_richness, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +

  # 2) Model-predicted means (ggeffects output; response scale)
  geom_point(
    data = pred_site_rich,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +

  # 3) 95% confidence intervals
  geom_errorbar(
    data = pred_site_rich,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +

  # Use identical site colours for both layers
  scale_colour_manual(values = site_cols) +

  # Labels
  labs(
    x = NULL,
    y = "Bee species richness",
    title = "Predicted Bee Richness per Plot by Site",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = raw observations"
  ) +

  # Theme
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  )

p_rich
```
```{r print plot to tiff}
tiff(
  file.path("..", "output", "figures", "predicted_bee_richness_by_site_600dpi.tiff"),
  width = 3000, height = 2250, res = 600, compression = "lzw"
)

print(p_rich)
dev.off()
```

