---
title: "05_site_unimodel"
output: html_document
---

```{r Load packages}

library(tidyverse)
library(vegan)
library(iNEXT)
library(lme4)
library(lmerTest)
library(broom.mixed)
```

```{r Load data}
# Compact helper function to read raw CSVs
bee_data <- read.csv(file.path("..", "data", "cleaned", "bee_data.csv"),
           stringsAsFactors = FALSE,
           check.names = FALSE)
```


```{r Data prep}
# 0) Make absolutely sure the names you use are present in bee_data
# (your range is correct and includes 145 species)
species_cols <-
  names(bee_data) |>
  (\(nm) nm[
    match("Andrena aeneiventris", nm) :
    match("Stelis punctulatissima", nm)
  ])()

# Sanity assert: all species columns exist in bee_data
missing_in_bee_data <- setdiff(species_cols, names(bee_data))
if (length(missing_in_bee_data) > 0) {
  stop("These species names are not in bee_data: ",
       paste(missing_in_bee_data, collapse = ", "))
}

# 1) Aggregate to **plot** level (sum across visits), keeping ALL species
bee_plot <-
  bee_data |>
  select(region, site, plotID, plot, all_of(species_cols)) |>
  group_by(region, site, plotID, plot) |>
  summarise(across(all_of(species_cols), ~ sum(.x, na.rm = TRUE)),
            .groups = "drop") |>
  arrange(region, site, plot)

# Quick check: the two Andrena species should be present now
grep("Andrena aeneiventris", names(bee_plot), value = TRUE)
grep("Andrena afzeliella", names(bee_plot), value = TRUE)

# 2) (Optional) Aggregate further to **site** level (8 rows)
bee_site <-
  bee_plot |>
  group_by(region, site) |>
  summarise(across(all_of(species_cols), ~ sum(.x, na.rm = TRUE)),
            .groups = "drop") |>
  arrange(region, site)

# Verify those species exist after site aggregation too
grep("Andrena aeneiventris", names(bee_site), value = TRUE)
grep("Andrena afzeliella", names(bee_site), value = TRUE)

# 3) Final community matrices (species only)
comm_plot <-
  bee_plot |>
  select(all_of(species_cols)) |>
  as.data.frame()

comm_site <-
  bee_site |>
  select(all_of(species_cols)) |>
  as.data.frame()

# A) Across the entire dataset (plot aggregation should keep species totals)
stopifnot(sum(bee_data$`Andrena aeneiventris`, na.rm = TRUE) ==
            sum(bee_plot$`Andrena aeneiventris`, na.rm = TRUE))

stopifnot(sum(bee_data$`Andrena afzeliella`, na.rm = TRUE) ==
            sum(bee_plot$`Andrena afzeliella`, na.rm = TRUE))

# B) Within each (region, site) pair (site aggregation preserves plot totals)
check_aeneiventris <-
  bee_plot |>
  group_by(region, site) |>
  summarise(x = sum(`Andrena aeneiventris`, na.rm = TRUE), .groups = "drop") |>
  left_join(
    bee_site |>
      select(region, site, `Andrena aeneiventris`) |>
      rename(y = `Andrena aeneiventris`),
    by = c("region", "site")
  ) |>
  mutate(diff = x - y)

stopifnot(max(abs(check_aeneiventris$diff), na.rm = TRUE) == 0)
```
# Rarefy on plot-scale
```{r}
# Convert plot-level community matrix to matrix format
comm_plot_mat <- comm_plot |> as.matrix()

# Total individuals per plot
n_total_plot <- rowSums(comm_plot_mat)

# Rarefaction size = minimum number of individuals across plots
m_plot <- min(n_total_plot)
message("Common rarefaction size for plot-level rarefaction m_plot = ", m_plot)
print(n_total_plot)
```

```{r Rerefied q0 computation}
# Rarefied richness (q0) at plot level
q0_m_plot <-
  vegan::rarefy(comm_plot_mat, sample = m_plot) |>
  as.numeric()

plot_results <-
  bee_plot |>
  select(region, site, plotID, plot) |>
  mutate(
    n_total = n_total_plot,
    q0_m = q0_m_plot
  )

plot_results
```




# Rarefy on site-scale
```{r Compute toals and common rarefaction size}
# Species-only site matrix (you already created comm_site)
# Ensure matrix format for vegan/iNEXT
comm_site_mat <- comm_site |> as.matrix()

# Total individuals per site (8 values)
n_total <- comm_site_mat |> rowSums()

# Common rarefaction size = min total across sites
m <- min(n_total)
message("Common rarefaction size m = ", m)
print(n_total)
```
```{r Rarefied q0 computation}
q0_m <-
  comm_site_mat |>
  vegan::rarefy(sample = m) |>
  as.numeric()


site_results <-
  bee_site |>
  select(region, site) |>
  mutate(
    n_total = n_total,
    q0_m = q0_m
  )

site_results
```
```{r Model q0}

m_q0 <- glmmTMB(q0_m ~ site + (1 | region), data = site_results, family = nbinom2)
summary(m_q0)
```
```{r}
plot(m_q0)
qqnorm(resid(m_q0)); qqline(resid(m_q0))

plot_q0 <-
  site_results |>
  mutate(site = factor(site, levels = c("control", "focal"))) |>
  ggplot(aes(x = site, y = q0_m, color = site, group = region)) +
  geom_point(size = 3) +
  geom_line(alpha = 0.7) +
  theme_minimal() +
  labs(
    x = "Type de site",
    y = "q0 (richesse rarefiée)",
    title = "Comparaison de la richesse rarefiée (q0) entre sites"
  ) +
  theme(legend.position = "none")

plot_q0
```






































Bee ab model prep:
```{r bee ab model prep: overdispersion test, response data distribution}
#baseline poisson model
m_pois <- glmer(
  bee_abundance ~ site + (1 | region) + (1 | plotID),
  data = data_all_scaled,
  family = poisson
)

summary(m_pois)

#test for overdispersion
overdisp_fun <- function(model) {
  rdf <- df.residual(model)
  rp <- residuals(model, type = "pearson")
  Pearson.chisq <- sum(rp^2)
  ratio <- Pearson.chisq / rdf
  pval <- pchisq(Pearson.chisq, df = rdf, lower.tail = FALSE)
  c(chisq = Pearson.chisq, ratio = ratio, rdf = rdf, p = pval)
}

overdisp_fun(m_pois) #clear overdispersion

#negatvie binomial modlel
m_nb <- glmmTMB(
  bee_abundance ~ site + (1 | region) + (1 | plotID),
  family = nbinom2,   # Negative binomial
  data = data_all_scaled)

summary(m_nb)

# Check diagnostics
sim_nb <- simulateResiduals(m_nb, n = 1000)
plot(sim_nb)
testDispersion(sim_nb) # no overdispersion with negbin
testZeroInflation(sim_nb) # no zero inflation

#negative binomial is the distribution used from now on
```

```{r bee ab model prep: test for collinearity among habitat explanatory variables}
# List of numeric environmental variables
env_vars <- c("visit", "bareground", "vegcover", "vegheight", "herb", "woody",
              "sand", "gravel", "cobble", "stone", "wood", "soil", "flat",
              "sloped", "steep", "canopy", "canopy_hp", "branchpile",
              "stonepile", "patchiness", "standingdeadwood", "lyingdeadwood",
              "hill_q0", "hill_q1", "hill_q2")
#visit and patchiness, although included in coming models as ordinal, are treated as numeric for Spearman computation

# Compute Spearman correlation matrix
cor_matrix <- cor(
  data_all_scaled[env_vars],
  method = "spearman",
  use = "pairwise.complete.obs")

# Identify pairs with |correlation| > 0.7
high_corr <- which(abs(cor_matrix) > 0.7 & abs(cor_matrix) <= 1, arr.ind = TRUE)

# Create a data frame of correlated pairs
correlated_pairs <- data.frame(
  var1 = rownames(cor_matrix)[high_corr[, 1]],
  var2 = colnames(cor_matrix)[high_corr[, 2]],
  correlation = cor_matrix[high_corr]
)

# Remove duplicates (var1-var2 vs var2-var1)
correlated_pairs[correlated_pairs$var1 < correlated_pairs$var2, ]

# visualize correlation
corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  addCoef.col = "black",
  number.cex = 0.6
)

# New variable dataframe (merged covariates with spearman above 0.7)
# bareground - vegcover -> bareground
# herb - woody -> herb
# flat - sloped -> sloped
# hill_q0 - hill_q1 -> hill_q1

# canopy_hp excluded as constant across visits
env_vars <- c("bareground", "vegheight", "herb", "sand", "gravel", "cobble",
              "stone", "wood", "soil", "sloped", "steep", "canopy",
              "branchpile", "stonepile", "standingdeadwood", "lyingdeadwood",
              "hill_q1", "hill_q2")
```

Can we include site as explanatory variable in a multivariate model? - No, this chunk shows that too many of the habitat variables confound with site, and should therefore not be included in the same multivariate model.
```{r bee ab model prep: diagnostic site vs habitat covariates}
num_vars <- c("visit", "bareground", "vegcover", "vegheight", "herb", "woody",
              "sand", "gravel", "cobble", "stone", "wood", "soil", "flat",
              "sloped", "steep", "canopy", "canopy_hp", "branchpile",
              "stonepile", "patchiness", "standingdeadwood", "lyingdeadwood",
              "hill_q1", "hill_q2")

site_tests <- lapply(num_vars, function(v) {
  wt <- wilcox.test(data_all_scaled[[v]] ~ data_all_scaled$site)
  data.frame(
    variable = v,
    W = wt$statistic,
    p_value = wt$p.value
  )
})

site_tests <- do.call(rbind, site_tests)

site_tests$p_adj <- p.adjust(site_tests$p_value, method = "BH")

# these variables confound with site and cannot be included in a multivariate model together
confounded_vars <- site_tests |>
  subset(p_adj < 0.05) |>
  arrange(p_adj)

confounded_vars$variable

# these variables do not confound with site and could be included in a multivariate model together
independent_vars <- site_tests |>
  subset(p_adj >= 0.05)

independent_vars$variable
```

Run a univariate model to see the effect of site on bee abundance.
```{r bee ab model: site}
NBab1 <- glmmTMB(
  bee_abundance ~ site + (1 | region) + (1|plotID),
  family = nbinom2,   # Negative binomial
  data = data_all_scaled)

summary(NBab1)

# Get predicted values for 'site'
pred_site <- ggpredict(NBab1, terms = "site")

# Plot
plot(pred_site) +
  ggtitle("Predicted Bee Abundance by Site") +
  ylab("Predicted Abundance") +
  xlab("Site")

#add raw data jitter

site_cols <- c(
  "focal" = "#6B7C93",
  "control"   = "#D08C60"
)

ggplot() +
  # raw data: barely there
  geom_jitter(
    data = data_all_scaled,
    aes(x = site, y = bee_abundance, colour = site),
    width = 0.12,
    alpha = 0.15,
    size = 1.6,
    show.legend = FALSE
  ) +
  # predicted means
  geom_point(
    data = pred_site,
    aes(x = x, y = predicted, colour = x),
    size = 4
  ) +
  # confidence intervals
  geom_errorbar(
    data = pred_site,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.12,
    linewidth = 1
  ) +
  scale_colour_manual(values = site_cols) +
  labs(
    x = NULL,
    y = "Predicted bee abundance",
    title = "Predicted bee abundance by site, raw data included"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    axis.title.y = element_text(margin = margin(r = 8))
  )
```

Run univariate models for ordinal covariates
```{r bee ab model: univarite models for ordinal variables}
# patchiness
p_ord <- glmmTMB(bee_abundance ~ patchiness_ord + (1 | region) + (1|plotID),
                 family = nbinom2, data = data_all_scaled)
summary(p_ord)

#visit
v_ord <- glmmTMB(bee_abundance ~ visit_ord + (1 | region) + (1|plotID),
                 family = nbinom2, data = data_all_scaled)
summary(v_ord)


#plot 
pred <- ggpredict(p_ord, terms = "patchiness_ord")
ggplot(pred, aes(x = x, y = predicted)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  theme_minimal() +
  labs(title = "Effect of Patchiness (Ordinal) on Bee Abundance",
       x = "Patchiness Level", y = "Predicted Abundance")


pred <- ggpredict(v_ord, terms = "visit_ord")
ggplot(pred, aes(x = x, y = predicted)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  theme_minimal() +
  labs(title = "Effect of visit (Ordinal) on Bee Abundance",
       x = "Visit Number", y = "Predicted Abundance")

```

Try interactive model between site * visit 
```{r}
m_sv <- glmmTMB(bee_abundance ~ site*visit + (1 | region) + (1|plotID),
                 family = nbinom2, data = data_all_scaled)
summary(m_sv)

pred_sv <- ggpredict(m_sv, terms = c("visit", "site"))


plot(pred_sv) +
  ggtitle("Predicted Bee Abundance across Visits by Site") +
  xlab("Visit") +
  ylab("Predicted abundance (Negative Binomial)") +
  theme_minimal(base_size = 13)

# Custom color palette
site_cols <- c("focal" = "#6B7C93", "control" = "#D08C60")

ggplot() +
  # Raw data
  geom_jitter(
    data = data_all_scaled,
    aes(x = visit, y = bee_abundance, colour = site),
    width = 0.2, alpha = 0.2, size = 1.8
  ) +
  # Predicted values
  geom_line(
    data = pred_sv,
    aes(x = x, y = predicted, colour = group),
    size = 1.2
  ) +
  # Confidence intervals
  geom_ribbon(
    data = pred_sv,
    aes(x = x, ymin = conf.low, ymax = conf.high, fill = group),
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  scale_colour_manual(values = site_cols) +
  scale_fill_manual(values = site_cols) +
  labs(
    x = "Visit (ordinal)",
    y = "Predicted Bee Abundance",
    title = "Predicted Bee Abundance Across Visits by Site"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```


Run univariate models for numeric covariates
```{r bee ab model: run univariate model for numeric variabels}
model_um_abundance <- tibble(
  Variable = env_vars,
  Model = map(env_vars, ~ glmmTMB(
    as.formula(paste("bee_abundance ~ `", .x, "` + (1 | region) + (1|plotID)", sep = "")),
    family = nbinom2,
    data = data_all_scaled
  ))
)
```


```{r bee ab model: plot univarite model predictions}
model_df <- model_um_abundance

# Extract model stats
model_df <- model_df |>
  mutate(
    Summary = map(Model, ~ summary(.x)$coefficients$cond),
    CoefName = map_chr(Summary, ~ setdiff(rownames(.x), "(Intercept)")),
    Estimate = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Estimate"]),
    StdError = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Std. Error"]),
    p_value = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Pr(>|z|)"]),
    CI_lower = Estimate - 1.96 * StdError,
    CI_upper = Estimate + 1.96 * StdError,
    AIC = map_dbl(Model, AIC)
  )

print(model_df)

# Plot all model output estimates, mark significant ones in different color
# Sort by AIC for convenience
model_df_am <- model_df |> arrange(AIC)

# Add significance flag
model_df_am <- model_df_am |>
  mutate(Significant = ifelse(p_value < 0.05, "Significant", "Not Significant"))

# Forest plot with color by significance
ggplot(model_df_am, aes(x = reorder(Variable, Estimate), y = Estimate, color = Significant)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  ylab("Effect Size (Estimate)") +
  xlab("Environmental Variable") +
  ggtitle("Univariate Effects on Bee Abundance") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(values = c("Significant" = "#D55E00", "Not Significant" = "gray"))


# plot significant univariate models
# Filter significant models
significant_df <- model_df |>
  filter(!is.na(p_value), p_value < 0.05) |>
  arrange(AIC)

# View summary table
print(significant_df |> select(Variable, AIC, Estimate, CI_lower, CI_upper, p_value))

# Plot predicted effects using ggpredict
plots <- map2(significant_df$Model, significant_df$Variable, ~ {
  pred <- ggpredict(.x, terms = .y, bias_correction = TRUE, transform = TRUE)
  ggplot(pred, aes(x = x, y = predicted)) +
    geom_line() +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
    theme_minimal() +
    labs(title = paste("Effect of", .y, "on Bee Abundance"),
         x = .y, y = "Predicted Abundance")
})

ggplot(significant_df, aes(x = reorder(Variable, Estimate), y = Estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  ylab("Effect Size (Estimate)") +
  xlab("Environmental Variable") +
  ggtitle("Significant Predictors of Bee Abundance (p < 0.05)") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray")
```


Build multivariate model for bee abundance
```{r}
AB_glmm <- glmmTMB(
  bee_abundance ~ hill_q1 + hill_q2 + sand + soil + vegheight + patchiness_ord +
    (1 | region) + (1|plotID),
  family = nbinom2,
  data = data_all_scaled
)
print(summary(AB_glmm))

vif(lm(bee_abundance ~ hill_q1 + hill_q2 + sand + soil + vegheight + patchiness_ord,
       data = data_all_scaled))
#all values below 3 - no multicollinearity 

```

```{r}

# Extract tidy summary
coef_df <- broom.mixed::tidy(AB_glmm, conf.int = TRUE)

# Filter only fixed effects
coef_df <- coef_df |>
  filter(effect == "fixed") |>
  mutate(significant = ifelse(conf.low * conf.high > 0, "Significant", "Not Significant"))



# Plot
ggplot(coef_df, aes(x = estimate, y = term, color = significant)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  scale_color_manual(values = c("Significant" = "orange", "Not Significant" = "gray")) +
  theme_minimal() +
  labs(title = "Falcon Plot of GLMM Effects",
       x = "Estimate (with 95% CI)",
       y = "Predictor")

```


```{r}
AB_glmm <- glmmTMB(
  bee_abundance ~ site + hill_q1 + hill_q2 + sloped + sand + patchiness +
    (1 | region/site/plotID),
  family = nbinom2,
  data = data_all_scaled
)
print(summary(AB_glmm))

```






Modelling bee richness


Richness modelling - test for overdispersion / zero inflation
```{r bee models richness}
#baseline poisson model
m_pois <- glmer(
  bee_richness ~ site + (1 | region/site/plotID), 
  data = bee_data,
  family = poisson
)

summary(m_pois)

#test for overdispersion
overdisp_fun <- function(model) {
  rdf <- df.residual(model)
  rp <- residuals(model, type = "pearson")
  Pearson.chisq <- sum(rp^2)
  ratio <- Pearson.chisq / rdf
  pval <- pchisq(Pearson.chisq, df = rdf, lower.tail = FALSE)
  c(chisq = Pearson.chisq, ratio = ratio, rdf = rdf, p = pval)
}

overdisp_fun(m_pois) #clear overdispersion

#negatvie binomial modlel
m_nb <- glmmTMB(
  bee_richness ~ site + (1 | region/site/plotID),
  family = nbinom2,   # Negative binomial
  data = bee_data)

# Check diagnostics
sim_nb <- simulateResiduals(m_nb, n = 1000)
plot(sim_nb)
testDispersion(sim_nb) # no overdispersion with negbin
testZeroInflation(sim_nb) # no zero inflation

```
```{r}
# Run univariate models for bee richness

env_vars <- c("bareground", "vegheight", "herb", "sand",
              "gravel", "cobble", "stone", "wood", "soil", "sloped",
              "steep", "canopy", "branchpile",
              "stonepile", "patchiness", "standingdeadwood", "lyingdeadwood", "hill_q0", "hill_q1", "hill_q2")

bee_um_richness <- tibble(
  Variable = env_vars,
  Model = map(env_vars, ~ glmmTMB(
    as.formula(paste("bee_richness ~ `", .x, "` + (1 | region/site/plotID)", sep = "")),
    family = nbinom2,  # still appropriate if richness is count data
    data = data_all_scaled
  ))
)
```
Due to warning: Model convergence problem; non-positive-definite Hessian matrix - run diagnostics
```{r}
# Check convergence for each model
convergence_flags <- map_lgl(bee_um_richness$Model, ~ {
  any(grepl("Model convergence problem", .x$fit$optinfo$conv$lme4$messages))
})

# Combine with variable names
tibble(Variable = bee_um_richness$Variable, ConvergenceIssue = convergence_flags)
```
models themselves do not have convergence issues, we can proceed. 
(possible explanation: glmmTMB often issues a temporary warning if the Hessian is singular at some iteration, even if the optimizer later resolves it.)


```{r}
model_df_richness <- bee_um_richness

# Extract model stats
model_df_richness <- model_df_richness |>
  mutate(
    Summary = map(Model, ~ summary(.x)$coefficients$cond),
    CoefName = map_chr(Summary, ~ setdiff(rownames(.x), "(Intercept)")),
    Estimate = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Estimate"]),
    StdError = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Std. Error"]),
    p_value = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Pr(>|z|)"]),
    CI_lower = Estimate - 1.96 * StdError,
    CI_upper = Estimate + 1.96 * StdError,
    AIC = map_dbl(Model, AIC)
  )

# Filter significant models
significant_df_richness <- model_df_richness |>
  filter(!is.na(p_value), p_value < 0.05) |>
  arrange(AIC)

# View summary table
print(significant_df_richness |>
        select(Variable, AIC, Estimate, CI_lower, CI_upper, p_value))

# Plot predicted effects using ggpredict
plots_richness <- map2(significant_df_richness$Model,
                       significant_df_richness$Variable, ~ {
  pred <- ggpredict(.x, terms = .y, bias_correction = TRUE, transform = TRUE)
  ggplot(pred, aes(x = x, y = predicted)) +
    geom_line() +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
    theme_minimal() +
    labs(title = paste("Effect of", .y, "on Bee Richness"),
         x = .y, y = "Predicted Richness")
})

plots_richness


ggplot(significant_df_richness,
       aes(x = reorder(Variable, Estimate), y = Estimate)) +
  geom_point(size = 3, color = "#D55E00") +
  geom_errorbar(
    aes(ymin = CI_lower, ymax = CI_upper),
    width = 0.2,
    color = "#D55E00") +
  coord_flip() +
  theme_minimal() +
  ylab("Effect Size (Estimate)") +
  xlab("Environmental Variable") +
  ggtitle("Significant Predictors of Bee Richness (p < 0.05)") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray")
```


```{r}
# Forest plot for all models
model_df_richness <- model_df_richness |>
  arrange(AIC) |>
  mutate(Significant = ifelse(p_value < 0.05, "Significant", "Not Significant"))

ggplot(model_df_richness,
       aes(x = reorder(Variable, Estimate),
           y = Estimate, color = Significant)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  ylab("Effect Size (Estimate)") +
  xlab("Environmental Variable") +
  ggtitle("Univariate Effects on Bee Richness") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(values = c("Significant" = "#D55E00",
                                "Not Significant" = "gray"))
```





```{r visit and site interactive model}

m_nb <- glmmTMB(bee_abundance ~ site * visit + (1 | region) + (1 | plotID),
                family = nbinom2, data = data_all_scaled)

# Newdata grid over site × visit (months 1..5)
newdata <- expand.grid(
  site  = levels(data_all_scaled$site),
  visit = sort(unique(data_all_scaled$visit))
)

# Marginal predictions on response scale (exclude random effects)
pred <- predict(m_nb, newdata = newdata,
                type = "response",
                se.fit = TRUE,
                re.form = NA)

newdata <- newdata |>
  mutate(
    fit = as.numeric(pred$fit),
    se  = as.numeric(pred$se.fit),
    lwr = pmax(fit - 1.96 * se, 0),
    upr = fit + 1.96 * se
  )

# Plot: raw points (transparent) + predicted line + 95% CI ribbon
p <- ggplot() +
  geom_point(
    data = data_all_scaled,
    aes(x = visit, y = bee_abundance, color = site),
    alpha = 0.25, size = 2,
    position = position_jitter(width = 0.05, height = 0)
  ) +
  geom_ribbon(
    data = newdata,
    aes(x = visit, ymin = lwr, ymax = upr, fill = site, group = site),
    alpha = 0.20
  ) +
  geom_line(
    data = newdata,
    aes(x = visit, y = fit, color = site, group = site),
    linewidth = 1.2
  ) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    x = "Visit (month)",
    y = "Predicted bee abundance (Negative Binomial, nbinom2)",
    color = "Site",
    fill  = "Site",
    title = "Bee abundance: raw counts with glmmTMB predictions and 95% CI"
  ) +
  theme_minimal(base_size = 13)

print(p)

```

