---
title: "06_habvariables_multimodel"
output: html_document
---

```{r Load packages}
library(tidyverse)
library(ggeffects)
library(glmmTMB)
library(car)
library(DHARMa)
library(MuMIn)
library(vegan)
library(iNEXT)
library(lme4)
library(corrplot)
library(lmerTest)
library(broom.mixed)
```

```{r Load data}
data_all_scaled <- readRDS(file.path("..", "data", "cleaned", "data_all_scaled.rds"))
```
Slope is at plot level because only plots have repeated measures
Region remains a random intercept to capture large-scale spatial structure
No need for site as fixed or random because you use plot-level predictors
 (1 | region) + (1 + visit | plotID) 
 
 
## Bee abundance models

Bee ab model prep:
```{r bee Poisson ab model prep: overdispersion test, response data distribution}
#baseline poisson null model
m_pois <- glmmTMB(
  bee_abundance ~ 1 +
    (1 | region) +
    (1 + visit | plotID),
  data = data_all_scaled,
  family = poisson
)

summary(m_pois)

#test for overdispersion
overdisp_fun <- function(model) {
  rdf <- df.residual(model)
  rp <- residuals(model, type = "pearson")
  Pearson.chisq <- sum(rp^2)
  ratio <- Pearson.chisq / rdf
  pval <- pchisq(Pearson.chisq, df = rdf, lower.tail = FALSE)
  c(chisq = Pearson.chisq, ratio = ratio, rdf = rdf, p = pval)
}

overdisp_fun(m_pois) #clear overdispersion

sim_pois <- simulateResiduals(m_pois)
plot(sim_pois)
testDispersion(sim_pois)
testZeroInflation(sim_pois)
```

```{r bee nbinom ab model prep: overdispersion test, response data distribution}
m_nb <- glmmTMB(
  bee_abundance ~ 1 +
    (1 | region) +
    (1 + visit | plotID),
  data = data_all_scaled,
  family = nbinom2
)
summary(m_nb)

sim_nb <- simulateResiduals(m_nb)
plot(sim_nb)
testDispersion(sim_nb)
testZeroInflation(sim_nb)
```
"It is common that the KS test shows significance even when the model is fine.
Minor deviations in QQ plots without patterns in other tests do not require fixing."


```{r bee ab model prep: test for collinearity among habitat explanatory variables}
# List of numeric environmental variables
env_vars <- c("visit", "bareground", "vegcover", "vegheight", "herb", "woody",
              "sand", "gravel", "cobble", "stone", "wood", "soil", "flat",
              "sloped", "steep", "canopy", "canopy_hp", "branchpile",
              "stonepile", "patchiness", "standingdeadwood", "lyingdeadwood",
              "hill_q0", "hill_q1", "hill_q2")
#visit and patchiness, although included in coming models as ordinal, are treated as numeric for Spearman computation

# Compute Spearman correlation matrix
cor_matrix <- cor(
  data_all_scaled[env_vars],
  method = "spearman",
  use = "pairwise.complete.obs")

# Identify pairs with |correlation| > 0.7
high_corr <- which(abs(cor_matrix) > 0.7 & abs(cor_matrix) <= 1, arr.ind = TRUE)

# Create a data frame of correlated pairs
correlated_pairs <- data.frame(
  var1 = rownames(cor_matrix)[high_corr[, 1]],
  var2 = colnames(cor_matrix)[high_corr[, 2]],
  correlation = cor_matrix[high_corr]
)

# Remove duplicates (var1-var2 vs var2-var1)
correlated_pairs[correlated_pairs$var1 < correlated_pairs$var2, ]

# visualize correlation
corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  addCoef.col = "black",
  number.cex = 0.6
)

# New variable dataframe (merged covariates with spearman above 0.7)
# bareground - vegcover -> bareground
# herb - woody -> herb
# flat - sloped -> sloped
# hill_q0 - hill_q1 -> hill_q1

# canopy_hp excluded as constant across visits
env_vars <- c("bareground", "vegheight", "herb", "sand", "gravel", "cobble",
              "stone", "wood", "soil", "sloped", "steep", "canopy",
              "branchpile", "stonepile", "standingdeadwood", "lyingdeadwood",
              "hill_q1", "hill_q2")
```

Can we include site as explanatory variable in a multivariate model? - No, this chunk shows that too many of the habitat variables confound with site, and should therefore not be included in the same multivariate model.
```{r bee ab model prep: diagnostic site vs habitat covariates}
num_vars <- c("visit", "bareground", "vegcover", "vegheight", "herb", "woody",
              "sand", "gravel", "cobble", "stone", "wood", "soil", "flat",
              "sloped", "steep", "canopy", "canopy_hp", "branchpile",
              "stonepile", "patchiness", "standingdeadwood", "lyingdeadwood",
              "hill_q1", "hill_q2")

site_tests <- lapply(num_vars, function(v) {
  wt <- wilcox.test(data_all_scaled[[v]] ~ data_all_scaled$site)
  data.frame(
    variable = v,
    W = wt$statistic,
    p_value = wt$p.value
  )
})

site_tests <- do.call(rbind, site_tests)

site_tests$p_adj <- p.adjust(site_tests$p_value, method = "BH")

# these variables confound with site and cannot be included in a multivariate model together
confounded_vars <- site_tests |>
  subset(p_adj < 0.05) |>
  arrange(p_adj)

confounded_vars$variable

# these variables do not confound with site and could be included in a multivariate model together
independent_vars <- site_tests |>
  subset(p_adj >= 0.05)

independent_vars$variable
```

Run univariate models for ordinal covariates
```{r bee ab model: univarite models for ordinal variables}
# patchiness
p_ord <- glmmTMB(bee_abundance ~ patchiness_ord + (1 | region) + (1 + visit | plotID),
                 family = nbinom2, data = data_all_scaled)
summary(p_ord)

#visit
v_ord <- glmmTMB(bee_abundance ~ visit_ord + (1 | region) + (1 + visit | plotID),
                 family = nbinom2, data = data_all_scaled)
summary(v_ord)


#plot 
pred <- ggpredict(p_ord, terms = "patchiness_ord")
ggplot(pred, aes(x = x, y = predicted)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  theme_minimal() +
  labs(title = "Effect of Patchiness (Ordinal) on Bee Abundance",
       x = "Patchiness Level", y = "Predicted Abundance")


pred <- ggpredict(v_ord, terms = "visit_ord")
ggplot(pred, aes(x = x, y = predicted)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  theme_minimal() +
  labs(title = "Effect of visit (Ordinal) on Bee Abundance",
       x = "Visit Number", y = "Predicted Abundance")

```

Run univariate models for numeric covariates
```{r bee ab model: run univariate model for numeric variabels}
model_um_abundance <- tibble(
  Variable = env_vars,
  Model = map(env_vars, ~ glmmTMB(
    as.formula(paste("bee_abundance ~ `", .x, "` + (1 | region) + (1 + visit | plotID)", sep = "")),
    family = nbinom2,
    data = data_all_scaled
  ))
)
```


```{r bee ab model: plot univarite model predictions}
model_df <- model_um_abundance

# Extract model stats
model_df <- model_df |>
  mutate(
    Summary = map(Model, ~ summary(.x)$coefficients$cond),
    CoefName = map_chr(Summary, ~ setdiff(rownames(.x), "(Intercept)")),
    Estimate = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Estimate"]),
    StdError = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Std. Error"]),
    p_value = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Pr(>|z|)"]),
    CI_lower = Estimate - 1.96 * StdError,
    CI_upper = Estimate + 1.96 * StdError,
    AIC = map_dbl(Model, AIC)
  )

print(model_df)

# Plot all model output estimates, mark significant ones in different color
# Sort by AIC for convenience
model_df_am <- model_df |> arrange(AIC)

# Add significance flag
model_df_am <- model_df_am |>
  mutate(Significant = ifelse(p_value < 0.05, "Significant", "Not Significant"))

# Forest plot with color by significance
ggplot(model_df_am, aes(x = reorder(Variable, Estimate), y = Estimate, color = Significant)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  ylab("Effect Size (Estimate)") +
  xlab("Environmental Variable") +
  ggtitle("Univariate Effects on Bee Abundance") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(values = c("Significant" = "#D55E00", "Not Significant" = "gray"))


# plot significant univariate models
# Filter significant models
significant_df <- model_df |>
  filter(!is.na(p_value), p_value < 0.05) |>
  arrange(AIC)

# View summary table
print(significant_df |> select(Variable, AIC, Estimate, CI_lower, CI_upper, p_value))

# Plot predicted effects using ggpredict
plots <- map2(significant_df$Model, significant_df$Variable, ~ {
  pred <- ggpredict(.x, terms = .y, bias_correction = TRUE, transform = TRUE)
  ggplot(pred, aes(x = x, y = predicted)) +
    geom_line() +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
    theme_minimal() +
    labs(title = paste("Effect of", .y, "on Bee Abundance"),
         x = .y, y = "Predicted Abundance")
})

ggplot(significant_df, aes(x = reorder(Variable, Estimate), y = Estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  ylab("Effect Size (Estimate)") +
  xlab("Environmental Variable") +
  ggtitle("Significant Predictors of Bee Abundance (p < 0.05)") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray")
```


Build multivariate model for bee abundance

model convergence issue with random slope (1 + visit | plotID) included. Will add after dredge 
```{r Bee abundance global model}
AB_glmm <- glmmTMB(
  bee_abundance ~
    hill_q1 + hill_q2 + sand + herb + canopy + patchiness_ord +
    (1 | region) + (1 | plotID),
  family = nbinom2,
  data = data_all_scaled,
  na.action = na.fail
)
print(summary(AB_glmm))

vif(lm(bee_abundance ~ hill_q1 + hill_q2 + sand + soil + vegheight +
    canopy + steep + patchiness_ord,
       data = data_all_scaled))
#all values below 3 - no multicollinearity 

```
```{r Bee abundance dredge}
dd <- dredge(AB_glmm, rank = "AICc", trace = 2)
dd

head(dd)
```

Best model abundance - add random intercept visit
```{r}

# If you kept only the best fixed-effects set:
final_abund <- glmmTMB(
  bee_abundance ~ hill_q1 + hill_q2 + patchiness_ord +
    (1 | region) +
    (1 + visit | plotID),
  family = nbinom2,
  data = data_all_scaled
)
summary(final_abund)

```

```{r model averaging abundance}
cand <- get.models(dd, subset = delta <= 2)

avg_mod <- model.avg(cand, revised.var = TRUE)
summary(avg_mod)
confint(avg_mod)

# Sum of weights can be calculated like that
sw(subset(dd, delta <= 2))
```

```{r falcon plot model average richness}

# -------------------------------------------------
# 1. PREP COEFFICIENTS (model.avg output)
# -------------------------------------------------
coef_df <- as.data.frame(summary(avg_mod)$coefmat.full)
coef_df$term <- rownames(coef_df)

# Remove intercept
coef_df <- coef_df[coef_df$term != "cond((Int))", ]

# Add CI
ci <- confint(avg_mod)
coef_df$lower <- ci[coef_df$term, 1]
coef_df$upper <- ci[coef_df$term, 2]

# -------------------------------------------------
# 2. MAP IMPORTANCE (from sw(top_set))
# -------------------------------------------------
imp <- sw(subset(dd, delta <= 2))

coef_df$importance <- NA
coef_df$importance[coef_df$term == "cond(hill_q1)"]             <- imp["cond(hill_q1)"]
coef_df$importance[coef_df$term == "cond(hill_q2)"]             <- imp["cond(hill_q2)"]
coef_df$importance[coef_df$term == "cond(sand)"]                <- imp["cond(sand)"]
coef_df$importance[coef_df$term == "cond(vegheight)"]          <- imp["cond(vegheight)"]
coef_df$importance[coef_df$term == "cond(patchiness_ord.L)"]    <- imp["cond(patchiness_ord)"]

# Drop Q and C contrasts
coef_df <- coef_df[!coef_df$term %in% c("cond(patchiness_ord.Q)", "cond(patchiness_ord.C)"), ]

# -------------------------------------------------
# 3. COLOR CODE BY IMPORTANCE THRESHOLD
# -------------------------------------------------
importance_threshold <- 0.8

coef_df$color <- ifelse(
  coef_df$importance >= importance_threshold,
  "#D55E00",   # orange
  "gray50"
)

# -------------------------------------------------
# 4. CLEAN LABELS
# -------------------------------------------------
coef_df$clean_label <- coef_df$term %>%
  str_replace("cond\\(", "") %>%
  str_replace("\\)", "") %>%
  str_replace("_ord.L", " (linear)") %>%
  str_replace("hill_q1", "Hill q=1") %>%
  str_replace("hill_q2", "Hill q=2") %>%
  str_replace("sand", "Sand") %>%
  str_replace("vegheight", "Vegetation height") %>%
  str_replace("patchiness", "Plot heterogeneity")

# -------------------------------------------------
# 5. FINAL PLOT (NO size legend, NO circles encoding)
# -------------------------------------------------
ggplot(coef_df, aes(x = Estimate, y = reorder(clean_label, Estimate))) +
  
  # CI bars
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 height = 0,
                 color = coef_df$color,
                 linewidth = 0.8) +
  
  # Points
  geom_point(color = coef_df$color,
             size = 3, stroke = 1) +
  
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  
  theme_minimal(base_size = 13) +
  labs(
    title = "Falcon Plot of Model-Averaged Effects",
    subtitle = "Predictors with importance ≥ 0.8 colored in orange",
    x = "Model-averaged estimate (95% CI)",
    y = NULL
  )

```
We performed an information‑theoretic model selection using AICc and averaged all models within ΔAICc ≤ 2 to account for model selection uncertainty. Model‑averaged estimates and variable importance (sum of AICc weights) are shown in the Falcon plot (Fig. X).
Three predictors were strongly supported across the model set, each with an importance of 1.0: patchiness (linear component), plant diversity Hill q=1, and Hill q=2. All three had positive model‑averaged effects, indicating that more heterogeneous habitat structure and higher plant diversity are associated with increased bee abundance. Among these, patchiness had the largest effect size.
In contrast, sand cover (importance 0.34) and bare ground (importance 0.18) received little support and contributed negligibly to explaining variation in bee abundance. Their model‑averaged confidence intervals overlapped zero, indicating weak evidence for any directional effect.
Overall, the model‑averaged results highlight that habitat structural heterogeneity and plant diversity—rather than bare soil characteristics—are the key predictors of bee abundance at the plot level.
gpt



# Bee richness

AUthor of qqplot KS deviation value is sensitive for glmmm with random slope and can tehrefore be disconsidered?
move on with poisson as rest of values look okay
```{r bee Poisson ri model prep: overdispersion test, response data distribution}
#baseline poisson null model
m_pois <- glmmTMB(
  bee_richness ~ 1 +
    (1 | region) +
    (1 + visit | plotID),
  data = data_all_scaled,
  family = poisson
)

summary(m_pois)

sim_pois <- simulateResiduals(m_pois)
plot(sim_pois)
testDispersion(sim_pois)
testZeroInflation(sim_pois)
```
CHatGPT explanation of using POisson vs nbinom here
We evaluated Poisson and negative binomial GLMMs for the richness response.
The Poisson model showed no evidence of overdispersion (DHARMa p = 0.488),
no zero inflation (p = 0.392), and stable random effects including a random
slope of visit within plot. The NB model exhibited singular convergence
(random slope–intercept correlation = ±1), indicating overparameterization.
We therefore used a Poisson GLMM for richness, as diagnostics supported the Poisson distribution.

```{r bee nbinom ab model prep: overdispersion test, response data distribution}
m_nb <- glmmTMB(
  bee_richness ~ 1 +
    (1 | region) +
    (1 + visit | plotID),
  data = data_all_scaled,
  family = nbinom2
)
summary(m_nb)

sim_nb <- simulateResiduals(m_nb)
plot(sim_nb)
testDispersion(sim_nb)
testZeroInflation(sim_nb)
```

```{r Bee richness global model}
RI_glmm <- glmmTMB(
  bee_richness ~
    hill_q1 + hill_q2 + sand + vegheight + steep + bareground + patchiness_ord +
    (1 | region) + (1 + visit | plotID),
  family = poisson,
  data = data_all_scaled,
  na.action = na.fail
)
print(summary(RI_glmm))

vif(lm(bee_richness ~ hill_q1 + hill_q2 + sand + soil + vegheight +
    canopy + steep + patchiness_ord,
       data = data_all_scaled))
#all values below 3 - no multicollinearity 

```
```{r Bee richness dredge}
dr <- dredge(RI_glmm, rank = "AICc", trace = 2)
dr

head(dr)
```

```{r model averaging richness}
cand <- get.models(dr, subset = delta <= 4)

avg_mod_ri <- model.avg(cand, revised.var = TRUE)
summary(avg_mod_ri)

```



```{r}

# Extract tidy summary
coef_df <- broom.mixed::tidy(AB_glmm, conf.int = TRUE)

# Filter only fixed effects
coef_df <- coef_df |>
  filter(effect == "fixed") |>
  mutate(significant = ifelse(conf.low * conf.high > 0, "Significant", "Not Significant"))



# Plot
ggplot(coef_df, aes(x = estimate, y = term, color = significant)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  scale_color_manual(values = c("Significant" = "orange", "Not Significant" = "gray")) +
  theme_minimal() +
  labs(title = "Falcon Plot of GLMM Effects",
       x = "Estimate (with 95% CI)",
       y = "Predictor")

```


```{r}
AB_glmm <- glmmTMB(
  bee_abundance ~ site + hill_q1 + hill_q2 + sloped + sand + patchiness +
    (1 | region/site/plotID),
  family = nbinom2,
  data = data_all_scaled
)
print(summary(AB_glmm))

```






Modelling bee richness


Richness modelling - test for overdispersion / zero inflation
```{r bee models richness}
#baseline poisson model
m_pois <- glmer(
  bee_richness ~ site + (1 | region/site/plotID), 
  data = bee_data,
  family = poisson
)

summary(m_pois)

#test for overdispersion
overdisp_fun <- function(model) {
  rdf <- df.residual(model)
  rp <- residuals(model, type = "pearson")
  Pearson.chisq <- sum(rp^2)
  ratio <- Pearson.chisq / rdf
  pval <- pchisq(Pearson.chisq, df = rdf, lower.tail = FALSE)
  c(chisq = Pearson.chisq, ratio = ratio, rdf = rdf, p = pval)
}

overdisp_fun(m_pois) #clear overdispersion

#negatvie binomial modlel
m_nb <- glmmTMB(
  bee_richness ~ site + (1 | region/site/plotID),
  family = nbinom2,   # Negative binomial
  data = bee_data)

# Check diagnostics
sim_nb <- simulateResiduals(m_nb, n = 1000)
plot(sim_nb)
testDispersion(sim_nb) # no overdispersion with negbin
testZeroInflation(sim_nb) # no zero inflation

```
```{r}
# Run univariate models for bee richness

env_vars <- c("bareground", "vegheight", "herb", "sand",
              "gravel", "cobble", "stone", "wood", "soil", "sloped",
              "steep", "canopy", "branchpile",
              "stonepile", "patchiness", "standingdeadwood", "lyingdeadwood", "hill_q0", "hill_q1", "hill_q2")

bee_um_richness <- tibble(
  Variable = env_vars,
  Model = map(env_vars, ~ glmmTMB(
    as.formula(paste("bee_richness ~ `", .x, "` + (1 | region/site/plotID)", sep = "")),
    family = nbinom2,  # still appropriate if richness is count data
    data = data_all_scaled
  ))
)
```
Due to warning: Model convergence problem; non-positive-definite Hessian matrix - run diagnostics
```{r}
# Check convergence for each model
convergence_flags <- map_lgl(bee_um_richness$Model, ~ {
  any(grepl("Model convergence problem", .x$fit$optinfo$conv$lme4$messages))
})

# Combine with variable names
tibble(Variable = bee_um_richness$Variable, ConvergenceIssue = convergence_flags)
```
models themselves do not have convergence issues, we can proceed. 
(possible explanation: glmmTMB often issues a temporary warning if the Hessian is singular at some iteration, even if the optimizer later resolves it.)


```{r}
model_df_richness <- bee_um_richness

# Extract model stats
model_df_richness <- model_df_richness |>
  mutate(
    Summary = map(Model, ~ summary(.x)$coefficients$cond),
    CoefName = map_chr(Summary, ~ setdiff(rownames(.x), "(Intercept)")),
    Estimate = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Estimate"]),
    StdError = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Std. Error"]),
    p_value = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Pr(>|z|)"]),
    CI_lower = Estimate - 1.96 * StdError,
    CI_upper = Estimate + 1.96 * StdError,
    AIC = map_dbl(Model, AIC)
  )

# Filter significant models
significant_df_richness <- model_df_richness |>
  filter(!is.na(p_value), p_value < 0.05) |>
  arrange(AIC)

# View summary table
print(significant_df_richness |>
        select(Variable, AIC, Estimate, CI_lower, CI_upper, p_value))

# Plot predicted effects using ggpredict
plots_richness <- map2(significant_df_richness$Model,
                       significant_df_richness$Variable, ~ {
  pred <- ggpredict(.x, terms = .y, bias_correction = TRUE, transform = TRUE)
  ggplot(pred, aes(x = x, y = predicted)) +
    geom_line() +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
    theme_minimal() +
    labs(title = paste("Effect of", .y, "on Bee Richness"),
         x = .y, y = "Predicted Richness")
})

plots_richness


ggplot(significant_df_richness,
       aes(x = reorder(Variable, Estimate), y = Estimate)) +
  geom_point(size = 3, color = "#D55E00") +
  geom_errorbar(
    aes(ymin = CI_lower, ymax = CI_upper),
    width = 0.2,
    color = "#D55E00") +
  coord_flip() +
  theme_minimal() +
  ylab("Effect Size (Estimate)") +
  xlab("Environmental Variable") +
  ggtitle("Significant Predictors of Bee Richness (p < 0.05)") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray")
```


```{r}
# Forest plot for all models
model_df_richness <- model_df_richness |>
  arrange(AIC) |>
  mutate(Significant = ifelse(p_value < 0.05, "Significant", "Not Significant"))

ggplot(model_df_richness,
       aes(x = reorder(Variable, Estimate),
           y = Estimate, color = Significant)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  ylab("Effect Size (Estimate)") +
  xlab("Environmental Variable") +
  ggtitle("Univariate Effects on Bee Richness") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(values = c("Significant" = "#D55E00",
                                "Not Significant" = "gray"))
```


