---
title: "06_habvariables_multimodel"
output: word_document
output_dir: "knit"
---

```{r Load packages}
library(tidyverse)
library(ggeffects)
library(glmmTMB)
library(gt)
library(car)
library(DHARMa)
library(patchwork)
library(performance)
library(MuMIn)
library(vegan)
library(iNEXT)
library(lme4)
library(corrplot)
library(lmerTest)
library(broom.mixed)
```

```{r Load data}
data_all_scaled <- readRDS(file.path("..", "data", "cleaned", "data_all_scaled.rds"))
```

Test correlation between differenct bee metrics
```{r}
# List of numeric environmental variables
bee_vars <- c("bee_abundance", "bee_richness", "bee_hill_q1", "bee_hill_q2")
#visit and patchiness, although included in coming models as ordinal, are treated as numeric for Spearman computation

# Compute Spearman correlation matrix
cor_matrix <- cor(
  data_all_scaled[bee_vars],
  method = "spearman",
  use = "pairwise.complete.obs")

cor_matrix
```
Because all plots were surveyed with identical sampling effort (same area, same method, same observer, same frequency), rarefaction was not applied. Raw abundance, richness, and Hill diversity values were directly comparable across samples. Given the high correlation between Hill q = 1 and HIll q = 2 values, I run abundance, richness, and HIil q = 1 models in parallel.
Rarefaction would artificially reduce information content and compromise the GLMM structure. Therefore, analyses were conducted using raw count-based response variables standardized by equal sample effort.

“Bee abundance, species richness, and Hill diversity metrics were strongly correlated (Spearman ρ = 0.75–0.99). To avoid redundant inference, we focused on abundance and richness as primary response variables, and used Hill q1 as a complementary diversity metric. Hill q2 was excluded due to near‑perfect correlation with Hill q1.”
gpt


Collinearity tests environmental variables§
```{r bee ab model prep: test for collinearity among habitat explanatory variables}
# List of numeric environmental variables
env_vars <- c("visit", "bareground", "vegcover", "vegheight", "herb", "woody",
              "sand", "gravel", "cobble", "stone", "wood", "soil", "flat",
              "sloped", "steep", "canopy", "canopy_hp", "branchpile",
              "stonepile", "patchiness", "standingdeadwood", "lyingdeadwood",
              "hill_q0", "hill_q1", "hill_q2")
#visit and patchiness, although included in coming models as ordinal, are treated as numeric for Spearman computation

# Compute Spearman correlation matrix
cor_matrix <- cor(
  data_all_scaled[env_vars],
  method = "spearman",
  use = "pairwise.complete.obs")

# Identify pairs with |correlation| > 0.7
high_corr <- which(abs(cor_matrix) > 0.7 & abs(cor_matrix) <= 1, arr.ind = TRUE)

# Create a data frame of correlated pairs
correlated_pairs <- data.frame(
  var1 = rownames(cor_matrix)[high_corr[, 1]],
  var2 = colnames(cor_matrix)[high_corr[, 2]],
  correlation = cor_matrix[high_corr]
)

# Remove duplicates (var1-var2 vs var2-var1)
correlated_pairs <- correlated_pairs[correlated_pairs$var1 < correlated_pairs$var2, ]

print(correlated_pairs)
# visualize correlation
corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  addCoef.col = "black",
  number.cex = 0.6
)

# New variable dataframe (merged covariates with spearman above 0.7)
# bareground - vegcover -> bareground
# herb - woody -> herb
# flat - sloped -> sloped
# hill_q0  hill_q1  hill_q2 -> hill_q0

# canopy_hp excluded as constant across visits
env_vars <- c("bareground", "vegheight", "herb", "sand", "gravel", "cobble",
              "stone", "wood", "soil", "sloped", "steep", "canopy",
              "branchpile", "stonepile", "standingdeadwood", "lyingdeadwood",
              "hill_q0")
```

Can we include site as explanatory variable in a multivariate model? - No, this chunk shows that too many of the habitat variables confound with site, and should therefore not be included in the same multivariate model.
```{r bee ab model prep: diagnostic site vs habitat covariates}
num_vars <- c("visit", "bareground", "vegcover", "vegheight", "herb", "woody",
              "sand", "gravel", "cobble", "stone", "wood", "soil", "flat",
              "sloped", "steep", "canopy", "canopy_hp", "branchpile",
              "stonepile", "patchiness", "standingdeadwood", "lyingdeadwood",
              "hill_q0")

site_tests <- lapply(num_vars, function(v) {
  wt <- wilcox.test(data_all_scaled[[v]] ~ data_all_scaled$site)
  data.frame(
    variable = v,
    W = wt$statistic,
    p_value = wt$p.value
  )
})

site_tests <- do.call(rbind, site_tests)

site_tests$p_adj <- p.adjust(site_tests$p_value, method = "BH")

# these variables confound with site and cannot be included in a multivariate model together
confounded_vars <- site_tests |>
  subset(p_adj < 0.05) |>
  arrange(p_adj)

confounded_vars$variable

# these variables do not confound with site and could be included in a multivariate model together
independent_vars <- site_tests |>
  subset(p_adj >= 0.05)

independent_vars$variable
```

Run univariate models for ordinal covariates
```{r bee ab model: univarite models for ordinal variables}
# patchiness
p_ord <- glmmTMB(bee_abundance ~ patchiness_ord + (1 + visit | plotID),
                 family = nbinom1, data = data_all_scaled)
summary(p_ord)

#visit
v_ord <- glmmTMB(bee_abundance ~ visit_ord + (1 + visit | plotID),
                 family = nbinom1, data = data_all_scaled)
summary(v_ord)


#plot 
pred <- ggpredict(p_ord, terms = "patchiness_ord")
ggplot(pred, aes(x = x, y = predicted)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  theme_minimal() +
  labs(title = "Effect of Patchiness (Ordinal) on Bee Abundance",
       x = "Patchiness Level", y = "Predicted Abundance")


pred <- ggpredict(v_ord, terms = "visit_ord")
ggplot(pred, aes(x = x, y = predicted)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  theme_minimal() +
  labs(title = "Effect of visit (Ordinal) on Bee Abundance",
       x = "Visit Number", y = "Predicted Abundance")

```

Run univariate models for numeric covariates
```{r bee ab model: run univariate model for numeric variabels}
model_um_abundance <- tibble(
  Variable = env_vars,
  Model = map(env_vars, ~ glmmTMB(
    as.formula(paste("bee_abundance ~ `", .x, "` + (1 + visit | plotID)", sep = "")),
    family = nbinom1,
    data = data_all_scaled
  ))
)
```

```{r bee ab model: plot univarite model predictions}
model_df <- model_um_abundance

# Extract model stats
model_df <- model_df |>
  mutate(
    Summary = map(Model, ~ summary(.x)$coefficients$cond),
    CoefName = map_chr(Summary, ~ setdiff(rownames(.x), "(Intercept)")),
    Estimate = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Estimate"]),
    StdError = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Std. Error"]),
    p_value = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Pr(>|z|)"]),
    CI_lower = Estimate - 1.96 * StdError,
    CI_upper = Estimate + 1.96 * StdError,
    AIC = map_dbl(Model, AIC)
  )

print(model_df)

# Plot all model output estimates, mark significant ones in different color
# Sort by AIC for convenience
model_df_am <- model_df |> arrange(AIC)

# Add significance flag
model_df_am <- model_df_am |>
  mutate(Significant = ifelse(p_value < 0.05, "Significant", "Not Significant"))

# Forest plot with color by significance
ggplot(model_df_am, aes(x = reorder(Variable, Estimate), y = Estimate, color = Significant)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  ylab("Effect Size (Estimate)") +
  xlab("Environmental Variable") +
  ggtitle("Univariate Effects on Bee Abundance") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(values = c("Significant" = "#D55E00", "Not Significant" = "gray"))


# plot significant univariate models
# Filter significant models
significant_df <- model_df |>
  filter(!is.na(p_value), p_value < 0.05) |>
  arrange(AIC)

# View summary table
print(significant_df |> select(Variable, AIC, Estimate, CI_lower, CI_upper, p_value))

# Plot predicted effects using ggpredict
plots <- map2(significant_df$Model, significant_df$Variable, ~ {
  pred <- ggpredict(.x, terms = .y, bias_correction = TRUE)
  ggplot(pred, aes(x = x, y = predicted)) +
    geom_line() +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
    theme_minimal() +
    labs(title = paste("Effect of", .y, "on Bee Abundance"),
         x = .y, y = "Predicted Abundance")
})

ggplot(significant_df, aes(x = reorder(Variable, Estimate), y = Estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  ylab("Effect Size (Estimate)") +
  xlab("Environmental Variable") +
  ggtitle("Significant Predictors of Bee Abundance (p < 0.05)") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray")
```


# Bee abundance
Bee ab model prep:
Base Poisson model shows clear zero inflation.
```{r bee Poisson ab model prep: overdispersion test, response data distribution}
#baseline poisson null model
m_pois <- glmmTMB(
  bee_abundance ~ 1 +
    (1 | region) +
    (1 + visit | plotID),
  data = data_all_scaled,
  family = poisson
)

summary(m_pois)

sim_pois <- simulateResiduals(m_pois)
plot(sim_pois)
testDispersion(sim_pois)
testZeroInflation(sim_pois)
check_model(m_pois)
```

Comparing nbinom2 and nbinom1 distributions
```{r bee nbinom2 ab model prep: overdispersion test, response data distribution}
m_nb <- glmmTMB(
  bee_abundance ~ 1 +
    (1 | region) +
    (1 + visit | plotID),
  data = data_all_scaled,
  family = nbinom2
)
summary(m_nb)

sim_nb <- simulateResiduals(m_nb)
plot(sim_nb)
testDispersion(sim_nb)
testZeroInflation(sim_nb)
check_model(m_nb)
```
KS test for nbinom2 shows significant deviation

```{r bee nbinom1 ab model prep: overdispersion test, response data distribution}
m_nb <- glmmTMB(
  bee_abundance ~ 1 +
    (1 | region) +
    (1 + visit | plotID),
  data = data_all_scaled,
  family = nbinom1
)
summary(m_nb)

sim_nb <- simulateResiduals(m_nb)
plot(sim_nb)
testDispersion(sim_nb)
testZeroInflation(sim_nb)
```
nbinom1 fits better - use this from now on

Try model with different random effects structure
```{r bee nbinom1 ab model with different random effects structure}
m_nb_simple <- glmmTMB(
  bee_abundance ~ 1 +
    (1 + visit | plotID),
  data = data_all_scaled,
  family = nbinom1
)
summary(m_nb_simple)
sim_nb_simple <- simulateResiduals(m_nb_simple)
plot(sim_nb_simple)
testDispersion(sim_nb_simple)
testZeroInflation(sim_nb_simple)
check_model(m_nb_simple)
```
We initially considered a random-effects structure reflecting the sampling design, including random intercepts for region, sites nested within region, and random slopes of visit within plots.
However, region exibited near-zero variance without improving model fit.
Following Bates et al. (2015) and Harrison et al. (2018), we retained only the random intercept for plotID, which was strongly supported by the data and reflected the clustering of repeated surveys within plots.
Additional hierarchical levels (region, site) were included as fixed effects where biologically relevant.
gpt
(Variance estimates showed that the region-level variance was near zero (1.2×10⁻⁷),
ICC values and conditional R² confirmed that these additional random effects explained no additional hierarchical variation.
Model diagnostics (DHARMa) indicated no improvement in dispersion, zero-inflation, or residual structure.
Following Bates et al. (2015) and Harrison et al. (2018), we therefore selected the more parsimonious random-effects structure (1|plotID), which was strongly supported by the data and avoided overparameterization.)
gpt


Build multivariate model for bee abundance

only adding plotID as random effect as
stable variance estimate
no singularity
same AIC
simpler structure (thus preferred by parsimony)
fixed effects behave cleanly and consistently
```{r Bee abundance global model}
AB_glmm <- glmmTMB(
  bee_abundance ~
    hill_q0 + sand + herb + bareground + vegheight + patchiness_ord+
    (1 + visit | plotID),
  family = nbinom1,
  data = data_all_scaled,
  na.action = na.fail
)
print(summary(AB_glmm))

check_model(AB_glmm)
```
1. Correlation = –0.98 → essentially singular
2. Slope variance tiny → slope not informative
3. Plots are too few (n=64) × visits too few (5 points each) → limited temporal resolution
4. Slope breaks in the full model (but not in null model)
5. Fixed effects unchanged with or without slope
6. Reviewer will accept dropping slope if you explain these diagnostics
So the correct random structure remains:

he random slope variance was very small (σ² = 0.039), and the intercept–slope correlation was extremely high (r = –0.98), indicating overparameterization and a near‑singular fit.
According to Bates et al. (2015) and Harrison et al. (2018), such correlations signal that the random slope is not supported by the data and should be removed.
Simplifying to (1 | plotID) produced stable models without affecting fixed-effect inference.
We therefore retained the simpler random-effects structure in all models.
gpt

Remove visit from random effects structure
```{r Bee abundance global model without visit as random slope}
AB_glmm <- glmmTMB(
  bee_abundance ~
    hill_q0 + sand + herb + bareground + vegheight + patchiness_ord +
    (1 | plotID),
  family = nbinom1,
  data = data_all_scaled,
  na.action = na.fail
)
print(summary(AB_glmm))

check_model(AB_glmm)
```

```{r Bee abundance dredge}
dd <- dredge(AB_glmm, rank = "AICc", trace = 2)
dd

head(dd)
```

```{r model averaging abundance}
cand_ab <- get.models(dd, subset = delta <= 2)

# best model (to be used later for ggpred)
bestmod_ab <- cand_ab[[1]]

avg_mod_ab <- model.avg(cand_ab, revised.var = TRUE)
summary(avg_mod_ab)
confint(avg_mod_ab)

# Sum of weights can be calculated like that
sw(subset(dd, delta <= 2))
```

```{r forest plot model average - abundance}
# 1. PREP COEFFICIENTS (model.avg output)
coef_df <- as.data.frame(summary(avg_mod_ab)$coefmat.full)
coef_df$term <- rownames(coef_df)

# Remove intercept
coef_df <- coef_df[coef_df$term != "cond((Int))", ]

# Add CI
ci <- confint(avg_mod_ab)
coef_df$lower <- ci[coef_df$term, 1]
coef_df$upper <- ci[coef_df$term, 2]

# 2. MAP IMPORTANCE (from sw(top_set))

imp <- sw(subset(dd, delta <= 2))

coef_df$importance <- NA
coef_df$importance[coef_df$term == "cond(hill_q0)"] <- imp["cond(hill_q0)"]
coef_df$importance[coef_df$term == "cond(sand)"]  <- imp["cond(sand)"]
coef_df$importance[coef_df$term == "cond(herb)"] <- imp["cond(herb)"]
coef_df$importance[coef_df$term == "cond(bareground)"] <- imp["cond(bareground)"]
coef_df$importance[coef_df$term == "cond(vegheight)"] <- imp["cond(vegheight)"]
coef_df$importance[coef_df$term == "cond(patchiness_ord.L)"] <- imp["cond(patchiness_ord)"]
# Drop Q and C contrasts
coef_df <- coef_df[!coef_df$term %in% c("cond(patchiness_ord.Q)", "cond(patchiness_ord.C)"), ]

# 3. COLOR CODE BY IMPORTANCE THRESHOLD

importance_threshold <- 0.8

coef_df$color <- ifelse(
  coef_df$importance >= importance_threshold,
  "#C71586",
  "gray40"
)

# 4. CLEAN LABELS

coef_df$clean_label <- coef_df$term %>%
  str_replace("cond\\(", "") %>%
  str_replace("\\)", "") %>%
  str_replace("_ord.L", "") %>%
  str_replace("hill_q0", "Floral Richness") %>%
  str_replace("sand", "Sand") %>%
  str_replace("herb", "Herb layer cover") %>%
  str_replace("bareground", "Bareground") %>%
  str_replace("vegheight", "Vegetation height") %>%
  str_replace("patchiness", "Plot heterogeneity")


# 5. plot

p_forest_ab <- ggplot(coef_df, aes(x = Estimate, y = reorder(clean_label, Estimate))) +
  
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 width = 0,
                 color = coef_df$color,
                 linewidth = 0.8) +
  
  geom_point(color = coef_df$color,
             size = 3, stroke = 1) +
  
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_line(color = "grey85"),   # vertical gridlines ON
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 14),
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 4)),
    plot.subtitle = element_text(size = 12, margin = margin(b = 8)),
    
    axis.title.y = element_text(margin = margin(r = 10))
  ) +
  
  labs(
    title = "Model-Averaged Effects on Bee Abundance",
    subtitle = "Predictors with importance ≥ 0.8 highlighted purple",
    x = "Model-averaged estimate (95% CI)",
    y = NULL
  )

p_forest_ab

```
Sand cover received moderate support in the model set (importance = 0.82), indicating that it may contribute to explaining variation in bee abundance. Its model‑averaged effect was positive but small (estimate ≈ 0.08–0.10), and the 95% confidence interval slightly overlapped zero, suggesting substantial uncertainty in the strength of this relationship. Thus, sand appears to play a secondary role relative to patchiness and plant diversity.
Model averaging identified two strongly supported predictors of bee abundance: habitat patchiness and plant species richness (Hill q0), both with importance = 1.0 and positive model‑averaged coefficients whose confidence intervals excluded zero. Sand cover received moderate support (importance = 0.82) and showed a small positive effect, although its confidence interval slightly overlapped zero, indicating some uncertainty. In contrast, herb cover, vegetation height, and bare ground had low importance values (≤ 0.17) and model‑averaged effects near zero, suggesting weak evidence that these variables influence bee abundance. Overall, habitat heterogeneity and floristic richness emerged as the primary predictors of bee abundance.
gpt

```{r print model avg forest plot abundance}
tiff(
  file.path("..", "output", "figures", "forest_mavgbee_ab_600dpi.tiff"),
  width = 3990, height = 3000, res = 600, compression = "lzw"
)

print(p_forest_ab)  
dev.off()
```

Best model bee abundance
```{r best model bee abundance}
bestmod_ab
bestmod_ab <- glmmTMB(bee_abundance ~ hill_q0 + patchiness_ord + sand + (1 | plotID),
                      family = nbinom1,
                      data = data_all_scaled)

summary(bestmod_ab)
```

Plot predictions. Backtransform x-axis plant abundance for interpretability
```{r bes model bee abundance plot predictions}
# Extract scaling attributes for hill and sand
hill_mean <- attr(data_all_scaled$hill_q0, "scaled:center")
hill_sd   <- attr(data_all_scaled$hill_q0, "scaled:scale")

sand_mean <- attr(data_all_scaled$sand, "scaled:center")
sand_sd   <- attr(data_all_scaled$sand, "scaled:scale")

# Generate predictions for each predictor
pred_hill   <- ggpredict(bestmod_ab, terms = "hill_q0")
pred_patch  <- ggpredict(bestmod_ab, terms = "patchiness_ord")
pred_sand   <- ggpredict(bestmod_ab, terms = "sand")

# Back-transform hill_q0 and sand to raw units
pred_hill$x_raw <- pred_hill$x * hill_sd + hill_mean
pred_sand$x_raw <- pred_sand$x * sand_sd + sand_mean

data_all_scaled$hill_q0_raw <- data_all_scaled$hill_q0 * hill_sd + hill_mean
data_all_scaled$sand_raw     <- data_all_scaled$sand     * sand_sd  + sand_mean

# Plot 1: Hill q0 (raw axis) + jittered raw data + predictions
p_hill <- ggplot() +
  geom_jitter(
    data = data_all_scaled,
    aes(x = hill_q0_raw, y = bee_abundance),
    width = 0.15, height = 0.05, alpha = 0.25, 
    size = 1.5, color = "grey50"
  ) +
  geom_ribbon(
    data = pred_hill,
    aes(x = x_raw, ymin = conf.low, ymax = conf.high),
    fill = "plum2", alpha = 0.35
  ) +
  geom_line(
    data = pred_hill,
    aes(x = x_raw, y = predicted, group = 1),
    color = "#7A0177", linewidth = 1.3
  ) +
  labs(
    title = "Effect of Plant Genus Richness (Hill q = 0)",
    subtitle = "Predicted bee abundance ± 95% CI with raw data",
    x = "Plant genus richness (Hill q0, raw units)",
    y = "Bee abundance"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13, color = "black"),
    plot.title = element_text(size = 16, face = "bold")
  )

# Plot 2: Patchiness (ordinal) + jittered raw data + predictions
p_patch <- ggplot() +
  geom_jitter(
    data = data_all_scaled,
    aes(x = patchiness_ord, y = bee_abundance),
    width = 0.1, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  geom_point(
    data = pred_patch,
    aes(x = x, y = predicted),
    color = "#7A0177", size = 3
  ) +
  geom_errorbar(
    data = pred_patch,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    color = "#7A0177", width = 0.1, linewidth = 1
  ) +
  labs(
    title = "Effect of Plot Heterogeneity",
    subtitle = "Predicted bee abundance ± 95% CI with raw data",
    x = "Plot heterogeneity (ordinal)",
    y = "Bee abundance"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13, color = "black"),
    plot.title = element_text(size = 16, face = "bold")
  )

# Plot 3: Sand (raw axis) + jittered raw points + predictions
p_sand <- ggplot() +
  geom_jitter(
    data = data_all_scaled,
    aes(x = sand_raw, y = bee_abundance),
    width = 0.15, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  geom_ribbon(
    data = pred_sand,
    aes(x = x_raw, ymin = conf.low, ymax = conf.high),
    fill = "plum2", alpha = 0.35
  ) +
  geom_line(
    data = pred_sand,
    aes(x = x_raw, y = predicted, group = 1),
    color = "#7A0177", linewidth = 1.3
  ) +
  labs(
    title = "Effect of Sand Cover",
    subtitle = "Predicted bee abundance ± 95% CI with raw data",
    x = "Sand cover (%)",
    y = "Bee abundance"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13, color = "black"),
    plot.title = element_text(size = 16, face = "bold")
  )

# Combine all three into a multi-panel figure

p_combined_abundance <- p_hill + p_patch + p_sand + plot_layout(ncol = 3)

p_combined_abundance
```
We performed an information‑theoretic model selection using AICc and averaged all models within ΔAICc ≤ 2 to account for model‑selection uncertainty. Model‑averaged estimates and variable importance (sum of AICc weights) are presented in the forest plot (Fig. X).
Two predictors received full support across the candidate model set (importance = 1.0): habitat patchiness (linear component) and plant diversity Hill q = 0. Both had positive model‑averaged coefficients whose 95% confidence intervals did not overlap zero, indicating that more heterogeneous habitat structure and higher plant species richness were consistently associated with higher bee abundance. Patchiness showed the largest effect size among all predictors.
Sand cover received intermediate support (importance = 0.78). Its conditional model‑averaged effect was moderately positive and approached statistical significance, suggesting that sand cover may contribute to explaining variation in bee abundance but with greater model uncertainty. In contrast, herb cover (importance = 0.21) and bare ground (importance = 0.16) had low importance values and model‑averaged confidence intervals overlapping zero, indicating weak evidence for any consistent directional effect.
Overall, the model‑averaged results highlight that habitat structural heterogeneity and plant species richness—rather than vegetation cover or soil surface characteristics—are the primary predictors of bee abundance at the plot level.
gpt



# Bee richness
Poisson base model does not fit
```{r bee Poisson ri model prep: overdispersion test, response data distribution}
#baseline poisson null model
m_pois <- glmmTMB(
  bee_richness ~ 1 +
    (1 | plotID),
  data = data_all_scaled,
  family = poisson
)

summary(m_pois)

sim_pois <- simulateResiduals(m_pois)
plot(sim_pois)
testDispersion(sim_pois)
testZeroInflation(sim_pois)
```

Nbinom2 base model does not fit as well as Nbinom1
```{r bee nbinom2 ri model prep: overdispersion test, response data distribution}
m_nb2 <- glmmTMB(
  bee_richness ~ 1 +
    (1 | plotID),
  data = data_all_scaled,
  family = nbinom2
)
summary(m_nb2)

sim_nb2 <- simulateResiduals(m_nb2)
plot(sim_nb2)
testDispersion(sim_nb2)
testZeroInflation(sim_nb2)
```

Nbinom1
```{r bee nbinom1 ri model prep: overdispersion test, response data distribution}
m_nb1 <- glmmTMB(
  bee_richness ~ 1 +
    (1 | plotID),
  data = data_all_scaled,
  family = nbinom1
)
summary(m_nb1)

sim_nb1 <- simulateResiduals(m_nb1)
plot(sim_nb1)
testDispersion(sim_nb1)
testZeroInflation(sim_nb1)
```
We proceed with nbinom1, as
better AIC
better dispersion
more appropriate variance structure
clean residuals
stable random effect
gpt

Compare random effects structure as in abundance models.
```{r bee nbinom ri model prep: overdispersion test, response data distribution}
m_nb <- glmmTMB(
  bee_richness ~ 1 +
    (1 | region) +
    (1 + visit | plotID),
  data = data_all_scaled,
  family = nbinom1
)
summary(m_nb)

sim_nb <- simulateResiduals(m_nb)
plot(sim_nb)
testDispersion(sim_nb)
testZeroInflation(sim_nb)
```
The maximal random-effects structure (1 | region) + (1 + visit | plotID) produced near-zero variance for region and a perfect correlation (±1) between random intercepts and slopes, indicating model singularity and overparameterization (Bates et al. 2015; Harrison et al. 2018). In addition, fixed-effect standard errors were undefined (NaN). We therefore simplified the random structure to a random intercept for plotID, which produced stable estimates and satisfactory diagnostic results.
gpt


```{r Bee richness global model complex}
RI_glmm <- glmmTMB(
  bee_richness ~
    hill_q0 + sand + bareground + herb + vegheight + patchiness_ord +
    (1 | plotID),
  family = nbinom2,
  data = data_all_scaled,
  na.action = na.fail
)

print(summary(RI_glmm))

vif(lm(bee_richness ~ hill_q0 + sand + bareground +
       herb + vegheight + patchiness_ord,
       data = data_all_scaled))
#all values below 3 - no multicollinearity 

```

Build global model with simple random effects structure
```{r Bee richness global model}
RI_glmm <- glmmTMB(
  bee_richness ~
    hill_q0 + sand + bareground + herb + vegheight + patchiness_ord +
    (1 | plotID),
  family = nbinom1,
  data = data_all_scaled,
  na.action = na.fail
)

print(summary(RI_glmm))

vif(lm(bee_richness ~ hill_q0 + sand + bareground +
       herb + vegheight + patchiness_ord,
       data = data_all_scaled))
#all values below 3 - no multicollinearity
```

```{r Bee richness dredge}
dr <- dredge(RI_glmm, rank = "AICc", trace = 2)
dr

head(dr)
```

```{r model averaging richness}
cand_ri <- get.models(dr, subset = delta <= 2)

# best model (to be used later for ggpred)
bestmod_ri <- cand_ri[[1]]

avg_mod_ri <- model.avg(cand_ri, revised.var = TRUE)
summary(avg_mod_ri)

# Sum of weights can be calculated like that
sw(subset(dr, delta <= 2))
```

```{r forest plot model average - richness}
# 1. PREP COEFFICIENTS
coef_df_ri <- as.data.frame(summary(avg_mod_ri)$coefmat.full)
coef_df_ri$term <- rownames(coef_df_ri)
coef_df_ri <- coef_df_ri[coef_df_ri$term != "cond((Int))", ]

# Add CI
ci_ri <- confint(avg_mod_ri)
coef_df_ri$lower <- ci_ri[coef_df_ri$term, 1]
coef_df_ri$upper <- ci_ri[coef_df_ri$term, 2]

# 2. MAP IMPORTANCE
coef_df_ri$importance <- NA
coef_df_ri$importance[coef_df_ri$term == "cond(hill_q0)"] <- imp["cond(hill_q0)"]
coef_df_ri$importance[coef_df_ri$term == "cond(sand)"]  <- imp["cond(sand)"]
coef_df_ri$importance[coef_df_ri$term == "cond(herb)"] <- imp["cond(herb)"]
coef_df_ri$importance[coef_df_ri$term == "cond(bareground)"] <- imp["cond(bareground)"]
coef_df_ri$importance[coef_df_ri$term == "cond(vegheight)"] <- imp["cond(vegheight)"]
coef_df_ri$importance[coef_df_ri$term == "cond(patchiness_ord.L)"] <- imp["cond(patchiness_ord)"]
# Drop Q and C contrasts
coef_df_ri <- coef_df_ri[!coef_df_ri$term %in% c("cond(patchiness_ord.Q)", "cond(patchiness_ord.C)"), ]

# 3. COLOR CODE
coef_df_ri$color <- ifelse(coef_df_ri$importance >= 0.8, "#C71586", "gray40")

# 4. CLEAN LABELS
coef_df_ri$clean_label <- coef_df$term %>%
  str_replace("cond\\(", "") %>%
  str_replace("\\)", "") %>%
  str_replace("_ord.L", "") %>%
  str_replace("hill_q0", "Floral Richness") %>%
  str_replace("sand", "Sand") %>%
  str_replace("herb", "Herb layer cover") %>%
  str_replace("bareground", "Bareground") %>%
  str_replace("vegheight", "Vegetation height") %>%
  str_replace("patchiness", "Plot heterogeneity")

# 5. PLOT
p_forest_ri <- ggplot(coef_df_ri, aes(x = Estimate, y = reorder(clean_label, Estimate))) +
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 width = 0, color = coef_df_ri$color, linewidth = 0.8) +
  geom_point(color = coef_df_ri$color, size = 3, stroke = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_line(color = "grey85"),
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 14),
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 4)),
    plot.subtitle = element_text(size = 12, margin = margin(b = 8)),
    axis.title.y = element_text(margin = margin(r = 10))
  ) +
  labs(
    title = "Model-Averaged Effects on Bee Species Richness",
    subtitle = "Predictors with importance ≥ 0.8 highlighted purple",
    x = "Model-averaged estimate (95% CI)",
    y = NULL
  )

p_forest_ri
```

```{r print model avg forest plot richness}
tiff(
  file.path("..", "output", "figures", "forest_mavgbee_ri_600dpi.tiff"),
  width = 3990, height = 3000, res = 600, compression = "lzw"
)

print(p_forest_ri)  
dev.off()
```


To visualise ecological relationships, we generated predicted effects from a final GLMM that included the strongly supported predictors identified through model averaging (hill_q0 and patchiness) and the moderately supported variable sand. Predicted marginal effects were obtained using ggpredict(), holding other predictors at their mean values.
gpt

Best model bee richness
```{r best model bee richness}
bestmod_ri
bestmod_ri <- glmmTMB(bee_richness ~ hill_q0 + patchiness_ord + (1 | plotID),
                      family = nbinom1,
                      data = data_all_scaled)
```

Plot predictions. Backtransform x-axis plant richness for interpretability
```{r bes model bee richness plot predictions}
# Generate predictions
pred_hill   <- ggpredict(bestmod_ri, terms = "hill_q0")
pred_patch  <- ggpredict(bestmod_ri, terms = "patchiness_ord")

# back-transform scaled hill values
hill_mean <- attr(data_all_scaled$hill_q0, "scaled:center")
hill_sd   <- attr(data_all_scaled$hill_q0, "scaled:scale")

pred_hill$x_raw <- pred_hill$x * hill_sd + hill_mean


# Also back-transform raw jitter points for Hill q0:
data_all_scaled$hill_q0_raw <- data_all_scaled$hill_q0 * hill_sd + hill_mean


p_hill <- ggplot() +
  
  geom_jitter(
    data = data_all_scaled,
    aes(x = hill_q0_raw, y = bee_richness),
    width = 0.15, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  
  geom_ribbon(
    data = pred_hill,
    aes(x = x_raw, ymin = conf.low, ymax = conf.high),
    fill = "plum2", alpha = 0.35
  ) +
  
  geom_line(
    data = pred_hill,
    aes(x = x_raw, y = predicted, group = 1),
    color = "#7A0177", linewidth = 1.3
  ) +
  
  labs(
    title = "Effect of Flowering Plant Genus Richness (Hill q = 0)",
    subtitle = "Predicted bee richness ± 95% CI with raw data",
    x = "Plant genus richness (Hill q = 0)",
    y = "Bee richness"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13, color = "black"),
    plot.title = element_text(size = 16, face = "bold")
  )

# ============================================================
# 5. Plot 2: Patchiness (ordinal) prediction with jitter
# ============================================================

p_patch <- ggplot() +
  
  geom_jitter(
    data = data_all_scaled,
    aes(x = patchiness_ord, y = bee_richness),
    width = 0.1, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  
  geom_point(
    data = pred_patch,
    aes(x = x, y = predicted),
    color = "#7A0177", size = 3
  ) +
  
  geom_errorbar(
    data = pred_patch,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    color = "#7A0177", width = 0.1, linewidth = 1
  ) +
  
  labs(
    title = "Effect of Plot Heterogeneity",
    subtitle = "Predicted bee richness ± 95% CI with raw data",
    x = "Plot heterogeneity",
    y = "Bee richness"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13, color = "black"),
    plot.title = element_text(size = 16, face = "bold")
  )

# ============================================================
# 6. Combine into a two-panel figure
# ============================================================

p_combined <- p_hill + p_patch + plot_layout(ncol = 2)

p_combined

```

```{r print predicted best model richness}
tiff(
  file.path("..", "output", "figures", "bestmod_pred_ri_600dpi.tiff"),
  width = 7000, height = 3000, res = 600, compression = "lzw"
)

print(p_combined)  
dev.off()
```




# Bee Hill Q = 1
“Bee abundance and species richness were treated as primary response variables. Hill diversity (q1), which was strongly correlated with richness, was analysed as a complementary metric using a Gamma GLMM. Model averaging was applied to Hillq1 to evaluate whether the same environmental predictors identified for abundance and richness also structured abundance‑weighted diversity. Because Hill q1 is a derived metric and showed minor residual deviations, its results are interpreted as confirmatory rather than independent evidence.”
gpt

Build base model
```{r bee nbinom2 q1 model prep: overdispersion test, response data distribution}
m_q1 <- glmmTMB(
  bee_hill_q1 ~ 1 +
    (1 | plotID),
  data = data_all_scaled,
  family = Gamma(link = "log")
)
summary(m_q1)

sim_q1 <- simulateResiduals(m_q1)
plot(sim_q1)
testDispersion(sim_q1)
testZeroInflation(sim_q1)
```

```{r Bee hill q1 global model}
Q1_glmm <- glmmTMB(
  bee_hill_q1 ~
    hill_q0 + sand + bareground + herb + vegheight + patchiness_ord +
    (1 | plotID),
  family = Gamma(link = "log"),
  data = data_all_scaled,
  na.action = na.fail
)

print(summary(Q1_glmm))

vif(lm(bee_hill_q1 ~ hill_q0 + sand + bareground +
       herb + vegheight + patchiness_ord,
       data = data_all_scaled))
#all values below 3 - no multicollinearity 

sim_Q1_glmm <- simulateResiduals(Q1_glmm)
plot(sim_Q1_glmm)
testDispersion(Q1_glmm)
testZeroInflation(Q1_glmm)

check_model(Q1_glmm)
```

```{r Bee hill q1 dredge}
dq <- dredge(Q1_glmm, rank = "AICc", trace = 2)
dq

head(dq)
```

```{r model averaging hill q1}
cand_q1 <- get.models(dq, subset = delta <= 2)

# best model (to be used later for ggpred)
bestmod_q1 <- cand_q1[[1]]

avg_mod_q1 <- model.avg(cand_q1, revised.var = TRUE)
summary(avg_mod_q1)

# Sum of weights can be calculated like that
sw(subset(dq, delta <= 2))
```

```{r forest plot model average - hillq1}
# 1. PREP COEFFICIENTS
coef_df_q1 <- as.data.frame(summary(avg_mod_q1)$coefmat.full)
coef_df_q1$term <- rownames(coef_df_q1)
coef_df_q1 <- coef_df_q1[coef_df_q1$term != "cond((Int))", ]

# Add CI
ci_q1 <- confint(avg_mod_q1)
coef_df_q1$lower <- ci_q1[coef_df_q1$term, 1]
coef_df_q1$upper <- ci_q1[coef_df_q1$term, 2]

# 2. MAP IMPORTANCE
coef_df_q1$importance <- NA
coef_df_q1$importance[coef_df_q1$term == "cond(hill_q0)"] <- imp["cond(hill_q0)"]
coef_df_q1$importance[coef_df_q1$term == "cond(sand)"]  <- imp["cond(sand)"]
coef_df_q1$importance[coef_df_q1$term == "cond(herb)"] <- imp["cond(herb)"]
coef_df_q1$importance[coef_df_q1$term == "cond(bareground)"] <- imp["cond(bareground)"]
coef_df_q1$importance[coef_df_q1$term == "cond(vegheight)"] <- imp["cond(vegheight)"]
coef_df_q1$importance[coef_df_q1$term == "cond(patchiness_ord.L)"] <- imp["cond(patchiness_ord)"]
# Drop Q and C contrasts
coef_df_q1 <- coef_df_q1[!coef_df_q1$term %in% c("cond(patchiness_ord.Q)", "cond(patchiness_ord.C)"), ]

# 3. COLOR CODE
coef_df_q1$color <- ifelse(coef_df_q1$importance >= 0.8, "#C71586", "gray40")

# 4. CLEAN LABELS
coef_df_q1$clean_label <- coef_df$term %>%
  str_replace("cond\\(", "") %>%
  str_replace("\\)", "") %>%
  str_replace("_ord.L", "") %>%
  str_replace("hill_q0", "Floral Richness") %>%
  str_replace("sand", "Sand") %>%
  str_replace("herb", "Herb layer cover") %>%
  str_replace("bareground", "Bareground") %>%
  str_replace("vegheight", "Vegetation height") %>%
  str_replace("patchiness", "Plot heterogeneity")

# 5. PLOT
p_forest_q1 <- ggplot(coef_df_q1, aes(x = Estimate, y = reorder(clean_label, Estimate))) +
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 width = 0, color = coef_df_q1$color, linewidth = 0.8) +
  geom_point(color = coef_df_q1$color, size = 3, stroke = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_line(color = "grey85"),
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 14),
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 4)),
    plot.subtitle = element_text(size = 12, margin = margin(b = 8)),
    axis.title.y = element_text(margin = margin(r = 10))
  ) +
  labs(
    title = "Model-Averaged Effects on Bee Shannon diversity",
    subtitle = "Predictors with importance ≥ 0.8 highlighted purple",
    x = "Model-averaged estimate (95% CI)",
    y = NULL
  )

p_forest_q1
```
```{r print model avg forest plot Hill q1}
tiff(
  file.path("..", "output", "figures", "forest_mavgbee_q1_600dpi.tiff"),
  width = 3990, height = 3000, res = 600, compression = "lzw"
)

print(p_forest_q1)  
dev.off()
```


# Table extraction
```{r table model averaging}
extract_avg_table <- function(avg_model, dredge_obj, response_name) {
  
  coef_df <- as.data.frame(summary(avg_model)$coefmat.full)
  coef_df$term <- rownames(coef_df)
  
  # Remove intercept
  coef_df <- coef_df %>%
    filter(term != "cond((Int))")
  
  ci <- confint(avg_model)
  coef_df$lower <- ci[coef_df$term, 1]
  coef_df$upper <- ci[coef_df$term, 2]
  
  # Get importance from dredge
  importance_raw <- sw(subset(dredge_obj, delta <= 2))
  
  # Convert safely to named numeric vector
  importance_vals <- as.numeric(importance_raw)
  names(importance_vals) <- names(importance_raw)
  
  # Clean base term (collapse polynomial suffix)
  coef_df$base_term <- coef_df$term %>%
    str_replace("\\.L|\\.Q|\\.C", "") %>%
    str_trim()
  
  # Match correctly (THIS FIXES YOUR ISSUE)
  coef_df$importance <- importance_vals[
    match(coef_df$base_term, names(importance_vals))
  ]
  
  # Keep only linear term for ordinal predictors
  coef_df <- coef_df %>%
    filter(!str_detect(term, "\\.Q|\\.C"))
  
  coef_df <- coef_df %>%
    mutate(
      Predictor = term %>%
        str_replace("cond\\(", "") %>%
        str_replace("\\)", "") %>%
        str_replace("_ord\\.L", "") %>%
        str_replace("_ord", "") %>%
        str_trim()
    )
  
  
  coef_df <- coef_df %>%
    transmute(
      Response = response_name,
      Predictor,
      Estimate = round(Estimate, 3),
      CI = sprintf("%.2f – %.2f", lower, upper),
      Importance = round(importance, 2),
      Significant = ifelse(lower * upper > 0, "yes", "no")
    )
  
  return(coef_df)
}

tab_ab <- extract_avg_table(avg_mod_ab, dd, "Abundance")
tab_ri <- extract_avg_table(avg_mod_ri, dr, "Richness")
tab_q1 <- extract_avg_table(avg_mod_q1, dq, "Shannon q1")

# Combine all responses
table_avg_full <- bind_rows(tab_ab, tab_ri, tab_q1)

table_avg_full

table_avg_main <- table_avg_full %>%
  filter(Importance >= 0.5)

table_avg_main

```

```{r save model avg tables}
# Create nicely formatted gt table
gt_table_avg_full <- table_avg_full %>%
  gt() %>%
  cols_label(
    Response   = "Response variable",
    Predictor  = "Predictor",
    Estimate   = "Estimate",
    CI         = "95% CI",
    Importance = "Importance (Σw)",
    Significant = "CI excludes 0"
  )

# Export to Word
gtsave(gt_table_avg_full,
       "../output/tables/model_avg_full.docx")

# Create nicely formatted gt table
gt_table_avg_main <- table_avg_main %>%
  gt() %>%
  cols_label(
    Response   = "Response variable",
    Predictor  = "Predictor",
    Estimate   = "Estimate",
    CI         = "95% CI",
    Importance = "Importance (Σw)",
    Significant = "CI excludes 0"
  )

# Export to Word
gtsave(gt_table_avg,
       "../output/tables/model_avg_main.docx")
```



# Richness model without patchiness
```{r Bee richness global model complex}
RI2_glmm <- glmmTMB(
  bee_richness ~
    hill_q0 + sand + bareground + herb + vegheight +
    (1 | plotID),
  family = nbinom2,
  data = data_all_scaled,
  na.action = na.fail
)

print(summary(RI2_glmm))

vif(lm(bee_richness ~ hill_q0 + sand + bareground +
       herb + vegheight,
       data = data_all_scaled))
#all values below 3 - no multicollinearity 

```

Build global model with simple random effects structure
```{r Bee richness global model}
RI2_glmm <- glmmTMB(
  bee_richness ~
    hill_q0 + sand + bareground + herb + vegheight +
    (1 | plotID),
  family = nbinom1,
  data = data_all_scaled,
  na.action = na.fail
)

print(summary(RI2_glmm))

vif(lm(bee_richness ~ hill_q0 + sand + bareground +
       herb + vegheight,
       data = data_all_scaled))
#all values below 3 - no multicollinearity
```

```{r Bee richness dredge}
dr2 <- dredge(RI2_glmm, rank = "AICc", trace = 2)
dr2

head(dr2)
```

```{r model averaging richness}
cand_ri2 <- get.models(dr2, subset = delta <= 2)

# best model (to be used later for ggpred)
bestmod_ri2 <- cand_ri2[[1]]

avg_mod_ri2 <- model.avg(cand_ri2, revised.var = TRUE)
summary(avg_mod_ri2)

# Sum of weights can be calculated like that
sw(subset(dr2, delta <= 2))
```

```{r forest plot model average - richness}
r# 1. PREP COEFFICIENTS (model.avg output)
coef_df_ri2 <- as.data.frame(summary(avg_mod_ri2)$coefmat.full)
coef_df_ri2$term <- rownames(coef_df_ri2)

# Remove intercept
coef_df_ri2 <- coef_df_ri2[coef_df_ri2$term != "cond((Int))", ]

# Add CI
ci_ri2 <- confint(avg_mod_ri2)
coef_df_ri2$lower <- ci_ri2[coef_df_ri2$term, 1]
coef_df_ri2$upper <- ci_ri2[coef_df_ri2$term, 2]

# 2. MAP IMPORTANCE (from sw(top_set))
imp_ri2 <- sw(subset(dr2, delta <= 2))
coef_df_ri2$importance <- NA
coef_df_ri2$importance[coef_df_ri2$term == "cond(hill_q0)"]    <- imp_ri2["cond(hill_q0)"]
coef_df_ri2$importance[coef_df_ri2$term == "cond(sand)"]        <- imp_ri2["cond(sand)"]
coef_df_ri2$importance[coef_df_ri2$term == "cond(herb)"]        <- imp_ri2["cond(herb)"]
coef_df_ri2$importance[coef_df_ri2$term == "cond(bareground)"]  <- imp_ri2["cond(bareground)"]
coef_df_ri2$importance[coef_df_ri2$term == "cond(vegheight)"]   <- imp_ri2["cond(vegheight)"]

# 3. COLOR CODE BY IMPORTANCE THRESHOLD
importance_threshold <- 0.8
coef_df_ri2$color <- ifelse(
  coef_df_ri2$importance >= importance_threshold,
  "#C71586",
  "gray40"
)

# 4. CLEAN LABELS
coef_df_ri2$clean_label <- coef_df_ri2$term %>%
  str_replace("cond\\(", "") %>%
  str_replace("\\)", "") %>%
  str_replace("hill_q0",    "Floral Richness") %>%
  str_replace("sand",       "Sand") %>%
  str_replace("herb",       "Herb layer cover") %>%
  str_replace("bareground", "Bareground") %>%
  str_replace("vegheight",  "Vegetation height")

# 5. PLOT
p_forest_ri2 <- ggplot(coef_df_ri2, aes(x = Estimate, y = reorder(clean_label, Estimate))) +
  
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 width = 0,
                 color = coef_df_ri2$color,
                 linewidth = 0.8) +
  
  geom_point(color = coef_df_ri2$color,
             size = 3, stroke = 1) +
  
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_line(color = "grey85"),
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor  = element_blank(),
    axis.text  = element_text(size = 13),
    axis.title = element_text(size = 14),
    plot.title    = element_text(face = "bold", size = 15, margin = margin(b = 4)),
    plot.subtitle = element_text(size = 12, margin = margin(b = 8)),
    axis.title.y  = element_text(margin = margin(r = 10))
  ) +
  
  labs(
    title    = "Model-Averaged Effects on Bee Richness",
    subtitle = "Predictors with importance ≥ 0.8 highlighted purple",
    x = "Model-averaged estimate (95% CI)",
    y = NULL
  )

p_forest_ri2
```

```{r print model avg forest plot richness}
tiff(
  file.path("..", "output", "figures", "forest_mavgbee_ri2_600dpi.tiff"),
  width = 3990, height = 3000, res = 600, compression = "lzw"
)

print(p_forest_ri2)  
dev.off()
```

```{r extract and print table of r2 model}
# Add ri2 to the extraction
tab_ri2 <- extract_avg_table(avg_mod_ri2, dr2, "Richness (no patchiness)")

# Combine all responses including the new model
table_avg_full <- bind_rows(tab_ri2)

# Filter main table
table_avg_main <- table_avg_full %>%
  filter(Importance >= 0.5)

# Full table
gt_table_avg_r2 <- tab_ri2 %>%
  gt() %>%
  cols_label(
    Response   = "Response variable",
    Predictor  = "Predictor",
    Estimate   = "Estimate",
    CI         = "95% CI",
    Importance = "Importance (Σw)",
    Significant = "CI excludes 0"
  )

# Export to Word
gtsave(gt_table_avg_r2,
       "../output/tables/model_avg_r2.docx")
```

# Pool visits 
```{r load and prepare data}
data_plot_scaled <- readRDS(
  file.path("..", "data", "cleaned", "data_plot_scaled.rds")
)

rarefied_size <- readRDS(
  file.path("..","data","cleaned","rarefied_size_plot.rds")
)

# merge size based and coverage based vaues
data_plot_scaled <- data_plot_scaled %>%
  left_join(
    rarefied_size %>%
      select(region, site, plotID, richness),
    by = c("region", "site", "plotID")
  ) %>%
  rename(richness_rarefied_size = richness)

# remove NAs
data_model <- data_plot_scaled %>%
  filter(!is.na(richness_rarefied_size))
```

Collinearity test enviornmental variables
```{r}
# Convert ordinal patchiness to numeric
data_plot_scaled <- data_plot_scaled %>%
  mutate(patchiness_num = as.numeric(patchiness_ord))

# List of variables INCLUDING patchiness
env_vars <- c("bareground", "vegcover", "vegheight", "herb", "woody",
              "sand", "gravel", "cobble", "stone", "wood", "soil", "flat",
              "sloped", "steep", "canopy", "canopy_hp", "branchpile",
              "stonepile", "standingdeadwood", "lyingdeadwood",
              "hill_q0", "hill_q1", "hill_q2",
              "patchiness_num")


# Compute Spearman correlation matrix
cor_matrix <- cor(
  data_plot_scaled[env_vars],
  method = "spearman",
  use = "pairwise.complete.obs"
)

# Identify pairs with |correlation| > 0.7
high_corr <- which(abs(cor_matrix) > 0.7 & abs(cor_matrix) < 1, arr.ind = TRUE)

# Create data frame
correlated_pairs <- data.frame(
  var1 = rownames(cor_matrix)[high_corr[, 1]],
  var2 = colnames(cor_matrix)[high_corr[, 2]],
  correlation = cor_matrix[high_corr]
)

# Remove duplicates
correlated_pairs <- correlated_pairs[correlated_pairs$var1 < correlated_pairs$var2, ]

print(correlated_pairs)

# Plot correlation matrix
corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  addCoef.col = "black",
  number.cex = 0.6
)
```
herb and vegheight highly correlated (0.79) - keep only herb

patchiness is correlated with hill_q0 (0.72), sand (0.71) - keep only patchiness?
VIF collinearity are not high - can we keep this?

Have to simpllify model from multivariate that it was prior to pooling visits


```{r chekc rarefied richness distribution}
hist(data_model$richness_rarefied_size)

```
# Pooled rarefied richness model (merged across plots)
```{r size based rarefaction model}
RI_glmm_rarefied <- glmmTMB(
  richness_rarefied_size ~
    herb + bareground + patchiness_ord +
    (1 | region),
  data = data_model,
  family = poisson(),
  na.action = na.fail
)

summary(RI_glmm_rarefied)
check_model(RI_glmm_rarefied)

RI2_glmm_rarefied <- glmmTMB(
  richness_rarefied_size ~
    hill_q0 + sand + herb + bareground + vegheight + patchiness_ord +
    (1 | region),
  data = data_model,
  family = nbinom1(),
  na.action = na.fail
)

summary(RI2_glmm_rarefied)
check_model(RI2_glmm_rarefied)

RI3_glmm_rarefied <- glmmTMB(
  richness_rarefied_size ~ herb + bareground +patchiness_ord +
    (1 | region),
  data = data_model,
  family = gaussian(link = "log"),
  na.action = na.fail
)
summary(RI3_glmm_rarefied)
check_model(RI3_glmm_rarefied)


RI4_glmm_rarefied <- glmmTMB(
  richness_rarefied_size ~ herb + bareground +patchiness_ord +
    (1 | region),
  data = data_model,
  family = gaussian(),
  na.action = na.fail
)

summary(RI4_glmm_rarefied)
check_model(RI4_glmm_rarefied)
```

```{r Rarefied bee abundance global model}
RAB_glmm <- glmmTMB(
  richness_rarefied_size ~ herb + bareground +patchiness_ord +
    (1 | region),
  data = data_model,
  family = gaussian(),
  na.action = na.fail
)

print(summary(RAB_glmm))
check_model(RAB_glmm)
```

```{r  bee abundance dredge}
dra <- dredge(RAB_glmm, rank = "AICc", trace = 2)
dra

head(dra)
```

```{r model averaging abundance}
cand_rab <- get.models(dra, subset = delta <= 2)

# best model (to be used later for ggpred)
bestmod_rab <- cand_rab[[1]]

avg_mod_rab <- model.avg(cand_rab, revised.var = TRUE)
summary(avg_mod_rab)
confint(avg_mod_rab)

# Sum of weights can be calculated like that
sw(subset(dra, delta <= 2))
```

```{r forest plot model average - abundance}
# 1. PREP COEFFICIENTS (model.avg output)
coef_df <- as.data.frame(summary(avg_mod_rab)$coefmat.full)
coef_df$term <- rownames(coef_df)
# Remove intercept
coef_df <- coef_df[coef_df$term != "cond((Int))", ]
# Add CI
ci <- confint(avg_mod_rab)
coef_df$lower <- ci[coef_df$term, 1]
coef_df$upper <- ci[coef_df$term, 2]
# 2. MAP IMPORTANCE (from sw(top_set))
imp <- sw(subset(dra, delta <= 2))
coef_df$importance <- NA
coef_df$importance[coef_df$term == "cond(herb)"] <- imp["cond(herb)"]
coef_df$importance[coef_df$term == "cond(patchiness_ord.L)"] <- imp["cond(patchiness_ord)"]
coef_df$importance[coef_df$term == "cond(patchiness_ord.Q)"] <- imp["cond(patchiness_ord)"]
# Drop Q and C contrasts and bareground
coef_df <- coef_df[!coef_df$term %in% c("cond(patchiness_ord.C)",
                                          "cond(bareground)"), ]
# 3. COLOR CODE BY IMPORTANCE THRESHOLD
importance_threshold <- 0.8
coef_df$color <- ifelse(
  coef_df$importance >= importance_threshold,
  "#C71586",
  "gray40"
)
# 4. CLEAN LABELS
coef_df$clean_label <- coef_df$term %>%
  str_replace("cond\\(", "") %>% 
  str_replace("\\)", "") %>%      
  str_replace("patchiness_ord.L", "Patchiness (linear)") %>%
  str_replace("patchiness_ord.Q", "Patchiness (quadratic)") %>%
  str_replace("herb", "Herbaceous vegetation")

# 5. plot
p_forest_rab <- ggplot(coef_df, aes(x = Estimate, y = reorder(clean_label, Estimate))) +
  
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 width = 0,
                 color = coef_df$color,
                 linewidth = 0.8) +
  
  geom_point(color = coef_df$color,
             size = 3, stroke = 1) +
  
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_line(color = "grey85"),
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 14),
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 4)),
    plot.subtitle = element_text(size = 12, margin = margin(b = 8)),
    axis.title.y = element_text(margin = margin(r = 10))
  ) +
  
  labs(
    title = "Model-Averaged Effects on Rarefied Bee Richness",
    subtitle = "Predictors with importance ≥ 0.8 highlighted purple",
    x = "Model-averaged estimate (95% CI)",
    y = NULL
  )
p_forest_rab

plot(ggpredict(bestmod_rab, terms = "patchiness_ord"))
ggpredict(bestmod_rab, terms = "patchiness_ord [all]")
```
Best model rarefied bee abundance
```{r best model POOLED bee abundance}
bestmod_rri
bestmod_rri <- glmmTMB(bee_abundance ~bareground + herb + patchiness_ord + (1 | region),
                      family = nbinom1,
                      data = data_plot_scaled)

summary(bestmod_pab)
```

# Pooled rarefied data - flower instead of patchiness
```{r Rarefied bee abundance global model}
RAB2_glmm <- glmmTMB(
  richness_rarefied_size ~ herb + bareground + hill_q0 +
    (1 | region),
  data = data_model,
  family = gaussian(),
  na.action = na.fail
)

print(summary(RAB2_glmm))
check_model(RAB2_glmm)
```

```{r  bee abundance dredge}
dra2 <- dredge(RAB2_glmm, rank = "AICc", trace = 2)
dra2

head(dra2)
```

```{r model averaging abundance}
cand_rab2 <- get.models(dra2, subset = delta <= 2)

# best model (to be used later for ggpred)
bestmod_rab2 <- cand_rab[[1]]

avg_mod_rab2 <- model.avg(cand_rab2, revised.var = TRUE)
summary(avg_mod_rab2)
confint(avg_mod_rab2)

# Sum of weights can be calculated like that
sw(subset(dra2, delta <= 2))
```

```{r forest plot model average - abundance}
# 1. PREP COEFFICIENTS (model.avg output)
coef_df <- as.data.frame(summary(avg_mod_rab2)$coefmat.full)
coef_df$term <- rownames(coef_df)
# Remove intercept
coef_df <- coef_df[coef_df$term != "cond((Int))", ]
# Add CI
ci <- confint(avg_mod_rab2)
coef_df$lower <- ci[coef_df$term, 1]
coef_df$upper <- ci[coef_df$term, 2]
# 2. MAP IMPORTANCE (from sw(top_set))
imp <- sw(subset(dra2, delta <= 2))
coef_df$importance <- NA
coef_df$importance[coef_df$term == "cond(herb)"] <- imp["cond(herb)"]
coef_df$importance[coef_df$term == "cond(hill_q0)"] <- imp["cond(hill_q0"]

# # Drop Q and C contrasts and bareground
# coef_df <- coef_df[!coef_df$term %in% c("cond(patchiness_ord.C)",
                                          # "cond(bareground)"), ]
# 3. COLOR CODE BY IMPORTANCE THRESHOLD
importance_threshold <- 0.8
coef_df$color <- ifelse(
  coef_df$importance >= importance_threshold,
  "#C71586",
  "gray40"
)
# 4. CLEAN LABELS
coef_df$clean_label <- coef_df$term %>%
  str_replace("cond\\(", "") %>% 
  str_replace("\\)", "") %>%      
  str_replace("patchiness_ord.L", "Patchiness (linear)") %>%
  str_replace("patchiness_ord.Q", "Patchiness (quadratic)") %>%
  str_replace("herb", "Herbaceous vegetation")

# 5. plot
p_forest_rab <- ggplot(coef_df, aes(x = Estimate, y = reorder(clean_label, Estimate))) +
  
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 width = 0,
                 color = coef_df$color,
                 linewidth = 0.8) +
  
  geom_point(color = coef_df$color,
             size = 3, stroke = 1) +
  
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_line(color = "grey85"),
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 14),
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 4)),
    plot.subtitle = element_text(size = 12, margin = margin(b = 8)),
    axis.title.y = element_text(margin = margin(r = 10))
  ) +
  
  labs(
    title = "Model-Averaged Effects on Rarefied Bee Richness",
    subtitle = "Predictors with importance ≥ 0.8 highlighted purple",
    x = "Model-averaged estimate (95% CI)",
    y = NULL
  )
p_forest_rab

plot(ggpredict(bestmod_rab, terms = "patchiness_ord"))
ggpredict(bestmod_rab, terms = "patchiness_ord [all]")
```





# Pooled raw data
```{r load and prepare data}
data_plot_scaled <- readRDS(
  file.path("..", "data", "cleaned", "data_plot_scaled.rds")
)
```

Collinearity test environmental variables
```{r}
# Convert ordinal patchiness to numeric
data_plot_scaled <- data_plot_scaled %>%
  mutate(patchiness_num = as.numeric(patchiness_ord))


# List of variables INCLUDING patchiness
env_vars <- c("bareground", "vegcover", "vegheight", "herb", "woody",
              "sand", "gravel", "cobble", "stone", "wood", "soil", "flat",
              "sloped", "steep", "canopy", "canopy_hp", "branchpile",
              "stonepile", "standingdeadwood", "lyingdeadwood",
              "hill_q0", "hill_q1", "hill_q2",
              "patchiness_num")


# Compute Spearman correlation matrix
cor_matrix <- cor(
  data_plot_scaled[env_vars],
  method = "spearman",
  use = "pairwise.complete.obs"
)

# Identify pairs with |correlation| > 0.7
high_corr <- which(abs(cor_matrix) > 0.7 & abs(cor_matrix) < 1, arr.ind = TRUE)

# Create data frame
correlated_pairs <- data.frame(
  var1 = rownames(cor_matrix)[high_corr[, 1]],
  var2 = colnames(cor_matrix)[high_corr[, 2]],
  correlation = cor_matrix[high_corr]
)

# Remove duplicates
correlated_pairs <- correlated_pairs[correlated_pairs$var1 < correlated_pairs$var2, ]

print(correlated_pairs)

# Plot correlation matrix
corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  addCoef.col = "black",
  number.cex = 0.6
)
```
herb and vegheight highly correlated (0.79) - keep only herb

patchiness is correlated with hill_q0 (0.72), sand (0.71) - keep only patchiness?
VIF collinearity are not high - can we keep this?

Have to simpllify model from multivariate that it was prior to pooling visits


# Pooled abundance
```{r check model distribution}
nb2_plot <- glmmTMB(
  bee_abundance ~
    herb + bareground + patchiness_ord +
    (1 |region),
  family = nbinom2,
  data = data_plot_scaled,
  na.action = na.fail
)
print(summary(nb2_plot))

check_model(nb2_plot)

nb1_plot <- glmmTMB(
  bee_abundance ~
    herb + bareground + patchiness_ord +
    (1 |region),
  family = nbinom1,
  data = data_plot_scaled,
  na.action = na.fail
)
print(summary(nb1_plot))

check_model(nb1_plot)

# Stick with nbinom1, better AIC
```


```{r Pooled bee abundance global model}
PAB_glmm <- glmmTMB(
  bee_abundance ~
   herb + bareground + patchiness_ord +
    (1 | region),
  family = nbinom1,
  data = data_plot_scaled,
  na.action = na.fail
)
print(summary(PAB_glmm))

check_model(PAB_glmm)
```
```{r Pooled bee abundance dredge}
dpa <- dredge(PAB_glmm, rank = "AICc", trace = 2)
dpa

head(dpa)
```

```{r model averaging abundance}
cand_pab <- get.models(dpa, subset = delta <= 2)

# best model (to be used later for ggpred)
bestmod_pab <- cand_pab[[1]]

avg_mod_pab <- model.avg(cand_pab, revised.var = TRUE)
summary(avg_mod_pab)
confint(avg_mod_pab)

# Sum of weights can be calculated like that
sw(subset(dpa, delta <= 2))
```

```{r forest plot model average - abundance}
# 1. PREP COEFFICIENTS (model.avg output)
coef_df <- as.data.frame(summary(avg_mod_pab)$coefmat.full)
coef_df$term <- rownames(coef_df)

# Remove intercept
coef_df <- coef_df[coef_df$term != "cond((Int))", ]

# Add CI
ci <- confint(avg_mod_pab)
coef_df$lower <- ci[coef_df$term, 1]
coef_df$upper <- ci[coef_df$term, 2]

# 2. MAP IMPORTANCE (from sw(top_set))

imp <- sw(subset(dpa, delta <= 2))

coef_df$importance <- NA
coef_df$importance[coef_df$term == "cond(bareground)"] <- imp["cond(bareground)"]
coef_df$importance[coef_df$term == "cond(herb)"] <- imp["cond(herb)"]
coef_df$importance[coef_df$term == "cond(patchiness_ord.L)"] <- imp["cond(patchiness_ord)"]
# Drop Q and C contrasts
coef_df <- coef_df[!coef_df$term %in% c("cond(patchiness_ord.Q)", "cond(patchiness_ord.C)"), ]

# 3. COLOR CODE BY IMPORTANCE THRESHOLD

importance_threshold <- 0.8

coef_df$color <- ifelse(
  coef_df$importance >= importance_threshold,
  "#C71586",
  "gray40"
)

# 4. CLEAN LABELS

coef_df$clean_label <- coef_df$term %>%
  str_replace("cond\\(", "") %>% 
  str_replace("\\)", "") %>%      
  str_replace("_ord.L", "") %>% 
  str_replace("bareground", "Bareground") %>%
  str_replace("herb", "Herbaceous vegetation") %>%
  str_replace("patchiness", "Plot heterogeneity")


# 5. plot

p_forest_pab <- ggplot(coef_df, aes(x = Estimate, y = reorder(clean_label, Estimate))) +
  
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 width = 0,
                 color = coef_df$color,
                 linewidth = 0.8) +
  
  geom_point(color = coef_df$color,
             size = 3, stroke = 1) +
  
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_line(color = "grey85"),   # vertical gridlines ON
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 14),
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 4)),
    plot.subtitle = element_text(size = 12, margin = margin(b = 8)),
    
    axis.title.y = element_text(margin = margin(r = 10))
  ) +
  
  labs(
    title = "Model-Averaged Effects on Raw Pooled Bee Abundance",
    subtitle = "Predictors with importance ≥ 0.8 highlighted purple",
    x = "Model-averaged estimate (95% CI)",
    y = NULL
  )

p_forest_pab
```

Best model POOLED bee abundance
```{r best model POOLED bee abundance}
bestmod_pab
bestmod_pab <- glmmTMB(bee_abundance ~bareground + herb + patchiness_ord + (1 | region),
                      family = nbinom1,
                      data = data_plot_scaled)

summary(bestmod_pab)
```

Plot predictions. Backtransform x-axis plant abundance for interpretability
```{r bes model bee abundance plot predictions}
# Generate predictions for each predictor
pred_herb      <- ggpredict(bestmod_pab, terms = "herb")
pred_bareground <- ggpredict(bestmod_pab, terms = "bareground")
pred_patch      <- ggpredict(bestmod_pab, terms = "patchiness_ord")

# Back-transform hill_q0 and sand to raw units
herb_mean <- attr(data_plot_scaled$herb, "scaled:center")
herb_sd   <- attr(data_plot_scaled$herb, "scaled:scale")

bareground_mean <- attr(data_plot_scaled$bareground, "scaled:center")
bareground_sd   <- attr(data_plot_scaled$bareground, "scaled:scale")


# plot predictions
p_herb <- ggplot() +
  geom_jitter(
    data = data_plot_scaled,
    aes(x = herb, y = bee_abundance),
    width = 0.15, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  geom_ribbon(
    data = pred_herb,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    fill = "plum2", alpha = 0.35
  ) +
  geom_line(
    data = pred_herb,
    aes(x = x, y = predicted),
    color = "#7A0177", linewidth = 1.3
  ) +
  labs(
    title = "Effect of herb Vegetation",
    x = "herb vegetation (scaled)",
    y = "Bee abundance"
  ) +
  theme_minimal(base_size = 14) +
  theme(panel.grid.minor = element_blank())

p_bare <- ggplot() +
  geom_jitter(
    data = data_plot_scaled,
    aes(x = bareground, y = bee_abundance),
    width = 0.15, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  geom_ribbon(
    data = pred_bareground,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    fill = "plum2", alpha = 0.35
  ) +
  geom_line(
    data = pred_bareground,
    aes(x = x, y = predicted),
    color = "#7A0177", linewidth = 1.3
  ) +
  labs(
    title = "Effect of Bare Ground",
    x = "Bare ground (scaled)",
    y = "Bee abundance"
  ) +
  theme_minimal(base_size = 14) +
  theme(panel.grid.minor = element_blank())

p_patch <- ggplot() +
  geom_jitter(
    data = data_plot_scaled,
    aes(x = patchiness_ord, y = bee_abundance),
    width = 0.1, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  geom_point(
    data = pred_patch,
    aes(x = x, y = predicted),
    color = "#7A0177", size = 3
  ) +
  geom_errorbar(
    data = pred_patch,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    color = "#7A0177", width = 0.1, linewidth = 1
  ) +
  labs(
    title = "Effect of Habitat Heterogeneity",
    x = "Plot heterogeneity (ordinal)",
    y = "Bee abundance"
  ) +
  theme_minimal(base_size = 14) +
  theme(panel.grid.minor = element_blank())

p_combined_abundance <- p_herb + p_bare + p_patch + plot_layout(ncol = 3)

p_combined_abundance
```



# Pooled richness
```{r check model distribution}
nb2_plot <- glmmTMB(
  bee_richness ~
    herb + bareground + patchiness_ord +
    (1 |region),
  family = nbinom2,
  data = data_plot_scaled,
  na.action = na.fail
)
print(summary(nb2_plot))
check_model(nb2_plot)
nb1_plot <- glmmTMB(
  bee_richness ~
    herb + bareground + patchiness_ord +
    (1 |region),
  family = nbinom1,
  data = data_plot_scaled,
  na.action = na.fail
)
print(summary(nb1_plot))
check_model(nb1_plot)
# Stick with nbinom1, better AIC
```

```{r Pooled bee richness global model}
PRI_glmm <- glmmTMB(
  bee_richness ~
   herb + bareground + patchiness_ord +
    (1 | region),
  family = nbinom1,
  data = data_plot_scaled,
  na.action = na.fail
)
print(summary(PRI_glmm))
check_model(PRI_glmm)
```

```{r Pooled bee richness dredge}
dpr <- dredge(PRI_glmm, rank = "AICc", trace = 2)
dpr
head(dpr)
```

```{r model averaging richness}
cand_pri <- get.models(dpr, subset = delta <= 2)
# best model (to be used later for ggpred)
bestmod_pri <- cand_pri[[1]]
avg_mod_pri <- model.avg(cand_pri, revised.var = TRUE)
summary(avg_mod_pri)
confint(avg_mod_pri)
# Sum of weights can be calculated like that
sw(subset(dpr, delta <= 2))
```

```{r forest plot model average - richness}
# 1. PREP COEFFICIENTS (model.avg output)
coef_df <- as.data.frame(summary(avg_mod_pri)$coefmat.full)
coef_df$term <- rownames(coef_df)
# Remove intercept
coef_df <- coef_df[coef_df$term != "cond((Int))", ]
# Add CI
ci <- confint(avg_mod_pri)
coef_df$lower <- ci[coef_df$term, 1]
coef_df$upper <- ci[coef_df$term, 2]
# 2. MAP IMPORTANCE (from sw(top_set))
imp <- sw(subset(dpr, delta <= 2))
coef_df$importance <- NA
coef_df$importance[coef_df$term == "cond(bareground)"] <- imp["cond(bareground)"]
coef_df$importance[coef_df$term == "cond(herb)"] <- imp["cond(herb)"]
coef_df$importance[coef_df$term == "cond(patchiness_ord.L)"] <- imp["cond(patchiness_ord)"]
coef_df$importance[coef_df$term == "cond(patchiness_ord.Q)"] <- imp["cond(patchiness_ord)"]
# Drop only C contrast
coef_df <- coef_df[!coef_df$term %in% c("cond(patchiness_ord.C)"), ]
# 3. COLOR CODE BY IMPORTANCE THRESHOLD
importance_threshold <- 0.8
coef_df$color <- ifelse(
  !is.na(coef_df$importance) & coef_df$importance >= importance_threshold,
  "#C71586",
  "gray40"
)
# 4. CLEAN LABELS
coef_df$clean_label <- coef_df$term %>%
  str_replace("cond\\(", "") %>% 
  str_replace("\\)", "") %>%
  str_replace("_ord\\.L", " (linear)") %>%
  str_replace("_ord\\.Q", " (quadratic)") %>%
  str_replace("bareground", "Bareground") %>%
  str_replace("herb", "Herbaceous vegetation") %>%
  str_replace("patchiness", "Plot heterogeneity")
# 5. plot
p_forest_pri <- ggplot(coef_df, aes(x = Estimate, y = reorder(clean_label, Estimate))) +
  
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 width = 0,
                 color = coef_df$color,
                 linewidth = 0.8) +
  
  geom_point(color = coef_df$color,
             size = 3, stroke = 1) +
  
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_line(color = "grey85"),
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 14),
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 4)),
    plot.subtitle = element_text(size = 12, margin = margin(b = 8)),
    axis.title.y = element_text(margin = margin(r = 10))
  ) +
  
  labs(
    title = "Model-Averaged Effects on Raw Pooled Bee Richness",
    subtitle = "Predictors with importance ≥ 0.8 highlighted purple",
    x = "Model-averaged estimate (95% CI)",
    y = NULL
  )
p_forest_pri
```

The changes from the previous version:
- Bareground kept in — only `patchiness_ord.C` is dropped
- Added `!is.na(coef_df$importance) &` to the color logic to safely handle any terms with NA importance without throwing an error
```

