---
title: "06_habvariables_multimodel"
output: html_document
output_dir: "knit"
---

```{r Load packages}
library(tidyverse)
library(ggeffects)
library(glmmTMB)
library(car)
library(DHARMa)
library(patchwork)
library(MuMIn)
library(vegan)
library(iNEXT)
library(lme4)
library(corrplot)
library(lmerTest)
library(broom.mixed)
```

```{r Load data}
data_all_scaled <- readRDS(file.path("..", "data", "cleaned", "data_all_scaled.rds"))
```
Slope is at plot level because only plots have repeated measures
Region remains a random intercept to capture large-scale spatial structure
No need for site as fixed or random because you use plot-level predictors
 (1 | region) + (1 + visit | plotID) 

Test correlation between differenct bee metrics
```{r}
# List of numeric environmental variables
bee_vars <- c("bee_abundance", "bee_richness", "bee_hill_q1", "bee_hill_q2")
#visit and patchiness, although included in coming models as ordinal, are treated as numeric for Spearman computation

# Compute Spearman correlation matrix
cor_matrix <- cor(
  data_all_scaled[bee_vars],
  method = "spearman",
  use = "pairwise.complete.obs")

cor_matrix
```
Because all plots were surveyed with identical sampling effort (same area, same method, same observer, same frequency), rarefaction was not applied. Raw abundance, richness, and Hill diversity values were directly comparable across samples. Given the high correlation between HIll q = 1 and HIll q = 2 values, I only run abudnance, richness, and HIll q = 1 models in parallel.
Rarefaction would artificially reduce information content and compromise the GLMM structure. Therefore, analyses were conducted using raw count-based response variables standardized by equal sample effort.
 
## Bee abundance models

Bee ab model prep:
```{r bee Poisson ab model prep: overdispersion test, response data distribution}
#baseline poisson null model
m_pois <- glmmTMB(
  bee_abundance ~ 1 +
    (1 | region) +
    (1 + visit | plotID),
  data = data_all_scaled,
  family = poisson
)

summary(m_pois)

#test for overdispersion
overdisp_fun <- function(model) {
  rdf <- df.residual(model)
  rp <- residuals(model, type = "pearson")
  Pearson.chisq <- sum(rp^2)
  ratio <- Pearson.chisq / rdf
  pval <- pchisq(Pearson.chisq, df = rdf, lower.tail = FALSE)
  c(chisq = Pearson.chisq, ratio = ratio, rdf = rdf, p = pval)
}

overdisp_fun(m_pois) #clear overdispersion

sim_pois <- simulateResiduals(m_pois)
plot(sim_pois)
testDispersion(sim_pois)
testZeroInflation(sim_pois)
```
```{r bee nbinom2 ab model prep: overdispersion test, response data distribution}
m_nb <- glmmTMB(
  bee_abundance ~ 1 +
    (1 | region) +
    (1 + visit | plotID),
  data = data_all_scaled,
  family = nbinom2
)
summary(m_nb)

sim_nb <- simulateResiduals(m_nb)
plot(sim_nb)
testDispersion(sim_nb)
testZeroInflation(sim_nb)
```

KS test for nbinom2 shows significant deviation, try nbinom1 
```{r bee nbinom1 ab model prep: overdispersion test, response data distribution}
m_nb <- glmmTMB(
  bee_abundance ~ 1 +
    (1 | region) +
    (1 + visit | plotID),
  data = data_all_scaled,
  family = nbinom1
)
summary(m_nb)

sim_nb <- simulateResiduals(m_nb)
plot(sim_nb)
testDispersion(sim_nb)
testZeroInflation(sim_nb)
```
nbinom1 fits better - use this from now on
"It is common that the KS test shows significance even when the model is fine.
Minor deviations in QQ plots without patterns in other tests do not require fixing."

```{r bee nbinom1 ab model with different random effects structure}
m_nb_simple <- glmmTMB(
  bee_abundance ~ 1 +
    (1 + visit | plotID),
  data = data_all_scaled,
  family = nbinom1
)
summary(m_nb_simple)

sim_nb_simple <- simulateResiduals(m_nb_simple)
plot(sim_nb_simple)
testDispersion(sim_nb_simple)
testZeroInflation(sim_nb_simple)
```
We initially considered a random-effects structure reflecting the sampling design, including random intercepts for region, sites nested within region, and random slopes of visit within plots.
However, region exibited near-zero variance without improving model fit.
Following Bates et al. (2015) and Harrison et al. (2018), we retained only the random intercept for plotID, which was strongly supported by the data and reflected the clustering of repeated surveys within plots.
Additional hierarchical levels (region, site) were included as fixed effects where biologically relevant.
gpt
(Variance estimates showed that the region-level variance was near zero (1.2×10⁻⁷),
ICC values and conditional R² confirmed that these additional random effects explained no additional hierarchical variation.
Model diagnostics (DHARMa) indicated no improvement in dispersion, zero-inflation, or residual structure.
Following Bates et al. (2015) and Harrison et al. (2018), we therefore selected the more parsimonious random-effects structure (1|plotID), which was strongly supported by the data and avoided overparameterization.)

```{r bee ab model prep: test for collinearity among habitat explanatory variables}
# List of numeric environmental variables
env_vars <- c("visit", "bareground", "vegcover", "vegheight", "herb", "woody",
              "sand", "gravel", "cobble", "stone", "wood", "soil", "flat",
              "sloped", "steep", "canopy", "canopy_hp", "branchpile",
              "stonepile", "patchiness", "standingdeadwood", "lyingdeadwood",
              "hill_q0", "hill_q1", "hill_q2")
#visit and patchiness, although included in coming models as ordinal, are treated as numeric for Spearman computation

# Compute Spearman correlation matrix
cor_matrix <- cor(
  data_all_scaled[env_vars],
  method = "spearman",
  use = "pairwise.complete.obs")

# Identify pairs with |correlation| > 0.7
high_corr <- which(abs(cor_matrix) > 0.7 & abs(cor_matrix) <= 1, arr.ind = TRUE)

# Create a data frame of correlated pairs
correlated_pairs <- data.frame(
  var1 = rownames(cor_matrix)[high_corr[, 1]],
  var2 = colnames(cor_matrix)[high_corr[, 2]],
  correlation = cor_matrix[high_corr]
)

# Remove duplicates (var1-var2 vs var2-var1)
correlated_pairs[correlated_pairs$var1 < correlated_pairs$var2, ]

# visualize correlation
corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  addCoef.col = "black",
  number.cex = 0.6
)

# New variable dataframe (merged covariates with spearman above 0.7)
# bareground - vegcover -> bareground
# herb - woody -> herb
# flat - sloped -> sloped
# hill_q0  hill_q1  hill_q2 -> hill_q0

# canopy_hp excluded as constant across visits
env_vars <- c("bareground", "vegheight", "herb", "sand", "gravel", "cobble",
              "stone", "wood", "soil", "sloped", "steep", "canopy",
              "branchpile", "stonepile", "standingdeadwood", "lyingdeadwood",
              "hill_q0")
```

Can we include site as explanatory variable in a multivariate model? - No, this chunk shows that too many of the habitat variables confound with site, and should therefore not be included in the same multivariate model.
```{r bee ab model prep: diagnostic site vs habitat covariates}
num_vars <- c("visit", "bareground", "vegcover", "vegheight", "herb", "woody",
              "sand", "gravel", "cobble", "stone", "wood", "soil", "flat",
              "sloped", "steep", "canopy", "canopy_hp", "branchpile",
              "stonepile", "patchiness", "standingdeadwood", "lyingdeadwood",
              "hill_q0")

site_tests <- lapply(num_vars, function(v) {
  wt <- wilcox.test(data_all_scaled[[v]] ~ data_all_scaled$site)
  data.frame(
    variable = v,
    W = wt$statistic,
    p_value = wt$p.value
  )
})

site_tests <- do.call(rbind, site_tests)

site_tests$p_adj <- p.adjust(site_tests$p_value, method = "BH")

# these variables confound with site and cannot be included in a multivariate model together
confounded_vars <- site_tests |>
  subset(p_adj < 0.05) |>
  arrange(p_adj)

confounded_vars$variable

# these variables do not confound with site and could be included in a multivariate model together
independent_vars <- site_tests |>
  subset(p_adj >= 0.05)

independent_vars$variable
```

Run univariate models for ordinal covariates
```{r bee ab model: univarite models for ordinal variables}
# patchiness
p_ord <- glmmTMB(bee_abundance ~ patchiness_ord + (1 + visit | plotID),
                 family = nbinom1, data = data_all_scaled)
summary(p_ord)

#visit
v_ord <- glmmTMB(bee_abundance ~ visit_ord + (1 + visit | plotID),
                 family = nbinom1, data = data_all_scaled)
summary(v_ord)


#plot 
pred <- ggpredict(p_ord, terms = "patchiness_ord")
ggplot(pred, aes(x = x, y = predicted)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  theme_minimal() +
  labs(title = "Effect of Patchiness (Ordinal) on Bee Abundance",
       x = "Patchiness Level", y = "Predicted Abundance")


pred <- ggpredict(v_ord, terms = "visit_ord")
ggplot(pred, aes(x = x, y = predicted)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  theme_minimal() +
  labs(title = "Effect of visit (Ordinal) on Bee Abundance",
       x = "Visit Number", y = "Predicted Abundance")

```

Run univariate models for numeric covariates
```{r bee ab model: run univariate model for numeric variabels}
model_um_abundance <- tibble(
  Variable = env_vars,
  Model = map(env_vars, ~ glmmTMB(
    as.formula(paste("bee_abundance ~ `", .x, "` + (1 + visit | plotID)", sep = "")),
    family = nbinom1,
    data = data_all_scaled
  ))
)
```

```{r bee ab model: plot univarite model predictions}
model_df <- model_um_abundance

# Extract model stats
model_df <- model_df |>
  mutate(
    Summary = map(Model, ~ summary(.x)$coefficients$cond),
    CoefName = map_chr(Summary, ~ setdiff(rownames(.x), "(Intercept)")),
    Estimate = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Estimate"]),
    StdError = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Std. Error"]),
    p_value = map_dbl(Summary, ~ .x[setdiff(rownames(.x), "(Intercept)"), "Pr(>|z|)"]),
    CI_lower = Estimate - 1.96 * StdError,
    CI_upper = Estimate + 1.96 * StdError,
    AIC = map_dbl(Model, AIC)
  )

print(model_df)

# Plot all model output estimates, mark significant ones in different color
# Sort by AIC for convenience
model_df_am <- model_df |> arrange(AIC)

# Add significance flag
model_df_am <- model_df_am |>
  mutate(Significant = ifelse(p_value < 0.05, "Significant", "Not Significant"))

# Forest plot with color by significance
ggplot(model_df_am, aes(x = reorder(Variable, Estimate), y = Estimate, color = Significant)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  ylab("Effect Size (Estimate)") +
  xlab("Environmental Variable") +
  ggtitle("Univariate Effects on Bee Abundance") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(values = c("Significant" = "#D55E00", "Not Significant" = "gray"))


# plot significant univariate models
# Filter significant models
significant_df <- model_df |>
  filter(!is.na(p_value), p_value < 0.05) |>
  arrange(AIC)

# View summary table
print(significant_df |> select(Variable, AIC, Estimate, CI_lower, CI_upper, p_value))

# Plot predicted effects using ggpredict
plots <- map2(significant_df$Model, significant_df$Variable, ~ {
  pred <- ggpredict(.x, terms = .y, bias_correction = TRUE)
  ggplot(pred, aes(x = x, y = predicted)) +
    geom_line() +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
    theme_minimal() +
    labs(title = paste("Effect of", .y, "on Bee Abundance"),
         x = .y, y = "Predicted Abundance")
})

ggplot(significant_df, aes(x = reorder(Variable, Estimate), y = Estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  ylab("Effect Size (Estimate)") +
  xlab("Environmental Variable") +
  ggtitle("Significant Predictors of Bee Abundance (p < 0.05)") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray")
```


Build multivariate model for bee abundance

only adding plotID as random effect as
stable variance estimate
no singularity
same AIC
simpler structure (thus preferred by parsimony)
fixed effects behave cleanly and consistently
```{r Bee abundance global model}
AB_glmm <- glmmTMB(
  bee_abundance ~
    hill_q0 + sand + herb + bareground + vegheight + patchiness_ord +
    (1 + visit | plotID),
  family = nbinom1,
  data = data_all_scaled,
  na.action = na.fail
)
print(summary(AB_glmm))

vif(lm(bee_abundance ~ hill_q0 + sand + soil + vegheight + patchiness_ord,
       data = data_all_scaled))
#all values below 3 - no multicollinearity 
```
1. Correlation = –0.98 → essentially singular
2. Slope variance tiny → slope not informative
3. Plots are too few (n=64) × visits too few (5 points each) → limited temporal resolution
4. Slope breaks in the full model (but not in null model)
5. Fixed effects unchanged with or without slope
6. Reviewer will accept dropping slope if you explain these diagnostics
So the correct random structure remains:

he random slope variance was very small (σ² = 0.039), and the intercept–slope correlation was extremely high (r = –0.98), indicating overparameterization and a near‑singular fit.
According to Bates et al. (2015) and Harrison et al. (2018), such correlations signal that the random slope is not supported by the data and should be removed.
Simplifying to (1 | plotID) produced stable models without affecting fixed-effect inference.
We therefore retained the simpler random-effects structure in all models.
gpt 

Remove visit from random effects structure
```{r Bee abundance global model without visit as random slope}
AB_glmm <- glmmTMB(
  bee_abundance ~
    hill_q0 + sand + herb + bareground + vegheight + patchiness_ord +
    (1 | plotID),
  family = nbinom1,
  data = data_all_scaled,
  na.action = na.fail
)
print(summary(AB_glmm))

vif(lm(bee_abundance ~ hill_q0 + sand + bareground + vegheight + patchiness_ord,
       data = data_all_scaled))
#all values below 3 - no multicollinearity 
```

```{r Bee abundance dredge}
dd <- dredge(AB_glmm, rank = "AICc", trace = 2)
dd

head(dd)

# best model (to be used later for ggpred)
bestmod_ab <- cand[[1]]
```

```{r model averaging abundance}
cand <- get.models(dd, subset = delta <= 2)

avg_mod <- model.avg(cand, revised.var = TRUE)
summary(avg_mod)
confint(avg_mod)

# Sum of weights can be calculated like that
sw(subset(dd, delta <= 2))
```

```{r falcon plot model average - abundance}

# 1. PREP COEFFICIENTS (model.avg output)
coef_df <- as.data.frame(summary(avg_mod)$coefmat.full)
coef_df$term <- rownames(coef_df)

# Remove intercept
coef_df <- coef_df[coef_df$term != "cond((Int))", ]

# Add CI
ci <- confint(avg_mod)
coef_df$lower <- ci[coef_df$term, 1]
coef_df$upper <- ci[coef_df$term, 2]

# 2. MAP IMPORTANCE (from sw(top_set))

imp <- sw(subset(dd, delta <= 2))

coef_df$importance <- NA
coef_df$importance[coef_df$term == "cond(hill_q0)"] <- imp["cond(hill_q0)"]
coef_df$importance[coef_df$term == "cond(sand)"]  <- imp["cond(sand)"]
coef_df$importance[coef_df$term == "cond(herb)"] <- imp["cond(herb)"]
coef_df$importance[coef_df$term == "cond(bareground)"] <- imp["cond(bareground)"]
coef_df$importance[coef_df$term == "cond(vegheight)"] <- imp["cond(vegheight)"]
coef_df$importance[coef_df$term == "cond(patchiness_ord.L)"] <- imp["cond(patchiness_ord)"]
# Drop Q and C contrasts
coef_df <- coef_df[!coef_df$term %in% c("cond(patchiness_ord.Q)", "cond(patchiness_ord.C)"), ]

# 3. COLOR CODE BY IMPORTANCE THRESHOLD

importance_threshold <- 0.8

coef_df$color <- ifelse(
  coef_df$importance >= importance_threshold,
  "#C71586",
  "gray40"
)

# 4. CLEAN LABELS

coef_df$clean_label <- coef_df$term %>%
  str_replace("cond\\(", "") %>%
  str_replace("\\)", "") %>%
  str_replace("_ord.L", "") %>%
  str_replace("hill_q0", "Floral Richness") %>%
  str_replace("sand", "Sand") %>%
  str_replace("herb", "Herb layer cover") %>%
  str_replace("bareground", "Bareground") %>%
  str_replace("vegheight", "Vegetation height") %>%
  str_replace("patchiness", "Plot heterogeneity")


# 5. plot

p_falcon_ab <- ggplot(coef_df, aes(x = Estimate, y = reorder(clean_label, Estimate))) +
  
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 width = 0,
                 color = coef_df$color,
                 linewidth = 0.8) +
  
  geom_point(color = coef_df$color,
             size = 3, stroke = 1) +
  
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_line(color = "grey85"),   # vertical gridlines ON
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 14),
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 4)),
    plot.subtitle = element_text(size = 12, margin = margin(b = 8)),
    
    axis.title.y = element_text(margin = margin(r = 10))
  ) +
  
  labs(
    title = "Model-Averaged Effects on Bee Abundance",
    subtitle = "Predictors with importance ≥ 0.8 colored in purple",
    x = "Model-averaged estimate (95% CI)",
    y = NULL
  )

p_falcon_ab

```
Sand cover received moderate support in the model set (importance = 0.82), indicating that it may contribute to explaining variation in bee abundance. Its model‑averaged effect was positive but small (estimate ≈ 0.08–0.10), and the 95% confidence interval slightly overlapped zero, suggesting substantial uncertainty in the strength of this relationship. Thus, sand appears to play a secondary role relative to patchiness and plant diversity.
Model averaging identified two strongly supported predictors of bee abundance: habitat patchiness and plant species richness (Hill q0), both with importance = 1.0 and positive model‑averaged coefficients whose confidence intervals excluded zero. Sand cover received moderate support (importance = 0.82) and showed a small positive effect, although its confidence interval slightly overlapped zero, indicating some uncertainty. In contrast, herb cover, vegetation height, and bare ground had low importance values (≤ 0.17) and model‑averaged effects near zero, suggesting weak evidence that these variables influence bee abundance. Overall, habitat heterogeneity and floristic richness emerged as the primary predictors of bee abundance.
gpt

```{r print model avg falcon plot abundance}
tiff(
  file.path("..", "output", "figures", "falcon_mavgbee_ab_600dpi.tiff"),
  width = 3990, height = 3000, res = 600, compression = "lzw"
)

print(p_falcon_ab)  
dev.off()
```
Best model bee abundance
```{r best model bee abundance}
bestmod_ab
bestmod_ab <- glmmTMB(bee_abundance ~ hill_q0 + patchiness_ord + sand + (1 | plotID),
                      family = nbinom1,
                      data = data_all_scaled)

summary(bestmod_ab)
```

Plot predictions. Backtransform x-axis plant abundance for interpretability
```{r bes model bee abundance plot predictions}
# Extract scaling attributes for hill and sand
hill_mean <- attr(data_all_scaled$hill_q0, "scaled:center")
hill_sd   <- attr(data_all_scaled$hill_q0, "scaled:scale")

sand_mean <- attr(data_all_scaled$sand, "scaled:center")
sand_sd   <- attr(data_all_scaled$sand, "scaled:scale")

# Generate predictions for each predictor
pred_hill   <- ggpredict(bestmod_ab, terms = "hill_q0")
pred_patch  <- ggpredict(bestmod_ab, terms = "patchiness_ord")
pred_sand   <- ggpredict(bestmod_ab, terms = "sand")

# Back-transform hill_q0 and sand to raw units
pred_hill$x_raw <- pred_hill$x * hill_sd + hill_mean
pred_sand$x_raw <- pred_sand$x * sand_sd + sand_mean

data_all_scaled$hill_q0_raw <- data_all_scaled$hill_q0 * hill_sd + hill_mean
data_all_scaled$sand_raw     <- data_all_scaled$sand     * sand_sd  + sand_mean

# Plot 1: Hill q0 (raw axis) + jittered raw data + predictions
p_hill <- ggplot() +
  geom_jitter(
    data = data_all_scaled,
    aes(x = hill_q0_raw, y = bee_abundance),
    width = 0.15, height = 0.05, alpha = 0.25, 
    size = 1.5, color = "grey50"
  ) +
  geom_ribbon(
    data = pred_hill,
    aes(x = x_raw, ymin = conf.low, ymax = conf.high),
    fill = "plum2", alpha = 0.35
  ) +
  geom_line(
    data = pred_hill,
    aes(x = x_raw, y = predicted, group = 1),
    color = "#7A0177", linewidth = 1.3
  ) +
  labs(
    title = "Effect of Plant Genus Richness (Hill q = 0)",
    subtitle = "Predicted bee abundance ± 95% CI with raw data",
    x = "Plant genus richness (Hill q0, raw units)",
    y = "Bee abundance"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13, color = "black"),
    plot.title = element_text(size = 16, face = "bold")
  )

# Plot 2: Patchiness (ordinal) + jittered raw data + predictions
p_patch <- ggplot() +
  geom_jitter(
    data = data_all_scaled,
    aes(x = patchiness_ord, y = bee_abundance),
    width = 0.1, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  geom_point(
    data = pred_patch,
    aes(x = x, y = predicted),
    color = "#7A0177", size = 3
  ) +
  geom_errorbar(
    data = pred_patch,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    color = "#7A0177", width = 0.1, linewidth = 1
  ) +
  labs(
    title = "Effect of Plot Heterogeneity",
    subtitle = "Predicted bee abundance ± 95% CI with raw data",
    x = "Plot heterogeneity (ordinal)",
    y = "Bee abundance"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13, color = "black"),
    plot.title = element_text(size = 16, face = "bold")
  )

# Plot 3: Sand (raw axis) + jittered raw points + predictions
p_sand <- ggplot() +
  geom_jitter(
    data = data_all_scaled,
    aes(x = sand_raw, y = bee_abundance),
    width = 0.15, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  geom_ribbon(
    data = pred_sand,
    aes(x = x_raw, ymin = conf.low, ymax = conf.high),
    fill = "plum2", alpha = 0.35
  ) +
  geom_line(
    data = pred_sand,
    aes(x = x_raw, y = predicted, group = 1),
    color = "#7A0177", linewidth = 1.3
  ) +
  labs(
    title = "Effect of Sand Cover",
    subtitle = "Predicted bee abundance ± 95% CI with raw data",
    x = "Sand cover (%)",
    y = "Bee abundance"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13, color = "black"),
    plot.title = element_text(size = 16, face = "bold")
  )

# Combine all three into a multi-panel figure

p_combined_abundance <- p_hill + p_patch + p_sand + plot_layout(ncol = 3)

p_combined_abundance
```
We performed an information‑theoretic model selection using AICc and averaged all models within ΔAICc ≤ 2 to account for model‑selection uncertainty. Model‑averaged estimates and variable importance (sum of AICc weights) are presented in the Falcon plot (Fig. X).
Two predictors received full support across the candidate model set (importance = 1.0): habitat patchiness (linear component) and plant diversity Hill q = 0. Both had positive model‑averaged coefficients whose 95% confidence intervals did not overlap zero, indicating that more heterogeneous habitat structure and higher plant species richness were consistently associated with higher bee abundance. Patchiness showed the largest effect size among all predictors.
Sand cover received intermediate support (importance = 0.78). Its conditional model‑averaged effect was moderately positive and approached statistical significance, suggesting that sand cover may contribute to explaining variation in bee abundance but with greater model uncertainty. In contrast, herb cover (importance = 0.21) and bare ground (importance = 0.16) had low importance values and model‑averaged confidence intervals overlapping zero, indicating weak evidence for any consistent directional effect.
Overall, the model‑averaged results highlight that habitat structural heterogeneity and plant species richness—rather than vegetation cover or soil surface characteristics—are the primary predictors of bee abundance at the plot level.
gpt



# Bee richness
Poisson base model does not fit
```{r bee Poisson ri model prep: overdispersion test, response data distribution}
#baseline poisson null model
m_pois <- glmmTMB(
  bee_richness ~ 1 +
    (1 | plotID),
  data = data_all_scaled,
  family = poisson
)

summary(m_pois)

sim_pois <- simulateResiduals(m_pois)
plot(sim_pois)
testDispersion(sim_pois)
testZeroInflation(sim_pois)
```

Nbinom2 base model does not fit as well as Nbinom1
```{r bee nbinom2 ri model prep: overdispersion test, response data distribution}
m_nb2 <- glmmTMB(
  bee_richness ~ 1 +
    (1 | plotID),
  data = data_all_scaled,
  family = nbinom2
)
summary(m_nb2)

sim_nb2 <- simulateResiduals(m_nb2)
plot(sim_nb2)
testDispersion(sim_nb2)
testZeroInflation(sim_nb2)
```

Nbinom1
```{r bee nbinom1 ri model prep: overdispersion test, response data distribution}
m_nb1 <- glmmTMB(
  bee_richness ~ 1 +
    (1 | plotID),
  data = data_all_scaled,
  family = nbinom1
)
summary(m_nb1)

sim_nb1 <- simulateResiduals(m_nb1)
plot(sim_nb1)
testDispersion(sim_nb1)
testZeroInflation(sim_nb1)
```
We proceed with nbinom1, as
better AIC
better dispersion
more appropriate variance structure
clean residuals
stable random effect
gpt

Compare random effects structure as in abundance models.
Given the  hierarchical structure of our data, we try this structure.

```{r bee nbinom ri model prep: overdispersion test, response data distribution}
m_nb <- glmmTMB(
  bee_richness ~ 1 +
    (1 | region) +
    (1 + visit | plotID),
  data = data_all_scaled,
  family = nbinom1
)
summary(m_nb)

sim_nb <- simulateResiduals(m_nb)
plot(sim_nb)
testDispersion(sim_nb)
testZeroInflation(sim_nb)
```
The maximal random-effects structure (1 | region) + (1 + visit | plotID) produced near-zero variance for region and a perfect correlation (±1) between random intercepts and slopes, indicating model singularity and overparameterization (Bates et al. 2015; Harrison et al. 2018). In addition, fixed-effect standard errors were undefined (NaN). We therefore simplified the random structure to a random intercept for plotID, which produced stable estimates and satisfactory diagnostic results.
gpt

```{r Bee richness global model}
RI_glmm <- glmmTMB(
  bee_richness ~
    hill_q0 + sand + bareground + herb + vegheight + patchiness_ord +
    (1 | plotID),
  family = nbinom1,
  data = data_all_scaled,
  na.action = na.fail
)

print(summary(RI_glmm))

vif(lm(bee_richness ~ hill_q0 + sand + bareground +
       herb + vegheight + patchiness_ord,
       data = data_all_scaled))
#all values below 3 - no multicollinearity 

```

```{r Bee richness dredge}
dr <- dredge(RI_glmm, rank = "AICc", trace = 2)
dr

head(dr)

# best model (to be used later for ggpred)
cand[[1]]
```

```{r model averaging richness}
cand <- get.models(dr, subset = delta <= 4)

avg_mod_ri <- model.avg(cand, revised.var = TRUE)
summary(avg_mod_ri)

# Sum of weights can be calculated like that
sw(subset(dr, delta <= 2))
```


```{r falcon plot model average - richness}
# 1. PREP COEFFICIENTS
coef_df_ri <- as.data.frame(summary(avg_mod_ri)$coefmat.full)
coef_df_ri$term <- rownames(coef_df_ri)
coef_df_ri <- coef_df_ri[coef_df_ri$term != "cond((Int))", ]

# Add CI
ci_ri <- confint(avg_mod_ri)
coef_df_ri$lower <- ci_ri[coef_df_ri$term, 1]
coef_df_ri$upper <- ci_ri[coef_df_ri$term, 2]

# 2. MAP IMPORTANCE
coef_df_ri$importance <- NA
coef_df_ri$importance[coef_df_ri$term == "cond(hill_q0)"] <- imp["cond(hill_q0)"]
coef_df_ri$importance[coef_df_ri$term == "cond(sand)"]  <- imp["cond(sand)"]
coef_df_ri$importance[coef_df_ri$term == "cond(herb)"] <- imp["cond(herb)"]
coef_df_ri$importance[coef_df_ri$term == "cond(bareground)"] <- imp["cond(bareground)"]
coef_df_ri$importance[coef_df_ri$term == "cond(vegheight)"] <- imp["cond(vegheight)"]
coef_df_ri$importance[coef_df_ri$term == "cond(patchiness_ord.L)"] <- imp["cond(patchiness_ord)"]
# Drop Q and C contrasts
coef_df_ri <- coef_df_ri[!coef_df_ri$term %in% c("cond(patchiness_ord.Q)", "cond(patchiness_ord.C)"), ]

# 3. COLOR CODE
coef_df_ri$color <- ifelse(coef_df_ri$importance >= 0.8, "#C71586", "gray40")

# 4. CLEAN LABELS
coef_df_ri$clean_label <- coef_df$term %>%
  str_replace("cond\\(", "") %>%
  str_replace("\\)", "") %>%
  str_replace("_ord.L", "") %>%
  str_replace("hill_q0", "Floral Richness") %>%
  str_replace("sand", "Sand") %>%
  str_replace("herb", "Herb layer cover") %>%
  str_replace("bareground", "Bareground") %>%
  str_replace("vegheight", "Vegetation height") %>%
  str_replace("patchiness", "Plot heterogeneity")

# 5. PLOT
p_falcon_ri <- ggplot(coef_df_ri, aes(x = Estimate, y = reorder(clean_label, Estimate))) +
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 width = 0, color = coef_df_ri$color, linewidth = 0.8) +
  geom_point(color = coef_df_ri$color, size = 3, stroke = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_line(color = "grey85"),
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 14),
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 4)),
    plot.subtitle = element_text(size = 12, margin = margin(b = 8)),
    axis.title.y = element_text(margin = margin(r = 10))
  ) +
  labs(
    title = "Model-Averaged Effects on Bee Species Richness",
    subtitle = "Predictors with importance ≥ 0.8 colored in pink",
    x = "Model-averaged estimate (95% CI)",
    y = NULL
  )

p_falcon_ri
```

```{r print model avg falcon plot richness}
tiff(
  file.path("..", "output", "figures", "falcon_mavgbee_ri_600dpi.tiff"),
  width = 3990, height = 3000, res = 600, compression = "lzw"
)

print(p_falcon_ab)  
dev.off()
```


To visualise ecological relationships, we generated predicted effects from a final GLMM that included the strongly supported predictors identified through model averaging (hill_q0 and patchiness) and the moderately supported variable sand. Predicted marginal effects were obtained using ggpredict(), holding other predictors at their mean values.
gpt

Best model bee richness
```{r best model bee richness}
bestmod_ri <- glmmTMB(bee_richness ~ hill_q0 + patchiness_ord + (1 | plotID),
                      family = nbinom1,
                      data = data_all_scaled)
```

Plot predictions. Backtransform x-axis plant richness for interpretability
```{r bes model bee richness plot predictions}
# Generate predictions
pred_hill   <- ggpredict(bestmod_ri, terms = "hill_q0")
pred_patch  <- ggpredict(bestmod_ri, terms = "patchiness_ord")

# back-transform scaled hill values
hill_mean <- attr(data_all_scaled$hill_q0, "scaled:center")
hill_sd   <- attr(data_all_scaled$hill_q0, "scaled:scale")

pred_hill$x_raw <- pred_hill$x * hill_sd + hill_mean


# Also back-transform raw jitter points for Hill q0:
data_all_scaled$hill_q0_raw <- data_all_scaled$hill_q0 * hill_sd + hill_mean


p_hill <- ggplot() +
  
  geom_jitter(
    data = data_all_scaled,
    aes(x = hill_q0_raw, y = bee_richness),
    width = 0.15, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  
  geom_ribbon(
    data = pred_hill,
    aes(x = x_raw, ymin = conf.low, ymax = conf.high),
    fill = "plum2", alpha = 0.35
  ) +
  
  geom_line(
    data = pred_hill,
    aes(x = x_raw, y = predicted, group = 1),
    color = "#7A0177", linewidth = 1.3
  ) +
  
  labs(
    title = "Effect of Flowering Plant Genus Richness (Hill q = 0)",
    subtitle = "Predicted bee richness ± 95% CI with raw data",
    x = "Plant genus richness (Hill q = 0)",
    y = "Bee richness"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13, color = "black"),
    plot.title = element_text(size = 16, face = "bold")
  )

# ============================================================
# 5. Plot 2: Patchiness (ordinal) prediction with jitter
# ============================================================

p_patch <- ggplot() +
  
  geom_jitter(
    data = data_all_scaled,
    aes(x = patchiness_ord, y = bee_richness),
    width = 0.1, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  
  geom_point(
    data = pred_patch,
    aes(x = x, y = predicted),
    color = "#7A0177", size = 3
  ) +
  
  geom_errorbar(
    data = pred_patch,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    color = "#7A0177", width = 0.1, linewidth = 1
  ) +
  
  labs(
    title = "Effect of Plot Heterogeneity",
    subtitle = "Predicted bee richness ± 95% CI with raw data",
    x = "Plot heterogeneity",
    y = "Bee richness"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13, color = "black"),
    plot.title = element_text(size = 16, face = "bold")
  )

# ============================================================
# 6. Combine into a two-panel figure
# ============================================================

p_combined <- p_hill + p_patch + plot_layout(ncol = 2)

p_combined

```

```{r print predicted best model richness}
tiff(
  file.path("..", "output", "figures", "bestmod_pred_ri_600dpi.tiff"),
  width = 7000, height = 3000, res = 600, compression = "lzw"
)

print(p_combined)  
dev.off()
```

# Bee Hill Q = 1
Build base model
```{r bee nbinom2 q1 model prep: overdispersion test, response data distribution}
m_q1 <- glmmTMB(
  bee_hill_q1 ~ 1 +
    (1 | plotID),
  data = data_all_scaled,
  family = nbinom1
)
summary(m_q1)

sim_q1 <- simulateResiduals(m_q1)
plot(sim_q1)
testDispersion(sim_q1)
testZeroInflation(sim_q1)
```