---
title: "06_site_unimodel"
author: Gregory B. Egloff
---

```{r Load packages}
library(DHARMa)
library(dplyr)
library(emmeans)
library(ggeffects)
library(ggplot2)
library(glmmTMB)
library(gt)
library(lmerTest)
library(patchwork)
library(performance)
library(tidyverse)
```

# Survey-level data (unpooled across visits)
```{r Load data}
data_all_scaled <- readRDS(file.path("..", "data", "cleaned", "data_all_scaled.rds"))
```

## Richness
```{r fit richness model on nbinom2}
nb2_rich <- glmmTMB(
  bee_richness ~ site + (1 | region) + (1 | plotID),
  family = nbinom2,
  data = data_all_scaled
)

summary(nb2_rich)

res_nb2 <- simulateResiduals(nb2_rich)
plot(res_nb2)
testDispersion(nb2_rich)
testZeroInflation(nb2_rich)

check_model(nb2_rich)
```

```{r fit richness model on nbinom1}
nb1_rich <- glmmTMB(
  bee_richness ~ site + (1 | region) + (1 | plotID),
  family = nbinom1,
  data = data_all_scaled
)

summary(nb1_rich)

res_nb1 <- simulateResiduals(nb1_rich)
plot(res_nb1)
testDispersion(nb1_rich)
testZeroInflation(nb1_rich)

check_model(nb1_rich)
```

```{r richness site model}
RI_site <- glmmTMB(
  bee_richness ~ site + (1 | region) + (1 | plotID),
  family = nbinom1,
  data = data_all_scaled
)

print(summary(RI_site))
```

```{r richness site model predicted values}
# Absolute predicted richness per site
pred_ri <- emmeans(RI_site, ~ site, type = "response")

# Absolute difference
contrast(pred_ri, "revpairwise")
```

```{r plot model predictions richness}
# Predicted marginal means for site on the response scale
pred_site_rich <- ggpredict(RI_site, terms = "site", type = "fixed")

# ensure plotting order (control, focal)
pred_site_rich$x <- factor(pred_site_rich$x, levels = c("control", "focal"))
data_all_scaled$site <- factor(data_all_scaled$site, levels = c("control", "focal"))

# Color palette
site_cols <- list(
  "focal" = "#2C7FB8",
  "control"   = "#D95F09",
  "focal_light"   = "cadetblue3",
  "control_light" = "#D95"
  )

# Plot: predicted means ± 95% CI with raw jittered points

p_rich <- ggplot() +
  # 1) Raw data (plot × visit), jittered in background
  geom_jitter(
    data = data_all_scaled,
    aes(x = site, y = bee_richness, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +

  # 2) Model-predicted means (ggeffects output; response scale)
  geom_point(
    data = pred_site_rich,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +

  # 3) 95% confidence intervals
  geom_errorbar(
    data = pred_site_rich,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +

  # Use identical site colours for both layers
  scale_colour_manual(values = site_cols) +
  
  scale_x_discrete(labels = c("control" = "Control", "focal" = "Focal")) +

  # Labels
  labs(
    x = NULL,
    y = "Bee richness",
    title = "Predicted Bee Richness per Plot by Site",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = raw observations"
  ) +
  
  # Theme
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  )

p_rich
```

## Abundance
```{r fit abundance model on nbinom2}
nb2_abun <- glmmTMB(
  bee_abundance ~ site + (1 | region) + (1 | plotID),
  family = nbinom2,
  data = data_all_scaled
)

summary(nb2_abun)

res_nb2 <- simulateResiduals(nb2_abun)
plot(res_nb2)
testDispersion(nb2_abun)
testZeroInflation(nb2_abun)

check_model(nb2_abun)
```

```{r fit abundance model on nbinom1}
nb1_abun <- glmmTMB(
  bee_abundance ~ site + (1 | region) + (1 | plotID),
  family = nbinom1,
  data = data_all_scaled
)

summary(nb1_abun)

res_nb1 <- simulateResiduals(nb1_abun)
plot(res_nb1)
testDispersion(nb1_abun)
testZeroInflation(nb1_abun)

check_model(nb1_abun)
```

```{r abundance site model}
AB_site <- glmmTMB(
  bee_abundance ~ site + (1 | region) + (1 | plotID),
  family = nbinom1,
  data = data_all_scaled
)

print(summary(AB_site))
```

```{r abundance site model predicted values}
# Absolute predicted richness per site
pred_ab <- emmeans(AB_site, ~ site, type = "response")

# Absolute difference
contrast(pred_ab, "revpairwise")
```

```{r plot model predictions abundance}
# Predicted marginal means for site on the response scale
pred_site_abun <- ggpredict(AB_site, terms = "site", type = "fixed")

# ensure plotting order (control, focal)
pred_site_abun$x <- factor(pred_site_abun$x, levels = c("control", "focal"))
data_all_scaled$site <- factor(data_all_scaled$site, levels = c("control", "focal"))

# Color palette
site_cols <- list(
  "focal" = "#2C7FB8",
  "control"   = "#D95F09",
  "focal_light"   = "cadetblue3",
  "control_light" = "#D95"
  )

# Plot predicted means ± 95% CI with raw jittered points

p_abun <- ggplot() +
  # 1) Raw data (plot × visit), jittered in background
  geom_jitter(
    data = data_all_scaled,
    aes(x = site, y = bee_abundance, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +

  # 2) Model-predicted means (ggeffects output; response scale)
  geom_point(
    data = pred_site_abun,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +

  # 3) 95% confidence intervals
  geom_errorbar(
    data = pred_site_abun,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +

  # Use identical site colours for both layers
  scale_colour_manual(values = site_cols) +
  
  scale_x_discrete(labels = c("control" = "Control", "focal" = "Focal")) +

  # Labels
  labs(
    x = NULL,
    y = "Bee abundance",
    title = "Predicted Bee Abundance per Plot by Site",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = raw observations"
  ) +

  # Theme
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  )

p_abun
```

## Hill q1
```{r Hillq1 site model}
Q1_site <- glmmTMB(log(bee_hill_q1) ~ site + (1 | region) + (1  | plotID),
        family = gaussian(),
        data = data_all_scaled)

summary(Q1_site)

# res_Q1 <- simulateResiduals(Q1_site) takes too long
check_model(Q1_site)
```

```{r Hillq1 site model predicted values}
pred_q1 <- emmeans(Q1_site, ~ site, type = "response")
contrast(pred_q1, "revpairwise")
```

```{r plot model predictions hillq1}
# Predicted marginal means - check scale first then back-transform if needed
pred_site_q1 <- ggpredict(Q1_site, terms = "site", type = "fixed")

# Ensure plotting order
pred_site_q1$x       <- factor(pred_site_q1$x,      levels = c("control", "focal"))
data_all_scaled$site <- factor(data_all_scaled$site, levels = c("control", "focal"))

# Colors
site_cols <- list(
  "focal"         = "#2C7FB8",
  "control"       = "#D95F09",
  "focal_light"   = "cadetblue3",
  "control_light" = "#D95"
)

# Plot
p_q1 <- ggplot() +
  # 1) Raw data on original scale
  geom_jitter(
    data = data_all_scaled,
    aes(x = site, y = bee_hill_q1, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +
  # 2) Model-predicted means (back-transformed)
  geom_point(
    data = pred_site_q1,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +
  # 3) 95% confidence intervals (back-transformed)
  geom_errorbar(
    data = pred_site_q1,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +
  scale_colour_manual(values = site_cols) +
  scale_x_discrete(labels = c("control" = "Control", "focal" = "Focal")) +
  labs(
    x = NULL,
    y = "Bee Shannon diversity",
    title = "Predicted Bee Shannon Diversity per Plot by Site",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = raw observations"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  ) +
  scale_y_continuous(
    breaks = scales::pretty_breaks(4),
    labels = scales::label_number(accuracy = 1)
  )

p_q1
```


### Model predictions patchwork plot
```{r Model predictions patchwork plot}
# Shared publication theme
theme_pub <- theme_minimal(base_size = 16) +
  theme(
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 18),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)
  )


# Strip individual titles and subtitles, add theme sizes
p_abun_clean <- p_abun +
  labs(title = NULL, subtitle = NULL) +
  theme_pub

p_rich_clean <- p_rich +
  labs(title = NULL, subtitle = NULL) +
  theme_pub

p_q1_clean <- p_q1 +
  labs(title = NULL, subtitle = NULL) +
  theme_pub

# Combine into one row
p_combined <- p_abun_clean + p_rich_clean + p_q1_clean +
  plot_layout(ncol = 3) +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")",
    )

p_combined
```

```{r print prediction patchwork plot}
# save tiff
tiff(
  file.path("..", "output", "figures", "preds_bee_raw_site_600dpi.tiff"),
  width = 7500, height = 2500, res = 600, compression = "lzw"
)
print(p_combined)
dev.off()
```

```{r print prediction patchwork plot}
# save tiff
tiff(
  file.path("..", "output", "figures", "Figure_E1.tiff"),
  width = 7500, height = 2500, res = 600, compression = "lzw"
)
print(p_combined)
dev.off()
```

### Model output tables
```{r model output table univariate models with R2 values}
extract_site_table <- function(model, response_name) {
  
  em <- emmeans(model, ~ site, type = "response")
  em_df <- as.data.frame(em)
  
  em_df <- em_df %>%
    rename(
      lower = any_of(c("asymp.LCL", "lower.CL")),
      upper = any_of(c("asymp.UCL", "upper.CL"))
    )
  
  means_wide <- em_df %>%
    mutate(site = as.character(site)) %>%
    transmute(
      site,
      mean = response,
      lower = lower,
      upper = upper
    ) %>%
    mutate(
      mean_ci = sprintf(
        "%.2f (%.2f–%.2f)",
        mean,
        lower,
        upper
      )
    ) %>%
    select(site, mean_ci) %>%
    pivot_wider(
      names_from = site,
      values_from = mean_ci
    )
  
  contr <- contrast(em, "revpairwise") %>%
    summary(infer = TRUE, type = "response") %>%
    as.data.frame() %>%
    rename(
      lower = any_of(c("asymp.LCL", "lower.CL")),
      upper = any_of(c("asymp.UCL", "upper.CL"))
    )
  
  contr_out <- contr %>%
    transmute(
      effect = sprintf(
        "%.2f× (%.2f–%.2f)",
        ratio,
        lower,
        upper
      ),
      p_value = ifelse(
        p.value < 0.001,
        "< 0.001",
        sprintf("%.3f", p.value)
      )
    )
  
  # R²
  r2_vals <- performance::r2(model)
  
  r2_marg <- ifelse(
    is.null(r2_vals$R2_marginal),
    NA,
    sprintf("%.3f", r2_vals$R2_marginal)
  )
  
  r2_cond <- ifelse(
    is.null(r2_vals$R2_conditional),
    NA,
    sprintf("%.3f", r2_vals$R2_conditional)
  )
  
  bind_cols(
    means_wide,
    contr_out
  ) %>%
    mutate(
      response = sprintf(
        "%s (R²m = %s, R²c = %s)",
        response_name,
        r2_marg,
        r2_cond
      ),
      .before = everything()
    )
}

tab_abun <- extract_site_table(AB_site, "Bee abundance")
tab_rich <- extract_site_table(RI_site, "Bee richness")
tab_q1   <- extract_site_table(Q1_site, "Bee Shannon diversity")

table_univsite <- bind_rows(tab_abun, tab_rich, tab_q1)
table_univsite
```

```{r save univsite model output table}
gt_table_univsite <- table_univsite %>%
  gt() %>%
  cols_label(
    response = "Response",
    control  = "Control \n(mean ± 95% CI)",
    focal    = "Focal \n(mean ± 95% CI)",
    effect = "Focal vs Control \n(ratio ± 95% CI)",
    p_value = "P-value"
  )

gtsave(gt_table_univsite, filename = "..", "output", "tables", "site_univ_effects.docx")
```

```{r save univsite model output table}
gt_table_univsite <- table_univsite %>%
  gt() %>%
  cols_label(
    response = "Response",
    control  = "Control \n(mean ± 95% CI)",
    focal    = "Focal \n(mean ± 95% CI)",
    effect = "Focal vs Control \n(ratio ± 95% CI)",
    p_value = "P-value"
  )

gtsave(gt_table_univsite, filename = "..", "output", "tables", "Table_E1.docx")
```


## Survey-level site and visit interactive
Run models for all metrics
```{r fit models}
# Abundance
AB_sitevisit <- glmmTMB(
  bee_abundance ~ site * as.factor(visit) + (1 | region) + (1 + visit | plotID),
  family = nbinom1,
  data = data_all_scaled
)

# Richness
RI_sitevisit <- glmmTMB(
  bee_richness ~ site * as.factor(visit) + (1 | region) + (1 + visit | plotID),
  family = nbinom1,
  data = data_all_scaled
)

# Hill q1
Q1_sitevisit <- glmmTMB(
  log(bee_hill_q1) ~ site * as.factor(visit) + (1 | region) + (1 + visit | plotID),
  family = gaussian(),
  data = data_all_scaled
)

summary(AB_sitevisit)
summary(RI_sitevisit)
summary(Q1_sitevisit)

check_model(AB_sitevisit)
check_model(RI_sitevisit)
check_model(Q1_sitevisit)
```

### Model predictions plot
```{r plot predictions}
plot_sitevisit <- function(model, response_var, y_label, back_transform_log = FALSE) {
  
  preds <- ggpredict(model, terms = c("visit", "site"))
  
  p <- ggplot(preds, aes(x = x, y = predicted, color = group, group = group)) +
    geom_jitter(
      data = data_all_scaled,
      aes(x = as.numeric(as.factor(visit)), y = .data[[response_var]], color = site),
      position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.3),
      alpha = 0.2, size = 1.2, inherit.aes = FALSE
    ) +
    geom_errorbar(aes(ymin = conf.low, ymax = conf.high), 
                  width = 0.15, position = position_dodge(0.3), linewidth = 0.6) +
    geom_line(position = position_dodge(0.3), linewidth = 0.7, linetype = "dashed", alpha = 0.6) +
    geom_point(position = position_dodge(0.3), size = 3) +
    scale_color_manual(values = c("control" = "#D95F09", "focal" = "#2C7FB8"),
                        labels = c("Control", "Focal")) +
    labs(x = "Survey visit", y = y_label, color = NULL) +
    theme_classic(base_size = 13) +
    theme(
      legend.position = c(0.15, 0.85),
      legend.background = element_rect(fill = "white", color = NA),
      axis.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
  
  return(p)
}

p_ab_sv <- plot_sitevisit(AB_sitevisit, "bee_abundance", "Bee abundance")
p_ri_sv <- plot_sitevisit(RI_sitevisit, "bee_richness", "Bee richness")
p_q1_sv <- plot_sitevisit(Q1_sitevisit, "bee_hill_q1", "Bee Shannon diversity")

p_ab_sv
p_ri_sv
p_q1_sv

theme_pub <- theme_minimal(base_size = 16) +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)
  )

p_combined_sv <- (
  (p_ab_sv + theme_pub + theme(axis.title.x = element_blank())) +
  (p_ri_sv + theme_pub + theme(axis.title.x = element_blank())) +
  (p_q1_sv + theme_pub)
) +
  plot_layout(
    nrow = 3,
    guides = "collect"
  ) +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")"
  ) &
  theme(
    legend.position = "bottom"
  )

p_combined_sv
```

```{r export figure}
# save tiff
tiff(
  file.path("..", "output", "figures", "sitevisit_600dpi.tiff"),
  width = 3000, height = 6000, res = 600, compression = "lzw"
)
print(p_combined_sv)
dev.off()
```

```{r export figure}
# save tiff
tiff(
  file.path("..", "output", "figures", "Figure_H1.tiff"),
  width = 3000, height = 6000, res = 600, compression = "lzw"
)
print(p_combined_sv)
dev.off()
```

### Model output table
```{r}
extract_sitevisit_table <- function(model, response_name) {
  
  # R² values
  r2_vals <- performance::r2(model)
  
  response_label <- sprintf(
    "%s (R²m = %.3f, R²c = %.3f)",
    response_name,
    r2_vals$R2_marginal,
    r2_vals$R2_conditional
  )
  
  # Estimated marginal means
  em <- emmeans(
    model,
    ~ site | visit,
    type = "response"
  )
  
  em_df <- as.data.frame(em)
  
  mean_col <- if ("response" %in% names(em_df)) "response" else "emmean"
  
  lower_col <- if ("asymp.LCL" %in% names(em_df)) {
    "asymp.LCL"
  } else {
    "lower.CL"
  }
  
  upper_col <- if ("asymp.UCL" %in% names(em_df)) {
    "asymp.UCL"
  } else {
    "upper.CL"
  }
  
  means_tbl <- em_df %>%
    mutate(
      visit = as.character(visit),
      mean_ci = sprintf(
        "%.2f (%.2f–%.2f)",
        .data[[mean_col]],
        .data[[lower_col]],
        .data[[upper_col]]
      )
    ) %>%
    select(
      visit,
      site,
      mean_ci
    ) %>%
    pivot_wider(
      names_from = site,
      values_from = mean_ci
    )
  
  # Contrasts
  contr <- contrast(
    em,
    "revpairwise",
    by = "visit"
  ) %>%
    summary(
      infer = TRUE,
      type = "response"
    ) %>%
    as.data.frame() %>%
    rename(
      lower = any_of(c("asymp.LCL", "lower.CL")),
      upper = any_of(c("asymp.UCL", "upper.CL"))
    )
  
  effect_col <- intersect(
    c("ratio", "estimate"),
    names(contr)
  )
  
  contrast_tbl <- contr %>%
    transmute(
      visit = as.character(visit),
      Effect = sprintf(
        "%.2f (%.2f–%.2f)",
        .data[[effect_col]],
        lower,
        upper
      ),
      `P-value` = ifelse(
        p.value < 0.001,
        "<0.001",
        sprintf("%.3f", p.value)
      )
    )
  
  left_join(
    means_tbl,
    contrast_tbl,
    by = "visit"
  ) %>%
    mutate(
      Response = response_label
    )
}

# Extract tables
tab_ab_sv <- extract_sitevisit_table(
  AB_sitevisit,
  "Bee abundance"
)

tab_ri_sv <- extract_sitevisit_table(
  RI_sitevisit,
  "Bee richness"
)

tab_q1_sv <- extract_sitevisit_table(
  Q1_sitevisit,
  "Bee Shannon diversity"
)

# Combine
table_sitevisit <- bind_rows(
  tab_ab_sv,
  tab_ri_sv,
  tab_q1_sv
)

# Helper function for section headers
add_section <- function(df, name) {
  
  header <- tibble(
    Response = name,
    visit = "",
    control = NA_character_,
    focal = NA_character_,
    Effect = NA_character_,
    `P-value` = NA_character_
  )
  
  bind_rows(
    header,
    df %>%
      select(
        visit,
        control,
        focal,
        Effect,
        `P-value`
      )
  )
}

# Build final table
table_sitevisit_final <- bind_rows(
  
  add_section(
    filter(
      table_sitevisit,
      grepl("^Bee abundance", Response)
    ),
    unique(
      table_sitevisit$Response[
        grepl("^Bee abundance", table_sitevisit$Response)
      ]
    )
  ),
  
  add_section(
    filter(
      table_sitevisit,
      grepl("^Bee richness", Response)
    ),
    unique(
      table_sitevisit$Response[
        grepl("^Bee richness", table_sitevisit$Response)
      ]
    )
  ),
  
  add_section(
    filter(
      table_sitevisit,
      grepl("^Bee Shannon diversity", Response)
    ),
    unique(
      table_sitevisit$Response[
        grepl("^Bee Shannon diversity", table_sitevisit$Response)
      ]
    )
  )
)

# GT table
gt_table_sitevisit <- table_sitevisit_final %>%
  gt() %>%
  cols_label(
    Response = "Response",
    visit = "Survey visit",
    control = "Control (mean ± 95% CI)",
    focal = "Focal (mean ± 95% CI)",
    Effect = "Focal vs Control",
    `P-value` = "P-value"
  ) %>%
  cols_align(
    align = "left",
    columns = Response
  ) %>%
  cols_align(
    align = "center",
    columns = c(
      visit,
      control,
      focal,
      Effect,
      `P-value`
    )
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      rows = is.na(control)
    )
  ) %>%
  fmt_missing(
    everything(),
    missing_text = ""
  )

gt_table_sitevisit
```

```{r export table }
# Export
gtsave(
  gt_table_sitevisit,
  filename = "..", "output", "tables", "sitevisit_effects.docx"
)
```

```{r export table }
# Export
gtsave(
  gt_table_sitevisit,
  filename = "..", "output", "tables", "Table_H1.docx"
)
```


# Rarefied data - pooled across visits
Load rarefied data
```{r load data}
data_model_cov <- readRDS(file.path(
  "..", "data", "cleaned", "data_model_cov.rds"))

data_model_size <- readRDS(file.path(
  "..", "data", "cleaned", "data_model_size.rds"))
```


# Coverage-rarefied models
## Richness
```{r Coverage-based rarefaction model}
RRIcov_site <- glmmTMB(
  richness_rarefied_cov ~ site + (1 | region),
  data = data_model_cov,
  family = Gamma(link = "log")
)

summary(RRIcov_site)

check_model(RRIcov_site)

# Predicted values
pred_cov_rich <- emmeans(RRIcov_site, ~ site, type = "response")
contrast(pred_cov_rich, "revpairwise")
r2(RRIcov_site)
```

```{r plot predicted coverage-based rarefaction}
# Predicted marginal means
pred_site_rrich_cov <- ggpredict(RRIcov_site, terms = "site", type = "fixed")

# ensure plotting order
pred_site_rrich_cov$x <- factor(pred_site_rrich_cov$x, levels = c("control", "focal"))
data_model_cov$site <- factor(data_model_cov$site, levels = c("control", "focal"))

# Color palette
site_cols <- list(
  "focal" = "#2C7FB8",
  "control" = "#D95F09",
  "focal_light" = "cadetblue3",
  "control_light" = "#D95"
)

# Plot
p_rrich_cov <- ggplot() +
  geom_jitter(
    data = data_model_cov,
    aes(x = site, y = richness_rarefied_cov, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +
  geom_point(
    data = pred_site_rrich_cov,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +
  geom_errorbar(
    data = pred_site_rrich_cov,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +
  scale_colour_manual(values = site_cols) +
  scale_x_discrete(labels = c("control" = "Control", "focal" = "Focal")) +
  labs(
    x = NULL,
    y = "Bee richness \n (coverage-rarefied)",
    title = "Predicted Bee Richness per Plot by Site (coverage-based rarefaction)",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = standardised richness at equal coverage"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  )

p_rrich_cov
```

## Hill q1
```{r Coverage-based rarefaction model (Hill q1)}
RQ1cov_site <- glmmTMB(
  hill_q1_rarefied_cov ~ site + (1 | region),
  data = data_model_cov,
  family = gaussian()
)

summary(RQ1cov_site)

check_model(RQ1cov_site)

# Predicted values
pred_cov_q1 <- emmeans(RQ1cov_site, ~ site, type = "response")
contrast(pred_cov_q1, "revpairwise")
```

```{r Plot predicted Hill q1 (coverage-based rarefaction)}
# Predicted marginal means
pred_site_q1_cov <- ggpredict(RQ1cov_site, terms = "site", type = "fixed")

# ensure plotting order
pred_site_q1_cov$x <- factor(pred_site_q1_cov$x, levels = c("control", "focal"))
data_model_cov$site <- factor(data_model_cov$site, levels = c("control", "focal"))

# Color palette
site_cols <- list(
  "focal" = "#2C7FB8",
  "control" = "#D95F09",
  "focal_light" = "cadetblue3",
  "control_light" = "#D95"
)

# Plot
p_q1_cov <- ggplot() +
  geom_jitter(
    data = data_model_cov,
    aes(x = site, y = hill_q1_rarefied_cov, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +
  geom_point(
    data = pred_site_q1_cov,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +
  geom_errorbar(
    data = pred_site_q1_cov,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +
  scale_colour_manual(values = site_cols) +
  scale_x_discrete(labels = c("control" = "Control", "focal" = "Focal")) +
  labs(
    x = NULL,
    y = "Bee Shannon diversity \n (coverage-rarefied)",
    title = "Predicted Bee Diversity per Plot by Site (coverage-based rarefaction, q = 1)",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = standardised Hill q1 at equal coverage"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  )

p_q1_cov
```


## Model predictions patchwork plot
```{r model predictions patchwork plor coverage based rarefaction}
# Shared publication theme
theme_pub <- theme_minimal(base_size = 16) +
  theme(
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 18),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)
  )
# clean plots
p_rrich_cov_clean <- p_rrich_cov +
  labs(title = NULL, subtitle = NULL) +
  theme_pub

p_q1_cov_clean <- p_q1_cov +
  labs(title = NULL, subtitle = NULL) +
  theme_pub

# Combine plots
p_cov_combined <- p_rrich_cov_clean + p_q1_cov_clean +
  patchwork::plot_layout(ncol = 2) +
  patchwork::plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")"
  )

p_cov_combined
```

```{r export plot}
# save tiff
tiff(
  file.path("..", "output", "figures", "preds_bee_covgrar_600dpi.tiff"),
  width = 5000, height = 2500, res = 600, compression = "lzw"
)

print(p_cov_combined)

dev.off()
```

```{r export plot}
tiff(
  file.path("..", "output", "figures", "Figure_4.tiff"),
  width = 5000, height = 2500, res = 600, compression = "lzw"
)

print(p_cov_combined)

dev.off()
```


## Model outputs table
```{r Model ouputs table coverage based rarefaction}
# helper function
extract_site_table_rarefied <- function(model, response_name, effect_type) {
  
  em <- emmeans(model, ~ site, type = "response")
  em_df <- as.data.frame(em)
  
  mean_col  <- if ("response" %in% names(em_df)) "response" else "emmean"
  lower_col <- if ("asymp.LCL" %in% names(em_df)) "asymp.LCL" else "lower.CL"
  upper_col <- if ("asymp.UCL" %in% names(em_df)) "asymp.UCL" else "upper.CL"
  
  means_wide <- em_df |>
    mutate(site = as.character(site)) |>
    transmute(
      site,
      mean = .data[[mean_col]],
      lower = .data[[lower_col]],
      upper = .data[[upper_col]]
    ) |>
    mutate(mean_ci = sprintf("%.2f (%.2f\u2013%.2f)", mean, lower, upper)) |>
    select(site, mean_ci) |>
    pivot_wider(names_from = site, values_from = mean_ci)
  
  # raw means (unformatted), needed to compute % change for additive-scale models
  control_mean <- em_df[[mean_col]][em_df$site == "control"]
  focal_mean   <- em_df[[mean_col]][em_df$site == "focal"]
  pct_change   <- (focal_mean - control_mean) / control_mean * 100
  
  contr <- contrast(em, "revpairwise") |>
    summary(infer = TRUE, type = "response") |>
    as.data.frame() |>
    rename(
      lower = any_of(c("asymp.LCL", "lower.CL")),
      upper = any_of(c("asymp.UCL", "upper.CL"))
    )
  
  if (effect_type == "ratio") {
    contr_out <- contr |>
      transmute(
        effect = sprintf("%.2f\u00d7 (%.2f\u2013%.2f)", ratio, lower, upper),
        p_value = ifelse(p.value < 0.001, "< 0.001", sprintf("%.3f", p.value))
      )
  } else {
    contr_out <- contr |>
      transmute(
        effect = sprintf("+%.2f (%.2f\u2013%.2f); +%.0f%%", estimate, lower, upper, pct_change),
        p_value = ifelse(p.value < 0.001, "< 0.001", sprintf("%.3f", p.value))
      )
  }
  
  # R2 extraction
  r2_vals <- tryCatch(r2(model), error = function(e) NULL)
  r2_marg <- if (!is.null(r2_vals$R2_marginal)) sprintf("%.3f", r2_vals$R2_marginal) else NA
  r2_cond <- if (!is.null(r2_vals$R2_conditional)) sprintf("%.3f", r2_vals$R2_conditional) else NA
  
  bind_cols(means_wide, contr_out) |>
    mutate(
      response = sprintf("%s (R\u00b2m = %s, R\u00b2c = %s)", response_name, r2_marg, r2_cond),
      .before = everything()
    )
}

tab_cov_rich <- extract_site_table_rarefied(RRIcov_site, "Bee richness \n(coverage-rarefied)", "ratio")
tab_cov_q1   <- extract_site_table_rarefied(RQ1cov_site, "Bee Shannon diversity \n(coverage-rarefied)", "additive")

table_cov_rarefied <- bind_rows(tab_cov_rich, tab_cov_q1)

gt_table_cov <- table_cov_rarefied |>
  gt() |>
  cols_label(
    response = "Response",
    control  = "Control (mean \u00b1 95% CI)",
    focal    = "Focal (mean \u00b1 95% CI)",
    effect   = "Focal vs Control Difference (\u00b1 95% CI)",
    p_value  = "P-value"
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  )

gt_table_cov
```

```{r export table}
gtsave(
  gt_table_cov,
  filename = "..", "output", "tables", "cov_rarefaction_effects.docx"
)
```

```{r export table}
gtsave(
  gt_table_cov,
  filename = "..", "output", "tables", "Table_1.docx"
)
```


# Size-rarefied models
## Richness
```{r size-based rarefaction model}
RRIsiz_site <- glmmTMB(
  richness_rarefied_size ~ site + (1 | region),
  data = data_model_size,
  family = gaussian()
)

summary(RRIsiz_site)


check_model(RRIsiz_site)

# Predicted values
pred_siz_rich <- emmeans(RRIsiz_site, ~ site, type = "response")
contrast(pred_siz_rich, "revpairwise")
```

```{r plot predicted size-based rarefaction }
# Predicted marginal means
pred_site_rrich_size <- ggpredict(RRIsiz_site, terms = "site", type = "fixed")

# ensure plotting order
pred_site_rrich_size$x <- factor(pred_site_rrich_size$x, levels = c("control", "focal"))
data_model_size$site <- factor(data_model_size$site, levels = c("control", "focal"))

# Color palette
site_cols <- list(
  "focal" = "#2C7FB8",
  "control" = "#D95F09",
  "focal_light" = "cadetblue3",
  "control_light" = "#D95"
)

# Plot
p_rrich_size <- ggplot() +
  geom_jitter(
    data = data_model_size,
    aes(x = site, y = richness_rarefied_size, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +
  geom_point(
    data = pred_site_rrich_size,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +
  geom_errorbar(
    data = pred_site_rrich_size,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +
  scale_colour_manual(values = site_cols) +
  scale_x_discrete(labels = c("control" = "Control", "focal" = "Focal")) +
  labs(
    x = NULL,
    y = "Bee richness \n (size-standardised)",
    title = "Predicted Bee Richness per Plot by Site (size-based rarefaction)",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = rarefied plot values"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  )

p_rrich_size
```


## Hill q1
```{r Size-based rarefaction model (Hill q1)}
RRIsiz_q1_site <- glmmTMB(
  hill_q1_rarefied_size ~ site + (1 | region),
  data = data_model_size,
  family = gaussian()
)

summary(RRIsiz_q1_site)

check_model(RRIsiz_q1_site)

# Predicted values
pred_siz_q1 <- emmeans(RRIsiz_q1_site, ~ site, type = "response")
contrast(pred_siz_rich, "revpairwise")
```

```{r Plot predicted size-based rarefaction (Hill q1)}
# Predicted marginal means
pred_site_q1_size <- ggpredict(RRIsiz_q1_site, terms = "site", type = "fixed")

# ensure plotting order
pred_site_q1_size$x <- factor(pred_site_q1_size$x, levels = c("control", "focal"))
data_model_size$site <- factor(data_model_size$site, levels = c("control", "focal"))

# Plot
p_q1_size <- ggplot() +
  geom_jitter(
    data = data_model_size,
    aes(x = site, y = hill_q1_rarefied_size, colour = site),
    width = 0.12, alpha = 0.15, size = 1.6, show.legend = FALSE
  ) +
  geom_point(
    data = pred_site_q1_size,
    aes(x = x, y = predicted, colour = x),
    size = 3.8, show.legend = FALSE
  ) +
  geom_errorbar(
    data = pred_site_q1_size,
    aes(x = x, ymin = conf.low, ymax = conf.high, colour = x),
    width = 0.10, linewidth = 1, show.legend = FALSE
  ) +
  scale_colour_manual(values = site_cols) +
  scale_x_discrete(labels = c("control" = "Control", "focal" = "Focal")) +
  labs(
    x = NULL,
    y = "Bee Shannon diversity \n (size-standardised)",
    title = "Predicted Bee Diversity per Plot by Site (size-based rarefaction)",
    subtitle = "Points = model-estimated means ± 95% CI \nJittered points = rarefied plot values"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    axis.title.y = element_text(margin = margin(r = 8))
  )

p_q1_size
```


## Model predictions patchwork plot
```{r model predictions patchwork plos size based rarefaction}
# Shared publication theme
theme_pub <- theme_minimal(base_size = 16) +
  theme(
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 18),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)
  )

# clean plots
p_rrich_size_clean <- p_rrich_size +
  labs(title = NULL, subtitle = NULL) +
  theme_pub

p_q1_size_clean <- p_q1_size +
  labs(title = NULL, subtitle = NULL) +
  theme_pub

# combine plots
p_size_combined <- p_rrich_size_clean + p_q1_size_clean +
  patchwork::plot_layout(ncol = 2) +
  patchwork::plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")"
  )

p_size_combined
```

```{r export figure}
# save tiff
tiff(
  file.path("..", "output", "figures", "preds_bee_sizerar_600dpi.tiff"),
  width = 5000, height = 2500, res = 600, compression = "lzw"
)
print(p_size_combined)
dev.off()
```

```{r export figure}
tiff(
  file.path("..", "output", "figures", "Figure_D1.tiff"),
  width = 5000, height = 2500, res = 600, compression = "lzw"
)
print(p_size_combined)
dev.off()
```


## Model outputs table
```{r Model outputs table size-based rarefation}
# helper function
extract_site_table_rarefied <- function(model, response_name, effect_type) {
  
  em <- emmeans(model, ~ site)
  em_df <- as.data.frame(em)
  
  lower_col <- if ("asymp.LCL" %in% names(em_df)) "asymp.LCL" else "lower.CL"
  upper_col <- if ("asymp.UCL" %in% names(em_df)) "asymp.UCL" else "upper.CL"
  
  means_wide <- em_df |>
    mutate(site = as.character(site)) |>
    transmute(
      site,
      mean = emmean,
      lower = .data[[lower_col]],
      upper = .data[[upper_col]]
    ) |>
    mutate(
      mean_ci = sprintf(
        "%.2f (%.2f–%.2f)",
        mean,
        lower,
        upper
      )
    ) |>
    select(site, mean_ci) |>
    pivot_wider(
      names_from = site,
      values_from = mean_ci
    )
  
  # percentage increase
  control_mean <- em_df$emmean[em_df$site == "control"]
  focal_mean   <- em_df$emmean[em_df$site == "focal"]
  pct_change   <- (focal_mean - control_mean) / control_mean * 100
  
  if (effect_type == "ratio") {
    
    contr <- contrast(
      em,
      "revpairwise"
    ) |>
      summary(
        infer = TRUE,
        type = "response"
      ) |>
      as.data.frame()
    
    contr <- contr |>
      rename(
        lower = any_of(c("asymp.LCL", "lower.CL")),
        upper = any_of(c("asymp.UCL", "upper.CL"))
      )
    
    contr_out <- contr |>
      transmute(
        effect = sprintf(
          "%.2f× (%.2f–%.2f)",
          ratio,
          lower,
          upper
        ),
        p_value = ifelse(
          p.value < 0.001,
          "< 0.001",
          sprintf("%.3f", p.value)
        )
      )
    
  } else {
    
    contr <- pairs(em) |>
      summary(infer = TRUE) |>
      as.data.frame()
    
    contr_out <- contr |>
      transmute(
        effect = sprintf(
          "+%.2f (%.2f–%.2f); +%.0f%%",
          -estimate,
          -upper.CL,
          -lower.CL,
          pct_change
        ),
        p_value = ifelse(
          p.value < 0.001,
          "< 0.001",
          sprintf("%.3f", p.value)
        )
      )
  }
  
  # R²
  r2_vals <- tryCatch(
    performance::r2(model),
    error = function(e) NULL
  )
  
  r2_marg <- if (!is.null(r2_vals$R2_marginal))
    sprintf("%.3f", r2_vals$R2_marginal)
  else
    NA
  
  r2_cond <- if (!is.null(r2_vals$R2_conditional))
    sprintf("%.3f", r2_vals$R2_conditional)
  else
    NA
  
  bind_cols(
    means_wide,
    contr_out
  ) |>
    mutate(
      response = sprintf(
        "%s (R²m = %s, R²c = %s)",
        response_name,
        r2_marg,
        r2_cond
      ),
      .before = everything()
    )
}

tab_size_rich <- extract_site_table_rarefied(RRIsiz_site, "Bee richness \n(size-rarefied)", "absolute")
tab_size_q1   <- extract_site_table_rarefied(RRIsiz_q1_site, "Bee Shannon diversity \n(size-rarefied)", "absolute")

table_size_rarefied <- bind_rows(tab_size_rich, tab_size_q1)

gt_table_size <- table_size_rarefied |>
  gt() |>
  cols_label(
    response = "Response",
    control  = "Control (mean ± 95% CI)",
    focal    = "Focal (mean ± 95% CI)",
    effect   = "Focal vs Control Difference (± 95% CI)",
    p_value  = "P-value"
  ) |>
  cols_align(
    align = "center",
    everything()
  )

gt_table_size
```

```{r export table}
gtsave(
  gt_table_size,
  filename = "..", "output", "tables", "size_rarefaction_effects.docx"
)
```

```{r export table}
gtsave(
  gt_table_size,
  filename = "..", "output", "tables", "Table_D1.docx"
)
```

