---
title: "06_habvariables_multimodel"
author: Gregory B. Egloff
---

```{r Load packages}
library(tidyverse)
library(ggeffects)
library(glmmTMB)
library(gt)
library(DHARMa)
library(dplyr)
library(patchwork)
library(performance)
library(MuMIn)
library(corrplot)
```

```{r Load data}
data_all_scaled <- readRDS(file.path("..", "data", "cleaned", "data_all_scaled.rds"))
```

# Unpooled - each visit separate per plot
Collinearity tests bee metrics
```{r}
# List of numeric environmental variables
bee_vars <- c("bee_abundance", "bee_richness", "bee_hill_q1", "bee_hill_q2")
#visit and patchiness, although included in coming models as ordinal, are treated as numeric for Spearman computation

# Compute Spearman correlation matrix
cor_matrix <- cor(
  data_all_scaled[bee_vars],
  method = "spearman",
  use = "pairwise.complete.obs")

cor_matrix
```
Hill q2 therefore not included in MS

Collinearity tests environmental variables
```{r collinearity test}
# Convert ordinal patchiness to numeric (consistent with plot-level approach)
data_all_scaled <- data_all_scaled |>
  mutate(patchiness_num = as.numeric(patchiness_ord))

# List of numeric environmental variables
env_vars <- c("bareground", "vegcover", "vegheight", "herb", "woody",
              "sand", "gravel", "cobble", "stone", "soil", "flat",
              "sloped", "steep", "canopy", "branchpile",
              "stonepile", "standingdeadwood", "lyingdeadwood",
              "hill_q0", "patchiness_num")
# visit and patchiness, although included in coming models as ordinal, are treated as numeric for Spearman computation

# Compute Spearman correlation matrix
cor_matrix <- cor(
  data_all_scaled[env_vars],
  method = "spearman",
  use = "pairwise.complete.obs"
)

# Identify pairs with |correlation| > 0.7
high_corr <- which(abs(cor_matrix) > 0.7 & abs(cor_matrix) <= 1, arr.ind = TRUE)

# Create a data frame of correlated pairs
correlated_pairs <- data.frame(
  var1 = rownames(cor_matrix)[high_corr[, 1]],
  var2 = colnames(cor_matrix)[high_corr[, 2]],
  correlation = round(cor_matrix[high_corr], 3)
)

# Remove duplicates (var1-var2 vs var2-var1)
correlated_pairs <- correlated_pairs[correlated_pairs$var1 < correlated_pairs$var2, ]
print(correlated_pairs)

# Human-readable labels for publication
clean_labels <- c(
  visit = "Survey visit",
  bareground = "Bareground cover", vegcover = "Vegetation cover",
  vegheight = "Vegetation height", herb = "Herbaceous veg. cover",
  woody = "Woody veg. cover", sand = "Sand cover", gravel = "Gravel cover",
  cobble = "Cobble cover", stone = "Stone cover",
  soil = "Soil cover", flat = "Flat", sloped = "Sloped",
  steep = "Steep", canopy = "Canopy cover",
  branchpile = "Branch piles", stonepile = "Stone piles",
  patchiness_num = "Plot heterogeneity",
  standingdeadwood = "Standing deadwood", lyingdeadwood = "Lying deadwood",
  hill_q0 = "Floral richness"
)

cor_matrix_labeled <- cor_matrix
colnames(cor_matrix_labeled) <- clean_labels[colnames(cor_matrix_labeled)]
rownames(cor_matrix_labeled) <- clean_labels[rownames(cor_matrix_labeled)]

# Plot correlation matrix (publication export)
tiff(
  file.path("..", "output", "figures", "habitat_corrplot_rawdata_600dpi.tiff"),
  width = 6000, height = 6000, res = 600, compression = "lzw"
)

corrplot(
  cor_matrix_labeled,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  addCoef.col = "black",
  number.cex = 0.6
)

dev.off()

# Also render inline
corrplot(
  cor_matrix_labeled,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  addCoef.col = "black",
  number.cex = 0.6
)

# New variable dataframe (merged covariates with spearman above 0.7)
# bareground - vegcover -> bareground
# herb - woody -> herb
# flat - sloped -> sloped
# hill_q0  hill_q1  hill_q2 -> hill_q0

env_vars <- c("bareground", "vegheight", "herb", "sand", "gravel", "cobble",
              "stone", "wood", "soil", "sloped", "steep", "canopy",
              "branchpile", "stonepile", "standingdeadwood", "lyingdeadwood",
              "hill_q0")
```

```{r export figure}
tiff(
  file.path("..", "output", "figures", "Figure_F2.tiff"),
  width = 6000, height = 6000, res = 600, compression = "lzw"
)

corrplot(
  cor_matrix_labeled,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  addCoef.col = "black",
  number.cex = 0.6
)

dev.off()
```


```{r bee ab model prep: test for collinearity among habitat explanatory variables}
# List of numeric environmental variables
env_vars <- c("visit", "bareground", "vegcover", "vegheight", "herb", "woody",
              "sand", "gravel", "cobble", "stone", "wood", "soil", "flat",
              "sloped", "steep", "canopy", "branchpile", "stonepile",
              "patchiness", "standingdeadwood", "lyingdeadwood",
              "hill_q0", "hill_q1", "hill_q2")
#visit and patchiness, although included in coming models as ordinal, are treated as numeric for Spearman computation

# Compute Spearman correlation matrix
cor_matrix <- cor(
  data_all_scaled[env_vars],
  method = "spearman",
  use = "pairwise.complete.obs")

# Identify pairs with |correlation| > 0.7
high_corr <- which(abs(cor_matrix) > 0.7 & abs(cor_matrix) <= 1, arr.ind = TRUE)

# Create a data frame of correlated pairs
correlated_pairs <- data.frame(
  var1 = rownames(cor_matrix)[high_corr[, 1]],
  var2 = colnames(cor_matrix)[high_corr[, 2]],
  correlation = cor_matrix[high_corr]
)

# Remove duplicates (var1-var2 vs var2-var1)
correlated_pairs <- correlated_pairs[correlated_pairs$var1 < correlated_pairs$var2, ]

print(correlated_pairs)
# visualize correlation
corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  addCoef.col = "black",
  number.cex = 0.6
)

# New variable dataframe (merged covariates with spearman above 0.7)
# bareground - vegcover -> bareground
# herb - woody -> herb
# flat - sloped -> sloped
# hill_q0  hill_q1  hill_q2 -> hill_q0

env_vars <- c("bareground", "vegheight", "herb", "sand", "gravel", "cobble",
              "stone", "wood", "soil", "sloped", "steep", "canopy",
              "branchpile", "stonepile", "standingdeadwood", "lyingdeadwood",
              "hill_q0")
```

Can we include site as explanatory variable in a multivariate model? - No, this chunk shows that too many of the habitat variables confound with site, and should therefore not be included in the same multivariate model.
```{r bee ab model prep: diagnostic site vs habitat covariates}
num_vars <- c("visit", "bareground", "vegcover", "vegheight", "herb", "woody",
              "sand", "gravel", "cobble", "stone", "wood", "soil", "flat",
              "sloped", "steep", "canopy", "branchpile",
              "stonepile", "patchiness", "standingdeadwood", "lyingdeadwood",
              "hill_q0")

site_tests <- lapply(num_vars, function(v) {
  wt <- wilcox.test(data_all_scaled[[v]] ~ data_all_scaled$site)
  data.frame(
    variable = v,
    W = wt$statistic,
    p_value = wt$p.value
  )
})

site_tests <- do.call(rbind, site_tests)

site_tests$p_adj <- p.adjust(site_tests$p_value, method = "BH")

# these variables confound with site and cannot be included in a multivariate model together
confounded_vars <- site_tests |>
  subset(p_adj < 0.05) |>
  arrange(p_adj)

confounded_vars$variable

# these variables do not confound with site and could be included in a multivariate model together
independent_vars <- site_tests |>
  subset(p_adj >= 0.05)

independent_vars$variable
```


## Bee abundance
```{r Bee abundance global model}
AB_glmm <- glmmTMB(
  bee_abundance ~
    hill_q0 + sand + herb + bareground + vegheight + patchiness_ord +
    (1 | plotID),
  family = nbinom1,
  data = data_all_scaled,
  na.action = na.fail
)
print(summary(AB_glmm))

check_model(AB_glmm)
```

```{r Bee abundance dredge}
dd <- dredge(AB_glmm, rank = "AICc", trace = 2)
dd

head(dd)
```

```{r model averaging abundance}
# filter only valid models
all_models <- get.models(dd, subset = TRUE)

valid_vec <- sapply(all_models, function(m) {
  vc <- try(vcov(m)$cond, silent = TRUE)
  !(inherits(vc, "try-error") || any(is.na(diag(vc))))
})

dd$valid <- valid_vec

dd_valid <- subset(dd, valid)

# filter by dAIC 2
cand_ab <- get.models(dd_valid, subset = delta <= 2)

# best model (to be used later for ggpred)
bestmod_ab <- cand_ab[[1]]

avg_mod_ab <- model.avg(cand_ab, revised.var = TRUE, fit = TRUE)

summary(avg_mod_ab)

# sum of weights
sw(avg_mod_ab)
```

Best model bee abundance
```{r best model bee abundance}
bestmod_ab
bestmod_ab <- glmmTMB(bee_abundance ~ hill_q0 + patchiness_ord + sand + (1 | plotID),
                      family = nbinom1,
                      data = data_all_scaled)

summary(bestmod_ab)
```

Plot predictions. Backtransform x-axis plant abundance for interpretability
```{r bes model bee abundance plot predictions}
# Extract scaling attributes for hill and sand
hill_mean <- attr(data_all_scaled$hill_q0, "scaled:center")
hill_sd   <- attr(data_all_scaled$hill_q0, "scaled:scale")

sand_mean <- attr(data_all_scaled$sand, "scaled:center")
sand_sd   <- attr(data_all_scaled$sand, "scaled:scale")

# Generate predictions for each predictor
pred_hill   <- ggpredict(bestmod_ab, terms = "hill_q0")
pred_patch  <- ggpredict(bestmod_ab, terms = "patchiness_ord")
pred_sand   <- ggpredict(bestmod_ab, terms = "sand")

# Back-transform hill_q0 and sand to raw units
pred_hill$x_raw <- pred_hill$x * hill_sd + hill_mean
pred_sand$x_raw <- pred_sand$x * sand_sd + sand_mean

data_all_scaled$hill_q0_raw <- data_all_scaled$hill_q0 * hill_sd + hill_mean
data_all_scaled$sand_raw     <- data_all_scaled$sand     * sand_sd  + sand_mean

# Plot 1: Hill q0 (raw axis) + jittered raw data + predictions
p_hill_ab <- ggplot() +
  geom_jitter(
    data = data_all_scaled,
    aes(x = hill_q0_raw, y = bee_abundance),
    width = 0.15, height = 0.05, alpha = 0.25, 
    size = 1.5, color = "grey50"
  ) +
  geom_ribbon(
    data = pred_hill,
    aes(x = x_raw, ymin = conf.low, ymax = conf.high),
    fill = "#D2A9D8", alpha = 0.35
  ) +
  geom_line(
    data = pred_hill,
    aes(x = x_raw, y = predicted, group = 1),
    color = "#B10DC9", linewidth = 1.3
  ) +
  labs(
    title = "Effect of Floral richness",
    subtitle = "Predicted bee abundance ± 95% CI with raw data",
    x = "Floral richness",
    y = "Bee abundance"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13, color = "black"),
    plot.title = element_text(size = 16, face = "bold")
  )

# Plot 2: Patchiness (ordinal) + jittered raw data + predictions
p_patch_ab <- ggplot() +
  geom_jitter(
    data = data_all_scaled,
    aes(x = patchiness_ord, y = bee_abundance),
    width = 0.1, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  geom_point(
    data = pred_patch,
    aes(x = x, y = predicted),
    color = "#B10DC9", size = 3
  ) +
  geom_errorbar(
    data = pred_patch,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    color = "#B10DC9", width = 0.1, linewidth = 1
  ) +
  labs(
    title = "Effect of Plot Heterogeneity",
    subtitle = "Predicted bee abundance ± 95% CI with raw data",
    x = "Plot heterogeneity",
    y = "Bee abundance"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13, color = "black"),
    plot.title = element_text(size = 16, face = "bold")
  )

# Plot 3: Sand (raw axis) + jittered raw points + predictions
p_sand_ab <- ggplot() +
  geom_jitter(
    data = data_all_scaled,
    aes(x = sand_raw, y = bee_abundance),
    width = 0.15, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  geom_ribbon(
    data = pred_sand,
    aes(x = x_raw, ymin = conf.low, ymax = conf.high),
    fill = "#D2A9D8", alpha = 0.35
  ) +
  geom_line(
    data = pred_sand,
    aes(x = x_raw, y = predicted, group = 1),
    color = "#B10DC9", linewidth = 1.3
  ) +
  labs(
    title = "Effect of Sand Cover",
    subtitle = "Predicted bee abundance ± 95% CI with raw data",
    x = "Sand cover (%)",
    y = "Bee abundance"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13, color = "black"),
    plot.title = element_text(size = 16, face = "bold")
  )

# Combine all three into a multi-panel figure

p_combined_abundance <- p_hill_ab + p_patch_ab + p_sand_ab + plot_layout(ncol = 3)

p_combined_abundance
```


## Bee richness
```{r Bee richness global model}
RI_glmm <- glmmTMB(
  bee_richness ~
    hill_q0 + sand + bareground + herb + vegheight + patchiness_ord +
    (1 | plotID),
  family = nbinom1,
  data = data_all_scaled,
  na.action = na.fail
)

print(summary(RI_glmm))
check_model(RI_glmm)
```

```{r Bee richness dredge}
dr <- dredge(RI_glmm, rank = "AICc", trace = 2)
dr

head(dr)
```

```{r model averaging richness}
# extract all models
all_models <- get.models(dr, subset = TRUE)

# robust validity check
valid_vec <- sapply(all_models, function(m) {

  # check Hessian
  pdHess_ok <- !is.null(m$sdr) && isTRUE(m$sdr$pdHess)

  # check vcov
  vc_ok <- FALSE
  vc <- try(vcov(m)$cond, silent = TRUE)

  if (!inherits(vc, "try-error")) {
    vc_ok <- !any(is.na(diag(vc)))
  }

  pdHess_ok & vc_ok
})

# add validity column
dr$valid <- valid_vec

# retain only valid models with finite AIC statistics
dr_valid <- subset(
  dr,
  valid &
    !is.na(AICc) &
    !is.na(delta)
)

# candidate set (ΔAICc ≤ 2)
cand_ri <- get.models(
  dr_valid,
  subset = delta <= 2
)

# check candidate set
length(cand_ri)

# best-ranked model for predictions
bestmod_ri <- cand_ri[[1]]

# model averaging
avg_mod_ri <- model.avg(
  cand_ri,
  revised.var = TRUE,
  fit = TRUE
)

summary(avg_mod_ri)

# variable importance
imp_ri <- sw(avg_mod_ri)

imp_ri
```

Best model bee richness
```{r best model bee richness}
bestmod_ri

bestmod_ri <- glmmTMB(bee_richness ~ hill_q0 + patchiness_ord + 
                      (1 | plotID),
                      family = nbinom1,
                      data = data_all_scaled)

summary(bestmod_ri)
```

Plot predictions. Backtransform x-axis plant richness for interpretability
```{r bes model bee richness plot predictions}
# Generate predictions
pred_hill   <- ggpredict(bestmod_ri, terms = "hill_q0")
pred_patch  <- ggpredict(bestmod_ri, terms = "patchiness_ord")

# back-transform scaled hill values
hill_mean <- attr(data_all_scaled$hill_q0, "scaled:center")
hill_sd   <- attr(data_all_scaled$hill_q0, "scaled:scale")

pred_hill$x_raw <- pred_hill$x * hill_sd + hill_mean


# Also back-transform raw jitter points for Hill q0:
data_all_scaled$hill_q0_raw <- data_all_scaled$hill_q0 * hill_sd + hill_mean


p_hill_ri <- ggplot() +
  
  geom_jitter(
    data = data_all_scaled,
    aes(x = hill_q0_raw, y = bee_richness),
    width = 0.15, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  
  geom_ribbon(
    data = pred_hill,
    aes(x = x_raw, ymin = conf.low, ymax = conf.high),
    fill = "#CE84A8", alpha = 0.35
  ) +
  
  geom_line(
    data = pred_hill,
    aes(x = x_raw, y = predicted, group = 1),
    color = "#85144B", linewidth = 1.3
  ) +
  
  labs(
    title = "Effect of Floral richness",
    subtitle = "Predicted bee richness ± 95% CI with raw data",
    x = "Floral richness",
    y = "Bee richness"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13, color = "black"),
    plot.title = element_text(size = 16, face = "bold")
  )

# 5. Plot 2: Patchiness (ordinal) prediction with jitter
p_patch_ri <- ggplot() +
  
  geom_jitter(
    data = data_all_scaled,
    aes(x = patchiness_ord, y = bee_richness),
    width = 0.1, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  
  geom_point(
    data = pred_patch,
    aes(x = x, y = predicted),
    color = "#85144B", size = 3
  ) +
  
  geom_errorbar(
    data = pred_patch,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    color = "#85144B", width = 0.1, linewidth = 1
  ) +
  
  labs(
    title = "Effect of Plot Heterogeneity",
    subtitle = "Predicted bee richness ± 95% CI with raw data",
    x = "Plot heterogeneity",
    y = "Bee richness"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13, color = "black"),
    plot.title = element_text(size = 16, face = "bold")
  )

# ============================================================
# 6. Combine into a two-panel figure
# ============================================================

p_combined <- p_hill_ri + p_patch_ri + plot_layout(ncol = 2)

p_combined
```

## Bee Hill Q = 1
Initially gamma log link was used but changed to log gaussian for consistency with site univ model distribution
```{r Bee hill q1 global model}
Q1_glmm <- glmmTMB(
  log(bee_hill_q1) ~
    hill_q0 + sand + bareground + herb + vegheight + patchiness_ord +
    (1 | plotID),
  family = gaussian(),
  data = data_all_scaled,
  na.action = na.fail
)

print(summary(Q1_glmm))

sim_Q1_glmm <- simulateResiduals(Q1_glmm)
plot(sim_Q1_glmm)
testDispersion(Q1_glmm)
testZeroInflation(Q1_glmm)

check_model(Q1_glmm)
```

```{r Bee hill q1 dredge}
dq <- dredge(Q1_glmm, rank = "AICc", trace = 2)
dq

head(dq)
```

```{r model averaging hill q1}
# filter only valid models
all_models <- get.models(dq, subset = TRUE)

valid_vec <- sapply(all_models, function(m) {
  vc <- try(vcov(m)$cond, silent = TRUE)
  !(inherits(vc, "try-error") || any(is.na(diag(vc))))
})

dq$valid <- valid_vec

dq_valid <- subset(dq, valid)

# filter by dAIC 2
cand_q1 <- get.models(dq_valid, subset = delta <= 2)

# best model (to be used later for ggpred)
bestmod_q1 <- cand_q1[[1]]

avg_mod_q1 <- model.avg(cand_q1, revised.var = TRUE, fit = TRUE)

summary(avg_mod_q1)

# sum of weights
sw(avg_mod_q1)
```

Best model bee hillq1
```{r best model bee hillq1}
bestmod_q1

bestmod_q1 <- glmmTMB(log(bee_hill_q1) ~ hill_q0 + patchiness_ord + (1 | plotID),
                      family = gaussian(),
                      data = data_all_scaled)

summary(bestmod_q1)
```

Plot predictions. Backtransform x-axis plant richness for interpretability
```{r Best model hillq1 prediction plot}
# Generate predictions
pred_hill  <- ggpredict(bestmod_q1, terms = "hill_q0")
pred_patch <- ggpredict(bestmod_q1, terms = "patchiness_ord")


# Back-transform hill_q0
hill_mean <- attr(data_all_scaled$hill_q0, "scaled:center")
hill_sd   <- attr(data_all_scaled$hill_q0, "scaled:scale")

pred_hill$x_raw <- pred_hill$x * hill_sd + hill_mean
data_all_scaled$hill_q0_raw <- data_all_scaled$hill_q0 * hill_sd + hill_mean

# Plot Hill q0 effect with raw data and predictions
p_hill_q1 <- ggplot() +
  geom_jitter(
    data = data_all_scaled,
    aes(x = hill_q0_raw, y = bee_hill_q1),
    width = 0.15, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  geom_ribbon(
    data = pred_hill,
    aes(x = x_raw, ymin = conf.low, ymax = conf.high),
    fill = "#E09691", alpha = 0.35
  ) +
  geom_line(
    data = pred_hill,
    aes(x = x_raw, y = predicted, group = 1),
    color = "#FF4136", linewidth = 1.3
  ) +
  labs(
    title = "Effect of Floral richness",
    subtitle = "Predicted bee Shannon diversity (Hill q = 1) ± 95% CI with raw data",
    x = "Floral richness",
    y = "Bee Shannon diversity"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13, color = "black"),
    plot.title = element_text(size = 16, face = "bold")
  ) +
  scale_y_continuous(
    breaks = scales::pretty_breaks(n = 4),
    labels = scales::label_number(accuracy = 1)
  )

# Plot patchiness (ordinal) with raw data and predictions
p_patch_q1 <- ggplot() +
  geom_jitter(
    data = data_all_scaled,
    aes(x = patchiness_ord, y = bee_hill_q1),
    width = 0.1, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  geom_point(
    data = pred_patch,
    aes(x = x, y = predicted),
    color = "#FF4136", size = 3
  ) +
  geom_errorbar(
    data = pred_patch,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    color = "#FF4136", width = 0.1, linewidth = 1
  ) +
  labs(
    title = "Effect of Plot Heterogeneity",
    subtitle = "Predicted bee Shannon diversity (Hill q = 1) ± 95% CI with raw data",
    x = "Plot heterogeneity",
    y = "Bee Shannon diversity"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13, color = "black"),
    plot.title = element_text(size = 16, face = "bold")
  ) +
  scale_y_continuous(
    breaks = scales::pretty_breaks(4),
    labels = scales::label_number(accuracy = 1)
  )

# Combine both panels into one figure
p_combined_q1 <- p_hill_q1 + p_patch_q1 + plot_layout(ncol = 2)

p_combined_q1
```


## Model avg table extraction
```{r table model averaging}
# extraction function
extract_avg_table <- function(avg_model, dredge_obj, response_name) {
  
  coef_df <- as.data.frame(summary(avg_model)$coefmat.full)
  coef_df$term <- rownames(coef_df)
  
  coef_df <- coef_df |>
    filter(term != "cond((Int))")
  
  ci <- confint(avg_model)
  coef_df$lower <- ci[coef_df$term, 1]
  coef_df$upper <- ci[coef_df$term, 2]
  
  coef_df$p_value <- coef_df[, "Pr(>|z|)"]
  
  coef_df$p_value_fmt <- ifelse(
    coef_df$p_value < 0.001,
    "<0.001",
    formatC(coef_df$p_value, format = "f", digits = 3)
  )
  
  importance_raw <- sw(avg_model)
  importance_vals <- as.numeric(importance_raw)
  names(importance_vals) <- names(importance_raw)
  
  coef_df$base_term <- str_replace(coef_df$term, "\\.L|\\.Q|\\.C", "")
  
  coef_df$importance <- importance_vals[
    match(coef_df$base_term, names(importance_vals))
  ]
  
  coef_df <- coef_df |>
    filter(!str_detect(term, "\\.Q|\\.C"))
  
  coef_df$Predictor <- coef_df$term |>
    str_replace("cond\\(", "") |>
    str_replace("\\)", "") |>
    str_replace("_ord\\.L", "") |>
    str_replace("_ord", "") |>
    str_replace("hill_q0", "Floral Richness") |>
    str_replace("sand", "Sand") |>
    str_replace("herb", "Herbaceous vegetation") |>
    str_replace("bareground", "Bareground") |>
    str_replace("vegheight", "Vegetation height") |>
    str_replace("patchiness", "Plot heterogeneity (linear)") |>
    str_trim()
  
  coef_df |>
    transmute(
      Response = response_name,
      Predictor,
      `Estimate (± 95% CI)` = sprintf("%.3f (%.2f – %.2f)", Estimate, lower, upper),
      `Importance (Σw)` = round(importance, 2),
      `P-value` = p_value_fmt
    )
}

# extract tables
tab_ab <- extract_avg_table(avg_mod_ab, dd, "Abundance")
tab_ri <- extract_avg_table(avg_mod_ri, dr, "Richness")
tab_q1 <- extract_avg_table(avg_mod_q1, dq, "Bee Shannon diversity")

table_avg_full <- bind_rows(tab_ab, tab_ri, tab_q1)

# predictor order
predictor_order <- c(
  "Plot heterogeneity (linear)",
  "Floral Richness",
  "Sand",
  "Vegetation height",
  "Bareground",
  "Herbaceous vegetation"
)

table_avg_full <- table_avg_full |>
  mutate(Predictor = factor(Predictor, levels = predictor_order)) |>
  arrange(Response, Predictor)

# FIXED section function (correct column name!)
add_section <- function(df, name) {
  header <- tibble(
    Predictor = name,
    `Estimate (± 95% CI)` = NA_character_,
    `Importance (Σw)` = NA_real_,
    `P-value` = NA_character_
  )
  bind_rows(header, df |> select(-Response))
}

table_avg_final <- bind_rows(
  add_section(filter(table_avg_full, Response == "Abundance"), "Abundance"),
  add_section(filter(table_avg_full, Response == "Richness"), "Richness"),
  add_section(filter(table_avg_full, Response == "Bee Shannon diversity"), "Bee Shannon diversity")
)

# create gt table (FIXED row check!)
gt_table_avg <- table_avg_final |>
  gt() |>
  cols_align(align = "left", columns = Predictor) |>
  cols_align(
    align = "center",
    columns = c(`Estimate (± 95% CI)`, `Importance (Σw)`, `P-value`)
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = is.na(`Estimate (± 95% CI)`))
  ) |>
  fmt_missing(everything(), missing_text = "")

gt_table_avg
```

```{r export table}
gtsave(
  gt_table_avg,
  "..", "output", "tables", "model_avg_raw.docx"
)
```

```{r export table}
gtsave(
  gt_table_avg,
  "..", "output", "tables", "Table_E2.1.docx"
)
```

## Model avg combined plot
```{r plot combined all plots}
extract_plot_df <- function(avg_model, dredge_obj, response_name) {
  
  coef_df <- as.data.frame(summary(avg_model)$coefmat.full)
  coef_df$term <- rownames(coef_df)
  
  coef_df <- coef_df |>
    filter(term != "cond((Int))")
  
  ci <- confint(avg_model)
  coef_df$lower <- ci[coef_df$term, 1]
  coef_df$upper <- ci[coef_df$term, 2]
  
  imp_raw <- imp_raw <- sw(avg_model)
  imp_vals <- as.numeric(imp_raw)
  names(imp_vals) <- names(imp_raw)
  
  coef_df$base_term <- str_replace(coef_df$term, "\\.L|\\.Q|\\.C", "")
  
  coef_df$importance <- imp_vals[
    match(coef_df$base_term, names(imp_vals))
  ]
  
  coef_df <- coef_df |>
    filter(!str_detect(term, "\\.Q|\\.C"))
  
  coef_df$Predictor <- coef_df$term |>
    str_replace("cond\\(", "") |>
    str_replace("\\)", "") |>
    str_replace("_ord\\.L", "") |>
    str_replace("_ord", "") |>
    str_replace("hill_q0", "Floral Richness") |>
    str_replace("sand", "Sand") |>
    str_replace("herb", "Herbaceous vegetation") |>
    str_replace("bareground", "Bareground") |>
    str_replace("vegheight", "Vegetation height") |>
    str_replace("patchiness", "Plot heterogeneity (linear)") |>
    str_trim()
  
  coef_df |>
    transmute(
      Response = response_name,
      Predictor,
      Estimate,
      lower,
      upper,
      importance
    )
}

df_combined <- bind_rows(
  extract_plot_df(avg_mod_ab, dd, "Abundance"),
  extract_plot_df(avg_mod_ri, dr, "Richness"),
  extract_plot_df(avg_mod_q1, dq, "Bee Shannon div.")
)

df_combined <- df_combined |>
  mutate(
    Predictor = factor(Predictor, levels = rev(predictor_order)),
    Response = factor(Response, levels = c(
      "Bee Shannon div.",
      "Richness",
      "Abundance"
    ))
  )

df_combined <- df_combined |>
  complete(Predictor, Response)


pd <- position_dodge(width = 0.6)

p_combined_dodged <- ggplot(
  df_combined,
  aes(x = Estimate, y = Predictor, color = Response, group = Response)
) +
  
  geom_errorbarh(
    aes(xmin = lower, xmax = upper, alpha = ifelse(is.na(importance), FALSE, importance >= 0.8)),
    position = pd,
    height = 0,
    linewidth = 1,
    na.rm = FALSE
  ) +
  
  geom_point(
    aes(alpha = ifelse(is.na(importance), FALSE, importance >= 0.8)),
    position = pd,
    size = 3,
    na.rm = FALSE
  ) +
  
  scale_alpha_manual(
    values = c("TRUE" = 1, "FALSE" = 0.7),
    guide = "none"
  ) +
  
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    scale_color_manual(
    values = c(
      "Abundance" = "#B10DC9",
      "Richness" = "#85144B",
      "Bee Shannon div." = "#FF4136"
    ),
    breaks = c("Abundance", "Richness", "Bee Shannon div."),
    labels = c(
      "Abundance" = "Bee abundance",
      "Richness" = "Bee richness",
      "Bee Shannon div." = "Bee Shannon diversity"
    )
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 14),
    legend.position = "right",
    legend.direction = "vertical",
    legend.title = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)
  ) +
  
  labs(
    x = "Model-averaged estimate (95% CI)",
    y = NULL,
    color = NULL
    )


p_combined_dodged
```
```{r export figure}
tiff(
  file.path("..", "output", "figures", "forest_rawcombined_600dpi.tiff"),
  width = 4500, height = 2600, res = 600, compression = "lzw"
)

print(p_combined_dodged)
dev.off()
```

```{r export figure}
tiff(
  file.path("..", "output", "figures", "Figure_E2.1.tiff"),
  width = 4500, height = 2600, res = 600, compression = "lzw"
)

print(p_combined_dodged)
dev.off()
```


## Bestmodel plots patchwork
```{r plot patchwork bestmodels all metrics}
# Shared publication theme
theme_pub <- theme_minimal(base_size = 16) +
  theme(
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 18),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)
  )

# Remove titles/subtitles and apply consistent theme
p_hill_ab_clean  <- p_hill_ab  + labs(title = NULL, subtitle = NULL) + theme_pub
p_patch_ab_clean <- p_patch_ab + labs(title = NULL, subtitle = NULL) + theme_pub
p_sand_ab_clean  <- p_sand_ab + labs(title = NULL, subtitle = NULL) + theme_pub

p_hill_ri_clean  <- p_hill_ri + labs(title = NULL, subtitle = NULL) + theme_pub
p_patch_ri_clean <- p_patch_ri + labs(title = NULL, subtitle = NULL) + theme_pub

p_hill_q1_clean  <- p_hill_q1 + labs(title = NULL, subtitle = NULL) + theme_pub
p_patch_q1_clean <- p_patch_q1 + labs(title = NULL, subtitle = NULL) + theme_pub

# Build rows (predictors per metric)
row_ab <- p_patch_ab_clean + p_hill_ab_clean
row_ri <- p_patch_ri_clean + p_hill_ri_clean
row_q1 <- p_patch_q1_clean + p_hill_q1_clean

# Combine into final multi-row figure
p_bestmod_all <-
  row_ab /
  row_ri /
  row_q1 +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")"
  ) +
  plot_layout(widths = c(1, 1))

p_bestmod_all
```

```{r export figure}
tiff(
  file.path("..", "output", "figures", "bestmod_pred_all_600dpi.tiff"),
  width = 5800, height = 7000, res = 600, compression = "lzw"
)

print(p_bestmod_all)
dev.off()
```

```{r export figure}
tiff(
  file.path("..", "output", "figures", "Figure_E2.2.tiff"),
  width = 5800, height = 7000, res = 600, compression = "lzw"
)

print(p_bestmod_all)
dev.off()
```


## Bestmodel table
```{r Bestmodel each }
# Function to extract best model results
extract_best_model_table <- function(model, response_name) {
  
  coef_df <- broom.mixed::tidy(
    model,
    effects = "fixed",
    conf.int = TRUE
  )
  
  coef_df <- coef_df |>
    filter(term != "(Intercept)") |>
    filter(
      !str_detect(term, "\\.Q") &
      !str_detect(term, "\\.C")
    ) |>
    transmute(
      Response = response_name,
      Predictor = term,
      `Estimate (± 95% CI)` = sprintf(
        "%.2f (%.2f–%.2f)",
        estimate,
        conf.low,
        conf.high
      ),
      `P-value` = case_when(
        p.value < 0.001 ~ "<0.001",
        TRUE ~ formatC(
          p.value,
          format = "f",
          digits = 3
        )
      )
    )
  
  # Clean predictor names
  coef_df$Predictor <- coef_df$Predictor |>
    str_replace("patchiness_ord.L", "Plot heterogeneity (linear)") |>
    str_replace("patchiness_ord", "Plot heterogeneity (linear)") |>
    str_replace("hill_q0", "Floral richness") |>
    str_replace("sand", "Sand") |>
    str_replace("bareground", "Bare ground") |>
    str_replace("vegheight", "Vegetation height") |>
    str_replace("herb", "Herb layer") |>
    str_trim()
  
  coef_df
}

# Extract model tables
tab_ab <- extract_best_model_table(
  bestmod_ab,
  "Abundance"
)

tab_ri <- extract_best_model_table(
  bestmod_ri,
  "Richness"
)

tab_q1 <- extract_best_model_table(
  bestmod_q1,
  "Hill q1"
)

# Extract R² values
r2_ab <- performance::r2(bestmod_ab)
r2_ri <- performance::r2(bestmod_ri)
r2_q1 <- performance::r2(bestmod_q1)

# Combine tables
table_best <- bind_rows(
  tab_ab,
  tab_ri,
  tab_q1
)

# Predictor order
table_best$Predictor <- factor(
  table_best$Predictor,
  levels = c(
    "Plot heterogeneity (linear)",
    "Floral richness",
    "Sand",
    "Vegetation height",
    "Bare ground",
    "Herb layer"
  )
)

table_best <- table_best |>
  arrange(Response, Predictor)

# Section helper
add_section <- function(df, name) {
  
  header <- tibble(
    Predictor = name,
    `Estimate (± 95% CI)` = NA_character_,
    `P-value` = NA_character_
  )
  
  bind_rows(
    header,
    df |> select(-Response)
  )
}

# Final table with R² values in section headers
table_final <- bind_rows(
  
  add_section(
    filter(table_best, Response == "Abundance"),
    sprintf(
      "Bee abundance (R²m = %.3f; R²c = %.3f)",
      r2_ab$R2_marginal,
      r2_ab$R2_conditional
    )
  ),
  
  add_section(
    filter(table_best, Response == "Richness"),
    sprintf(
      "Bee richness (R²m = %.3f; R²c = %.3f)",
      r2_ri$R2_marginal,
      r2_ri$R2_conditional
    )
  ),
  
  add_section(
    filter(table_best, Response == "Hill q1"),
    sprintf(
      "Bee Shannon diversity (R²m = %.3f; R²c = %.3f)",
      r2_q1$R2_marginal,
      r2_q1$R2_conditional
    )
  )
)

# Create gt table
gt_table <- table_final |>
  gt() |>
  cols_label(
    Predictor = "Predictor",
    `Estimate (± 95% CI)` = "Estimate (± 95% CI)",
    `P-value` = "P-value"
  ) |>
  cols_align(
    align = "left",
    columns = Predictor
  ) |>
  cols_align(
    align = "center",
    columns = c(
      `Estimate (± 95% CI)`,
      `P-value`
    )
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      rows = is.na(`Estimate (± 95% CI)`)
    )
  ) |>
  fmt_missing(
    everything(),
    missing_text = ""
  )

gt_table
```


```{r save bestmodel table}
gtsave(
  gt_table,
  "..", "output", "tables", "bestmodel_table.docx"
)
```

```{r save bestmodel table}
gtsave(
  gt_table,
  "..", "output", "tables", "Table_E2.2.docx"
)
```


# Rarefied data - pooled across visits
```{r load data}
data_model_cov <- readRDS(file.path(
  "..", "data", "cleaned", "data_model_cov.rds"))

data_model_size <- readRDS(file.path(
  "..", "data", "cleaned", "data_model_size.rds"))
```

Collinearity test environmental variables
```{r correlation matrix habitat variables rarefied data}
# Convert ordinal patchiness to numeric
data_model_cov <- data_model_cov |>
  mutate(patchiness_num = as.numeric(patchiness_ord))

# List of variables INCLUDING patchiness
env_vars <- c("bareground", "vegcover", "vegheight", "herb", "woody",
              "sand", "gravel", "cobble", "stone", "soil", "flat",
              "sloped", "steep", "canopy", "branchpile",
              "stonepile", "standingdeadwood", "lyingdeadwood", "hill_q0",
              "patchiness_num")

# Compute Spearman correlation matrix
cor_matrix <- cor(
  data_model_cov[env_vars],
  method = "spearman",
  use = "pairwise.complete.obs"
)

# Identify pairs with |correlation| > 0.7
high_corr <- which(abs(cor_matrix) > 0.7 & abs(cor_matrix) < 1, arr.ind = TRUE)

# Create data frame
correlated_pairs <- data.frame(
  var1 = rownames(cor_matrix)[high_corr[, 1]],
  var2 = colnames(cor_matrix)[high_corr[, 2]],
  correlation = round(cor_matrix[high_corr], 3)
)

# Remove duplicates
correlated_pairs <- correlated_pairs[correlated_pairs$var1 < correlated_pairs$var2, ]
print(correlated_pairs)

# Human-readable labels for publication
clean_labels <- c(
  bareground = "Bareground cover", vegcover = "Vegetation cover",
  vegheight = "Vegetation height", herb = "Herbaceous veg. cover",
  woody = "Woody veg. cover", sand = "Sand cover", gravel = "Gravel cover",
  cobble = "Cobble cover", stone = "Stone cover",
  soil = "Soil cover", flat = "Flat", sloped = "Sloped",
  steep = "Steep", canopy = "Canopy cover", branchpile = "Branch piles",
  stonepile = "Stone piles", standingdeadwood = "Standing deadwood",
  lyingdeadwood = "Lying deadwood", hill_q0 = "Floral richness",
  patchiness_num = "Plot heterogeneity")

cor_matrix_labeled <- cor_matrix
colnames(cor_matrix_labeled) <- clean_labels[colnames(cor_matrix_labeled)]
rownames(cor_matrix_labeled) <- clean_labels[rownames(cor_matrix_labeled)]

# Plot correlation matrix (same structure as your original, publication export added)
tiff(
  file.path("..", "output", "figures", "habitat_corrplot_rarefdata_600dpi.tiff"),
  width = 6000, height = 6000, res = 600, compression = "lzw"
)

corrplot(
  cor_matrix_labeled,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  addCoef.col = "black",
  number.cex = 0.6
)

dev.off()

# Also render inline
corrplot(
  cor_matrix_labeled,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  addCoef.col = "black",
  number.cex = 0.6
)
```
```{r export figure}
tiff(
  file.path("..", "output", "figures", "Figure_F1.tiff"),
  width = 6000, height = 6000, res = 600, compression = "lzw"
)

corrplot(
  cor_matrix_labeled,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  addCoef.col = "black",
  number.cex = 0.6
)

dev.off()
```

herb and vegheight highly correlated (0.80) - keep only herb

patchiness is correlated with hill_q0 (0.73), sand (0.71) - keep only patchiness

# Coverage-based rarefaction
## Bee richness
```{r coverage rarefied bee richness global model}
RRI_glmm_cov <- glmmTMB(
  richness_rarefied_cov ~ herb + bareground + patchiness_ord +
    (1 | region),
  data = data_model_cov,
  family = Gamma(link = 'log'),
  na.action = na.fail
)

summary(RRI_glmm_cov)
check_model(RRI_glmm_cov)
```

```{r coverage rarefied bee richness dredge}
dri_cov <- dredge(RRI_glmm_cov, rank = "AICc", trace = 2)

head(dri_cov)
```

only one model within dAIC = 2 - no averaging needed
```{r size rarefied model averaging richness}
cand_ri_cov <- get.models(dri_cov, subset = delta <= 2)

bestmod_ri_cov <- cand_ri_cov[[1]]
```

Best model coverage rarefied richness
```{r best model coverage based rarefied richness}
bestmod_ri_cov <- glmmTMB(richness_rarefied_cov ~ patchiness_ord + (1 | region),
                          data = data_model_cov,
                          family = Gamma(link = 'log'),
                          na.action = na.fail)

summary(bestmod_ri_cov)
```

```{r plot best model coverage based rarefied richness}
# predictions
pred_patch_cov_ri <- ggpredict(bestmod_ri_cov, terms = "patchiness_ord")

# plot
p_patch_ricov <- ggplot() +
  
  geom_jitter(
    data = data_model_cov,
    aes(x = patchiness_ord, y = richness_rarefied_cov),
    width = 0.1, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  
  geom_point(
    data = pred_patch_cov_ri,
    aes(x = x, y = predicted),
    color = "#85144B", size = 3
  ) +
  
  geom_errorbar(
    data = pred_patch_cov_ri,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    color = "#85144B",
    width = 0.1, linewidth = 1
  ) +
  
  labs(
    title = "Effect of Plot Heterogeneity",
    subtitle = "Predicted bee richness (coverage-rarefied)",
    x = "Plot heterogeneity",
    y = "Bee richness"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13)
  )

p_patch_ricov
```


## Bee HIll Q = 1
```{r coverage rarefied bee hill q1 global model}
Q1_glmm_cov <- glmmTMB(
  hill_q1_rarefied_cov ~ herb + bareground + patchiness_ord +
    (1 | region),
  family = gaussian(),
  data = data_model_cov,
  na.action = na.fail
)

summary(Q1_glmm_cov)
check_model(Q1_glmm_cov)
```

```{r coverage rarefied bee hillq1 dredge}
dq1_cov <- dredge(Q1_glmm_cov, rank = "AICc", trace = 2)

head(dq1_cov)
```

```{r coverage rarefied model averaging hillq1}
cand_q1_cov <- get.models(dq1_cov, subset = delta <= 2)

bestmod_q1_cov <- cand_q1_cov[[1]]

avg_mod_q1_cov <- model.avg(cand_q1_cov, revised.var = TRUE)

summary(avg_mod_q1_cov)
confint(avg_mod_q1_cov)

# importance
sw(subset(dq1_cov, delta <= 2))
```

Best model coverage based rarefied Hill q1
```{r}
bestmod_q1_cov

bestmod_q1_cov <- glmmTMB(hill_q1_rarefied_cov ~ patchiness_ord + (1 | region),
                          data = data_model_cov,
                          family = gaussian(),
                          na.action = na.fail)

summary(bestmod_q1_cov)
```

Plot best model covrage based hill q1
```{r best model coverage hill q1 prediction}
# predictions
pred_patch_cov_q1 <- ggpredict(bestmod_q1_cov, terms = "patchiness_ord")

p_patch_q1cov <- ggplot() +
  
  geom_jitter(
    data = data_model_cov,
    aes(x = patchiness_ord, y = hill_q1_rarefied_cov),
    width = 0.1, height = 0.05,
    alpha = 0.25, size = 1.5, color = "grey50"
  ) +
  
  geom_point(
    data = pred_patch_cov_q1,
    aes(x = x, y = predicted),
    color = "#FF4136", size = 3
  ) +
  
  geom_errorbar(
    data = pred_patch_cov_q1,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    color = "#FF4136",
    width = 0.1, linewidth = 1
  ) +
  
  labs(
    title = "Effect of Plot Heterogeneity",
    subtitle = "Predicted bee Shannon diversity (coverage-rarefied)",
    x = "Plot heterogeneity",
    y = "Bee Shannon diversity"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13)
  )

p_patch_q1cov
```

## Model avg table extraction coverage-based rarefaction
```{r model avg table extraction coverage-based rarefaction}
## Model avg table extraction coverage-based rarefaction

# function for richness single model
extract_table_single_model <- function(model_obj, response_name) {

  coef_df <- as.data.frame(summary(model_obj)$coefficients$cond)
  coef_df$term <- rownames(coef_df)

  coef_df <- coef_df |>
    filter(term != "(Intercept)")

  # confidence intervals
  ci <- confint(model_obj)
  match_idx <- match(coef_df$term, rownames(ci))

  coef_df$lower <- ci[match_idx, 1]
  coef_df$upper <- ci[match_idx, 2]

  # p-values
  coef_df$p_value <- coef_df[, "Pr(>|z|)"]

  coef_df$p_value_fmt <- ifelse(
    coef_df$p_value < 0.001,
    "<0.001",
    formatC(coef_df$p_value, format = "f", digits = 3)
  )

  # importance = 1 (single model)
  coef_df$importance <- 1

  # clean labels (INCLUDING cubic)
  coef_df$Predictor <- coef_df$term |>
    str_replace("patchiness_ord.L", "Plot heterogeneity (linear)") |>
    str_replace("patchiness_ord.Q", "Plot heterogeneity (quadratic)") |>
    str_replace("patchiness_ord.C", "Plot heterogeneity (cubic)") |>
    str_replace("herb", "Herbaceous vegetation") |>
    str_replace("bareground", "Bareground") |>
    str_trim()

  coef_df |>
    transmute(
      Response = response_name,
      Predictor,
      `Estimate (± 95% CI)` = sprintf("%.3f (%.2f – %.2f)", Estimate, lower, upper),
      `Importance (Σw)` = importance,
      `P-value` = p_value_fmt
    )
}


# function model averaging Hill q1
extract_avg_table_cov <- function(avg_model, response_name) {

  coef_df <- as.data.frame(summary(avg_model)$coefmat.full)
  coef_df$term <- rownames(coef_df)

  coef_df <- coef_df |>
    filter(term != "cond((Int))")

  ci <- confint(avg_model)
  coef_df$lower <- ci[coef_df$term, 1]
  coef_df$upper <- ci[coef_df$term, 2]
  
  coef_df$p_value <- coef_df[, "Pr(>|z|)"]

  coef_df$p_value_fmt <- ifelse(
    coef_df$p_value < 0.001,
    "<0.001",
    formatC(coef_df$p_value, format = "f", digits = 3)
  )

  # importance
  imp_raw <- sw(avg_model)
  imp_vals <- as.numeric(imp_raw)
  names(imp_vals) <- names(imp_raw)
  
  coef_df$base_term <- str_replace(coef_df$term, "\\.L|\\.Q|\\.C", "")

  coef_df$importance <- imp_vals[
    match(coef_df$base_term, names(imp_vals))
  ]

  # clean labels (WITH cubic support)
  coef_df$Predictor <- coef_df$term |>
    str_replace("cond\\(", "") |>
    str_replace("\\)", "") |>
    str_replace("patchiness_ord.L", "Plot heterogeneity (linear)") |>
    str_replace("patchiness_ord.Q", "Plot heterogeneity (quadratic)") |>
    str_replace("patchiness_ord.C", "Plot heterogeneity (cubic)") |>
    str_replace("herb", "Herbaceous vegetation") |>
    str_replace("bareground", "Bareground") |>
    str_trim()

  coef_df |>
    transmute(
      Response = response_name,
      Predictor,
      `Estimate (± 95% CI)` = sprintf("%.3f (%.2f – %.2f)", Estimate, lower, upper),
      `Importance (Σw)` = round(importance, 2),
      `P-value` = p_value_fmt
    )
}

# apply function
tab_ri_cov <- extract_table_single_model(bestmod_ri_cov, "Bee richness (coverage-rarefied)")
tab_q1_cov <- extract_avg_table_cov(avg_mod_q1_cov,"Bee Shannon diversity (coverage-rarefied)")

# bind tables
table_avg_cov <- bind_rows(tab_ri_cov, tab_q1_cov)

# order
predictor_order <- c(
  "Plot heterogeneity (linear)",
  "Plot heterogeneity (quadratic)",
  "Plot heterogeneity (cubic)",
  "Herbaceous vegetation",
  "Bareground"
)

table_avg_cov <- table_avg_cov |>
  mutate(Predictor = factor(Predictor, levels = predictor_order)) |>
  arrange(Response, Predictor)

# add headers
add_section <- function(df, name) {
  header <- tibble(
    Predictor = name,
    `Estimate (± 95% CI)` = NA_character_,
    `Importance (Σw)` = NA_real_,
    `P-value` = NA_character_
  )
  bind_rows(header, df |> select(-Response))
}

table_avg_final_cov <- bind_rows(
  add_section(filter(table_avg_cov, Response == "Bee richness (coverage-rarefied)"),
              "Bee richness (coverage-rarefied)"),
  add_section(filter(table_avg_cov, Response == "Bee Shannon diversity (coverage-rarefied)"),
              "Bee Shannon diversity (coverage-rarefied)")
)


# gt table 
gt_table_cov <- table_avg_final_cov |>
  gt() |>
  cols_align(align = "left", columns = Predictor) |>
  cols_align(
    align = "center",
    columns = c(`Estimate (± 95% CI)`, `Importance (Σw)`, `P-value`)
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = is.na(`Estimate (± 95% CI)`))
  ) |>
  fmt_missing(everything(), missing_text = "")

gt_table_cov
```

```{r export table|>}
gtsave(
  gt_table_cov,
  file.path("..", "output", "tables", "model_avg_cov_rarefaction.docx")
)
```

```{r export table|>}
gtsave(
  gt_table_cov,
  file.path("..", "output", "tables", "Table_2.docx")
)
```


## Model avg patchwork plot coverage-based rarefaction
```{r Model avg patchwork plot coverage-based rarefaction}
## Combined forest plot coverage-based rarefaction (faceted, free x-scales)

# extraction function (handles BOTH cases) - unchanged from before
extract_plot_df_cov <- function(model_obj, response_name, is_avg = TRUE) {
  
  if (is_avg) {
    # model.avg case (Hill q1)
    coef_df <- as.data.frame(summary(model_obj)$coefmat.full)
    coef_df$term <- rownames(coef_df)
    
    coef_df <- coef_df |>
      filter(term != "cond((Int))")
    
    ci <- confint(model_obj)
    coef_df$lower <- ci[coef_df$term, 1]
    coef_df$upper <- ci[coef_df$term, 2]
    
    imp_raw <- sw(model_obj)
    
    coef_df$base_term <- str_replace(coef_df$term, "\\.L|\\.Q|\\.C", "")
    coef_df$importance <- imp_raw[
      match(coef_df$base_term, names(imp_raw))
    ]
    
    # clean labels
    coef_df$Predictor <- coef_df$term |>
      str_replace("cond\\(", "") |>
      str_replace("\\)", "") |>
      str_replace("patchiness_ord.L", "Plot heterogeneity (linear)") |>
      str_replace("patchiness_ord.Q", "Plot heterogeneity (quadratic)") |>
      str_replace("patchiness_ord.C", "Plot heterogeneity (cubic)") |>
      str_replace("herb", "Herbaceous vegetation") |>
      str_replace("^bareground$", "Bareground") |>
      str_trim()
    
  } else {
    # ----- single model case (richness)
    coef_df <- as.data.frame(summary(model_obj)$coefficients$cond)
    coef_df$term <- rownames(coef_df)
    
    coef_df <- coef_df |>
      filter(term != "(Intercept)")
    
    ci <- confint(model_obj)
    match_idx <- match(coef_df$term, rownames(ci))
    
    coef_df$lower <- ci[match_idx, 1]
    coef_df$upper <- ci[match_idx, 2]
    
    # assign importance = 1
    coef_df$importance <- 1
    
    # clean labels
    coef_df$Predictor <- coef_df$term |>
      str_replace("patchiness_ord.L", "Plot heterogeneity (linear)") |>
      str_replace("patchiness_ord.Q", "Plot heterogeneity (quadratic)") |>
      str_replace("patchiness_ord.C", "Plot heterogeneity (cubic)") |>
      str_replace("herb", "Herbaceous vegetation") |>
      str_replace("^bareground$", "Bareground") |>
      str_trim()
  }
  
  coef_df |>
    transmute(
      Response = response_name,
      Predictor,
      Estimate,
      lower,
      upper,
      importance
    )
}

# build combined dataframe
df_combined_cov <- bind_rows(
  extract_plot_df_cov(bestmod_ri_cov, "Richness", is_avg = FALSE),
  extract_plot_df_cov(avg_mod_q1_cov, "Bee Shannon div.", is_avg = TRUE)
)

# predictor order (include cubic!)
predictor_order <- c(
  "Plot heterogeneity (linear)",
  "Plot heterogeneity (quadratic)",
  "Plot heterogeneity (cubic)",
  "Herbaceous vegetation",
  "Bareground"
)

df_combined_cov <- df_combined_cov |>
  mutate(
    Predictor = factor(Predictor, levels = rev(predictor_order)),
    Response = factor(Response, levels = c(
      "Bee Shannon div.",
      "Richness"
    ))
  ) |>
  # NOTE: no complete(Predictor, Response) here - only plot predictors that
  # actually have a coefficient for that response (i.e. were retained in
  # that response's top/averaged model set). This drops Bareground from the
  # Richness panel (never in its Δ AICc <= 2 set) while keeping it for
  # Shannon diversity (Σw = 0.34, shown at low alpha).
  filter(!is.na(Estimate))

pd <- position_dodge(width = 0.6)


## plot in facet as richenss and Shannon are on different data scales (Gamma log link and Gaussian)

# flip order of richness and Shannon for plot
df_combined_cov <- df_combined_cov |>
  mutate(
    Predictor = factor(Predictor, levels = rev(predictor_order)),
    Response = factor(Response, levels = c(
      "Richness",
      "Bee Shannon div."
    ))
  ) |>
  filter(!is.na(Estimate))

p_combined_cov_faceted <- ggplot(
  df_combined_cov,
  aes(x = Estimate, y = Predictor,
      color = Response,
      alpha = ifelse(is.na(importance), FALSE, importance >= 0.8))
) +

  geom_errorbarh(
    aes(xmin = lower, xmax = upper),
    height = 0,
    linewidth = 1
  ) +

  geom_point(size = 3) +

  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.6), guide = "none") +

  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +

  scale_color_manual(
    values = c(
      "Richness" = "#85144B",
      "Bee Shannon div." = "#FF4136"
    ),
    guide = "none"  # redundant once faceted by Response
  ) +

  facet_wrap(
    ~ Response,
    scales = "free_x",
    labeller = as_labeller(c(
      "Richness" = "Bee richness\n(coverage-rarefied)",
      "Bee Shannon div." = "Bee Shannon diversity\n(coverage-rarefied)"
    ))
  ) +

  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 14),
    strip.text = element_text(size = 13),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
    panel.spacing = unit(1.2, "lines")
  ) +

  labs(
    x = "Model estimate (95% CI)",
    y = NULL
  )

p_combined_cov_faceted
```

```{r export figure|>}
tiff(
  file.path("..", "output", "figures", "forest_covcombined_600dpi.tiff"),
  width = 5200, height = 2600, res = 600, compression = "lzw"
)
print(p_combined_cov_faceted)
dev.off()
```

```{r export figure|>}
tiff(
  file.path("..", "output", "figures", "Figure_5.tiff"),
  width = 5200, height = 2600, res = 600, compression = "lzw"
)
print(p_combined_cov_faceted)
dev.off()
```


## Bestmodel plots
```{r Best model combined prediction plot}
theme_pub <- theme_minimal(base_size = 16) +
  theme(
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 18),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)
  )

# clean plots
p_patch_ricov_clean <- p_patch_ricov +
  labs(title = NULL,
       subtitle = NULL,
       y = "Bee richness \n(coverage-rarefied)") +
  theme_pub

p_patch_q1cov_clean <- p_patch_q1cov +
  labs(title = NULL,
       subtitle = NULL,
       y = "Bee Shannon diversity \n(coverage-rarefied)") +
  theme_pub

# combine
p_bestmod_cov <- p_patch_ricov_clean /
  p_patch_q1cov_clean +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")"
  )

p_bestmod_cov
```

```{r export figure|>}
tiff(
  file.path("..", "output", "figures", "bestmod_pred_cov_600dpi.tiff"),
  width = 3300, height = 5000, res = 600, compression = "lzw"
)

print(p_bestmod_cov)
dev.off()
```

```{r export figure|>}
tiff(
  file.path("..", "output", "figures", "Figure_6.tiff"),
  width = 3300, height = 5000, res = 600, compression = "lzw"
)

print(p_bestmod_cov)
dev.off()
```

## Bestmodel table
```{r Best model table (coverage-rarefied)}
# R² values
r2_ri_cov <- performance::r2(bestmod_ri_cov)
r2_q1_cov <- performance::r2(bestmod_q1_cov)

extract_best_model_cov <- function(model, response_name) {
  
  coef_df <- as.data.frame(summary(model)$coefficients$cond)
  coef_df$term <- rownames(coef_df)
  
  coef_df <- coef_df |>
    filter(term != "(Intercept)")
  
  coef_df <- coef_df |>
    transmute(
      Response = response_name,
      Predictor = term,
      Estimate = round(Estimate, 3),
      SE = round(`Std. Error`, 3),
      p = case_when(
        `Pr(>|z|)` < 0.001 ~ "<0.001",
        TRUE ~ formatC(
          `Pr(>|z|)`,
          format = "f",
          digits = 3
        )
      )
    )
  
  coef_df$Predictor <- coef_df$Predictor |>
    str_replace("patchiness_ord.L", "Plot heterogeneity (linear)") |>
    str_replace("patchiness_ord.Q", "Plot heterogeneity (quadratic)") |>
    str_replace("patchiness_ord.C", "Plot heterogeneity (cubic)") |>
    str_replace("herb", "Herbaceous vegetation") |>
    str_replace("bareground", "Bareground") |>
    str_trim()
  
  coef_df
}

# extract
tab_ri_cov_best <- extract_best_model_cov(
  bestmod_ri_cov,
  "Bee richness (coverage-rarefied)"
)

tab_q1_cov_best <- extract_best_model_cov(
  bestmod_q1_cov,
  "Bee Shannon diversity (coverage-rarefied)"
)

# combine
table_best_cov <- bind_rows(
  tab_ri_cov_best,
  tab_q1_cov_best
)

# predictor order
predictor_order <- c(
  "Plot heterogeneity (linear)",
  "Plot heterogeneity (quadratic)",
  "Plot heterogeneity (cubic)",
  "Herbaceous vegetation",
  "Bareground"
)

table_best_cov <- table_best_cov |>
  mutate(
    Predictor = factor(
      Predictor,
      levels = predictor_order
    )
  ) |>
  arrange(
    Response,
    Predictor
  )

# add section headers
add_section <- function(df, name) {
  
  header <- tibble(
    Predictor = name,
    Estimate = NA_real_,
    SE = NA_real_,
    p = NA_character_
  )
  
  bind_rows(
    header,
    df |> select(-Response)
  )
}

# final table with R² values
table_final_cov <- bind_rows(

  add_section(
    filter(
      table_best_cov,
      Response == "Bee richness (coverage-rarefied)"
    ),
    sprintf(
      "Bee richness (coverage-rarefied) (R²m = %.3f; R²c = %.3f)",
      r2_ri_cov$R2_marginal,
      r2_ri_cov$R2_conditional
    )
  ),

  add_section(
    filter(
      table_best_cov,
      Response == "Bee Shannon diversity (coverage-rarefied)"
    ),
    sprintf(
      "Bee Shannon diversity (coverage-rarefied) (R²m = %.3f; R²c = %.3f)",
      r2_q1_cov$R2_marginal,
      r2_q1_cov$R2_conditional
    )
  )
)

# gt table
gt_table_cov_best <- table_final_cov |>
  gt() |>
  cols_align(
    align = "left",
    columns = Predictor
  ) |>
  cols_align(
    align = "center",
    columns = c(
      Estimate,
      SE,
      p
    )
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      rows = is.na(Estimate)
    )
  ) |>
  fmt_missing(
    everything(),
    missing_text = ""
  )

gt_table_cov_best
```

```{r save betsmodel coverage table|>}
gtsave(
  gt_table_cov_best,
  file.path("..", "output", "tables", "bestmodel_table_cov.docx")
)
```

```{r save betsmodel coverage table|>}
gtsave(
  gt_table_cov_best,
  file.path("..", "output", "tables", "Table_3.docx")
)
```


# Size-based rarefied richness model
## Bee richness
```{r Size rarefied bee richness global model}
# global model (size-based)
RRI_glmm_size <- glmmTMB(
  richness_rarefied_size ~ herb + bareground + patchiness_ord +
    (1 | region),
  data = data_model_size,
  family = gaussian(),
  na.action = na.fail
)

summary(RRI_glmm_size)
check_model(RRI_glmm_size)
```

```{r size rarefied bee richness dredge}
dri_size <- dredge(RRI_glmm_size, rank = "AICc", trace = 2)

head(dri_size)
```

```{r size rarefied model averaging richness}
cand_ri_size <- get.models(dri_size, subset = delta <= 2)

bestmod_ri_size <- cand_ri_size[[1]]

avg_mod_ri_size <- model.avg(cand_ri_size, revised.var = TRUE)

summary(avg_mod_ri_size)
confint(avg_mod_ri_size)

# importance
sw(subset(dri_size, delta <= 2))
```

Best model size rarefied richness
```{r}
bestmod_ri_size

bestmod_ri_site <- glmmTMB(
  richness_rarefied_size ~ herb + + patchiness_ord +
    (1 | region),
  data = data_model_size,
  family = gaussian(),
  na.action = na.fail
)

summary(RRI_glmm_size)
```


Best model prediction — richness (size-rarefied)
```{r plot best model size rarefied richness}
# predictions (patchiness effect, herb averaged)
pred_patch_size_ri <- ggpredict(
  bestmod_ri_size,
  terms = "patchiness_ord"
)

# plot
p_patch_risize <- ggplot() +
  
  geom_jitter(
    data = data_model_size,
    aes(x = patchiness_ord, y = richness_rarefied_size),
    width = 0.1, height = 0.1,
    alpha = 0.25, size = 1.5,
    color = "grey50"
  ) +
  
  geom_point(
    data = pred_patch_size_ri,
    aes(x = x, y = predicted),
    color = "#85144B",
    size = 3
  ) +
  
  geom_errorbar(
    data = pred_patch_size_ri,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    color = "#85144B",
    width = 0.1,
    linewidth = 1
  ) +
  
  labs(
    title = "Effect of Plot heterogeneity",
    subtitle = "Predicted bee richness (size-rarefied)",
    x = "Plot heterogeneity",
    y = "Bee richness"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13)
  )

p_patch_risize
```


## Bee Hill Q = 1
```{r Size rarefied bee hill q1 global model}
# hill q1 size-based global model
Q1_glmm_size <- glmmTMB(
  hill_q1_rarefied_size ~ herb + bareground + patchiness_ord +
    (1 | region),
  family = gaussian(),
  data = data_model_size,
  na.action = na.fail
)

summary(Q1_glmm_size)
check_model(Q1_glmm_size)
```
```{r size rarefied bee hillq1 dredge}
dq1_size <- dredge(Q1_glmm_size, rank = "AICc", trace = 2)

head(dq1_size)
```

```{r size rarefied model averaging hillq1}
cand_q1_size <- get.models(dq1_size, subset = delta <= 2)

bestmod_q1_size <- cand_q1_size[[1]]

avg_mod_q1_size <- model.avg(cand_q1_size, revised.var = TRUE)

summary(avg_mod_q1_size)
confint(avg_mod_q1_size)

# importance
sw(subset(dq1_size, delta <= 2))
```


Best model size rarefied Hill Q1
```{r Best model size based hillq1}
bestmod_q1_size

bestmod_q1_size <- glmmTMB(hill_q1_rarefied_size ~ herb + patchiness_ord + (1 | region),
                           family = gaussian(),
                           data = data_model_size,
                           na.action = na.fail)

summary(bestmod_q1_size)
```


Best model prediction — Hill q1 (size-rarefied)
```{r Best model size rarefied Hill Q1}
# predictions (patchiness effect, herb averaged)
pred_patch_size_q1 <- ggpredict(
  bestmod_q1_size,
  terms = "patchiness_ord"
)

# plot
p_patch_q1size <- ggplot() +
  
  geom_jitter(
    data = data_model_size,
    aes(x = patchiness_ord, y = hill_q1_rarefied_size),
    width = 0.1, height = 0.05,
    alpha = 0.25, size = 1.5,
    color = "grey50"
  ) +
  
  geom_point(
    data = pred_patch_size_q1,
    aes(x = x, y = predicted),
    color = "#FF4136",
    size = 3
  ) +
  
  geom_errorbar(
    data = pred_patch_size_q1,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    color = "#FF4136",
    width = 0.1,
    linewidth = 1
  ) +
  
  labs(
    title = "Effect of Plot heterogeneity",
    subtitle = "Predicted bee Shannon diversity (size-rarefied)",
    x = "Plot heterogeneity",
    y = "Bee Shannon diversity"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13)
  )

p_patch_q1size
```

## Model avg table extraction size-based rarefaction
```{r model avg table extraction size-based rarefaction}
extract_avg_table_size <- function(avg_model, response_name) {

  coef_df <- as.data.frame(summary(avg_model)$coefmat.full)
  coef_df$term <- rownames(coef_df)

  coef_df <- coef_df |>
    filter(term != "cond((Int))")

  ci <- confint(avg_model)
  coef_df$lower <- ci[coef_df$term, 1]
  coef_df$upper <- ci[coef_df$term, 2]

  coef_df$p_value <- coef_df[, "Pr(>|z|)"]

  coef_df$p_value_fmt <- ifelse(
    coef_df$p_value < 0.001,
    "<0.001",
    formatC(coef_df$p_value, format = "f", digits = 3)
  )

  # importance (model.avg directly)
  imp_raw <- sw(avg_model)
  imp_vals <- as.numeric(imp_raw)
  names(imp_vals) <- names(imp_raw)

  # base term grouping (important for patchiness)
  coef_df$base_term <- str_replace(coef_df$term, "\\.L|\\.Q|\\.C", "")

  coef_df$importance <- imp_vals[
    match(coef_df$base_term, names(imp_vals))
  ]

  # clean labels (keep L and Q distinct)
  coef_df$Predictor <- coef_df$term |>
    str_replace("cond\\(", "") |>
    str_replace("\\)", "") |>
    str_replace("patchiness_ord.L", "Plot heterogeneity (linear)") |>
    str_replace("patchiness_ord.Q", "Plot heterogeneity (quadratic)") |>
    str_replace("patchiness_ord.C", "Plot heterogeneity (cubic)") |>
    str_replace("hill_q0", "Floral Richness") |>
    str_replace("sand", "Sand") |>
    str_replace("herb", "Herbaceous vegetation") |>
    str_replace("bareground", "Bareground") |>
    str_replace("vegheight", "Vegetation height") |>
    str_trim()

  coef_df |>
    transmute(
      Response = response_name,
      Predictor,
      `Estimate (± 95% CI)` = sprintf("%.3f (%.2f – %.2f)", Estimate, lower, upper),
      `Importance (Σw)` = round(importance, 2),
      `P-value` = p_value_fmt
    )
}

# extract tables
tab_ri_size <- extract_avg_table_size(avg_mod_ri_size, "Bee richness (size-rarefied)")
tab_q1_size <- extract_avg_table_size(avg_mod_q1_size, "Bee Shannon diversity (size-rarefied)")

table_avg_size <- bind_rows(tab_ri_size, tab_q1_size)

# predictor order (adapted to rarefied models)
predictor_order <- c(
  "Plot heterogeneity (linear)",
  "Plot heterogeneity (quadratic)",
  "Plot heterogeneity (cubic)",
  "Herbaceous vegetation",
  "Bareground"
)

table_avg_size <- table_avg_size |>
  mutate(Predictor = factor(Predictor, levels = predictor_order)) |>
  arrange(Response, Predictor)

# section headers
add_section <- function(df, name) {
  header <- tibble(
    Predictor = name,
    `Estimate (± 95% CI)` = NA_character_,
    `Importance (Σw)` = NA_real_,
    `P-value` = NA_character_
  )
  bind_rows(header, df |> select(-Response))
}

table_avg_final_size <- bind_rows(
  add_section(filter(
    table_avg_size,
    Response == "Bee richness (size-rarefied)"), "Bee richness (size-rarefied)"),
  add_section(filter(
    table_avg_size,
    Response == "Bee Shannon diversity (size-rarefied)"), "Bee Shannon diversity (size-rarefied)")
)

# create gt table
gt_table_size <- table_avg_final_size |>
  gt() |>
  cols_align(align = "left", columns = Predictor) |>
  cols_align(
    align = "center",
    columns = c(`Estimate (± 95% CI)`, `Importance (Σw)`, `P-value`)
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = is.na(`Estimate (± 95% CI)`))
  ) |>
  fmt_missing(everything(), missing_text = "")

gt_table_size
```

```{r save size-based model avg tables|>}
gtsave(gt_table_size,
       file.path("..", "output", "tables", "model_avg_size_rarefaction.docx"))
```

```{r save size-based model avg tables|>}
gtsave(gt_table_size,
       file.path("..", "output", "tables", "Table_D2.1.docx"))
```


## Model avg patchwork plot size-based rarefaction
```{r combined forest plot size-based rarefaction}
# extraction function adapted for plotting
extract_plot_df_size <- function(avg_model, response_name) {

  coef_df <- as.data.frame(summary(avg_model)$coefmat.full)
  coef_df$term <- rownames(coef_df)

  coef_df <- coef_df |>
    filter(term != "cond((Int))")

  ci <- confint(avg_model)
  coef_df$lower <- ci[coef_df$term, 1]
  coef_df$upper <- ci[coef_df$term, 2]

  imp_raw <- sw(avg_model)

  coef_df$base_term <- str_replace(coef_df$term, "\\.L|\\.Q|\\.C", "")
  coef_df$importance <- imp_raw[
    match(coef_df$base_term, names(imp_raw))
  ]

  coef_df$Predictor <- coef_df$term |>
    str_replace("cond\\(", "") |>
    str_replace("\\)", "") |>
    str_replace("patchiness_ord.L", "Plot heterogeneity (linear)") |>
    str_replace("patchiness_ord.Q", "Plot heterogeneity (quadratic)") |>
    str_replace("patchiness_ord.C", "Plot heterogeneity (cubic)") |>
    str_replace("herb", "Herbaceous vegetation") |>
    str_replace("^bareground$", "Bareground") |>
    str_trim()

  coef_df |>
    transmute(
      Response = response_name,
      Predictor,
      Estimate,
      lower,
      upper,
      importance
    )
}

# build combined dataframe
df_combined_size <- bind_rows(
  extract_plot_df_size(avg_mod_ri_size, "Richness"),
  extract_plot_df_size(avg_mod_q1_size, "Bee Shannon div.")
)

# predictor order (include quadratic)
predictor_order <- c(
  "Plot heterogeneity (linear)",
  "Plot heterogeneity (quadratic)",
  "Plot heterogeneity (cubic)",
  "Herbaceous vegetation",
  "Bareground"
)

df_combined_size <- df_combined_size |>
  mutate(
    Predictor = factor(Predictor, levels = rev(predictor_order)),
    Response = factor(Response, levels = c(
      "Bee Shannon div.",
      "Richness"
    ))
  )

pd <- position_dodge(width = 0.6)

# plot
p_combined_size <- ggplot(
  df_combined_size,
  aes(x = Estimate, y = Predictor, color = Response, group = Response)
) +

  geom_errorbarh(
    aes(xmin = lower, xmax = upper,
        alpha = ifelse(is.na(importance), FALSE, importance >= 0.8)),
    position = pd,
    height = 0,
    linewidth = 1,
    na.rm = FALSE
  ) +

  geom_point(
    aes(alpha = ifelse(is.na(importance), FALSE, importance >= 0.8)),
    position = pd,
    size = 3,
    na.rm = FALSE
  ) +

  scale_alpha_manual(
    values = c("TRUE" = 1, "FALSE" = 0.7),
    guide = "none"
  ) +

  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +

  scale_color_manual(
    values = c(
      "Richness" = "#85144B",
      "Bee Shannon div." = "#FF4136"
    ),
    breaks = c("Richness", "Bee Shannon div."),
    labels = c(
      "Richness" = "Bee richness \n(size-rarefied)",
      "Bee Shannon div." = "Bee Shannon diversity \n(size-rarefied)"
    )
  ) +

  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 14),
    legend.position = "right",
    legend.direction = "vertical",
    legend.title = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)
  ) +

  labs(
    x = "Model-averaged estimate (95% CI)",
    y = NULL,
    color = NULL
  )

p_combined_size
```

```{r export figure|>}
tiff(
  file.path("..", "output", "figures", "forest_sizecombined_600dpi.tiff"),
  width = 4500, height = 2600, res = 600, compression = "lzw"
)

print(p_combined_size)
dev.off()
```

```{r export figure|>}
tiff(
  file.path("..", "output", "figures", "Figure_D2.1.tiff"),
  width = 4500, height = 2600, res = 600, compression = "lzw"
)

print(p_combined_size)
dev.off()
```


## Best model combined prediction plot (size-rarefied)
```{r Best model comined prediction plot}
theme_pub <- theme_minimal(base_size = 16) +
  theme(
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 18),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)
  )

# clean plots
p_patch_risize_clean <- p_patch_risize +
  labs(title = NULL,
       subtitle = NULL,
       y = "Bee richness \n(size-rarefied)") +
  theme_pub

p_patch_q1size_clean <- p_patch_q1size +
  labs(title = NULL,
       subtitle = NULL,
       y = "Bee Shannon diversity \n(size-rarefied)") +
  theme_pub

# combine
p_bestmod_size <- p_patch_risize_clean /
  p_patch_q1size_clean +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")"
  )

p_bestmod_size
``` 

```{r export plot|>}
tiff(
  file.path("..", "output", "figures", "bestmod_pred_size_600dpi.tiff"),
  width = 3300, height = 5000, res = 600, compression = "lzw"
)

print(p_bestmod_size)
dev.off()
```

```{r export plot|>}
tiff(
  file.path("..", "output", "figures", "Figure_D2.2.tiff"),
  width = 3300, height = 5000, res = 600, compression = "lzw"
)

print(p_bestmod_size)
dev.off()
```


## Best model table (size-rarefied)
```{r Best model table size rarfied}
# R² values
r2_ri_size <- performance::r2(bestmod_ri_size)
r2_q1_size <- performance::r2(bestmod_q1_size)

# reuse function extract_best_model_cov()

# extract
tab_ri_size_best <- extract_best_model_cov(
  bestmod_ri_size,
  "Bee richness (size-rarefied)"
)

tab_q1_size_best <- extract_best_model_cov(
  bestmod_q1_size,
  "Bee Shannon diversity (size-rarefied)"
)

# combine
table_best_size <- bind_rows(
  tab_ri_size_best,
  tab_q1_size_best
)

# order predictors
predictor_order <- c(
  "Plot heterogeneity (linear)",
  "Plot heterogeneity (quadratic)",
  "Plot heterogeneity (cubic)",
  "Herbaceous vegetation",
  "Bareground"
)

table_best_size <- table_best_size |>
  mutate(
    Predictor = factor(
      Predictor,
      levels = predictor_order
    )
  ) |>
  arrange(Response, Predictor)

# section headers
add_section <- function(df, name) {

  header <- tibble(
    Predictor = name,
    Estimate = NA_real_,
    SE = NA_real_,
    p = NA_character_
  )

  bind_rows(
    header,
    df |> select(-Response)
  )
}

# add R² values to section headers
table_final_size <- bind_rows(

  add_section(
    filter(
      table_best_size,
      Response == "Bee richness (size-rarefied)"
    ),
    sprintf(
      "Bee richness (size-rarefied) (R²m = %.3f; R²c = %.3f)",
      r2_ri_size$R2_marginal,
      r2_ri_size$R2_conditional
    )
  ),

  add_section(
    filter(
      table_best_size,
      Response == "Bee Shannon diversity (size-rarefied)"
    ),
    sprintf(
      "Bee Shannon diversity (size-rarefied) (R²m = %.3f; R²c = %.3f)",
      r2_q1_size$R2_marginal,
      r2_q1_size$R2_conditional
    )
  )
)

# gt table
gt_table_size <- table_final_size |>
  gt() |>
  cols_align(
    align = "left",
    columns = Predictor
  ) |>
  cols_align(
    align = "center",
    columns = c(
      Estimate,
      SE,
      p
    )
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      rows = is.na(Estimate)
    )
  ) |>
  fmt_missing(
    everything(),
    missing_text = ""
  )

gt_table_size
```

```{r export table|>}
gtsave(gt_table_size,
       file.path("..", "output", "tables", "bestmodel_table_size.docx"))
```

```{r export table|>}
gtsave(gt_table_size,
       file.path("..", "output", "tables", "Table_D2.2.docx"))
```
