---
title: "08_plot_heterogeneity_PCA"
output: html_document
---
```{r load packages}
library(tidyverse)
library(vegan)
library(ggplot2)
library(ggrepel)
library(dunn.test)
```

```{r Load data}
data_all <- read.csv(file.path("..", "data", "cleaned", "data_all.csv"))
```

```{r prepare data}
# Select variables of interest
hab <- data_all %>%
  select(
    patchiness,
    bareground, sand, soil,
    herb, woody, vegheight, flat,
    sloped, steep,
    hill_q0
  ) %>%
  drop_na()

cat("Complete rows for PCA:", nrow(hab), "\n")
cat("Patchiness distribution:\n")
print(table(hab$patchiness))

# Recode vairable names for simpler output interpretation
recode_vars <- function(x) case_match(x,
  "bareground" ~ "Bareground",
  "sand"       ~ "Sand",
  "soil"       ~ "Soil",
  "herb"       ~ "Herbaceous vegetation",
  "woody"      ~ "Woody vegetation",
  "vegheight"  ~ "Vegetation height",
  "sloped"     ~ "Sloped",
  "flat"       ~ "Flat",
  "steep"      ~ "Steep",
  "hill_q0"    ~ "Floral richness",
  .default = x
)
```

PCA on standardized habitat variables, with patchiness excluded from axis construction and treated as a passive variable.
```{r PCA habitat variables only}
hab_vars   <- hab %>% select(-patchiness)
patchiness <- hab$patchiness

pca_result  <- prcomp(hab_vars, scale. = TRUE, center = TRUE)
pca_summary <- summary(pca_result)
print(pca_summary)
screeplot(pca_result, type = "lines", main = "Scree plot")
```
Plot PCAhabitat variables with points colored by patchiness
```{r Biplot colored by patchiness}
var_exp    <- pca_summary$importance[2, 1:2] * 100
pca_scores <- as.data.frame(pca_result$x[, 1:2])
pca_scores$patchiness <- as.factor(patchiness)

arrow_scale <- 3
loadings <- as.data.frame(pca_result$rotation[, 1:2]) %>%
  rownames_to_column("variable") %>%
  mutate(
    PC1s  = PC1 * arrow_scale,
    PC2s  = PC2 * arrow_scale,
    label = recode_vars(variable)
  )

p_biplot <- ggplot(pca_scores, aes(x = PC1, y = PC2, colour = patchiness)) +
  geom_point(alpha = 0.7, size = 2.5) +
  scale_colour_manual(
    values = c("0" = "#d73027", "1" = "#fc8d59",
               "2" = "#91bfdb", "3" = "#4575b4"),
    name = "Plot heterogeneity"
  ) +
  geom_segment(
    data = loadings,
    aes(x = 0, y = 0, xend = PC1s, yend = PC2s),
    arrow = arrow(length = unit(0.25, "cm")),
    colour = "grey30", linewidth = 0.6,
    inherit.aes = FALSE
  ) +
  geom_text_repel(
    data = loadings,
    aes(x = PC1s, y = PC2s, label = label),
    colour = "grey15", size = 3.2,
    inherit.aes = FALSE,
    box.padding = 0.4
  ) +
  labs(
    x     = paste0("PC1 (", round(var_exp[1], 1), "% variance)"),
    y     = paste0("PC2 (", round(var_exp[2], 1), "% variance)"),
    title = "PCA of habitat variables — plot heterogeneity as passive variable"
  ) +
  theme_bw(base_size = 12)

print(p_biplot)
```
PC1 reflects a gradient from closed, herb‑dominated, homogeneous vegetation to more structurally complex habitats.
PC2 captures particularly gradient of substrate and floral resources.


Fit patchiness post-hoc onto PCA space to quantify association with multivariate habitat structure
```{r}
# envfit with patchiness as environemntal vector
ef <- envfit(
  ord          = pca_result$x[, 1:2],
  env          = data.frame(patchiness = patchiness),
  permutations = 999
)
print(ef)

ef_coords <- as.data.frame(scores(ef, display = "vectors")) * arrow_scale
ef_coords$variable <- rownames(ef_coords)

p_envfit <- p_biplot +
  geom_segment(
    data = ef_coords,
    aes(x = 0, y = 0, xend = PC1, yend = PC2),
    arrow = arrow(length = unit(0.3, "cm")),
    colour = "black", linewidth = 1,
    inherit.aes = FALSE
  ) +
  geom_text_repel(
    data = ef_coords,
    aes(x = PC1, y = PC2, label = "Plot heterogeneity"),
    colour = "black", size = 3.8, fontface = "bold",
    inherit.aes = FALSE
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_blank()  # remove title cleanly
  )

p_envfit
```
Export figure
```{r export figure}

tiff(
  file.path("..", "output", "figures", "pca_envfit_600dpi.tiff"),
  width = 4000, height = 2250,
  res = 600,
  compression = "lzw"
)

print(p_envfit)
dev.off()

```

