---
title: "08_plot_heterogeneity"
output: html_document
---
```{r load packages}
library(tidyverse)
library(vegan)
library(ggplot2)
library(ggrepel)
library(dunn.test)
```

```{r Load data}
data_all <- read.csv(file.path("..", "data", "cleaned", "data_all.csv"))
```

```{r prepare data}
# Select variables of interest
hab <- data_all %>%
  select(
    patchiness,
    bareground, sand, soil,
    herb, woody, vegheight, flat,
    sloped, steep,
    hill_q0
  ) %>%
  drop_na()

cat("Complete rows for PCA:", nrow(hab), "\n")
cat("Patchiness distribution:\n")
print(table(hab$patchiness))

# Recode vairable names for simpler output interpretation
recode_vars <- function(x) case_match(x,
  "bareground" ~ "Bareground",
  "sand"       ~ "Sand",
  "soil"       ~ "Soil",
  "herb"       ~ "Herbaceous vegetation",
  "woody"      ~ "Woody vegetation",
  "vegheight"  ~ "Vegetation height",
  "sloped"     ~ "Sloped",
  "flat"       ~ "Flat",
  "steep"      ~ "Steep",
  "hill_q0"    ~ "Floral richness",
  .default = x
)
```

PCA on standardized habitat variables, with patchiness excluded from axis construction and treated as a passive variable.
```{r PCA habitat variables only}
hab_vars   <- hab %>% select(-patchiness)
patchiness <- hab$patchiness

pca_result  <- prcomp(hab_vars, scale. = TRUE, center = TRUE)
pca_summary <- summary(pca_result)
print(pca_summary)
screeplot(pca_result, type = "lines", main = "Scree plot")
```
Plot PCAhabitat variables with points colored by patchiness
```{r Biplot colored by patchiness}
var_exp    <- pca_summary$importance[2, 1:2] * 100
pca_scores <- as.data.frame(pca_result$x[, 1:2])
pca_scores$patchiness <- as.factor(patchiness)

arrow_scale <- 3
loadings <- as.data.frame(pca_result$rotation[, 1:2]) %>%
  rownames_to_column("variable") %>%
  mutate(
    PC1s  = PC1 * arrow_scale,
    PC2s  = PC2 * arrow_scale,
    label = recode_vars(variable)
  )

p_biplot <- ggplot(pca_scores, aes(x = PC1, y = PC2, colour = patchiness)) +
  geom_point(alpha = 0.7, size = 2.5) +
  scale_colour_manual(
    values = c("0" = "#d73027", "1" = "#fc8d59",
               "2" = "#91bfdb", "3" = "#4575b4"),
    name = "Plot heterogeneity"
  ) +
  geom_segment(
    data = loadings,
    aes(x = 0, y = 0, xend = PC1s, yend = PC2s),
    arrow = arrow(length = unit(0.25, "cm")),
    colour = "grey30", linewidth = 0.6,
    inherit.aes = FALSE
  ) +
  geom_text_repel(
    data = loadings,
    aes(x = PC1s, y = PC2s, label = label),
    colour = "grey15", size = 3.2,
    inherit.aes = FALSE,
    box.padding = 0.4
  ) +
  labs(
    x     = paste0("PC1 (", round(var_exp[1], 1), "% variance)"),
    y     = paste0("PC2 (", round(var_exp[2], 1), "% variance)"),
    title = "PCA of habitat variables — plot heterogeneity as passive variable"
  ) +
  theme_bw(base_size = 12)

print(p_biplot)
```
PC1 reflects a gradient from closed, herb‑dominated, homogeneous vegetation to more structurally complex habitats.
PC2 captures particularly gradient of substrate and floral resources.


Fit patchiness post-hoc onto PCA space to quantify association with multivariate habitat structure
```{r}
# envfit with patchiness as environemntal vector
ef <- envfit(
  ord          = pca_result$x[, 1:2],
  env          = data.frame(patchiness = patchiness),
  permutations = 999
)
print(ef)

ef_coords <- as.data.frame(scores(ef, display = "vectors")) * arrow_scale
ef_coords$variable <- rownames(ef_coords)

p_envfit <- p_biplot +
  geom_segment(
    data = ef_coords,
    aes(x = 0, y = 0, xend = PC1, yend = PC2),
    arrow = arrow(length = unit(0.3, "cm")),
    colour = "black", linewidth = 1,
    inherit.aes = FALSE
  ) +
  geom_text_repel(
    data = ef_coords,
    aes(x = PC1, y = PC2, label = "Plot heterogeneity"),
    colour = "black", size = 3.8, fontface = "bold",
    inherit.aes = FALSE
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_blank()  # remove title cleanly
  )

p_envfit
```
Nearly 47% of the variation in patchiness is explained by the position of plots in PCA space (PC1–PC2).
In ecological multivariate analyses, this is a strong effect size, especially for a passive variable not used to build the PCA.

Low patchiness plots (red, class 0) cluster:
At high PC1
Often low PC2
Associated with dense herb cover and soil continuity

High patchiness plots (dark blue, class 3) cluster:
At negative PC1
At positive PC2
Aligned with sand, floral richness, woody elements, bare ground

Patchiness is strongly and significantly aligned with the main multivariate habitat gradients, increasing toward plots characterized by higher structural complexity, floral richness, and substrate heterogeneity, and decreasing in homogeneous, herb‑dominated habitats.
gpt

```{r print PCA}

tiff(
  file.path("..", "output", "figures", "pca_envfit_600dpi.tiff"),
  width = 4000, height = 2250,
  res = 600,
  compression = "lzw"
)

print(p_envfit)
dev.off()

```

