---
title: "08_patchiness"
output: html_document
---
```{r load packages}
library(tidyverse)
library(vegan)
library(ggplot2)
library(ggrepel)
library(dunn.test)
```

```{r Load data}
data_all <- read.csv(file.path("..", "data", "cleaned", "data_all.csv"))
```

```{r prepare data}
# Select variables of interest
hab <- data_all %>%
  select(
    patchiness,
    bareground, sand, soil,
    herb, woody, vegheight, flat,
    sloped, steep,
    hill_q0
  ) %>%
  drop_na()

cat("Complete rows for PCA:", nrow(hab), "\n")
cat("Patchiness distribution:\n")
print(table(hab$patchiness))

# Recode vairable names for simpler output interpretation
recode_vars <- function(x) case_match(x,
  "bareground" ~ "Bare ground",
  "sand"       ~ "Sand",
  "soil"       ~ "Soil",
  "herb"       ~ "Herb cover",
  "woody"      ~ "Woody cover",
  "vegheight"  ~ "Veg height",
  "sloped"     ~ "Sloped",
  "flat"       ~ "Flat",
  "steep"      ~ "Steep",
  "hill_q0"    ~ "Floral richness",
  .default = x
)
```

PCA on standardized habitat variables, with patchiness excluded from axis construction and treated as a passive variable.
```{r PCA habitat variables only}
hab_vars   <- hab %>% select(-patchiness)
patchiness <- hab$patchiness

pca_result  <- prcomp(hab_vars, scale. = TRUE, center = TRUE)
pca_summary <- summary(pca_result)
print(pca_summary)
screeplot(pca_result, type = "lines", main = "Scree plot")
```
Plot PCAhabitat variables with points colored by patchiness
```{r Biplot colored by patchiness}
var_exp    <- pca_summary$importance[2, 1:2] * 100
pca_scores <- as.data.frame(pca_result$x[, 1:2])
pca_scores$patchiness <- as.factor(patchiness)

arrow_scale <- 3
loadings <- as.data.frame(pca_result$rotation[, 1:2]) %>%
  rownames_to_column("variable") %>%
  mutate(
    PC1s  = PC1 * arrow_scale,
    PC2s  = PC2 * arrow_scale,
    label = recode_vars(variable)
  )

p_biplot <- ggplot(pca_scores, aes(x = PC1, y = PC2, colour = patchiness)) +
  geom_point(alpha = 0.7, size = 2.5) +
  scale_colour_manual(
    values = c("0" = "#d73027", "1" = "#fc8d59",
               "2" = "#91bfdb", "3" = "#4575b4"),
    name = "Plot heterogeneity"
  ) +
  geom_segment(
    data = loadings,
    aes(x = 0, y = 0, xend = PC1s, yend = PC2s),
    arrow = arrow(length = unit(0.25, "cm")),
    colour = "grey30", linewidth = 0.6,
    inherit.aes = FALSE
  ) +
  geom_text_repel(
    data = loadings,
    aes(x = PC1s, y = PC2s, label = label),
    colour = "grey15", size = 3.2,
    inherit.aes = FALSE,
    box.padding = 0.4
  ) +
  labs(
    x     = paste0("PC1 (", round(var_exp[1], 1), "% variance)"),
    y     = paste0("PC2 (", round(var_exp[2], 1), "% variance)"),
    title = "PCA of habitat variables — plot heterogeneity as passive variable"
  ) +
  theme_bw(base_size = 12)

print(p_biplot)
```
PC1 reflects a gradient from closed, herb‑dominated, homogeneous vegetation to more structurally complex habitats.
PC2 captures particularly gradient of substrate and floral resources.


Fit patchiness post-hoc onto PCA space to quantify association with multivariate habitat structure
```{r}
# envfit with patchiness as environemntal vector
ef <- envfit(
  ord          = pca_result$x[, 1:2],
  env          = data.frame(patchiness = patchiness),
  permutations = 999
)
print(ef)

ef_coords <- as.data.frame(scores(ef, display = "vectors")) * arrow_scale
ef_coords$variable <- rownames(ef_coords)

p_biplot +
  geom_segment(
    data = ef_coords,
    aes(x = 0, y = 0, xend = PC1, yend = PC2),
    arrow = arrow(length = unit(0.3, "cm")),
    colour = "black", linewidth = 1,
    inherit.aes = FALSE
  ) +
  geom_text_repel(
    data = ef_coords,
    aes(x = PC1, y = PC2, label = "Plot heterogeneity"),
    colour = "black", size = 3.8, fontface = "bold",
    inherit.aes = FALSE
  ) +
  labs(title = "PCA with Plot heterogeneity fitted as environmental vector (envfit)")
```
Nearly 47% of the variation in patchiness is explained by the position of plots in PCA space (PC1–PC2).
In ecological multivariate analyses, this is a strong effect size, especially for a passive variable not used to build the PCA.

Low patchiness plots (red, class 0) cluster:
At high PC1
Often low PC2
Associated with dense herb cover and soil continuity

High patchiness plots (dark blue, class 3) cluster:
At negative PC1
At positive PC2
Aligned with sand, floral richness, woody elements, bare ground

Patchiness is strongly and significantly aligned with the main multivariate habitat gradients, increasing toward plots characterized by higher structural complexity, floral richness, and substrate heterogeneity, and decreasing in homogeneous, herb‑dominated habitats.
gpt

```{r Correlation tests}
# Spearman
cor_results <- hab_vars %>%
  summarise(across(everything(),
                   ~ cor(.x, patchiness,
                         method = "spearman",
                         use    = "complete.obs"))) %>%
  pivot_longer(everything(),
               names_to  = "variable",
               values_to = "spearman_rho") %>%
  mutate(variable = recode_vars(variable)) %>%
  arrange(desc(abs(spearman_rho)))

print(cor_results)

# Kruskal Wallis
kw_results <- hab_vars %>%
  summarise(across(everything(),
                   ~ kruskal.test(.x,
                                  g = as.factor(patchiness))$p.value)) %>%
  pivot_longer(everything(),
               names_to  = "variable",
               values_to = "KW_pvalue") %>%
  mutate(
    sig      = case_when(
      KW_pvalue < 0.001 ~ "***",
      KW_pvalue < 0.01  ~ "**",
      KW_pvalue < 0.05  ~ "*",
      TRUE              ~ "ns"
    ),
    variable = recode_vars(variable)
  ) %>%
  arrange(KW_pvalue)

print(kw_results)

# Dunn Post-Hoc
key_vars <- c("hill_q0", "bareground", "vegheight", "herb", "woody")

for (v in key_vars) {
  cat("\n── Dunn test:", recode_vars(v), "──\n")
  dunn.test(hab_vars[[v]],
            g      = as.factor(patchiness),
            method = "bonferroni",
            kw     = TRUE)
}

# Summary table
summary_table <- left_join(cor_results, kw_results, by = "variable") %>%
  arrange(desc(abs(spearman_rho)))

print(summary_table)

# Boxplots

hab %>%
  select(patchiness, all_of(key_vars)) %>%
  pivot_longer(-patchiness,
               names_to  = "variable",
               values_to = "value") %>%
  mutate(
    patchiness = as.factor(patchiness),
    variable   = recode_vars(variable)
  ) %>%
  ggplot(aes(x = patchiness, y = value, fill = patchiness)) +
  geom_boxplot(alpha = 0.7, outlier.size = 1) +
  scale_fill_manual(values = c("0" = "#d73027", "1" = "#fc8d59",
                               "2" = "#91bfdb", "3" = "#4575b4")) +
  facet_wrap(~variable, scales = "free_y") +
  labs(x     = "Plot heterogeneity score",
       y     = "Value",
       title = "Habitat variables by plot heterogeneity score") +
  theme_bw(base_size = 11) +
  theme(legend.position = "none")
```
Spearman rank correlations revealed strong and directionally consistent associations between patchiness and several key habitat components (Table X). Patchiness increased with sand cover (ρ = 0.61), floral richness (ρ = 0.61), woody cover (ρ = 0.49), and bare ground availability (ρ = 0.47), while decreasing with herbaceous cover (ρ = −0.49) and homogeneous soil surfaces (ρ = −0.36). These relationships correspond closely to the ecological criteria used during field scoring, indicating that higher patchiness values consistently reflect increased structural and compositional heterogeneity. Variables only marginally related to patchiness (e.g. vegetation height, steepness) showed weaker correlations, while variables not central to the patchiness assessment (flat and sloped surfaces) were not significantly associated.

Kruskal–Wallis tests demonstrated that most habitat variables differed significantly among patchiness classes (p < 0.001 for 8 out of 10 variables, except for sloped and flat), indicating that patchiness categories represent distinct habitat configurations rather than arbitrary or noisy groupings. The absence of significant differences for variables unrelated to the scoring protocol reinforces that patchiness captures targeted aspects of habitat structure rather than acting as a general surrogate for all environmental variation.

Post‑hoc Dunn tests revealed that habitat variables responded to patchiness in distinct but ecologically interpretable ways. While floral richness increased monotonically across all patchiness classes, other variables exhibited threshold‑like patterns. Bare ground showed strong differentiation between the lowest and highest patchiness classes, with weaker contrasts at intermediate levels, indicating both entry and expansion thresholds. Herbaceous and woody cover displayed pronounced low‑end thresholds, with sharp separation between patchiness 0 and all higher classes, followed by saturation at higher patchiness levels. Vegetation height differed primarily at intermediate patchiness values, reflecting the emergence of vertical structural heterogeneity rather than a continuous gradient. These complementary response patterns indicate that patchiness integrates multiple habitat components that contribute differentially along the heterogeneity gradient.

Taken together, these results demonstrate that patchiness is a reproducible, ecologically interpretable index that aligns with independently measured habitat characteristics in a predictable and structured manner. Its strong correspondence with variables explicitly targeted during field assessment, combined with its limited association with irrelevant variables, supports its construct validity. Patchiness therefore represents a meaningful summary of habitat heterogeneity and is appropriate for inclusion as an explanatory variable in subsequent statistical models.


gpt
